# 0. Setup

**Module 00 — Project setup.** Load packages, read metadata, and create the PNA object. Run this module first, then continue with `01_quality_control.qmd` in the same R session. Look for **Decision checkpoint** callouts when editing the sample table.

## 0.1. Load packages and helper functions

A set of commonly used packages and helper functions are loaded here. The core packages include `tidyverse` for data manipulation and visualization, `Seurat` for single-cell analysis infrastructure, `pixelatorR` for PNA-specific analysis functions, `harmony` for batch correction and data harmonization, `ComplexHeatmap` for advanced heatmap visualizations, and `here` for robust file path management. Small helper functions used in later modules are defined in §0.3 below.

```{r}
#| label: setup_packages

# Load commonly used packages
library(tidyverse)
library(Seurat)
library(pixelatorR)
library(harmony)
library(ComplexHeatmap)
library(ggplotify)
library(here)
library(Matrix)
library(ggraph)

# Check pixelatorR version
if (packageVersion("pixelatorR") < "0.17.1") {
  stop(
    "Please update the pixelatorR package to version 0.17.1 or higher to run this analysis workflow."
  )
}

# Anchor the project root directory
here::i_am("modules/00_project_setup.qmd")

# Set global options for saving plots
options(
  # Overwrite existing files when saving plots - set to FALSE if you want to keep old versions of the plots
  export_plot.overwrite = TRUE
)
# Set global options for file formats to save plots in both PNG and PDF
options(export_plot.file_formats = c("png", "pdf"))

# Minimum p-value to use in -log10 transformations for volcano plots.
# When p_val_adj == 0 (floating-point underflow), -log10(0) becomes Inf and
# data-point labels are clipped. Coerce any p-value below this threshold to
# this value before computing -log10.
min_p_value_threshold <- 1e-300
```

## 0.2. Create the PNA object

### 0.2.1. Read metadata file

> **Decision checkpoint** (sample table)  
> Edit `data/metadata.csv` (paths, `condition` levels, row order for plot order) before running the next chunks. The template reads `metadata_path` below.

To handle file paths and sample-specific data, a metadata file must be provided as a CSV file. The metadata file must contain the following required columns:

- **`sample_id`**: A short sample identifier (e.g., `S1`, `S2`, `S3`)
- **`sample_alias`**: A more descriptive sample name (e.g., `S1_resting`, `S2_PHA`, `S3_IL2`)
- **`file_path`**: The filename of the PXL file located in the `data` folder (e.g., `S1_resting.pxl`)
- **`condition`**: The sample treatment or experimental condition (e.g., `resting`, `stimulated`, `IL2_treated`)

Additional columns can be added as needed (e.g., `donor`, `time_point`, `batch`, `replicate`) and will automatically be incorporated into the PNA object as metadata columns for downstream analysis.

**Example metadata file:**

```csv
sample_id,sample_alias,condition,file_path
S1,S1_resting,resting,PNA064_Sample_1_S1.layout.pxl
S2,S2_PHA,PHA,PNA064_Sample_2_S2.layout.pxl
```

```{r}
#| label: setup_read_metadata

# Read metadata file into a dataframe.
# Supports both comma-separated (,) and semicolon-separated (;) CSV files.
metadata_path <- here::here("data", "metadata.csv")
if (!file.exists(metadata_path)) {
  stop(
    "metadata.csv not found in the data/ folder. ",
    "Please create the file and try again."
  )
}
metadata_raw <- readLines(metadata_path, n = 1)
metadata_sep <- if (grepl(";", metadata_raw)) ";" else ","
metadata <- read.csv(metadata_path, sep = metadata_sep) |> as_tibble()

# Check that the metadata file contains the required columns - otherwise stop the script
if (
  !all(
    c("sample_id", "sample_alias", "file_path", "condition") %in%
      colnames(metadata)
  )
) {
  stop(
    "Metadata file must contain the columns: sample_id, sample_alias, file_path and condition"
  )
}

# Print the metadata
print(metadata)
```

### 0.2.2. Load and merge data

This step loads each PXL file specified in the metadata and merges them into a single combined PNA object. First, each PXL file is read using `ReadPNA_Seurat()`, which creates a Seurat-compatible PNA object for each sample. Proximity scores are not loaded at this stage (`load_proximity_scores = FALSE`) to reduce memory usage, as they can be computed later if needed. The individual PNA objects are then merged using Seurat's `merge()` function, which combines all samples into a single object. During merging, the `sample_id` from the metadata is automatically appended to each cell's identifier to ensure unique cell names across samples. Finally, `JoinLayers()` consolidates any split layers in the merged object, ensuring all data is accessible in a unified structure.

```{r}
#| label: setup_merge_data

# Load the data from the file paths into a list of PNA objects
pg_data <-
  here::here("data", metadata$file_path) |>
  lapply(ReadPNA_Seurat, load_proximity_scores = FALSE)

# Merge the PNA objects, adding the sample_id to the cell names
pg_data <-
  merge(
    pg_data[[1]],
    y = pg_data[-1],
    add.cell.ids = metadata$sample_id
  ) |>
  JoinLayers()
```

### 0.2.3. Add metadata

This step adds sample-level metadata from the metadata file to each cell in the PNA object. The process extracts the existing cell metadata from the PNA object and converts it to a tibble format. The `sample_id` is then extracted from each cell's identifier by removing the last 18 characters. The metadata from the CSV file (excluding the `file_path` column, which is not needed in the cell metadata) is then left-joined to the cell metadata based on matching `sample_id` values. Finally, the enriched metadata is converted back to a data frame with cell identifiers as rownames and assigned back to the PNA object's metadata slot. A sanity check is performed to verify that the metadata was correctly added by displaying cross-tabulations of metadata columns.

```{r}
#| label: setup_add_metadata

# Add metadata to the PNA object
pg_data[[]] <-
  pg_data[[]] |>
  as_tibble(rownames = "cell_id") |>
  mutate(sample_id = str_sub(cell_id, end = -18)) |> # Remove last 18 characters from the cell_id to get the sample_id
  left_join(
    metadata |>
      select(-file_path),
    by = "sample_id"
  ) |> # Do not add the file path to the PNA object
  column_to_rownames("cell_id")


# Set the sample_id, sample_alias and condition as factors with levels defined by the metadata file
# to ensure the correct order of the samples and conditions in the plots
pg_data$sample_id <-
  factor(
    pg_data$sample_id,
    metadata$sample_id
  )
pg_data$sample_alias <-
  factor(
    pg_data$sample_alias,
    metadata$sample_alias
  )
pg_data$condition <-
  factor(
    pg_data$condition,
    unique(metadata$condition)
  )

# Sanity check the metadata
for (col in colnames(metadata |> select(-file_path))) {
  print(col)
  print(table(pg_data[[]]$sample_alias, pg_data[[]][, col]))
}
```

## 0.3. Theme

> **Optional — plot appearance**  
> Adjust `cluster_palette`, `condition_palette`, `sample_palette`, or `gradient_palette` if you want different colors in exported figures. Defaults work out of the box.

Here we specify the color palettes for clusters and conditions. These can be adjusted to the color combinations that you want for your project, by replacing either palette with a named vector containing color codes.  The color palettes are created using the `PixelgenAccentColors()` function from the `pixelatorR` package, which creates a set of 12 colors for each hue. The color palettes are then used to color the clusters and conditions in the plots.

```{r}
#| label: setup_theme

cluster_palette <-
  c(
    pixelatorR::PixelgenAccentColors(
      level = 7,
      hue = c(
        "blues",
        "reds",
        "cyans",
        "oranges",
        "pinks",
        "greens",
        "purples",
        "yellows",
        "greys",
        "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 9,
      hue = c(
        "blues",
        "pinks",
        "reds",
        "yellows",
        "greys",
        "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 5,
      hue = c(
        "purples",
        "blues",
        "cyans",
        "greens",
        "reds",
        "oranges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 3,
      hue = c("blues", "reds")
    ),
    pixelatorR::PixelgenAccentColors(
      level = 12,
      hue = c("blues", "reds")
    )
  ) |>
  unname() |>
  rep(10)

# Create a condition palette for the conditions
unique_conditions <- unique(metadata$condition)
condition_palette <-
  set_names(
    create_discrete_palette(
      unique_conditions
    )[seq_along(unique_conditions)],
    unique_conditions
  )

# Create a sample palette for the samples
unique_samples <- unique(metadata$sample_alias)
sample_palette <-
  set_names(
    create_discrete_palette(
      unique_samples
    )[seq_along(unique_samples)],
    unique_samples
  )

# Use pixelatorRs palette for coloring cell types
cell_palette <-
  Pixelgen_cell_palette

gradient_palette <-
  c(
    "#1F395F",
    "#44628E",
    "#718BB2",
    "#A8B9D1",
    "#DBE1EA",
    "#F1EEE9",
    "#F0D7E0",
    "#E3A6B8",
    "#CB6E8B",
    "#A23F5E",
    "#781534"
  )

# Helper function to format large numbers compactly (e.g., 11234 -> "11.2k")
format_thousands <- function(x) {
  ifelse(
    abs(x) >= 1000,
    paste0(round(x / 1000, 1), "k"),
    as.character(round(x, 1))
  )
}
```
