# 5. Statistical testing

**Module 05 — Statistical testing (optional).** Differential abundance, clustering, and colocalization tests (cell-level Wilcoxon by default; optional limma pseudobulk in §§5.5–5.7). Skip this module if you do not need formal comparisons. Run the first chunk below after module `02` has finished.

```{r}
#| label: module05_setup

library(here)
here::i_am("modules/05_statistical_testing.qmd")

source(here::here("modules", "common_setup.R"))

if (!exists("pg_data", inherits = TRUE)) {
  pg_data <- load_checkpoint("annotated_seurat_object.rds")
}
if (!exists("markers_to_test", inherits = TRUE)) {
  markers_to_test <- load_checkpoint("markers_to_test.rds")
}
```

## 5.0. Set Up Statistical Testing

> **Decision checkpoint** (statistical contrasts)  
> Set `reference_condition` to one of the values in `metadata$condition`. All `{target}_vs_{reference}` outputs in module `05` use this baseline.  
> Optionally set `run_pseudobulk <- TRUE` to also run sample-level limma tests (requires ≥2 `sample_alias` values per `condition`).

By default, in this analysis we compare all conditions to a reference condition, which is typically the "resting" or "control" condition. This allows us to identify markers that are differentially abundant or spatially clustered or colocalized in stimulated or treated conditions compared to the baseline state.

Sections 5.1–5.3 use **cell-level** Wilcoxon tests (cells pooled by `condition`). These are the default outputs consumed by module 07 (under `{modality}/cell_level/`). With many cells they can overstate significance; when you have biological replicates, enable the optional **pseudobulk** limma path (setup in §5.4; tests in §§5.5–5.7). That path aggregates to `sample_alias` × cell type, then tests across samples; outputs go under `{modality}/pseudobulk/`.

```{r}
#| label: statistical_testing_set_reference

# Baseline / control condition. Must match a value in metadata$condition.
# All tests compare each other condition against this one (target − reference).
# Output folders are named {target}_vs_{reference}.
reference_condition <- "donor1"
if (!reference_condition %in% pg_data$condition) stop("The reference condition is not present in the Seurat object")

# If TRUE, run cell-level Wilcoxon tests (sections 5.1–5.3). Set FALSE to skip.
run_cell_level <- TRUE
# If TRUE, run sample-level limma tests (sections 5.5–5.7). Set FALSE to skip.
run_pseudobulk <- FALSE

# Optional continuous-integration overrides. Loaded after the flags above so
# smoke tests can change them (see .github/smoke-test/apply_module05_overrides.R).
smoke_hook <- here::here(".github/smoke-test/apply_module05_overrides.R")
if (file.exists(smoke_hook)) {
  source(smoke_hook)
}

# Optional batch / donor columns from metadata.csv to include in limma models,
# e.g. c("donor", "batch"). Leave as NULL for none. Do not add `condition` here.
pseudobulk_covariate_cols <- NULL
```

Helpers used by the analysis and plotting chunks below, grouped by role. Usually you will not need to modify these.

### Helper Functions

Helpers for creating the correct input and output tables for the analysis and plotting functions.

```{r}
#| label: statistical_testing_result_helpers
#| include: false

# Add columns to dataframes used by the volcano and heatmap functions
prepare_diff_plot_data <- function(df, effect_col, feature_col = NULL) {
  out <- df |>
    mutate(
      neg_log10_p_adj = -log10(pmax(p_adj, min_p_value_threshold)),
      significant = p_adj < 0.05,
      effect_size = !!sym(effect_col)
    )
  if (!is.null(feature_col)) {
    out <- out |> mutate(marker = !!sym(feature_col))
  }
  out
}

# Save the results to the results/05_statistical_testing/{modality}/{level} directory
save_diff_results <- function(df, modality, level, basename) {
  out_dir <- here::here("results/05_statistical_testing", modality, level)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  saveRDS(df, file.path(out_dir, paste0(basename, ".rds")))
  write.csv(df, file.path(out_dir, paste0(basename, ".csv")), row.names = FALSE)
  invisible(out_dir)
}

# Create an empty table that module 07 expects when no results are available; depending on the modality, the table will have different columns.
empty_diff_results <- function(kind = c("abundance", "proximity")) {
  kind <- match.arg(kind)
  shared <- tibble(
    p = numeric(),
    target = character(),
    reference = character(),
    method = character(),
    group = character(),
    p_adj = numeric()
  )
  if (identical(kind, "abundance")) {
    bind_cols(tibble(marker = character(), difference = numeric()), shared)
  } else {
    bind_cols(
      tibble(marker_1 = character(), marker_2 = character(), diff_median = numeric()),
      shared
    )
  }
}

# Ensure this empty dataframe has the expected columns; this is a safety check to ensure that the empty table has the correct columns for the given modality
ensure_diff_result_schema <- function(df, kind, label = "differential results") {
  if (is.data.frame(df) && ncol(df) > 0L) {
    return(df)
  }
  message("No ", label, "; writing an empty table with the expected columns.")
  empty_diff_results(kind)
}

# Helper function for BH p-value adjustment within cell type × contrast (not across the whole table)
adjust_global_fdr <- function(df) {
  if (!is.data.frame(df) || nrow(df) == 0L || !"p" %in% names(df)) {
    return(df)
  }
  df |>
    group_by(across(any_of(c("group", "target", "reference")))) |>
    mutate(p_adj = p.adjust(.data$p, method = "fdr")) |>
    ungroup()
}
```

### Sample-level Aggregation

Per-sample means for volcano side heatmaps and pseudobulk analyses.

#### Abundance

Create the table of per-sample means for the abundance modality here already; we will use it in the volcano and heatmap functions of the single cell level analysis as well as for the pseudobulk level analysis.

```{r}
#| label: statistical_testing_aggregate_abundance
#| include: false

# Mean CLR per sample × cell type × marker. Used for volcano side heatmaps
# (§5.1.1) and limma (§5.5). Saved under cell_level/; §5.5 also writes the
# same object under pseudobulk/ when limma runs.
summary_abundance_data <-
  names(markers_to_test) |>
  map(function(cell_type) {
    dat <- FetchData(
      subset(pg_data, cell_annotation == cell_type),
      vars = c(
        "sample_alias",
        "condition",
        as.character(unname(markers_to_test[[cell_type]]))
      ),
      layer = "data"
    )
    marker_cols <- setdiff(colnames(dat), c("sample_alias", "condition"))
    dat |>
      pivot_longer(all_of(marker_cols), names_to = "marker", values_to = "clr") |>
      group_by(.data$sample_alias, .data$condition, .data$marker) |>
      summarise(mean_clr = mean(clr, na.rm = TRUE), .groups = "drop") |>
      mutate(group = cell_type)
  }) |>
  bind_rows()

save_diff_results(
  summary_abundance_data,
  "abundance",
  "cell_level",
  "summary_abundance_data"
)
```

#### Proximity

Build the per-sample proximity table once here, the same way as abundance above, so the cell-level plots (§§5.2.1, 5.3.1) and the pseudobulk fits (§§5.6–5.7) all show and test the same numbers. `FilterProximityScores()` drops pairs below an isotype-derived background, then `SummarizeProximityScores()` takes the mean `log2_ratio` per `sample_alias` × cell type.

```{r}
#| label: statistical_testing_proximity_helpers
#| include: false

# Mean log2_ratio per sample × pair. Clustering and colocalization both use this table.
# Weak pairs are dropped with FilterProximityScores: median isotype fraction + 2 MAD
# (minimum 0.001), matching the usual pixelatorR spatial-analysis rule.
aggregate_sample_proximity <- function(
  obj,
  markers_vec,
  group_vars = c("sample_alias", "condition")
) {
  background_threshold_pct <- if ("isotype_fraction" %in% colnames(obj[[]])) {
    max(
      median(obj$isotype_fraction, na.rm = TRUE) +
        2 * mad(obj$isotype_fraction, na.rm = TRUE),
      0.001
    )
  } else {
    0.001
  }
  proximity_data <- ProximityScores(
    obj,
    lazy = TRUE,
    meta_data_columns = group_vars,
    add_marker_proportions = TRUE
  ) |>
    filter(marker_1 %in% markers_vec, marker_2 %in% markers_vec) |>
    FilterProximityScores(background_threshold_pct = background_threshold_pct)

  SummarizeProximityScores(
    proximity_data,
    proximity_metric = "log2_ratio",
    group_vars = group_vars,
    summary_stat = "mean",
    include_missing_obs = TRUE
  ) |>
    mutate(feature = paste(marker_1, marker_2, sep = "__"))
}
```

```{r}
#| label: statistical_testing_aggregate_proximity

# One table for clustering and colocalization, at both levels: mean log2_ratio
# per sample × cell type × pair. Self-proximity is the marker_1 == marker_2
# subset. Saved under cell_level/; §5.6 also writes it under pseudobulk/.
summary_proximity_data <-
  names(markers_to_test) |>
  map(function(cell_type) {
    tryCatch(
      aggregate_sample_proximity(
        subset(pg_data, cell_annotation == cell_type),
        markers_to_test[[cell_type]]
      ) |>
        mutate(group = cell_type),
      error = function(e) {
        message(paste("Proximity summary error for", cell_type, ":", e$message))
        NULL
      }
    )
  }) |>
  bind_rows()

if (nrow(summary_proximity_data) > 0) {
  # Clustering and colocalization read the same table, so write it under both
  save_diff_results(summary_proximity_data, "clustering", "cell_level", "summary_proximity_data")
  save_diff_results(summary_proximity_data, "colocalization", "cell_level", "summary_proximity_data")
} else {
  message("No sample-level proximity summaries; proximity side heatmaps will be skipped.")
}
```

### Plotting Functions

Volcanoes, sample heatmaps, and effect heatmaps. These functions are used to create the plots for the differential abundance and clustering analyses.

```{r}
#| label: statistical_testing_plot_helpers
#| include: false

top_features_by_significance <- function(df, n, group_cols = NULL) {
  if (!is.null(group_cols)) {
    df <- df |> group_by(across(all_of(group_cols)))
  }
  df |>
    slice_max(order_by = .data$neg_log10_p_adj, n = n, with_ties = FALSE) |>
    ungroup()
}

sample_zscore_heatmap <- function(mat, sample_info, legend_title = "Z-score") {
  sample_info <- sample_info |>
    mutate(
      sample_alias = as.character(.data$sample_alias),
      condition = as.character(.data$condition)
    ) |>
    arrange(.data$condition, .data$sample_alias)
  mat <- mat[, sample_info$sample_alias, drop = FALSE]
  mat_z <- t(scale(t(mat)))
  mat_z[!is.finite(mat_z)] <- 0
  z_lim <- max(2, max(abs(mat_z), na.rm = TRUE), na.rm = TRUE)
  ann_col <- data.frame(
    condition = factor(sample_info$condition, levels = unique(sample_info$condition)),
    row.names = sample_info$sample_alias
  )
  ComplexHeatmap::pheatmap(
    mat_z,
    color = PixelgenGradient(100, "BluesCherry"),
    breaks = seq(-z_lim, z_lim, length.out = 101),
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    annotation_col = ann_col,
    annotation_colors = list(condition = condition_palette[levels(ann_col$condition)]),
    cellwidth = 12,
    cellheight = max(3.5, min(8, 350 / max(nrow(mat_z), 1))),
    heatmap_legend_param = list(title = legend_title),
    fontsize_row = 6,
    fontsize_col = 7,
    angle_col = "45"
  ) |>
    as.ggplot()
}

export_volcano_with_optional_heatmap <- function(volcano, sample_hm, expr_mat, filename) {
  if (is.null(sample_hm) || is.null(expr_mat)) {
    export_plot(filename = filename, plot = volcano, width = 5.5, height = 4.5)
    return(invisible(NULL))
  }
  export_plot(
    filename = filename,
    plot = patchwork::wrap_plots(volcano, sample_hm, nrow = 1, widths = c(1, 1.2)),
    width = 5.5 + max(4, ncol(expr_mat) * 0.35 + 2.5),
    height = max(4.5, nrow(expr_mat) * 0.12 + 2)
  )
}

# Sample means heatmap for the labeled volcano features. summary_df is
# summary_abundance_data or summary_proximity_data; both levels pass the same
# table so the cell-level and pseudobulk heatmaps show identical values.
build_volcano_side_heatmap <- function(cell_type, top_df, modality, summary_df = NULL) {
  empty <- list(sample_hm = NULL, expr_mat = NULL)
  feature_order <- top_df |>
    arrange(desc(neg_log10_p_adj)) |>
    pull(marker) |>
    unique()
  if (length(feature_order) < 1 || is.null(summary_df) || nrow(summary_df) < 1) {
    return(empty)
  }

  if (identical(modality, "abundance")) {
    agg <- summary_df |>
      filter(group == cell_type, marker %in% feature_order)
    value_col <- "mean_clr"
    legend_title <- "CLR\nZ-score"
    values_fill <- NULL
  } else {
    agg <- summary_df |>
      filter(group == cell_type) |>
      mutate(marker = if (identical(modality, "clustering")) marker_1 else feature)
    agg <- if (identical(modality, "clustering")) {
      agg |> filter(marker_1 == marker_2, marker %in% feature_order)
    } else {
      agg |> filter(marker %in% feature_order)
    }
    value_col <- "mean_log2_ratio"
    legend_title <- "log2_ratio\nZ-score"
    values_fill <- 0
  }

  if (is.null(agg) || nrow(agg) < 1) {
    return(empty)
  }
  sample_info <- agg |>
    distinct(.data$sample_alias, .data$condition) |>
    arrange(.data$condition, .data$sample_alias)
  expr_mat <- agg |>
    mutate(sample_alias = as.character(sample_alias)) |>
    select(all_of(c("marker", "sample_alias", value_col))) |>
    pivot_wider(
      names_from = sample_alias,
      values_from = all_of(value_col),
      values_fill = values_fill,
      values_fn = mean
    ) |>
    column_to_rownames("marker") |>
    as.matrix()
  expr_mat <- expr_mat[, as.character(sample_info$sample_alias), drop = FALSE]
  keep <- intersect(feature_order, rownames(expr_mat))
  if (length(keep) < 1) {
    return(empty)
  }
  expr_mat <- expr_mat[keep, , drop = FALSE]
  list(
    sample_hm = sample_zscore_heatmap(expr_mat, sample_info, legend_title),
    expr_mat = expr_mat
  )
}

diff_volcano <- function(
  plot_data,
  label_data,
  title,
  subtitle,
  x_lab = "Effect size",
  label_size = 2,
  facet = TRUE
) {
  p <- plot_data |>
    ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
    geom_point(size = 0.5) +
    geom_vline(xintercept = 0) +
    geom_vline(xintercept = c(-1, 1), col = "grey") +
    geom_text(
      data = label_data,
      aes(label = marker),
      hjust = 0.5,
      vjust = -0.5,
      size = label_size
    )
  if (isTRUE(facet)) {
    p <- p + facet_wrap(~group, scales = "free")
  }
  p +
    scale_x_continuous(
      limits = range(plot_data$effect_size, na.rm = TRUE),
      expand = expansion(0.1)
    ) +
    scale_y_continuous(expand = expansion(c(0, 0.1))) +
    scale_color_gradientn(
      colors = gradient_palette,
      limits = c(-1, 1) * max(abs(plot_data$effect_size), na.rm = TRUE)
    ) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    labs(
      title = title,
      subtitle = subtitle,
      x = x_lab,
      y = "-log10(adjusted p-value)",
      color = "Effect size"
    )
}

effect_heatmap <- function(mat) {
  max_abs_effect <- max(abs(quantile(as.numeric(mat), c(0.01, 0.99), na.rm = TRUE)))
  mat |>
    pheatmap(
      color = PixelgenGradient(100, "BluesCherry"),
      breaks = seq(-max_abs_effect, max_abs_effect, length.out = 101),
      cellwidth = 8,
      cellheight = 8,
      cluster_rows = TRUE,
      cluster_cols = TRUE,
      heatmap_legend_param = list(title = "Effect size"),
      fontsize = 8
    ) |>
    as.ggplot()
}

# Faceted volcano, per–cell-type volcano + sample heatmap, and an effect heatmap.
# Colocalization also writes a pair-dot heatmap per cell type.
export_diff_plots <- function(
  results_df,
  modality,
  effect_col,
  analysis_label,
  feature_col = "marker",
  level = "cell_level",
  volcano_subtitle = "Effect size vs statistical significance by cell type",
  heatmap_summary_df = NULL,
  fill_lab = "Difference\nin medians",
  n_volcano_labels = 10L
) {
  if (identical(modality, "colocalization")) {
    results_df <- results_df |>
      mutate(marker = paste(marker_1, marker_2, sep = "__"))
    feature_col <- "marker"
  }
  plot_data_base <- prepare_diff_plot_data(results_df, effect_col, feature_col)
  plot_subdir <- paste0("differential_", modality, "_plots")
  is_coloc <- identical(modality, "colocalization")

  for (target_condition in unique(stats::na.omit(plot_data_base$target))) {
    plot_data <- plot_data_base |> filter(target == target_condition)
    if (nrow(plot_data) == 0) {
      next
    }
    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here(
      "results/05_statistical_testing",
      modality,
      level,
      plot_subdir,
      comparison_dir
    )
    contrast_lab <- paste0(target_condition, " vs ", reference_condition)
    export_plot(
      filename = file.path(base_output_dir, paste0("differential_", modality, "_volcano")),
      plot = diff_volcano(
        plot_data,
        top_features_by_significance(plot_data, n_volcano_labels, "group"),
        title = paste0(analysis_label, " (", contrast_lab, ")"),
        subtitle = volcano_subtitle,
        x_lab = paste0("Effect size (", effect_col, ")")
      ),
      width = 12,
      height = 8
    )

    for (cell_type in unique(plot_data$group)) {
      cell_data <- plot_data |> filter(group == cell_type)
      if (nrow(cell_data) < 1) {
        next
      }
      cell_type_dir <- file.path(base_output_dir, cell_type)
      dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)
      top_hits <- top_features_by_significance(cell_data, n_volcano_labels)
      volcano_cell <- diff_volcano(
        cell_data,
        top_hits,
        title = analysis_label,
        subtitle = paste0("Cell type: ", cell_type, " (", contrast_lab, ")"),
        x_lab = paste0("Effect size (", effect_col, ")"),
        label_size = 3,
        facet = FALSE
      )
      hm <- build_volcano_side_heatmap(
        cell_type,
        top_hits,
        modality = modality,
        summary_df = heatmap_summary_df
      )
      export_volcano_with_optional_heatmap(
        volcano_cell,
        hm$sample_hm,
        hm$expr_mat,
        file.path(cell_type_dir, "volcano_plot")
      )

      if (is_coloc) {
        plot_data_hm <- cell_data |>
          filter(p_adj < 0.05) |>
          arrange(desc(neg_log10_p_adj))
        # ColocalizationHeatmap errors with fewer than two rows, so skip
        # a cell type that has 0 or 1 significant pairs instead of aborting
        # export_diff_plots after results have already been written.
        if (nrow(plot_data_hm) < 2) {
          next
        }
        plot_markers <- unique(c(plot_data_hm$marker_1, plot_data_hm$marker_2))
        export_plot(
          filename = file.path(cell_type_dir, "colocalization_heatmap"),
          plot = ColocalizationHeatmap(
            plot_data_hm,
            marker1_col = "marker_1",
            marker2_col = "marker_2",
            value_col = "diff_median",
            type = "dots",
            colors = gradient_palette,
            size_col_transform = \(x) -log10(x)
          ) &
            labs(size = "-log10(p_adj)") &
            scale_size(range = c(1, 5), limits = c(0, max(-log10(plot_data_hm$p_adj)))) &
            labs(
              fill = fill_lab,
              title = analysis_label,
              subtitle = paste0("Cell type: ", cell_type, " (", contrast_lab, ")")
            ),
          width = max(c(4, length(plot_markers) * 0.15 + 1.5)),
          height = max(c(4, length(plot_markers) * 0.15 + 3))
        )
      }
    }

    sig <- if (is_coloc) {
      plot_data |> filter(significant)
    } else {
      plot_data |>
        group_by(.data$marker) |>
        filter(any(significant) & any(abs(effect_size) > 0.5)) |>
        ungroup()
    }
    if (nrow(sig) < 1) {
      next
    }
    if (is_coloc) {
      keep_pairs <- sig |>
        group_by(marker) |>
        summarise(max_neg_log10_p = max(neg_log10_p_adj), .groups = "drop") |>
        slice_max(order_by = max_neg_log10_p, n = 50L, with_ties = FALSE)
      sig <- sig |> filter(marker %in% keep_pairs$marker)
    }
    heatmap_mat <- sig |>
      select(marker, group, effect_size) |>
      pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
      column_to_rownames("marker") |>
      as.matrix()
    if (!is_coloc) {
      heatmap_mat <- t(heatmap_mat)
    }
    if (nrow(heatmap_mat) > 0 && ncol(heatmap_mat) > 0) {
      export_plot(
        filename = file.path(base_output_dir, paste0("differential_", modality, "_heatmap")),
        plot = effect_heatmap(heatmap_mat),
        width = max(6, ncol(heatmap_mat) * 0.15 + 1.5),
        height = max(4, nrow(heatmap_mat) * 0.2 + 1.5)
      )
    }
  }
  invisible(NULL)
}
```


## 5.1. Differential abundance analysis (cell-level)

Differential abundance analysis identifies markers that exhibit statistically significant differences in expression levels between experimental conditions. This analysis is essential for understanding how experimental treatments, disease states, or other conditions affect protein expression patterns within specific cell types.

The analysis is performed separately for each annotated cell type, comparing marker abundance between a target condition (e.g., stimulated, treated) and a reference condition (e.g., resting, control). For each marker within each cell type, a statistical test (typically the Wilcoxon rank-sum test, also known as the Mann-Whitney U test) is used to assess whether the distribution of marker abundance differs significantly between conditions. The Wilcoxon test is a non-parametric test that does not assume normal distribution of the data, making it robust for single-cell data which often exhibits non-normal distributions and high variability.

The test compares marker expression values (typically CLR-normalized values) between cells in the target condition versus cells in the reference condition. For each marker, the analysis reports an effect size (typically the difference in mean or median expression) and a p-value indicating the statistical significance of the difference. Multiple testing correction is applied using FDR within each cell type and contrast (not across all cell types together).

Results are organized by cell type and include both the raw statistical outputs and summary visualizations. Markers with significant differences (typically p_adj < 0.05) and large effect sizes are of particular biological interest, as they represent markers whose expression is substantially altered by the experimental condition. This analysis enables identification of condition-specific markers, understanding of cellular responses to treatments, and discovery of biomarkers associated with different experimental states.

```{r}
#| label: statistical_testing_differential_abundance_analysis
#| eval: !expr isTRUE(run_cell_level)

differential_abundance_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      {
        daa_res <- RunDAA(subset(pg_data, cell_annotation == x),
          contrast_column = "condition",
          reference = reference_condition,
          features = markers_to_test[[x]]
        )
        daa_res$group <- x
        return(daa_res)
      },
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  ensure_diff_result_schema(
    "abundance",
    label = "cell-level abundance results"
  ) |>
  adjust_global_fdr()

save_diff_results(
  differential_abundance_analysis,
  "abundance",
  "cell_level",
  "differential_abundance_analysis"
)
```    

### 5.1.1. Summarizing plots

Here we create summarizing plots for the differential abundance analysis results:

1. **Volcano plot**: Effect size vs −log10 adjusted p-value, faceted by cell type. The top markers labeled on each panel are selected by **adjusted significance** (not largest |effect|).

2. **Per–cell-type volcano + sample heatmap**: For each cell type, the volcano is exported beside a row-z-scored heatmap of sample-mean CLR for the labeled markers, using `summary_abundance_data` (same table as §5.5).

3. **Effect-size heatmap**: Significant markers (p_adj < 0.05 and |effect| > 0.5) across cell types.

```{r}
#| label: statistical_testing_differential_abundance_plots
#| eval: !expr isTRUE(run_cell_level)

# Very small p-values are floored before −log10 so volcano axes stay finite
export_diff_plots(
  differential_abundance_analysis,
  modality = "abundance",
  effect_col = "difference",
  analysis_label = "Differential abundance analysis",
  heatmap_summary_df = summary_abundance_data
)
```

## 5.2. Differential Clustering Analysis (cell-level)

Differential clustering analysis identifies markers that exhibit statistically significant differences in spatial clustering patterns between experimental conditions. In this context, "clustering" refers to the spatial aggregation of a marker with itself—that is, the extent to which a marker tends to form clusters or aggregates within cells rather than being uniformly distributed.

The analysis is performed separately for each cell type, comparing clustering patterns between target and reference conditions. A statistical test (typically the Wilcoxon rank-sum test) is used to assess whether the distribution of clustering scores (log2 ratio of observed versus expected self-proximity) differs significantly between conditions. Multiple testing correction is applied using FDR within each cell type and contrast.

```{r}
#| label: statistical_testing_differential_proximity
#| eval: !expr isTRUE(run_cell_level)

differential_clustering_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      {
        res <- DifferentialProximityAnalysis(
          subset(pg_data, cell_annotation == x),
          contrast_column = "condition",
          reference = reference_condition,
          proximity_metric = "log2_ratio",
          metric_type = "self",
          method = "seurat",
          min_cells_per_group = 10L,
          diff_threshold = 0.01,
          min_pct = 0,
          min_diff_pct = -Inf,
          lazy = TRUE
        ) |>
          filter(
            marker_1 %in% markers_to_test[[x]],
            marker_2 %in% markers_to_test[[x]]
          )
        if (nrow(res) == 0L) {
          return(NULL)
        }
        res$group <- x
        res
      },
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  ensure_diff_result_schema(
    "proximity",
    label = "cell-level clustering results"
  ) |>
  adjust_global_fdr()

save_diff_results(
  differential_clustering_analysis,
  "clustering",
  "cell_level",
  "differential_clustering_analysis"
)
```


### 5.2.1. Summarizing plots

Here we create summarizing plots for the differential clustering analysis results:

1. **Volcano plot**: Effect size vs −log10 adjusted p-value, faceted by cell type. Labeled markers are the top hits by **adjusted significance**.

2. **Per–cell-type volcano + sample heatmap**: Volcano beside a row-z-scored heatmap of sample-mean self-proximity (`log2_ratio`) for the labeled markers.

3. **Effect-size heatmap**: Significant markers across cell types.

```{r}
#| label: statistical_testing_differential_clustering_plots
#| eval: !expr isTRUE(run_cell_level)

if (nrow(differential_clustering_analysis) > 0) {
  export_diff_plots(
    differential_clustering_analysis,
    modality = "clustering",
    effect_col = "diff_median",
    feature_col = "marker_1",
    analysis_label = "Differential clustering analysis",
    heatmap_summary_df = summary_proximity_data
  )
} else {
  message("Skipping cell-level clustering plots: no Wilcoxon results.")
}
```

## 5.3. Differential Colocalization Analysis (cell-level)

Differential colocalization analysis identifies markers that exhibit statistically significant differences in spatial proximity patterns between experimental conditions. The analysis is performed separately for each cell type, comparing **cross-marker** pairs (`metric_type = "co"`) between target and reference conditions — self-proximity is tested in §5.2, not here. A statistical test is used to assess whether the distribution of colocalization scores (log2 ratio of observed versus expected proximity) differs significantly between conditions. Multiple testing correction is applied using FDR within each cell type and contrast.


```{r}
#| label: statistical_testing_differential_colocalization
#| eval: !expr isTRUE(run_cell_level)

differential_colocalization_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      {
        res <- DifferentialProximityAnalysis(
          subset(pg_data, cell_annotation == x),
          contrast_column = "condition",
          reference = reference_condition,
          proximity_metric = "log2_ratio",
          metric_type = "co",
          method = "seurat",
          min_cells_per_group = 10L,
          diff_threshold = 0.01,
          min_pct = 0,
          min_diff_pct = -Inf,
          lazy = TRUE
        ) |>
          filter(
            marker_1 %in% markers_to_test[[x]],
            marker_2 %in% markers_to_test[[x]]
          )
        if (nrow(res) == 0L) {
          return(NULL)
        }
        res$group <- x
        res
      },
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  ensure_diff_result_schema(
    "proximity",
    label = "cell-level colocalization results"
  ) |>
  adjust_global_fdr()

save_diff_results(
  differential_colocalization_analysis,
  "colocalization",
  "cell_level",
  "differential_colocalization_analysis"
)
```

### 5.3.1. Summarizing plots

Here we create summarizing plots for the differential colocalization analysis results:

1. **Volcano plot**: Effect size vs −log10 adjusted p-value for marker pairs. Labeled pairs are the top hits by **adjusted significance**.
2. **Per–cell-type volcano + sample heatmap**: Volcano beside a row-z-scored heatmap of sample-mean pair `log2_ratio` for the labeled pairs.
3. **Colocalization heatmap**: Difference in medians for marker pairs with p_adj < 0.05 (dot size = −log10 adjusted p-value).

```{r}
#| label: statistical_testing_differential_colocalization_plot
#| eval: !expr isTRUE(run_cell_level)

if (nrow(differential_colocalization_analysis) > 0) {
  export_diff_plots(
    differential_colocalization_analysis,
    modality = "colocalization",
    effect_col = "diff_median",
    analysis_label = "Differential Colocalization Analysis",
    heatmap_summary_df = summary_proximity_data
  )
} else {
  message("Skipping cell-level colocalization plots: no Wilcoxon results.")
}
```


## 5.4. Pseudobulk setup

Limma helpers for this section only. Usually you will not need to modify these.

```{r}
#| label: statistical_testing_limma_helpers
#| include: false

# Sample-level means → limma vs reference_condition. Uses metadata and
# pseudobulk_covariate_cols from the setup chunk.
run_pseudobulk_limma <- function(agg, feature_col, value_col, values_fill = NULL) {
  covariate_cols <- setdiff(as.character(pseudobulk_covariate_cols), c("condition", "sample_alias"))
  sample_info <- agg |>
    distinct(.data$sample_alias, .data$condition) |>
    mutate(sample_alias = as.character(.data$sample_alias))
  if (length(covariate_cols) > 0) {
    sample_info <- sample_info |>
      left_join(
        metadata |>
          transmute(
            sample_alias = as.character(.data$sample_alias),
            across(all_of(covariate_cols))
          ) |>
          distinct(sample_alias, .keep_all = TRUE),
        by = "sample_alias"
      ) |>
      drop_na(all_of(covariate_cols))
  }
  sample_info <- sample_info |>
    mutate(
      condition = droplevels(relevel(
        factor(as.character(.data$condition)),
        ref = reference_condition
      ))
    ) |>
    arrange(.data$sample_alias)

  expr_mat <- agg |>
    mutate(sample_alias = as.character(sample_alias)) |>
    filter(.data$sample_alias %in% sample_info$sample_alias) |>
    select(all_of(c(feature_col, "sample_alias", value_col))) |>
    pivot_wider(
      names_from = sample_alias,
      values_from = all_of(value_col),
      values_fill = values_fill,
      values_fn = mean
    ) |>
    column_to_rownames(feature_col) |>
    as.matrix()
  expr_mat <- expr_mat[, sample_info$sample_alias, drop = FALSE]

  rhs <- if (length(covariate_cols) > 0) {
    paste("~ condition +", paste(covariate_cols, collapse = " + "))
  } else {
    "~ condition"
  }
  design <- model.matrix(stats::as.formula(rhs), data = sample_info)
  fit <- eBayes(lmFit(expr_mat, design), trend = TRUE, robust = TRUE)
  map_dfr(
    setdiff(levels(sample_info$condition), reference_condition),
    function(target) {
      topTable(fit, coef = paste0("condition", target), number = Inf, sort.by = "none") |>
        as_tibble(rownames = "feature") |>
        transmute(
          feature,
          difference = .data$logFC,
          p = .data$P.Value,
          target = target,
          reference = reference_condition,
          method = "limma_pseudobulk"
        )
    }
  )
}
```

Run pre-flight checks for the pseudobulk analysis.

```{r}
#| label: statistical_testing_pseudobulk_setup

# Limma needs ≥2 samples per condition. Aggregation itself does not filter.
pseudobulk_rep_counts <- pg_data[[]] |>
  as.data.frame(stringsAsFactors = FALSE) |>
  dplyr::distinct(.data$sample_alias, .data$condition) |>
  dplyr::count(.data$condition, name = "n_samples")
pseudobulk_replicates_ok <- nrow(pseudobulk_rep_counts) > 0 &&
  all(pseudobulk_rep_counts$n_samples >= 2L)
pseudobulk_ok <- isTRUE(run_pseudobulk) && isTRUE(pseudobulk_replicates_ok)

if (isTRUE(run_pseudobulk) && !isTRUE(pseudobulk_replicates_ok)) {
  warning(
    "Skipping limma pseudobulk: need ≥2 sample_alias per condition.\n",
    paste(capture.output(print(pseudobulk_rep_counts)), collapse = "\n"),
    call. = FALSE
  )
} else if (!isTRUE(run_pseudobulk)) {
  message("Skipping limma pseudobulk (run_pseudobulk = FALSE).")
}

if (isTRUE(pseudobulk_ok)) {
  thin_cts <- as_tibble(pg_data[[]], rownames = "component") |>
    filter(.data$cell_annotation %in% names(markers_to_test)) |>
    distinct(.data$cell_annotation, .data$sample_alias, .data$condition) |>
    count(.data$cell_annotation, .data$condition, name = "n_samples") |>
    filter(n_samples < 2)
  if (nrow(thin_cts) > 0) {
    warning(
      "These cell types have <2 samples in a condition; limma may skip them:\n",
      paste(capture.output(print(thin_cts)), collapse = "\n"),
      call. = FALSE
    )
  }
  if (!is.null(pseudobulk_covariate_cols) && length(pseudobulk_covariate_cols) > 0) {
    missing_cols <- setdiff(as.character(pseudobulk_covariate_cols), colnames(metadata))
    if (length(missing_cols) > 0) {
      stop(
        "Missing covariate columns in metadata.csv: ",
        paste(missing_cols, collapse = ", ")
      )
    }
  }
  suppressPackageStartupMessages(library(limma))
}
```

## 5.5. Differential abundance analysis (pseudobulk-level)

> **Decision checkpoint** (enable above: `run_pseudobulk`)  
> Limma is fit on **`summary_abundance_data`** (mean CLR per `sample_alias` × cell type × marker, built in the aggregation chunk). Needs ≥2 samples per `condition`. The same table is also written under `pseudobulk/`; limma results as `differential_abundance_analysis.rds`. In module 07, set `stats_method` to `"pseudobulk"` to feed these tables into enrichment (default `"cell_level"` uses §5.1 Wilcoxon results).

```{r}
#| label: statistical_testing_pseudobulk_abundance

if (isTRUE(pseudobulk_ok)) {
  # Same sample-mean CLR table as §5.1.1 heatmaps; write it under pseudobulk/
  # so module 07 / plots can find it next to the limma results.
  save_diff_results(
    summary_abundance_data,
    "abundance",
    "pseudobulk",
    "summary_abundance_data"
  )

  differential_abundance_analysis_pseudobulk <-
    names(markers_to_test) |>
    map(function(cell_type) {
      tryCatch(
        {
          res <- run_pseudobulk_limma(
            summary_abundance_data |> filter(group == cell_type),
            "marker",
            "mean_clr"
          )
          if (nrow(res) == 0) {
            NULL
          } else {
            res |>
              rename(marker = feature) |>
              mutate(group = cell_type)
          }
        },
        error = function(e) {
          message(paste("Pseudobulk abundance error for", cell_type, ":", e$message))
          NULL
        }
      )
    }) |>
    bind_rows() |>
    ensure_diff_result_schema(
      "abundance",
      label = "pseudobulk abundance results"
    ) |>
    adjust_global_fdr()

  save_diff_results(
    differential_abundance_analysis_pseudobulk,
    "abundance",
    "pseudobulk",
    "differential_abundance_analysis"
  )
}
```

### 5.5.1. Pseudobulk abundance plots

```{r}
#| label: statistical_testing_pseudobulk_abundance_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_abundance_analysis_pseudobulk") &&
    nrow(differential_abundance_analysis_pseudobulk) > 0
) {
  export_diff_plots(
    differential_abundance_analysis_pseudobulk,
    modality = "abundance",
    effect_col = "difference",
    analysis_label = "Pseudobulk differential abundance",
    level = "pseudobulk",
    volcano_subtitle = "limma on mean CLR per sample × cell type",
    heatmap_summary_df = summary_abundance_data
  )
}
```

## 5.6. Differential clustering analysis (pseudobulk-level)

Self-proximity (`marker_1 == marker_2`) from `summary_proximity_data` (built with the aggregation helpers above, shared with §5.2), then limma. Column `diff_median` stores the limma logFC (mean difference) for schema compatibility with §5.2 / module 07.

```{r}
#| label: statistical_testing_save_pseudobulk_proximity

if (isTRUE(pseudobulk_ok) && nrow(summary_proximity_data) > 0) {
  # Same table as the cell-level sections; write it next to the limma results
  save_diff_results(summary_proximity_data, "clustering", "pseudobulk", "summary_proximity_data")
  save_diff_results(summary_proximity_data, "colocalization", "pseudobulk", "summary_proximity_data")
}
```

```{r}
#| label: statistical_testing_pseudobulk_clustering

if (isTRUE(pseudobulk_ok)) {
  # Abundance always writes a schema-complete RDS when the replicate gate
  # passes. Do the same here if the shared proximity summary is empty.
  differential_clustering_analysis_pseudobulk <-
    (
      if (nrow(summary_proximity_data) > 0) {
        names(markers_to_test) |>
          map(function(cell_type) {
            tryCatch(
              {
                cell_agg <- summary_proximity_data |>
                  filter(group == cell_type, marker_1 == marker_2)
                run_pseudobulk_limma(cell_agg, "feature", "mean_log2_ratio", values_fill = 0) |>
                  left_join(distinct(cell_agg, feature, marker_1, marker_2), by = "feature") |>
                  transmute(
                    marker_1,
                    marker_2,
                    diff_median = difference,
                    p,
                    target,
                    reference,
                    method,
                    group = cell_type
                  )
              },
              error = function(e) {
                message(paste("Pseudobulk clustering error for", cell_type, ":", e$message))
                NULL
              }
            )
          }) |>
          bind_rows()
      } else {
        tibble()
      }
    ) |>
    ensure_diff_result_schema(
      "proximity",
      label = "pseudobulk clustering results"
    ) |>
    adjust_global_fdr()

  save_diff_results(
    differential_clustering_analysis_pseudobulk,
    "clustering",
    "pseudobulk",
    "differential_clustering_analysis"
  )
}
```

### 5.6.1. Pseudobulk clustering plots

```{r}
#| label: statistical_testing_pseudobulk_clustering_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_clustering_analysis_pseudobulk") &&
    nrow(differential_clustering_analysis_pseudobulk) > 0
) {
  export_diff_plots(
    differential_clustering_analysis_pseudobulk,
    modality = "clustering",
    effect_col = "diff_median",
    feature_col = "marker_1",
    analysis_label = "Pseudobulk differential clustering",
    level = "pseudobulk",
    volcano_subtitle = "limma on mean log2_ratio per sample × cell type",
    heatmap_summary_df = summary_proximity_data
  )
}
```

## 5.7. Differential colocalization analysis (pseudobulk-level)

Off-diagonal pairs (`marker_1 != marker_2`) from the same sample-level table as §5.6, then limma. Same `diff_median` naming as §5.3 for schema compatibility. In module 07, set `stats_method` to `"pseudobulk"` to use these results (and the matching abundance tables) for enrichment.

```{r}
#| label: statistical_testing_pseudobulk_colocalization

if (isTRUE(pseudobulk_ok)) {
  # Same empty-summary contract as clustering / abundance: always write the RDS.
  differential_colocalization_analysis_pseudobulk <-
    (
      if (nrow(summary_proximity_data) > 0) {
        names(markers_to_test) |>
          map(function(cell_type) {
            tryCatch(
              {
                cell_agg <- summary_proximity_data |>
                  filter(group == cell_type, marker_1 != marker_2)
                run_pseudobulk_limma(cell_agg, "feature", "mean_log2_ratio", values_fill = 0) |>
                  left_join(distinct(cell_agg, feature, marker_1, marker_2), by = "feature") |>
                  transmute(
                    marker_1,
                    marker_2,
                    diff_median = difference,
                    p,
                    target,
                    reference,
                    method,
                    group = cell_type
                  )
              },
              error = function(e) {
                message(paste("Pseudobulk coloc error for", cell_type, ":", e$message))
                NULL
              }
            )
          }) |>
          bind_rows()
      } else {
        tibble()
      }
    ) |>
    ensure_diff_result_schema(
      "proximity",
      label = "pseudobulk coloc results"
    ) |>
    adjust_global_fdr()

  save_diff_results(
    differential_colocalization_analysis_pseudobulk,
    "colocalization",
    "pseudobulk",
    "differential_colocalization_analysis"
  )
}
```

### 5.7.1. Pseudobulk colocalization plots

```{r}
#| label: statistical_testing_pseudobulk_colocalization_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_colocalization_analysis_pseudobulk") &&
    nrow(differential_colocalization_analysis_pseudobulk) > 0
) {
  export_diff_plots(
    differential_colocalization_analysis_pseudobulk,
    modality = "colocalization",
    effect_col = "diff_median",
    analysis_label = "Pseudobulk differential colocalization",
    level = "pseudobulk",
    volcano_subtitle = "limma on mean log2_ratio per sample (pairs)",
    fill_lab = "Difference\n(mean)",
    heatmap_summary_df = summary_proximity_data
  )
}
```
