# 5. Statistical testing

**Module 05 — Statistical testing (optional).** Differential abundance, clustering, and colocalization tests (cell-level Wilcoxon by default; optional limma pseudobulk in §§5.5–5.7). Skip this module if you do not need formal comparisons. Run the first chunk below after module `02` has finished.

```{r}
#| label: module05_setup

library(here)
here::i_am("modules/05_statistical_testing.qmd")

source(here::here("modules", "common_setup.R"))

if (!exists("pg_data", inherits = TRUE)) {
  pg_data <- load_checkpoint("annotated_seurat_object.rds")
}
if (!exists("markers_to_test", inherits = TRUE)) {
  markers_to_test <- load_checkpoint("markers_to_test.rds")
}
```

## 5.0. Set reference sample

> **Decision checkpoint** (statistical contrasts)  
> Set `reference_condition` to one of the values in `metadata$condition`. All `{target}_vs_{reference}` outputs in module `05` use this baseline.  
> Optionally set `run_pseudobulk <- TRUE` to also run sample-level limma tests (requires ≥2 `sample_alias` values per `condition`).

By default, in this analysis we compare all conditions to a reference condition, which is typically the "resting" or "control" condition. This allows us to identify markers that are differentially abundant or spatially clustered or colocalized in stimulated or treated conditions compared to the baseline state.

Sections 5.1–5.3 use **cell-level** Wilcoxon tests (cells pooled by `condition`). These are the default outputs consumed by module 07 (under `{modality}/cell_level/`). With many cells they can overstate significance; when you have biological replicates, enable the optional **pseudobulk** limma path (setup in §5.4; tests in §§5.5–5.7). That path aggregates to `sample_alias` × cell type, then tests across samples; outputs go under `{modality}/pseudobulk/`.

```{r}
#| label: statistical_testing_set_reference

reference_condition <- "resting"

# Optional sample-level limma path (sections 5.5–5.7). Requires ≥2 samples
# (sample_alias) per condition. Default FALSE so the bundled example still renders.
run_pseudobulk <- FALSE

# Cell-level Wilcoxon path (sections 5.1–5.3). Set FALSE to skip those chunks.
run_cell_level <- TRUE

# Optional CI smoke overrides (see .github/smoke-test/apply_module05_overrides.R)
smoke_hook <- here::here(".github/smoke-test/apply_module05_overrides.R")
if (file.exists(smoke_hook)) {
  source(smoke_hook)
}

# Minimum cells per sample × cell type before including that pseudobulk observation
pseudobulk_min_cells <- 5L

# Optional limma covariates (column names from `metadata.csv`).
# These are added to the pseudobulk design matrix as additional fixed effects.
# Example: c("donor", "batch")
pseudobulk_covariate_cols <- NULL

if (!reference_condition %in% metadata$condition) stop("The reference condition is not present in the sample metadata")
```


## 5.1. Differential abundance analysis (cell-level)

Differential abundance analysis identifies markers that exhibit statistically significant differences in expression levels between experimental conditions. This analysis is essential for understanding how experimental treatments, disease states, or other conditions affect protein expression patterns within specific cell types.

The analysis is performed separately for each annotated cell type, comparing marker abundance between a target condition (e.g., stimulated, treated) and a reference condition (e.g., resting, control). For each marker within each cell type, a statistical test (typically the Wilcoxon rank-sum test, also known as the Mann-Whitney U test) is used to assess whether the distribution of marker abundance differs significantly between conditions. The Wilcoxon test is a non-parametric test that does not assume normal distribution of the data, making it robust for single-cell data which often exhibits non-normal distributions and high variability.

The test compares marker expression values (typically CLR-normalized values) between cells in the target condition versus cells in the reference condition. For each marker, the analysis reports an effect size (typically the difference in mean or median expression) and a p-value indicating the statistical significance of the difference. Multiple testing correction is applied using the false discovery rate (FDR) method to control for the large number of statistical tests performed across multiple markers and cell types.

Results are organized by cell type and include both the raw statistical outputs and summary visualizations. Markers with significant differences (typically p_adj < 0.05) and large effect sizes are of particular biological interest, as they represent markers whose expression is substantially altered by the experimental condition. This analysis enables identification of condition-specific markers, understanding of cellular responses to treatments, and discovery of biomarkers associated with different experimental states.

```{r}
#| label: statistical_testing_differential_abundance_analysis
#| eval: !expr isTRUE(run_cell_level)

differential_abundance_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      {
        daa_res <- RunDAA(subset(pg_data, cell_annotation == x),
          contrast_column = "condition",
          reference = reference_condition,
          features = markers_to_test[[x]]
        )
        daa_res$group <- x
        return(daa_res)
      },
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  mutate(p_adj = p.adjust(p, method = "fdr")) # recalculate the adjusted p values after multiple testing correction for all comparisons, not only per cell type

dir.create(here::here("results/05_statistical_testing/abundance/cell_level"), recursive = TRUE, showWarnings = FALSE)

# Save the differential abundance analysis results to a file
saveRDS(differential_abundance_analysis, here::here("results/05_statistical_testing/abundance/cell_level/differential_abundance_analysis.rds"))

# Save the differential abundance analysis results to a csv file
write.csv(differential_abundance_analysis, here::here("results/05_statistical_testing/abundance/cell_level/differential_abundance_analysis.csv"), row.names = FALSE)
```    

### 5.1.1. Summarizing plots

Here we create two types of summarizing plots for the differential abundance analysis results:

1. **Volcano plot**: Shows the relationship between effect size (log fold change) and statistical significance (-log10 adjusted p-value) for each marker, faceted by cell type. This helps identify markers that are both significantly different and have large effect sizes.

2. **Heatmap**: Displays significant markers (adjusted p-value < 0.05) across all cell types, with colors representing the effect size. This provides an overview of which markers are differentially abundant in which cell types.

```{r}
#| label: statistical_testing_differential_abundance_plots
#| eval: !expr isTRUE(run_cell_level)

# Determine the effect size column name
effect_size_col <- "difference"

# Determine the feature/marker column name
feature_col <- "marker"

# Determine target conditions - check if dataframe has target/contrast columns
if ("target" %in% colnames(differential_abundance_analysis)) {
  # If target column exists, get all unique targets
  target_conditions <- unique(differential_abundance_analysis$target)
  target_conditions <- target_conditions[!is.na(target_conditions)]
} else {
  stop("Could not find target column. Available columns: ", paste(colnames(differential_abundance_analysis), collapse = ", "))
}

# Prepare base data for plotting (add target column if it doesn't exist)
# p_adj values of zero (floating-point underflow) are coerced to
# min_p_value_threshold before the -log10 transform to prevent Inf values
# that would clip data-point labels in volcano plots.
plot_data_base <- differential_abundance_analysis |>
  mutate(
    neg_log10_p_adj = -log10(pmax(p_adj, min_p_value_threshold)),
    significant = p_adj < 0.05,
    effect_size = !!sym(effect_size_col),
    marker = !!sym(feature_col)
  )


# Loop through each target condition and create plots
for (target_condition in target_conditions) {
  # Filter data for this target (if target column exists)
  if ("target" %in% colnames(plot_data_base)) {
    plot_data <- plot_data_base |>
      filter(target == target_condition)
  }

  # Skip if no data for this target
  if (nrow(plot_data) == 0) {
    warning(paste("No data found for target condition:", target_condition))
    next
  }

  # Create directory name based on target and reference
  comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
  base_output_dir <- here::here("results/05_statistical_testing/abundance/cell_level/differential_abundance_plots", comparison_dir)

  # Plot 1: Volcano plot showing effect size vs significance, faceted by cell type
  # This plot helps identify markers that are both significantly different and have large effect sizes
  # Identify top 5 most extreme effect sizes per cell type for labeling
  top_markers_faceted <- plot_data |>
    group_by(group) |>
    slice_max(order_by = abs(effect_size), n = 5) |>
    ungroup()

  volcano_plot_differential_abundance <-
    plot_data |>
    ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
    geom_point(size = 0.5) +
    geom_vline(xintercept = 0) +
    geom_vline(xintercept = c(-1, 1), col = "grey") +
    geom_text(
      data = top_markers_faceted,
      aes(label = marker),
      hjust = 0.5,
      vjust = -0.5,
      size = 2
    ) +
    facet_wrap(~group, scales = "free") +
    scale_x_continuous(
      limits = c(min(plot_data$effect_size, na.rm = TRUE), max(plot_data$effect_size, na.rm = TRUE)),
      expand = expansion(0.1)
    ) +
    scale_y_continuous(
      expand = expansion(c(0, 0.1))
    ) +
    scale_color_gradientn(
      colors = gradient_palette,
      limits = c(-1, 1) * max(abs(plot_data$effect_size), na.rm = TRUE)
    ) +
    theme_bw() +
    theme(
      panel.grid = element_blank()
    ) +
    labs(
      title = paste("Differential abundance analysis: Volcano plot (", target_condition, " vs ", reference_condition, ")", sep = ""),
      subtitle = "Effect size vs statistical significance by cell type",
      x = paste("Effect size (", effect_size_col, ")", sep = ""),
      y = "-log10(adjusted p-value)",
      color = "Effect size"
    )

  # Save faceted volcano plot in the comparison directory
  export_plot(
    filename = file.path(base_output_dir, "differential_abundance_volcano"),
    plot = volcano_plot_differential_abundance,
    width = 12,
    height = 8
  )

  # Plot 2: Non-faceted volcano plots per cell type
  # Create individual folders per cell type and save non-faceted volcano plots for each cell type
  # Get unique cell types (groups)
  unique_cell_types <- unique(plot_data$group)

  # For each cell type, create a non-faceted volcano plot
  for (cell_type in unique_cell_types) {
    # Create directory for this cell type
    cell_type_dir <- file.path(base_output_dir, cell_type)
    dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

    # Filter data for this cell type
    cell_type_plot_data <- plot_data |>
      filter(group == cell_type)

    # Create non-faceted volcano plot for this cell type
    # Identify top 5 most extreme effect sizes for labeling
    top_markers_cell_type <- cell_type_plot_data |>
      slice_max(order_by = abs(effect_size), n = 5)

    volcano_plot_cell_type <-
      cell_type_plot_data |>
      ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
      geom_point(size = 0.5) +
      geom_vline(xintercept = 0) +
      geom_vline(xintercept = c(-1, 1), col = "grey") +
      geom_text(
        data = top_markers_cell_type,
        aes(label = marker),
        hjust = 0.5,
        vjust = -0.5,
        size = 3
      ) +
      facet_wrap(~group, scales = "free") +
      scale_x_continuous(
        limits = c(
          min(cell_type_plot_data$effect_size, na.rm = TRUE),
          max(cell_type_plot_data$effect_size, na.rm = TRUE)
        ),
        expand = expansion(0.1)
      ) +
      scale_y_continuous(
        expand = expansion(c(0, 0.1))
      ) +
      scale_color_gradientn(
        colors = gradient_palette,
        limits = c(-1, 1) * max(abs(cell_type_plot_data$effect_size), na.rm = TRUE)
      ) +
      theme_bw() +
      theme(
        panel.grid = element_blank()
      ) +
      labs(
        title = paste("Differential abundance analysis: Volcano plot", sep = ""),
        subtitle = paste("Cell type: ", cell_type, " (", target_condition, " vs ", reference_condition, ")", sep = ""),
        x = paste("Effect size (", effect_size_col, ")", sep = ""),
        y = "-log10(adjusted p-value)",
        color = "Effect size"
      )

    # Save non-faceted volcano plot
    export_plot(
      filename = file.path(cell_type_dir, "volcano_plot"),
      plot = volcano_plot_cell_type,
      width = 5.5,
      height = 4.5
    )
  }

  # Plot 3: Heatmap showing significant markers across cell types
  # This provides an overview of which markers are differentially abundant in which cell types
  # Filter for significant markers and prepare data for heatmap
  significant_markers <-
    plot_data |>
    group_by(marker) |>
    filter(any(significant) & any(abs(effect_size) > 0.5)) |>
    select(marker, group, effect_size) |>
    # Pivot to wide format for heatmap
    pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
    column_to_rownames("marker") |>
    as.matrix()

  # Only create heatmap if there are significant markers
  if (nrow(significant_markers) > 0 && ncol(significant_markers) > 0) {
    # Determine color scale breaks based on robust data range (use 1st and 99th percentiles to reduce outlier influence)
    effect_vals <- as.numeric(significant_markers)
    effect_quantiles <- quantile(effect_vals, probs = c(0.01, 0.99), na.rm = TRUE)
    max_abs_effect <- max(abs(effect_quantiles))
    breaks_range <- seq(-max_abs_effect, max_abs_effect, length.out = 101)

    heatmap_differential_abundance <-
      significant_markers |>
      t() |>
      pheatmap(
        color = PixelgenGradient(100, "BluesCherry"),
        breaks = breaks_range,
        cellwidth = 8,
        cellheight = 8,
        cluster_rows = TRUE,
        cluster_cols = TRUE,
        heatmap_legend_param = list(title = "Effect size"),
        fontsize = 8
      ) |>
      as.ggplot()

    export_plot(
      filename = file.path(base_output_dir, "differential_abundance_heatmap"),
      plot = heatmap_differential_abundance,

      # ~0.3 in per marker column + 4 in base margin; ~0.5 in per cell-type row + 2 in base margin.
      width = max(6, nrow(significant_markers) * 0.15 + 1.5),
      height = max(4, ncol(significant_markers) * 0.2 + 1.5)
    )
  } else {
    warning(paste("No significant markers found for heatmap visualization (p_adj < 0.05) for", target_condition, "vs", reference_condition))
  }
}
```

## 5.2. Differential Clustering Analysis (cell-level)

Differential clustering analysis identifies markers that exhibit statistically significant differences in spatial clustering patterns between experimental conditions. In this context, "clustering" refers to the spatial aggregation of a marker with itself—that is, the extent to which a marker tends to form clusters or aggregates within cells rather than being uniformly distributed.

The analysis is performed separately for each cell type, comparing clustering patterns between target and reference conditions. A statistical test (typically the Wilcoxon rank-sum test) is used to assess whether the distribution of clustering scores (log2 ratio of observed versus expected self-proximity) differs significantly between conditions. Multiple testing correction is applied using FDR to control for the number of statistical tests performed across markers and cell types.

```{r}
#| label: statistical_testing_differential_proximity
#| eval: !expr isTRUE(run_cell_level)

differential_clustering_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      {
        # Store markers in a variable to ensure proper evaluation in lazy query context
        # Convert to character vector to ensure proper SQL translation
        markers_vec <- as.character(markers_to_test[[x]])
        proximity_data <- ProximityScores(subset(pg_data, cell_annotation == x),
          lazy = TRUE,
          meta_data_columns = c("sample_alias", "condition")
        ) |>
          filter(marker_1 == marker_2) |>
          filter(marker_1 %in% markers_vec) |>
          filter(marker_2 %in% markers_vec) |>
          collect()

        res <- DifferentialProximityAnalysis(proximity_data,
          contrast_column = "condition",
          reference = reference_condition,
          proximity_metric = "log2_ratio",
          min_n_obs = 5
        )
        res$group <- x
        return(res)
      },
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  mutate(p_adj = p.adjust(p, method = "fdr")) # recalculate the adjusted p values after multiple testing correction for all comparisons, not only per cell type

dir.create(here::here("results/05_statistical_testing/clustering/cell_level"), recursive = TRUE, showWarnings = FALSE)

# Save the differential abundance analysis results to a file
saveRDS(differential_clustering_analysis, here::here("results/05_statistical_testing/clustering/cell_level/differential_clustering_analysis.rds"))

# Save the differential abundance analysis results to a csv file
write.csv(differential_clustering_analysis, here::here("results/05_statistical_testing/clustering/cell_level/differential_clustering_analysis.csv"), row.names = FALSE)
```


### 5.2.1. Summarizing plots

Here we create two types of summarizing plots for the differential clustering analysis results:

1. **Volcano plot**: Shows the relationship between effect size (log fold change) and statistical significance (-log10 adjusted p-value) for each marker, faceted by cell type. This helps identify markers that are both significantly different and have large effect sizes.

2. **Heatmap**: Displays significant markers (adjusted p-value < 0.05) across all cell types, with colors representing the effect size. This provides an overview of which markers are differentially clustered in which cell types.

```{r}
#| label: statistical_testing_differential_clustering_plots
#| eval: !expr isTRUE(run_cell_level)

# Determine the effect size column name
effect_size_col <- "diff_median"

# Determine the feature/marker column name
feature_col <- "marker_1"

# Determine target conditions - check if dataframe has target/contrast columns
if ("target" %in% colnames(differential_clustering_analysis)) {
  # If target column exists, get all unique targets
  target_conditions <- unique(differential_clustering_analysis$target)
  target_conditions <- target_conditions[!is.na(target_conditions)]
} else {
  stop("Could not find target column. Available columns: ", paste(colnames(differential_clustering_analysis), collapse = ", "))
}

# Prepare base data for plotting (add target column if it doesn't exist)
# p_adj values of zero are coerced to min_p_value_threshold before the
# -log10 transform to prevent Inf values that would clip data-point labels.
plot_data_base <- differential_clustering_analysis |>
  mutate(
    neg_log10_p_adj = -log10(pmax(p_adj, min_p_value_threshold)),
    significant = p_adj < 0.05,
    effect_size = !!sym(effect_size_col),
    marker = !!sym(feature_col)
  )


# Loop through each target condition and create plots
for (target_condition in target_conditions) {
  # Filter data for this target (if target column exists)
  if ("target" %in% colnames(plot_data_base)) {
    plot_data <- plot_data_base |>
      filter(target == target_condition)
  }

  # Skip if no data for this target
  if (nrow(plot_data) == 0) {
    warning(paste("No data found for target condition:", target_condition))
    next
  }

  # Create directory name based on target and reference
  comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
  base_output_dir <- here::here("results/05_statistical_testing/clustering/cell_level/differential_clustering_plots", comparison_dir)

  # Plot 1: Volcano plot showing effect size vs significance, faceted by cell type
  # This plot helps identify markers that are both significantly different and have large effect sizes
  # Identify top 5 most extreme effect sizes per cell type for labeling
  top_markers_faceted <- plot_data |>
    group_by(group) |>
    slice_max(order_by = abs(effect_size), n = 5) |>
    ungroup()

  volcano_plot_differential_clustering <-
    plot_data |>
    ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
    geom_point(size = 0.5) +
    geom_vline(xintercept = 0) +
    geom_vline(xintercept = c(-1, 1), col = "grey") +
    geom_text(
      data = top_markers_faceted,
      aes(label = marker),
      hjust = 0.5,
      vjust = -0.5,
      size = 2
    ) +
    facet_wrap(~group, scales = "free") +
    scale_x_continuous(
      limits = c(min(plot_data$effect_size, na.rm = TRUE), max(plot_data$effect_size, na.rm = TRUE)),
      expand = expansion(0.1)
    ) +
    scale_y_continuous(
      expand = expansion(c(0, 0.1))
    ) +
    scale_color_gradientn(
      colors = gradient_palette,
      limits = c(-1, 1) * max(abs(plot_data$effect_size), na.rm = TRUE)
    ) +
    theme_bw() +
    theme(
      panel.grid = element_blank()
    ) +
    labs(
      title = paste("Differential clustering analysis: Volcano plot (", target_condition, " vs ", reference_condition, ")", sep = ""),
      subtitle = "Effect size vs statistical significance by cell type",
      x = paste("Effect size (", effect_size_col, ")", sep = ""),
      y = "-log10(adjusted p-value)",
      color = "Effect size"
    )

  # Save faceted volcano plot in the comparison directory
  export_plot(
    filename = file.path(base_output_dir, "differential_clustering_volcano"),
    plot = volcano_plot_differential_clustering,
    width = 12,
    height = 8
  )

  # Plot 2: Non-faceted volcano plots per cell type
  # Create individual folders per cell type and save non-faceted volcano plots for each cell type
  # Get unique cell types (groups)
  unique_cell_types <- unique(plot_data$group)

  # For each cell type, create a non-faceted volcano plot
  for (cell_type in unique_cell_types) {
    # Create directory for this cell type
    cell_type_dir <- file.path(base_output_dir, cell_type)
    dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

    # Filter data for this cell type
    cell_type_plot_data <- plot_data |>
      filter(group == cell_type)

    # Create non-faceted volcano plot for this cell type
    # Identify top 5 most extreme effect sizes for labeling
    top_markers_cell_type <- cell_type_plot_data |>
      slice_max(order_by = abs(effect_size), n = 5)

    volcano_plot_cell_type <-
      cell_type_plot_data |>
      ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
      geom_point(size = 0.5) +
      geom_vline(xintercept = 0) +
      geom_vline(xintercept = c(-1, 1), col = "grey") +
      geom_text(
        data = top_markers_cell_type,
        aes(label = marker),
        hjust = 0.5,
        vjust = -0.5,
        size = 3
      ) +
      facet_wrap(~group, scales = "free") +
      scale_x_continuous(
        limits = c(
          min(cell_type_plot_data$effect_size, na.rm = TRUE),
          max(cell_type_plot_data$effect_size, na.rm = TRUE)
        ),
        expand = expansion(0.1)
      ) +
      scale_y_continuous(
        expand = expansion(c(0, 0.1))
      ) +
      scale_color_gradientn(
        colors = gradient_palette,
        limits = c(-1, 1) * max(abs(cell_type_plot_data$effect_size), na.rm = TRUE)
      ) +
      theme_bw() +
      theme(
        panel.grid = element_blank()
      ) +
      labs(
        title = paste("Differential clustering analysis: Volcano plot", sep = ""),
        subtitle = paste("Cell type: ", cell_type, " (", target_condition, " vs ", reference_condition, ")", sep = ""),
        x = paste("Effect size (", effect_size_col, ")", sep = ""),
        y = "-log10(adjusted p-value)",
        color = "Effect size"
      )

    # Save non-faceted volcano plot
    export_plot(
      filename = file.path(cell_type_dir, "volcano_plot"),
      plot = volcano_plot_cell_type,
      width = 5.5,
      height = 4.5
    )
  }

  # Plot 3: Heatmap showing significant markers across cell types
  # This provides an overview of which markers are differentially clustered in which cell types
  # Filter for significant markers and prepare data for heatmap
  significant_markers <-
    plot_data |>
    group_by(marker) |>
    filter(any(significant) & any(abs(effect_size) > 0.5)) |>
    select(marker, group, effect_size) |>
    # Pivot to wide format for heatmap
    pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
    column_to_rownames("marker") |>
    as.matrix()

  # Only create heatmap if there are significant markers
  if (nrow(significant_markers) > 0 && ncol(significant_markers) > 0) {
    # Determine color scale breaks based on robust data range (use 1st and 99th percentiles to reduce outlier influence)
    effect_vals <- as.numeric(significant_markers)
    effect_quantiles <- quantile(effect_vals, probs = c(0.01, 0.99), na.rm = TRUE)
    max_abs_effect <- max(abs(effect_quantiles))
    breaks_range <- seq(-max_abs_effect, max_abs_effect, length.out = 101)

    heatmap_differential_clustering <-
      significant_markers |>
      t() |>
      pheatmap(
        color = PixelgenGradient(100, "BluesCherry"),
        breaks = breaks_range,
        cellwidth = 8,
        cellheight = 8,
        cluster_rows = TRUE,
        cluster_cols = TRUE,
        heatmap_legend_param = list(title = "Effect size"),
        fontsize = 8
      ) |>
      as.ggplot()

    export_plot(
      filename = file.path(base_output_dir, "differential_clustering_heatmap"),
      plot = heatmap_differential_clustering,

      # ~0.3 in per marker column + 4 in base margin; ~0.5 in per cell-type row + 2 in base margin.
      width = max(6, nrow(significant_markers) * 0.15 + 1.5),
      height = max(4, ncol(significant_markers) * 0.2 + 1.5)
    )
  } else {
    warning(paste("No significant markers found for heatmap visualization (p_adj < 0.05) for", target_condition, "vs", reference_condition))
  }
}
```

## 5.3. Differential Colocalization Analysis (cell-level)

Differential colocalization analysis identifies markers that exhibit statistically significant differences in spatial proximity patterns between experimental conditions. The analysis is performed separately for each cell type, comparing colocalization patterns between target and reference conditions. A statistical test is used to assess whether the distribution of colocalization scores (log2 ratio of observed versus expected proximity) differs significantly between conditions. Multiple testing correction is applied using FDR to control for the number of statistical tests performed across markers and cell types.


```{r}
#| label: statistical_testing_differential_colocalization
#| eval: !expr isTRUE(run_cell_level)

differential_colocalization_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      {
        # Store markers in a variable to ensure proper evaluation in lazy query context
        # Convert to character vector to ensure proper SQL translation
        markers_vec <- as.character(markers_to_test[[x]])
        proximity_data <- ProximityScores(subset(pg_data, cell_annotation == x),
          lazy = TRUE,
          meta_data_columns = c("sample_alias", "condition")
        ) |>
          filter(marker_1 %in% markers_vec) |>
          filter(marker_2 %in% markers_vec) |>
          collect()

        res <- DifferentialProximityAnalysis(proximity_data,
          contrast_column = "condition",
          reference = reference_condition,
          proximity_metric = "log2_ratio",
          min_n_obs = 5
        )
        res$group <- x
        return(res)
      },
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  mutate(p_adj = p.adjust(p, method = "fdr")) # recalculate the adjusted p values after multiple testing correction for all comparisons, not only per cell type

dir.create(here::here("results/05_statistical_testing/colocalization/cell_level"), recursive = TRUE, showWarnings = FALSE)

# Save the differential abundance analysis results to a file
saveRDS(differential_colocalization_analysis, here::here("results/05_statistical_testing/colocalization/cell_level/differential_colocalization_analysis.rds"))

# Save the differential abundance analysis results to a csv file
write.csv(differential_colocalization_analysis, here::here("results/05_statistical_testing/colocalization/cell_level/differential_colocalization_analysis.csv"), row.names = FALSE)
```

### 5.3.1. Summarizing plots

Here we create summarizing plots for the differential colocalization analysis results:

1. **Colocalization Heatmap**: Displays the difference in medians for marker pairs, with dot size representing statistical significance (-log10 adjusted p-value). This provides an overview of which marker pairs are differentially colocalized for each cell type and condition comparison.

```{r}
#| label: statistical_testing_differential_colocalization_plot
#| eval: !expr isTRUE(run_cell_level)

# Determine target conditions - check if dataframe has target/contrast columns
if ("target" %in% colnames(differential_colocalization_analysis)) {
  # If target column exists, get all unique targets
  target_conditions <- unique(differential_colocalization_analysis$target)
  target_conditions <- target_conditions[!is.na(target_conditions)]
} else {
  stop("Could not find target column. Available columns: ", paste(colnames(differential_colocalization_analysis), collapse = ", "))
}

# Get unique cell types (groups)
unique_cell_types <- unique(differential_colocalization_analysis$group)
unique_cell_types <- unique_cell_types[!is.na(unique_cell_types)]

# Loop through each target condition and cell type combination
for (target_condition in target_conditions) {
  # Create directory name based on target and reference
  comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
  base_output_dir <- here::here("results/05_statistical_testing/colocalization/cell_level/differential_colocalization_plots", comparison_dir)

  for (cell_type in unique_cell_types) {
    # Create directory for this cell type
    cell_type_dir <- file.path(base_output_dir, cell_type)
    dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

    # Filter data for this target and cell type
    # Set diff_median to 0 for non-significant results (p_adj >= 0.001) to focus on highly significant changes
    plot_data <-
      differential_colocalization_analysis %>%
      filter(group == cell_type & target == target_condition)

    plot_markers <-
      plot_data %>%
      filter(p_adj < 0.001) %>%
      select(marker_1, marker_2) %>%
      deframe() %>%
      unique() %>%
      head(14)

    plot_data <-
      plot_data %>%
      filter(
        marker_1 %in% plot_markers,
        marker_2 %in% plot_markers
      )

    # Skip if insufficient data for this combination
    if (nrow(plot_data) < 2) {
      warning(paste("Insufficient data for plotting (", nrow(plot_data), "rows) for target condition:", target_condition, "and cell type:", cell_type))
      next
    }

    # Create the colocalization heatmap plot
    # Add title indicating the target and cell type
    colocalization_heatmap <-
      ColocalizationHeatmap(
        plot_data,
        marker1_col = "marker_1",
        marker2_col = "marker_2",
        value_col = "diff_median",
        type = "dots",
        colors = gradient_palette,
        size_col_transform = \(x) {
          -log10(x)
        }
      ) &
        labs(size = "-log10(p_adj)") &
        scale_size(range = c(1, 5), limits = c(0, max(-log10(plot_data$p_adj)))) &
        labs(fill = "Difference\nin medians") &
        labs(
          title = paste("Differential Colocalization Analysis", sep = ""),
          subtitle = paste("Cell type: ", cell_type, " (", target_condition, " vs ", reference_condition, ")", sep = "")
        )

    # Save the plot
    export_plot(
      filename = file.path(cell_type_dir, "colocalization_heatmap"),
      plot = colocalization_heatmap,
      width = max(c(4, length(plot_markers) * 0.15 + 1.5)),
      height = max(c(4, length(plot_markers) * 0.15 + 3))
    )
  }
}
```

## 5.4. Pseudobulk setup

Shared helpers for the optional limma path in §§5.5–5.7. Run this chunk before the abundance / clustering / colocalization tests. If `run_pseudobulk` is `FALSE`, or there are fewer than two `sample_alias` values per `condition`, later chunks skip silently. Clustering and colocalization share one sample-level proximity table built at the start of §5.6.

```{r}
#| label: statistical_testing_pseudobulk_helpers

# Number of top features labeled on volcano plots
pseudobulk_n_labels <- 5L

# Count sample_alias values per condition and decide whether limma can run
check_pseudobulk_replicates <- function(meta_df, min_n = 2L) {
  counts <- meta_df |>
    distinct(.data$sample_alias, .data$condition) |>
    dplyr::count(.data$condition, name = "n_samples")
  list(ok = nrow(counts) > 0 && all(counts$n_samples >= min_n), counts = counts)
}

# Fail-fast: stop if any requested covariate is missing from metadata.csv
validate_pseudobulk_covariates <- function(covariate_cols, metadata) {
  if (is.null(covariate_cols) || length(covariate_cols) == 0) {
    return(invisible(NULL))
  }
  missing_cols <- setdiff(as.character(covariate_cols), colnames(metadata))
  if (length(missing_cols) > 0) {
    stop(
      "Missing covariate columns in metadata.csv: ",
      paste(missing_cols, collapse = ", ")
    )
  }
  invisible(NULL)
}

# Join covariate columns from metadata.csv onto the per-sample table used by limma.
# Never join `condition` here — it is the limma contrast variable and must stay on
# sample_info as a single column (a join can otherwise create condition.x/y).
attach_pseudobulk_covariates <- function(sample_info, metadata, covariate_cols) {
  if (is.null(covariate_cols) || length(covariate_cols) == 0) {
    return(sample_info)
  }

  covariate_cols <- setdiff(
    as.character(covariate_cols),
    c("condition", "sample_alias")
  )
  if (length(covariate_cols) == 0) {
    return(sample_info)
  }

  out <- sample_info |>
    left_join(
      metadata |> select(sample_alias, all_of(covariate_cols)),
      by = "sample_alias"
    )

  if (!"condition" %in% names(out) && "condition.x" %in% names(out)) {
    out$condition <- out$condition.x
    out <- out |> select(-starts_with("condition."))
  }

  out
}

# Samples (sample_alias × condition) with at least min_cells in this cell type
pseudobulk_keep_samples <- function(obj, min_cells) {
  obj[[]] |>
    as_tibble(rownames = "component") |>
    dplyr::count(.data$sample_alias, .data$condition, name = "n_cells") |>
    filter(n_cells >= min_cells)
}

# CLR abundance + sample metadata. Fetch meta and assay separately, then merge
# on cell id. Do not cbind() — Seurat 5 FetchData can leak/drop `condition`,
# after which dplyr/plyr looks up a free object named condition and errors.
pseudobulk_fetch_abundance <- function(obj, markers_vec) {
  markers_vec <- setdiff(
    as.character(unname(markers_vec)),
    c("sample_alias", "condition", "cell_id")
  )
  missing_meta <- setdiff(c("sample_alias", "condition"), colnames(obj[[]]))
  if (length(missing_meta) > 0) {
    stop(
      "Missing columns in object metadata: ",
      paste(missing_meta, collapse = ", ")
    )
  }
  if (length(markers_vec) == 0L) {
    stop("No assay markers left to fetch after dropping metadata column names.")
  }

  # [[ ]] not $: $ can return NULL and tibble() then silently drops the column.
  md <- as.data.frame(obj[[]], stringsAsFactors = FALSE)
  meta <- data.frame(
    cell_id = if (!is.null(rownames(md)) && !identical(rownames(md), as.character(seq_len(nrow(md))))) {
      rownames(md)
    } else {
      Cells(obj)
    },
    sample_alias = md[["sample_alias"]],
    condition = md[["condition"]],
    stringsAsFactors = FALSE
  )

  assay <- FetchData(obj, vars = markers_vec, layer = "data")
  assay <- assay[, intersect(colnames(assay), markers_vec), drop = FALSE]
  assay <- data.frame(
    cell_id = rownames(assay),
    assay,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  merged <- merge(meta, assay, by = "cell_id", all.x = TRUE, sort = FALSE)
  if (!all(c("sample_alias", "condition") %in% names(merged))) {
    stop(
      "Lost sample_alias/condition after merge. Columns: ",
      paste(names(merged), collapse = ", ")
    )
  }
  merged
}

# Sample-level proximity summaries for one cell-type object.
# Clustering (§5.6) and colocalization (§5.7) both start from this table;
# they only differ in whether they keep self-pairs (marker_1 == marker_2)
# or off-diagonal pairs (marker_1 != marker_2).
# Group by sample_alias (not condition) so biological replicates stay separate
# for limma. FilterProximityScores requires p1/p2 from add_marker_proportions.
pseudobulk_summarize_proximity <- function(
  obj,
  markers_vec,
  min_cells = pseudobulk_min_cells,
  group_vars = c("sample_alias", "condition"),
  background_threshold_pct = 0.001
) {
  markers_vec <- as.character(markers_vec)
  keep_samples <- pseudobulk_keep_samples(obj, min_cells)
  if (nrow(keep_samples) < 2L) {
    return(NULL)
  }

  proximity_data <- ProximityScores(
    obj,
    lazy = TRUE,
    meta_data_columns = group_vars,
    add_marker_proportions = TRUE
  ) |>
    filter(marker_1 %in% markers_vec) |>
    filter(marker_2 %in% markers_vec) |>
    filter(sample_alias %in% keep_samples$sample_alias) |>
    FilterProximityScores(background_threshold_pct = background_threshold_pct)

  SummarizeProximityScores(
    proximity_data,
    proximity_metric = "log2_ratio",
    group_vars = group_vars,
    summary_stat = "mean",
    include_missing_obs = TRUE
  ) |>
    mutate(feature = paste(marker_1, marker_2, sep = "__"))
}

# Convert long sample × feature aggregates into a features × samples matrix
agg_to_expr_mat <- function(
  agg,
  feature_col,
  value_col,
  sample_info,
  values_fill = NULL
) {
  sample_ids <- as.character(sample_info$sample_alias)
  mat <- agg |>
    mutate(sample_alias = as.character(sample_alias)) |>
    select(all_of(c(feature_col, "sample_alias", value_col))) |>
    pivot_wider(
      names_from = sample_alias,
      values_from = all_of(value_col),
      values_fill = values_fill,
      values_fn = mean,
      names_repair = "minimal"
    ) |>
    column_to_rownames(feature_col) |>
    as.matrix()
  missing_samples <- setdiff(sample_ids, colnames(mat))
  if (length(missing_samples) > 0) {
    stop(
      "Pseudobulk matrix is missing samples: ",
      paste(missing_samples, collapse = ", ")
    )
  }
  mat[, sample_ids, drop = FALSE]
}

# Write RDS + CSV under results/05_statistical_testing/{modality}/pseudobulk/
save_pseudobulk_results <- function(df, modality, basename) {
  out_dir <- here::here("results/05_statistical_testing", modality, "pseudobulk")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  saveRDS(df, file.path(out_dir, paste0(basename, ".rds")))
  write.csv(df, file.path(out_dir, paste0(basename, ".csv")), row.names = FALSE)
  invisible(out_dir)
}

# Fit limma on a features × samples matrix; contrasts are target − reference
# (conditional on covariates when covariate_cols is set).
# expr_mat: features × samples; sample_info rows match colnames(expr_mat)
pseudobulk_limma_fit <- function(
  expr_mat,
  sample_info,
  reference_condition,
  covariate_cols = NULL
) {
  sample_info <- as.data.frame(sample_info, stringsAsFactors = FALSE)
  if (!"condition" %in% names(sample_info)) {
    stop(
      "sample_info has no condition column. Columns: ",
      paste(names(sample_info), collapse = ", ")
    )
  }
  sample_info$sample_alias <- as.character(sample_info[["sample_alias"]])
  sample_info$condition <- factor(as.character(sample_info[["condition"]]))
  stopifnot(identical(colnames(expr_mat), sample_info$sample_alias))
  if (!reference_condition %in% levels(sample_info$condition)) {
    stop("reference_condition not present in sample_info$condition")
  }
  sample_info$condition <- relevel(sample_info$condition, ref = reference_condition)

  # Condition fixed effects (one coefficient per condition level).
  # Using `~ 0 + condition` ensures coefficient names are the condition
  # levels themselves, which makes contrasts like `target-reference` easy.
  condition_design <- model.matrix(~ 0 + condition, data = sample_info)
  colnames(condition_design) <- levels(sample_info$condition)

  design <- condition_design
  if (!is.null(covariate_cols) && length(covariate_cols) > 0) {
    covariate_cols <- as.character(covariate_cols)
    missing_cols <- setdiff(covariate_cols, colnames(sample_info))
    if (length(missing_cols) > 0) {
      stop(
        "Missing pseudobulk covariate columns in sample_info: ",
        paste(missing_cols, collapse = ", ")
      )
    }

    # Drop the intercept from ~ covariates: condition already uses ~ 0 +
    # condition, so keeping (Intercept) would make the design rank-deficient.
    covar_formula <- as.formula(paste0("~ ", paste(covariate_cols, collapse = " + ")))
    covar_design <- model.matrix(covar_formula, data = sample_info)[, -1, drop = FALSE]
    design <- cbind(condition_design, covar_design)
  }

  targets <- setdiff(levels(sample_info$condition), reference_condition)
  if (length(targets) == 0) {
    return(tibble())
  }
  # Need residual DF after condition (+ covariate) coefficients.
  if (nrow(sample_info) <= ncol(design)) {
    return(tibble())
  }
  contrast_strings <- paste0(targets, "-", reference_condition)
  # Levels must match lmFit's design (conditions + covariates). Contrasts
  # only involve condition columns; covariate coefficients stay at 0.
  contrast_matrix <- limma::makeContrasts(
    contrasts = contrast_strings,
    levels = design
  )
  fit <- limma::lmFit(expr_mat, design)
  fit2 <- limma::eBayes(limma::contrasts.fit(fit, contrast_matrix))
  map_dfr(seq_along(targets), function(i) {
    tt <- limma::topTable(
      fit2,
      coef = i,
      number = Inf,
      sort.by = "none"
    ) |>
      as_tibble(rownames = "feature")
    tt |>
      transmute(
        feature,
        difference = logFC,
        p = P.Value,
        target = targets[[i]],
        reference = reference_condition,
        method = "limma_pseudobulk"
      )
  })
}

# Attach covariates, build the expression matrix, and run limma
run_pseudobulk_limma <- function(
  agg,
  feature_col,
  value_col,
  reference_condition,
  metadata,
  covariate_cols = NULL,
  values_fill = NULL
) {
  sample_info <- agg |>
    distinct(.data$sample_alias, .data$condition) |>
    attach_pseudobulk_covariates(metadata, covariate_cols) |>
    arrange(.data$sample_alias)
  expr_mat <- agg_to_expr_mat(
    agg,
    feature_col,
    value_col,
    sample_info,
    values_fill = values_fill
  )
  list(
    res = pseudobulk_limma_fit(
      expr_mat,
      sample_info,
      reference_condition,
      covariate_cols = covariate_cols
    ),
    sample_info = sample_info,
    expr_mat = expr_mat
  )
}

# Limma on sample-level proximity pairs; restores marker_1 / marker_2 for
# schema compatibility with cell-level DifferentialProximityAnalysis.
run_pseudobulk_proximity_limma <- function(agg, cell_type) {
  if (is.null(agg) || nrow(agg) == 0L) {
    return(NULL)
  }
  fit <- run_pseudobulk_limma(
    agg,
    feature_col = "feature",
    value_col = "mean_log2_ratio",
    reference_condition = reference_condition,
    metadata = metadata,
    covariate_cols = pseudobulk_covariate_cols,
    values_fill = 0
  )
  if (nrow(fit$res) == 0L) {
    return(NULL)
  }
  pair_map <- agg |> distinct(feature, marker_1, marker_2)
  fit$res |>
    left_join(pair_map, by = "feature") |>
    transmute(
      marker_1,
      marker_2,
      diff_median = difference,
      p,
      target,
      reference,
      method,
      group = cell_type
    )
}

# Row-z-scored sample heatmap (features × samples) as a ggplot
pseudobulk_sample_zscore_heatmap <- function(mat, sample_info, legend_title = "Z-score") {
  sample_info <- sample_info |>
    mutate(
      sample_alias = as.character(.data$sample_alias),
      condition = as.character(.data$condition)
    ) |>
    arrange(.data$condition, .data$sample_alias)
  stopifnot(all(sample_info$sample_alias %in% colnames(mat)))
  mat <- mat[, sample_info$sample_alias, drop = FALSE]
  mat_z <- t(scale(t(mat)))
  mat_z[!is.finite(mat_z)] <- 0
  z_max <- max(abs(mat_z), na.rm = TRUE)
  if (!is.finite(z_max) || z_max == 0) {
    z_max <- 1
  }
  z_lim <- max(2, ceiling(z_max * 10) / 10)
  ann_col <- data.frame(
    condition = factor(sample_info$condition, levels = unique(sample_info$condition)),
    row.names = sample_info$sample_alias
  )
  cond_levels <- levels(ann_col$condition)
  cond_cols <- condition_palette[cond_levels]
  if (any(is.na(cond_cols))) {
    fallback <- create_discrete_palette(cond_levels)[seq_along(cond_levels)]
    names(fallback) <- cond_levels
    cond_cols[is.na(cond_cols)] <- fallback[is.na(cond_cols)]
  }
  ComplexHeatmap::pheatmap(
    mat_z,
    color = PixelgenGradient(100, "BluesCherry"),
    breaks = seq(-z_lim, z_lim, length.out = 101),
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    annotation_col = ann_col,
    annotation_colors = list(condition = cond_cols),
    cellwidth = 12,
    cellheight = max(3.5, min(8, 350 / max(nrow(mat_z), 1))),
    heatmap_legend_param = list(title = legend_title),
    fontsize_row = 6,
    fontsize_col = 7,
    angle_col = "45"
  ) |>
    as.ggplot()
}

# Volcano of effect size vs −log10(adjusted p); optionally faceted by cell type
pseudobulk_volcano <- function(
  plot_data,
  label_data,
  label_col,
  title,
  subtitle,
  x_lab = "Effect size",
  label_size = 2,
  facet = TRUE,
  grey_guides = FALSE,
  expand_axes = FALSE
) {
  p <- plot_data |>
    ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
    geom_point(size = 0.5) +
    geom_vline(xintercept = 0)
  if (isTRUE(grey_guides)) {
    p <- p + geom_vline(xintercept = c(-1, 1), col = "grey")
  }
  p <- p +
    geom_text(
      data = label_data,
      aes(label = !!sym(label_col)),
      hjust = 0.5,
      vjust = -0.5,
      size = label_size
    )
  if (isTRUE(facet)) {
    p <- p + facet_wrap(~group, scales = "free")
  }
  if (isTRUE(expand_axes)) {
    p <- p +
      scale_x_continuous(
        limits = c(
          min(plot_data$effect_size, na.rm = TRUE),
          max(plot_data$effect_size, na.rm = TRUE)
        ),
        expand = expansion(0.1)
      ) +
      scale_y_continuous(expand = expansion(c(0, 0.1)))
  }
  p +
    scale_color_gradientn(
      colors = gradient_palette,
      limits = c(-1, 1) * max(abs(plot_data$effect_size), na.rm = TRUE)
    ) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    labs(
      title = title,
      subtitle = subtitle,
      x = x_lab,
      y = "-log10(adjusted p-value)",
      color = "Effect size"
    )
}

# Clustered heatmap of effect sizes (used for cross–cell-type summaries)
pseudobulk_effect_heatmap <- function(mat) {
  effect_vals <- as.numeric(mat)
  effect_quantiles <- quantile(effect_vals, probs = c(0.01, 0.99), na.rm = TRUE)
  max_abs_effect <- max(abs(effect_quantiles))
  breaks_range <- seq(-max_abs_effect, max_abs_effect, length.out = 101)
  mat |>
    pheatmap(
      color = PixelgenGradient(100, "BluesCherry"),
      breaks = breaks_range,
      cellwidth = 8,
      cellheight = 8,
      cluster_rows = TRUE,
      cluster_cols = TRUE,
      heatmap_legend_param = list(title = "Effect size"),
      fontsize = 8
    ) |>
    as.ggplot()
}

# Export a volcano, optionally combined with a sample heatmap beside it
export_volcano_with_optional_heatmap <- function(
  volcano,
  sample_hm,
  expr_mat,
  filename
) {
  if (!is.null(sample_hm) && !is.null(expr_mat)) {
    combined <- patchwork::wrap_plots(
      volcano,
      sample_hm,
      nrow = 1,
      widths = c(1, 1.2)
    )
    export_plot(
      filename = filename,
      plot = combined,
      width = 5.5 + max(4, ncol(expr_mat) * 0.35 + 2.5),
      height = max(4.5, nrow(expr_mat) * 0.12 + 2)
    )
  } else {
    export_plot(
      filename = filename,
      plot = volcano,
      width = 5.5,
      height = 4.5
    )
  }
}

# Features × cell types matrix of effect sizes for significant hits
pseudobulk_significant_effect_matrix <- function(
  plot_data,
  feature_col = "marker",
  abs_effect_min = 0.5
) {
  plot_data |>
    group_by(.data[[feature_col]]) |>
    filter(any(significant) & any(abs(effect_size) > abs_effect_min)) |>
    select(all_of(c(feature_col, "group", "effect_size"))) |>
    pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
    column_to_rownames(feature_col) |>
    as.matrix()
}

# Gate later chunks: run_pseudobulk AND ≥2 sample_alias per condition
pseudobulk_rep_check <- check_pseudobulk_replicates(
  pg_data[[]],
  min_n = 2L
)
pseudobulk_ok <- isTRUE(run_pseudobulk) && isTRUE(pseudobulk_rep_check$ok)

if (isTRUE(run_pseudobulk) && !isTRUE(pseudobulk_rep_check$ok)) {
  message(
    "Skipping limma pseudobulk: need ≥2 sample_alias per condition. ",
    "Current counts:\n",
    paste(
      capture.output(print(pseudobulk_rep_check$counts)),
      collapse = "\n"
    )
  )
} else if (!isTRUE(run_pseudobulk)) {
  message("Skipping limma pseudobulk (run_pseudobulk = FALSE).")
}

if (isTRUE(pseudobulk_ok)) {
  validate_pseudobulk_covariates(pseudobulk_covariate_cols, metadata)
  suppressPackageStartupMessages(library(limma))
}
```

## 5.5. Differential abundance analysis (pseudobulk-level)

> **Decision checkpoint** (enable above: `run_pseudobulk`)  
> Pseudobulk aggregates CLR abundance to **mean per `sample_alias` × cell type**, then fits limma across samples. Needs ≥2 samples per `condition`. Module 07 still reads the Wilcoxon results from §5.1 (`abundance/cell_level/`) unless you point it at `abundance/pseudobulk/`.

```{r}
#| label: statistical_testing_pseudobulk_abundance

if (isTRUE(pseudobulk_ok)) {
  differential_abundance_analysis_pseudobulk <-
    names(markers_to_test) |>
    map(function(cell_type) {
      tryCatch(
        {
          markers_vec <- setdiff(
            as.character(unname(markers_to_test[[cell_type]])),
            c("sample_alias", "condition", "cell_id")
          )
          obj <- subset(pg_data, cell_annotation == cell_type)
          dat <- pseudobulk_fetch_abundance(obj, markers_vec)
          n_cells <- dat |>
            dplyr::count(.data$sample_alias, .data$condition, name = "n_cells") |>
            filter(n_cells >= pseudobulk_min_cells)
          if (nrow(n_cells) < 2) {
            return(NULL)
          }
          agg <- dat |>
            filter(.data$sample_alias %in% n_cells$sample_alias) |>
            pivot_longer(
              all_of(markers_vec),
              names_to = "marker",
              values_to = "clr"
            ) |>
            group_by(.data$sample_alias, .data$condition, .data$marker) |>
            summarise(mean_clr = mean(clr), .groups = "drop")
          fit <- run_pseudobulk_limma(
            agg,
            feature_col = "marker",
            value_col = "mean_clr",
            reference_condition = reference_condition,
            metadata = metadata,
            covariate_cols = pseudobulk_covariate_cols
          )
          if (nrow(fit$res) == 0) {
            return(NULL)
          }
          fit$res |>
            rename(marker = feature) |>
            mutate(group = cell_type)
        },
        error = function(e) {
          message(paste("Pseudobulk abundance error for", cell_type, ":", e$message))
          NULL
        }
      )
    }) |>
    bind_rows()

  if (nrow(differential_abundance_analysis_pseudobulk) > 0) {
    differential_abundance_analysis_pseudobulk <-
      differential_abundance_analysis_pseudobulk |>
      mutate(p_adj = p.adjust(p, method = "fdr"))
  }

  save_pseudobulk_results(
    differential_abundance_analysis_pseudobulk,
    "abundance",
    "differential_abundance_analysis"
  )
}
```

### 5.5.1. Pseudobulk abundance plots

```{r}
#| label: statistical_testing_pseudobulk_abundance_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_abundance_analysis_pseudobulk") &&
    nrow(differential_abundance_analysis_pseudobulk) > 0
) {
  plot_data_base <- differential_abundance_analysis_pseudobulk |>
    mutate(
      neg_log10_p_adj = -log10(pmax(p_adj, min_p_value_threshold)),
      significant = p_adj < 0.05,
      effect_size = difference
    )
  target_conditions <- unique(plot_data_base$target)
  target_conditions <- target_conditions[!is.na(target_conditions)]

  for (target_condition in target_conditions) {
    plot_data <- plot_data_base |> filter(target == target_condition)
    if (nrow(plot_data) == 0) next

    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here(
      "results/05_statistical_testing/abundance/pseudobulk/differential_abundance_plots",
      comparison_dir
    )

    top_markers_faceted <- plot_data |>
      group_by(group) |>
      slice_max(order_by = abs(effect_size), n = pseudobulk_n_labels) |>
      ungroup()

    volcano_plot <- pseudobulk_volcano(
      plot_data,
      top_markers_faceted,
      label_col = "marker",
      title = paste0(
        "Pseudobulk differential abundance (", target_condition,
        " vs ", reference_condition, ")"
      ),
      subtitle = "limma on mean CLR per sample × cell type",
      x_lab = "Effect size (difference)"
    )

    export_plot(
      filename = file.path(base_output_dir, "differential_abundance_volcano"),
      plot = volcano_plot,
      width = 12,
      height = 8
    )

    for (cell_type in unique(plot_data$group)) {
      cell_type_dir <- file.path(base_output_dir, cell_type)
      dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

      cell_type_plot_data <- plot_data |> filter(group == cell_type)
      top_markers_cell_type <- cell_type_plot_data |>
        slice_max(order_by = abs(effect_size), n = pseudobulk_n_labels)

      volcano_plot_cell_type <- pseudobulk_volcano(
        cell_type_plot_data,
        top_markers_cell_type,
        label_col = "marker",
        title = "Pseudobulk differential abundance",
        subtitle = paste0(
          "Cell type: ", cell_type, " (", target_condition,
          " vs ", reference_condition, ")"
        ),
        x_lab = "Effect size (difference)",
        label_size = 3,
        grey_guides = TRUE,
        expand_axes = TRUE
      )

      # Mean CLR per sample (same grain as limma), row-z-scored; labeled markers only
      markers_vec <- unique(top_markers_cell_type$marker)
      sample_hm <- NULL
      expr_mat <- NULL
      if (length(markers_vec) >= 1) {
        obj <- subset(pg_data, cell_annotation == cell_type)
        dat <- pseudobulk_fetch_abundance(obj, markers_vec)
        n_keep <- dat |>
          dplyr::count(.data$sample_alias, .data$condition, name = "n_cells") |>
          filter(n_cells >= pseudobulk_min_cells)
        if (nrow(n_keep) >= 2) {
          agg <- dat |>
            filter(.data$sample_alias %in% n_keep$sample_alias) |>
            pivot_longer(
              all_of(markers_vec),
              names_to = "marker",
              values_to = "clr"
            ) |>
            group_by(.data$sample_alias, .data$condition, .data$marker) |>
            summarise(mean_clr = mean(clr), .groups = "drop")
          sample_info_hm <- agg |>
            distinct(.data$sample_alias, .data$condition) |>
            arrange(.data$condition, .data$sample_alias)
          marker_order <- top_markers_cell_type |>
            arrange(desc(abs(effect_size))) |>
            pull(marker) |>
            unique()
          expr_mat <- agg_to_expr_mat(
            agg,
            "marker",
            "mean_clr",
            sample_info_hm
          )
          expr_mat <- expr_mat[marker_order, , drop = FALSE]
          sample_hm <- pseudobulk_sample_zscore_heatmap(
            expr_mat,
            sample_info_hm,
            legend_title = "CLR\nZ-score"
          )
        }
      }

      export_volcano_with_optional_heatmap(
        volcano_plot_cell_type,
        sample_hm,
        expr_mat,
        file.path(cell_type_dir, "volcano_plot")
      )
    }

    significant_markers <- pseudobulk_significant_effect_matrix(plot_data)
    if (nrow(significant_markers) > 0 && ncol(significant_markers) > 0) {
      export_plot(
        filename = file.path(base_output_dir, "differential_abundance_heatmap"),
        plot = pseudobulk_effect_heatmap(t(significant_markers)),
        width = max(6, nrow(significant_markers) * 0.15 + 1.5),
        height = max(4, ncol(significant_markers) * 0.2 + 1.5)
      )
    }
  }
}
```

## 5.6. Differential clustering analysis (pseudobulk-level)

Self-proximity (`marker_1 == marker_2`) from the shared sample-level table below, then limma. Column `diff_median` stores the limma logFC (mean difference) for schema compatibility with §5.2 / module 07.

The next chunk builds **one** proximity summary for both clustering and colocalization: `FilterProximityScores()` drops pairs below an isotype-derived background, then `SummarizeProximityScores()` takes the mean `log2_ratio` per `sample_alias` × cell type (missing cells as 0).

```{r}
#| label: statistical_testing_pseudobulk_proximity_summary

if (isTRUE(pseudobulk_ok)) {
  # Isotype-based pair filter (same rule as the pixelatorR spatial-analysis
  # tutorial): median isotype fraction + 2 MAD, floored at 0.001.
  if ("isotype_fraction" %in% colnames(pg_data[[]])) {
    pseudobulk_proximity_background_pct <- max(
      median(pg_data$isotype_fraction, na.rm = TRUE) +
        2 * mad(pg_data$isotype_fraction, na.rm = TRUE),
      0.001
    )
  } else {
    message(
      "isotype_fraction missing from object metadata; ",
      "using FilterProximityScores background_threshold_pct = 0.001"
    )
    pseudobulk_proximity_background_pct <- 0.001
  }
  message(
    "FilterProximityScores background_threshold_pct = ",
    signif(pseudobulk_proximity_background_pct, 3)
  )

  summary_proximity_data_pseudobulk <-
    names(markers_to_test) |>
    map(function(cell_type) {
      tryCatch(
        {
          agg <- pseudobulk_summarize_proximity(
            subset(pg_data, cell_annotation == cell_type),
            markers_to_test[[cell_type]],
            background_threshold_pct = pseudobulk_proximity_background_pct
          )
          if (is.null(agg)) {
            return(NULL)
          }
          agg |> mutate(group = cell_type)
        },
        error = function(e) {
          message(
            paste("Pseudobulk proximity summary error for", cell_type, ":", e$message)
          )
          NULL
        }
      )
    }) |>
    bind_rows()

  if (nrow(summary_proximity_data_pseudobulk) == 0) {
    message(
      "No sample-level proximity summaries; ",
      "clustering and colocalization pseudobulk tests will be skipped."
    )
  }
}
```

```{r}
#| label: statistical_testing_pseudobulk_clustering

if (
  isTRUE(pseudobulk_ok) &&
    exists("summary_proximity_data_pseudobulk") &&
    nrow(summary_proximity_data_pseudobulk) > 0
) {
  differential_clustering_analysis_pseudobulk <-
    names(markers_to_test) |>
    map(function(cell_type) {
      tryCatch(
        {
          agg <- summary_proximity_data_pseudobulk |>
            filter(group == cell_type, marker_1 == marker_2)
          run_pseudobulk_proximity_limma(agg, cell_type)
        },
        error = function(e) {
          message(paste("Pseudobulk clustering error for", cell_type, ":", e$message))
          NULL
        }
      )
    }) |>
    bind_rows()

  if (nrow(differential_clustering_analysis_pseudobulk) > 0) {
    differential_clustering_analysis_pseudobulk <-
      differential_clustering_analysis_pseudobulk |>
      mutate(p_adj = p.adjust(p, method = "fdr"))
  }

  save_pseudobulk_results(
    differential_clustering_analysis_pseudobulk,
    "clustering",
    "differential_clustering_analysis"
  )
}
```

### 5.6.1. Pseudobulk clustering plots

```{r}
#| label: statistical_testing_pseudobulk_clustering_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_clustering_analysis_pseudobulk") &&
    nrow(differential_clustering_analysis_pseudobulk) > 0
) {
  plot_data_base <- differential_clustering_analysis_pseudobulk |>
    mutate(
      neg_log10_p_adj = -log10(pmax(p_adj, min_p_value_threshold)),
      significant = p_adj < 0.05,
      effect_size = diff_median,
      marker = marker_1
    )
  target_conditions <- unique(plot_data_base$target)
  target_conditions <- target_conditions[!is.na(target_conditions)]

  for (target_condition in target_conditions) {
    plot_data <- plot_data_base |> filter(target == target_condition)
    if (nrow(plot_data) == 0) next

    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here(
      "results/05_statistical_testing/clustering/pseudobulk/differential_clustering_plots",
      comparison_dir
    )

    top_markers_faceted <- plot_data |>
      group_by(group) |>
      slice_max(order_by = abs(effect_size), n = pseudobulk_n_labels) |>
      ungroup()

    volcano_plot <- pseudobulk_volcano(
      plot_data,
      top_markers_faceted,
      label_col = "marker",
      title = paste0(
        "Pseudobulk differential clustering (", target_condition,
        " vs ", reference_condition, ")"
      ),
      subtitle = "limma on mean log2_ratio per sample × cell type",
      x_lab = "Effect size (diff_median)"
    )

    export_plot(
      filename = file.path(base_output_dir, "differential_clustering_volcano"),
      plot = volcano_plot,
      width = 12,
      height = 8
    )

    for (cell_type in unique(plot_data$group)) {
      cell_type_dir <- file.path(base_output_dir, cell_type)
      dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

      cell_type_plot_data <- plot_data |> filter(group == cell_type)
      top_markers_cell_type <- cell_type_plot_data |>
        slice_max(order_by = abs(effect_size), n = pseudobulk_n_labels)

      volcano_plot_cell_type <- pseudobulk_volcano(
        cell_type_plot_data,
        top_markers_cell_type,
        label_col = "marker",
        title = "Pseudobulk differential clustering",
        subtitle = paste0(
          "Cell type: ", cell_type, " (", target_condition,
          " vs ", reference_condition, ")"
        ),
        x_lab = "Effect size (diff_median)",
        label_size = 3,
        grey_guides = TRUE,
        expand_axes = TRUE
      )

      markers_vec <- unique(top_markers_cell_type$marker)
      sample_hm <- NULL
      expr_mat <- NULL
      if (
        length(markers_vec) >= 1 &&
          exists("summary_proximity_data_pseudobulk")
      ) {
        agg <- summary_proximity_data_pseudobulk |>
          filter(
            group == cell_type,
            marker_1 == marker_2,
            marker_1 %in% markers_vec
          ) |>
          rename(marker = marker_1)
        if (nrow(agg) >= 1 && n_distinct(agg$sample_alias) >= 2) {
          sample_info_hm <- agg |>
            distinct(.data$sample_alias, .data$condition) |>
            arrange(.data$condition, .data$sample_alias)
          marker_order <- top_markers_cell_type |>
            filter(marker %in% unique(agg$marker)) |>
            arrange(desc(abs(effect_size))) |>
            pull(marker) |>
            unique()
          if (length(marker_order) >= 1) {
            expr_mat <- agg_to_expr_mat(
              agg,
              "marker",
              "mean_log2_ratio",
              sample_info_hm,
              values_fill = 0
            )
            expr_mat <- expr_mat[marker_order, , drop = FALSE]
            sample_hm <- pseudobulk_sample_zscore_heatmap(
              expr_mat,
              sample_info_hm,
              legend_title = "log2_ratio\nZ-score"
            )
          }
        }
      }

      export_volcano_with_optional_heatmap(
        volcano_plot_cell_type,
        sample_hm,
        expr_mat,
        file.path(cell_type_dir, "volcano_plot")
      )
    }

    significant_markers <- pseudobulk_significant_effect_matrix(plot_data)
    if (nrow(significant_markers) > 0 && ncol(significant_markers) > 0) {
      export_plot(
        filename = file.path(base_output_dir, "differential_clustering_heatmap"),
        plot = pseudobulk_effect_heatmap(t(significant_markers)),
        width = max(6, nrow(significant_markers) * 0.15 + 1.5),
        height = max(4, ncol(significant_markers) * 0.2 + 1.5)
      )
    }
  }
}
```

## 5.7. Differential colocalization analysis (pseudobulk-level)

Off-diagonal pairs (`marker_1 != marker_2`) from the same sample-level table as §5.6, then limma. Same `diff_median` naming as §5.3 for schema compatibility. To use these results in module 07, point its coloc RDS path to `results/05_statistical_testing/colocalization/pseudobulk/differential_colocalization_analysis.rds` (and the abundance path similarly under `abundance/pseudobulk/`).

```{r}
#| label: statistical_testing_pseudobulk_colocalization

if (
  isTRUE(pseudobulk_ok) &&
    exists("summary_proximity_data_pseudobulk") &&
    nrow(summary_proximity_data_pseudobulk) > 0
) {
  differential_colocalization_analysis_pseudobulk <-
    names(markers_to_test) |>
    map(function(cell_type) {
      tryCatch(
        {
          agg <- summary_proximity_data_pseudobulk |>
            filter(group == cell_type, marker_1 != marker_2)
          run_pseudobulk_proximity_limma(agg, cell_type)
        },
        error = function(e) {
          message(paste("Pseudobulk coloc error for", cell_type, ":", e$message))
          NULL
        }
      )
    }) |>
    bind_rows()

  if (nrow(differential_colocalization_analysis_pseudobulk) > 0) {
    differential_colocalization_analysis_pseudobulk <-
      differential_colocalization_analysis_pseudobulk |>
      mutate(p_adj = p.adjust(p, method = "fdr"))
  }

  save_pseudobulk_results(
    differential_colocalization_analysis_pseudobulk,
    "colocalization",
    "differential_colocalization_analysis"
  )
}
```

### 5.7.1. Pseudobulk colocalization plots

```{r}
#| label: statistical_testing_pseudobulk_colocalization_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_colocalization_analysis_pseudobulk") &&
    nrow(differential_colocalization_analysis_pseudobulk) > 0
) {
  coloc_base <-
    differential_colocalization_analysis_pseudobulk |>
    mutate(
      pair_id = paste(marker_1, marker_2, sep = "__"),
      neg_log10_p_adj = -log10(pmax(p_adj, min_p_value_threshold)),
      significant = p_adj < 0.05,
      effect_size = diff_median
    )

  target_conditions <- unique(coloc_base$target)
  target_conditions <- target_conditions[!is.na(target_conditions)]
  unique_cell_types <- unique(coloc_base$group)
  unique_cell_types <- unique_cell_types[!is.na(unique_cell_types)]

  for (target_condition in target_conditions) {
    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here(
      "results/05_statistical_testing/colocalization/pseudobulk/differential_colocalization_plots",
      comparison_dir
    )

    plot_data_target <- coloc_base |> filter(target == target_condition)

    top_pairs_faceted <- plot_data_target |>
      group_by(group) |>
      slice_max(order_by = abs(effect_size), n = pseudobulk_n_labels) |>
      ungroup()

    volcano_plot_coloc <- pseudobulk_volcano(
      plot_data_target,
      top_pairs_faceted,
      label_col = "pair_id",
      title = paste0(
        "Pseudobulk differential colocalization (",
        target_condition, " vs ", reference_condition, ")"
      ),
      subtitle = "limma on mean log2_ratio per sample (pairs)",
      x_lab = "Effect size (diff_median)",
      grey_guides = TRUE
    )

    export_plot(
      filename = file.path(base_output_dir, "differential_colocalization_volcano"),
      plot = volcano_plot_coloc,
      width = 12,
      height = 8
    )

    # Heatmap across cell types: top significant pairs × cell type
    sig_pairs <- plot_data_target |> filter(p_adj < 0.05)
    if (nrow(sig_pairs) > 0) {
      pair_strength <- sig_pairs |>
        group_by(pair_id) |>
        summarise(max_abs_effect = max(abs(effect_size)), .groups = "drop") |>
        slice_max(order_by = max_abs_effect, n = 50)

      heatmap_mat <- sig_pairs |>
        filter(pair_id %in% pair_strength$pair_id) |>
        select(pair_id, group, effect_size) |>
        pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
        column_to_rownames("pair_id") |>
        as.matrix()

      if (nrow(heatmap_mat) > 0 && ncol(heatmap_mat) > 0) {
        export_plot(
          filename = file.path(base_output_dir, "differential_colocalization_heatmap"),
          plot = pseudobulk_effect_heatmap(heatmap_mat),
          width = max(6, nrow(heatmap_mat) * 0.15 + 1.5),
          height = max(4, ncol(heatmap_mat) * 0.2 + 1.5)
        )
      }
    }

    for (cell_type in unique_cell_types) {
      cell_type_dir <- file.path(base_output_dir, cell_type)
      dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

      plot_data <- plot_data_target |> filter(group == cell_type)
      if (nrow(plot_data) < 1) next

      top_pairs_cell <- plot_data |>
        slice_max(order_by = abs(effect_size), n = pseudobulk_n_labels)

      volcano_plot_cell <- pseudobulk_volcano(
        plot_data,
        top_pairs_cell,
        label_col = "pair_id",
        title = "Pseudobulk differential colocalization: Volcano plot",
        subtitle = paste0(
          "Cell type: ", cell_type, " (", target_condition,
          " vs ", reference_condition, ")"
        ),
        x_lab = "Effect size (diff_median)",
        label_size = 3,
        facet = FALSE,
        grey_guides = TRUE
      )

      sample_hm <- NULL
      expr_mat <- NULL
      if (
        nrow(top_pairs_cell) > 0 &&
          exists("summary_proximity_data_pseudobulk")
      ) {
        pair_info <- top_pairs_cell |> distinct(pair_id, marker_1, marker_2)
        agg <- summary_proximity_data_pseudobulk |>
          filter(group == cell_type, marker_1 != marker_2) |>
          mutate(pair_id = feature) |>
          filter(pair_id %in% pair_info$pair_id)

        if (nrow(agg) >= 1 && n_distinct(agg$sample_alias) >= 2) {
          sample_info_hm <- agg |>
            distinct(.data$sample_alias, .data$condition) |>
            arrange(.data$condition, .data$sample_alias)

          row_order <- pair_info |>
            mutate(order_idx = match(pair_id, top_pairs_cell$pair_id)) |>
            arrange(order_idx) |>
            pull(pair_id) |>
            unique()

          expr_mat <- agg_to_expr_mat(
            agg,
            "pair_id",
            "mean_log2_ratio",
            sample_info_hm,
            values_fill = 0
          )
          row_order <- intersect(row_order, rownames(expr_mat))
          if (length(row_order) >= 1) {
            expr_mat <- expr_mat[row_order, , drop = FALSE]
            sample_hm <- pseudobulk_sample_zscore_heatmap(
              expr_mat,
              sample_info_hm,
              legend_title = "log2_ratio\nZ-score"
            )
          } else {
            expr_mat <- NULL
          }
        }
      }

      export_volcano_with_optional_heatmap(
        volcano_plot_cell,
        sample_hm,
        expr_mat,
        file.path(cell_type_dir, "volcano_plot")
      )

      # ColocalizationHeatmap: up to 14 markers from significant pairs
      plot_markers <- plot_data |>
        filter(p_adj < 0.05) |>
        select(marker_1, marker_2) |>
        pivot_longer(cols = c(marker_1, marker_2), values_to = "marker") |>
        pull(marker) |>
        unique() |>
        head(14)

      plot_data_heatmap <- plot_data |>
        filter(marker_1 %in% plot_markers, marker_2 %in% plot_markers)

      if (nrow(plot_data_heatmap) < 1) next

      colocalization_heatmap <-
        ColocalizationHeatmap(
          plot_data_heatmap,
          marker1_col = "marker_1",
          marker2_col = "marker_2",
          value_col = "diff_median",
          type = "dots",
          colors = gradient_palette,
          size_col_transform = \(x) {
            -log10(x)
          }
        ) &
          labs(size = "-log10(p_adj)") &
          scale_size(
            range = c(1, 5),
            limits = c(0, max(-log10(plot_data_heatmap$p_adj)))
          ) &
          labs(fill = "Difference\n(mean)") &
          labs(
            title = "Pseudobulk differential colocalization",
            subtitle = paste0(
              "Cell type: ", cell_type, " (", target_condition,
              " vs ", reference_condition, ")"
            )
          )

      export_plot(
        filename = file.path(cell_type_dir, "colocalization_heatmap"),
        plot = colocalization_heatmap,
        width = max(c(4, length(plot_markers) * 0.15 + 1.5)),
        height = max(c(4, length(plot_markers) * 0.15 + 3))
      )
    }
  }
}
```
