# 5. Statistical testing

**Module 05 — Statistical testing (optional).** Differential abundance, clustering, and colocalization tests (cell-level Wilcoxon by default; optional limma pseudobulk in §§5.5–5.7). Skip this module if you do not need formal comparisons. Run the first chunk below after module `02` has finished.

```{r}
#| label: module05_setup

library(here)
here::i_am("modules/05_statistical_testing.qmd")

source(here::here("modules", "common_setup.R"))

if (!exists("pg_data", inherits = TRUE)) {
  pg_data <- load_checkpoint("annotated_seurat_object.rds")
}
if (!exists("markers_to_test", inherits = TRUE)) {
  markers_to_test <- load_checkpoint("markers_to_test.rds")
}
```

## 5.0. Set reference sample

> **Decision checkpoint** (statistical contrasts)  
> Set `reference_condition` to one of the values in `metadata$condition`. All `{target}_vs_{reference}` outputs in module `05` use this baseline.  
> Optionally set `run_pseudobulk <- TRUE` to also run sample-level limma tests (requires ≥2 `sample_alias` values per `condition`).

By default, in this analysis we compare all conditions to a reference condition, which is typically the "resting" or "control" condition. This allows us to identify markers that are differentially abundant or spatially clustered or colocalized in stimulated or treated conditions compared to the baseline state.

Sections 5.1–5.3 use **cell-level** Wilcoxon tests (cells pooled by `condition`). These are the default outputs consumed by module 07 (under `{modality}/cell_level/`). With many cells they can overstate significance; when you have biological replicates, enable the optional **pseudobulk** limma path (setup in §5.4; tests in §§5.5–5.7). That path aggregates to `sample_alias` × cell type, then tests across samples; outputs go under `{modality}/pseudobulk/`.

```{r}
#| label: statistical_testing_set_reference

# Baseline / control condition. Must match a value in metadata$condition.
# All tests compare each other condition against this one (target − reference).
# Output folders are named {target}_vs_{reference}.
# Bundled example metadata uses donor1 / donor3 (both samples are resting).
reference_condition <- "donor1"

# If TRUE, also run sample-level limma tests (sections 5.5–5.7). Requires at
# least two samples (sample_alias) per condition. Results are written under
# results/05_statistical_testing/{modality}/pseudobulk/.
run_pseudobulk <- TRUE

# If TRUE, run cell-level Wilcoxon tests (sections 5.1–5.3). These are the
# default inputs for module 07. Set FALSE to skip that path.
run_cell_level <- TRUE

# Optional continuous-integration overrides. Loaded after the flags above so
# smoke tests can change them (see .github/smoke-test/apply_module05_overrides.R).
smoke_hook <- here::here(".github/smoke-test/apply_module05_overrides.R")
if (file.exists(smoke_hook)) {
  source(smoke_hook)
}

# --- Plotting and filtering options (do not re-run the statistical tests) ---

# Number of top hits labeled on each volcano plot (ranked by adjusted p-value).
n_volcano_labels <- 10L

# Adjusted p-value cutoff used to mark results as significant in plots, and to
# decide which pairs appear on colocalization heatmaps and effect heatmaps.
diff_p_adj_cutoff <- 0.05

# Minimum absolute effect size for markers on the cross–cell-type effect
# heatmaps (abundance and clustering). Markers must also pass the p-value
# cutoff above in at least one cell type.
diff_abs_effect_min <- 0.5

# Maximum number of marker pairs on the pseudobulk summary colocalization
# heatmap across cell types (strongest significant pairs). Does not limit the
# per–cell-type colocalization heatmaps.
coloc_cross_ct_heatmap_n <- 50L

# Minimum cells required for a sample within a cell type to be included in
# pseudobulk means and in the sample heatmaps next to volcanoes.
pseudobulk_min_cells <- 5L

# Optional batch / donor columns from metadata.csv to include in limma models,
# e.g. c("donor", "batch"). Leave as NULL for none. Do not list condition here.
pseudobulk_covariate_cols <- NULL

# Filters for cell-level clustering and colocalization tests only (Wilcoxon).
# They do not affect limma pseudobulk.
# - proximity_min_cells_per_group: minimum cells per condition in a cell type
# - proximity_diff_threshold: minimum absolute effect to keep a pair
# - proximity_min_pct / proximity_min_diff_pct: how widely a marker must be
#   detected; defaults keep almost all pairs
proximity_min_cells_per_group <- 10L
proximity_diff_threshold <- 0.01
proximity_min_pct <- 0
proximity_min_diff_pct <- -Inf

if (!reference_condition %in% metadata$condition) stop("The reference condition is not present in the sample metadata")
```

Shared helper functions used by the analysis and plotting chunks below.

```{r}
#| label: statistical_testing_shared_helpers

# Return the non-reference conditions present in a results table
get_target_conditions <- function(df) {
  if (!"target" %in% colnames(df)) {
    stop(
      "Could not find target column. Available columns: ",
      paste(colnames(df), collapse = ", ")
    )
  }
  targets <- unique(df$target)
  targets[!is.na(targets)]
}

# Add columns used by volcanoes and heatmaps (significance, effect size, −log10 p)
prepare_diff_plot_data <- function(df, effect_col, feature_col = NULL) {
  out <- df |>
    mutate(
      neg_log10_p_adj = -log10(pmax(p_adj, min_p_value_threshold)),
      significant = p_adj < diff_p_adj_cutoff,
      effect_size = !!sym(effect_col)
    )
  if (!is.null(feature_col)) {
    out <- out |> mutate(marker = !!sym(feature_col))
  }
  out
}

# Write results as RDS and CSV under results/05_statistical_testing/{modality}/{level}/
save_diff_results <- function(df, modality, level, basename) {
  out_dir <- here::here("results/05_statistical_testing", modality, level)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  saveRDS(df, file.path(out_dir, paste0(basename, ".rds")))
  write.csv(df, file.path(out_dir, paste0(basename, ".csv")), row.names = FALSE)
  invisible(out_dir)
}

# Recalculate FDR across the full results table (usually all cell types together).
# hard_fail = TRUE stops with a clear error if there are no rows to adjust.
adjust_global_fdr <- function(df, hard_fail = FALSE) {
  if (!is.data.frame(df) || nrow(df) == 0L || !"p" %in% names(df)) {
    if (isTRUE(hard_fail)) {
      stop(
        "No differential results to FDR-adjust. ",
        "All cell-type tests failed or returned empty; see messages above.",
        call. = FALSE
      )
    }
    return(df)
  }
  dplyr::mutate(df, p_adj = p.adjust(.data$p, method = "fdr"))
}

# Same as adjust_global_fdr, but errors if the proximity tests returned nothing
adjust_proximity_fdr <- function(df) {
  adjust_global_fdr(df, hard_fail = TRUE)
}

# Cell-level differential proximity (clustering or colocalization) for one cell
# type. Uses a Wilcoxon test on proximity scores. Keep only markers in
# markers_to_test. Call adjust_proximity_fdr() after combining cell types.
run_cell_level_proximity_da <- function(
  cell_type,
  markers_vec,
  pair_mode = c("self", "all", "co"),
  min_cells_per_group = proximity_min_cells_per_group,
  diff_threshold = proximity_diff_threshold,
  min_pct = proximity_min_pct,
  min_diff_pct = proximity_min_diff_pct
) {
  pair_mode <- match.arg(pair_mode)
  markers_vec <- as.character(markers_vec)
  obj <- subset(pg_data, cell_annotation == cell_type)

  res <- DifferentialProximityAnalysis(
    obj,
    contrast_column = "condition",
    reference = reference_condition,
    proximity_metric = "log2_ratio",
    metric_type = pair_mode,
    method = "seurat",
    min_cells_per_group = min_cells_per_group,
    diff_threshold = diff_threshold,
    min_pct = min_pct,
    min_diff_pct = min_diff_pct,
    lazy = TRUE
  )

  res <- res |>
    filter(
      .data$marker_1 %in% markers_vec,
      .data$marker_2 %in% markers_vec
    )
  if (nrow(res) == 0L) {
    return(NULL)
  }
  res$group <- cell_type
  res
}


# Select the n most significant rows (lowest adjusted p-value). Optionally
# pick the top n within each group (e.g. each cell type).
top_features_by_significance <- function(df, n, group_cols = NULL) {
  out <- df
  if (!"neg_log10_p_adj" %in% names(out)) {
    if (!"p_adj" %in% names(out)) {
      stop("top_features_by_significance needs neg_log10_p_adj or p_adj")
    }
    out <- out |>
      mutate(
        neg_log10_p_adj = -log10(pmax(.data$p_adj, min_p_value_threshold))
      )
  }
  if (!is.null(group_cols) && length(group_cols) > 0) {
    out <- out |> group_by(across(all_of(group_cols)))
  }
  out |>
    slice_max(order_by = .data$neg_log10_p_adj, n = n, with_ties = FALSE) |>
    ungroup()
}

# Fetch CLR abundance and sample metadata for one cell-type object.
# Metadata and assay values are merged carefully so sample labels are preserved.
pseudobulk_fetch_abundance <- function(obj, markers_vec) {
  markers_vec <- setdiff(
    as.character(unname(markers_vec)),
    c("sample_alias", "condition", "cell_id")
  )
  missing_meta <- setdiff(c("sample_alias", "condition"), colnames(obj[[]]))
  if (length(missing_meta) > 0) {
    stop(
      "Missing columns in object metadata: ",
      paste(missing_meta, collapse = ", ")
    )
  }
  if (length(markers_vec) == 0L) {
    stop("No assay markers left to fetch after dropping metadata column names.")
  }

  # Read metadata with [[ ]]; $ can drop columns silently
  md <- as.data.frame(obj[[]], stringsAsFactors = FALSE)
  meta <- data.frame(
    cell_id = if (!is.null(rownames(md)) && !identical(rownames(md), as.character(seq_len(nrow(md))))) {
      rownames(md)
    } else {
      Cells(obj)
    },
    sample_alias = md[["sample_alias"]],
    condition = md[["condition"]],
    stringsAsFactors = FALSE
  )

  assay <- FetchData(obj, vars = markers_vec, layer = "data")
  assay <- assay[, intersect(colnames(assay), markers_vec), drop = FALSE]
  assay <- data.frame(
    cell_id = rownames(assay),
    assay,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  merged <- merge(meta, assay, by = "cell_id", all.x = TRUE, sort = FALSE)
  if (!all(c("sample_alias", "condition") %in% names(merged))) {
    stop(
      "Lost sample_alias/condition after merge. Columns: ",
      paste(names(merged), collapse = ", ")
    )
  }
  merged
}

# Turn long sample-level summaries into a features × samples matrix for limma / heatmaps
agg_to_expr_mat <- function(
  agg,
  feature_col,
  value_col,
  sample_info,
  values_fill = NULL
) {
  sample_ids <- as.character(sample_info$sample_alias)
  mat <- agg |>
    mutate(sample_alias = as.character(sample_alias)) |>
    select(all_of(c(feature_col, "sample_alias", value_col))) |>
    pivot_wider(
      names_from = sample_alias,
      values_from = all_of(value_col),
      values_fill = values_fill,
      values_fn = mean,
      names_repair = "minimal"
    ) |>
    column_to_rownames(feature_col) |>
    as.matrix()
  missing_samples <- setdiff(sample_ids, colnames(mat))
  if (length(missing_samples) > 0) {
    stop(
      "Pseudobulk matrix is missing samples: ",
      paste(missing_samples, collapse = ", ")
    )
  }
  mat[, sample_ids, drop = FALSE]
}

# Heatmap of sample means (rows z-scored), annotated by condition
pseudobulk_sample_zscore_heatmap <- function(mat, sample_info, legend_title = "Z-score") {
  sample_info <- sample_info |>
    mutate(
      sample_alias = as.character(.data$sample_alias),
      condition = as.character(.data$condition)
    ) |>
    arrange(.data$condition, .data$sample_alias)
  stopifnot(all(sample_info$sample_alias %in% colnames(mat)))
  mat <- mat[, sample_info$sample_alias, drop = FALSE]
  mat_z <- t(scale(t(mat)))
  mat_z[!is.finite(mat_z)] <- 0
  z_max <- max(abs(mat_z), na.rm = TRUE)
  if (!is.finite(z_max) || z_max == 0) {
    z_max <- 1
  }
  z_lim <- max(2, ceiling(z_max * 10) / 10)
  ann_col <- data.frame(
    condition = factor(sample_info$condition, levels = unique(sample_info$condition)),
    row.names = sample_info$sample_alias
  )
  cond_levels <- levels(ann_col$condition)
  cond_cols <- condition_palette[cond_levels]
  if (any(is.na(cond_cols))) {
    fallback <- create_discrete_palette(cond_levels)[seq_along(cond_levels)]
    names(fallback) <- cond_levels
    cond_cols[is.na(cond_cols)] <- fallback[is.na(cond_cols)]
  }
  ComplexHeatmap::pheatmap(
    mat_z,
    color = PixelgenGradient(100, "BluesCherry"),
    breaks = seq(-z_lim, z_lim, length.out = 101),
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    annotation_col = ann_col,
    annotation_colors = list(condition = cond_cols),
    cellwidth = 12,
    cellheight = max(3.5, min(8, 350 / max(nrow(mat_z), 1))),
    heatmap_legend_param = list(title = legend_title),
    fontsize_row = 6,
    fontsize_col = 7,
    angle_col = "45"
  ) |>
    as.ggplot()
}

# Save a volcano plot, with an optional sample heatmap beside it when available
export_volcano_with_optional_heatmap <- function(
  volcano,
  sample_hm,
  expr_mat,
  filename
) {
  if (!is.null(sample_hm) && !is.null(expr_mat)) {
    combined <- patchwork::wrap_plots(
      volcano,
      sample_hm,
      nrow = 1,
      widths = c(1, 1.2)
    )
    export_plot(
      filename = filename,
      plot = combined,
      width = 5.5 + max(4, ncol(expr_mat) * 0.35 + 2.5),
      height = max(4.5, nrow(expr_mat) * 0.12 + 2)
    )
  } else {
    export_plot(
      filename = filename,
      plot = volcano,
      width = 5.5,
      height = 4.5
    )
  }
}


# Build the sample heatmap shown next to abundance or clustering volcanoes
# (mean per sample for the labeled markers)
build_volcano_sample_heatmap <- function(
  cell_type,
  top_features_df,
  modality = c("abundance", "clustering"),
  min_cells = pseudobulk_min_cells
) {
  modality <- match.arg(modality)
  marker_order <- top_features_df |>
    arrange(desc(neg_log10_p_adj)) |>
    pull(marker) |>
    unique()
  if (length(marker_order) < 1) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }

  obj <- subset(pg_data, cell_annotation == cell_type)

  if (identical(modality, "abundance")) {
    dat <- pseudobulk_fetch_abundance(obj, marker_order)
    n_keep <- dat |>
      dplyr::count(.data$sample_alias, .data$condition, name = "n_cells") |>
      filter(n_cells >= min_cells)
    if (nrow(n_keep) < 2) {
      return(list(sample_hm = NULL, expr_mat = NULL))
    }
    agg <- dat |>
      filter(.data$sample_alias %in% n_keep$sample_alias) |>
      pivot_longer(
        all_of(marker_order),
        names_to = "marker",
        values_to = "clr"
      ) |>
      group_by(.data$sample_alias, .data$condition, .data$marker) |>
      summarise(mean_clr = mean(clr), .groups = "drop")
    value_col <- "mean_clr"
    legend_title <- "CLR\nZ-score"
    values_fill <- NULL
  } else {
    n_keep <- as.data.frame(obj[[]], stringsAsFactors = FALSE) |>
      dplyr::count(.data$sample_alias, .data$condition, name = "n_cells") |>
      filter(n_cells >= min_cells)
    if (nrow(n_keep) < 2) {
      return(list(sample_hm = NULL, expr_mat = NULL))
    }
    prox <- ProximityScores(
      obj,
      lazy = TRUE,
      meta_data_columns = c("sample_alias", "condition")
    ) |>
      filter(marker_1 == marker_2) |>
      filter(marker_1 %in% marker_order) |>
      collect()
    if (nrow(prox) < 1) {
      return(list(sample_hm = NULL, expr_mat = NULL))
    }
    agg <- prox |>
      filter(.data$sample_alias %in% n_keep$sample_alias) |>
      group_by(.data$sample_alias, .data$condition, marker = .data$marker_1) |>
      summarise(
        mean_log2_ratio = mean(.data$log2_ratio, na.rm = TRUE),
        .groups = "drop"
      )
    value_col <- "mean_log2_ratio"
    legend_title <- "log2_ratio\nZ-score"
    values_fill <- 0
  }

  sample_info_hm <- agg |>
    distinct(.data$sample_alias, .data$condition) |>
    arrange(.data$condition, .data$sample_alias)
  if (dplyr::n_distinct(sample_info_hm$sample_alias) < 2) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }

  # Keep all labeled markers as rows; 0-fill missing sample × marker values
  expr_mat <- agg_to_expr_mat(
    agg,
    "marker",
    value_col,
    sample_info_hm,
    values_fill = values_fill
  )
  missing_markers <- setdiff(marker_order, rownames(expr_mat))
  if (length(missing_markers) > 0) {
    pad <- matrix(
      if (is.null(values_fill)) NA_real_ else values_fill,
      nrow = length(missing_markers),
      ncol = ncol(expr_mat),
      dimnames = list(missing_markers, colnames(expr_mat))
    )
    expr_mat <- rbind(expr_mat, pad)
  }
  expr_mat <- expr_mat[marker_order, , drop = FALSE]
  list(
    sample_hm = pseudobulk_sample_zscore_heatmap(
      expr_mat,
      sample_info_hm,
      legend_title = legend_title
    ),
    expr_mat = expr_mat
  )
}


# Build the sample heatmap shown next to colocalization volcanoes (marker pairs).
# Rows match the labeled pair_id values exactly (including self-pairs when labeled).
build_volcano_sample_heatmap_pairs <- function(
  cell_type,
  top_pairs_df,
  min_cells = pseudobulk_min_cells
) {
  pair_order <- top_pairs_df |>
    arrange(desc(neg_log10_p_adj)) |>
    pull(pair_id) |>
    unique()
  if (length(pair_order) < 1) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }

  markers_needed <- unique(c(top_pairs_df$marker_1, top_pairs_df$marker_2))
  obj <- subset(pg_data, cell_annotation == cell_type)
  n_keep <- as.data.frame(obj[[]], stringsAsFactors = FALSE) |>
    dplyr::count(.data$sample_alias, .data$condition, name = "n_cells") |>
    filter(n_cells >= min_cells)
  if (nrow(n_keep) < 2) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }

  # Keep self-pairs and cross-pairs so labeled volcano features can appear
  prox <- ProximityScores(
    obj,
    lazy = TRUE,
    meta_data_columns = c("sample_alias", "condition")
  ) |>
    filter(marker_1 %in% markers_needed) |>
    filter(marker_2 %in% markers_needed) |>
    collect()
  if (nrow(prox) < 1) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }

  agg <- prox |>
    filter(.data$sample_alias %in% n_keep$sample_alias) |>
    mutate(pair_id = paste(.data$marker_1, .data$marker_2, sep = "__")) |>
    filter(.data$pair_id %in% pair_order) |>
    group_by(.data$sample_alias, .data$condition, .data$pair_id) |>
    summarise(
      mean_log2_ratio = mean(.data$log2_ratio, na.rm = TRUE),
      .groups = "drop"
    )

  sample_info_hm <- agg |>
    distinct(.data$sample_alias, .data$condition) |>
    arrange(.data$condition, .data$sample_alias)
  if (dplyr::n_distinct(sample_info_hm$sample_alias) < 2) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }

  # Ensure every labeled pair is a heatmap row (0-fill if absent in a sample)
  if (nrow(agg) < 1) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }
  expr_mat <- agg_to_expr_mat(
    agg,
    "pair_id",
    "mean_log2_ratio",
    sample_info_hm,
    values_fill = 0
  )
  missing_pairs <- setdiff(pair_order, rownames(expr_mat))
  if (length(missing_pairs) > 0) {
    pad <- matrix(
      0,
      nrow = length(missing_pairs),
      ncol = ncol(expr_mat),
      dimnames = list(missing_pairs, colnames(expr_mat))
    )
    expr_mat <- rbind(expr_mat, pad)
  }
  expr_mat <- expr_mat[pair_order, , drop = FALSE]
  list(
    sample_hm = pseudobulk_sample_zscore_heatmap(
      expr_mat,
      sample_info_hm,
      legend_title = "log2_ratio\nZ-score"
    ),
    expr_mat = expr_mat
  )
}


# Standard volcano: effect size vs −log10(adjusted p-value)
pseudobulk_volcano <- function(
  plot_data,
  label_data,
  label_col,
  title,
  subtitle,
  x_lab = "Effect size",
  label_size = 2,
  facet = TRUE,
  grey_guides = FALSE,
  expand_axes = FALSE
) {
  p <- plot_data |>
    ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
    geom_point(size = 0.5) +
    geom_vline(xintercept = 0)
  if (isTRUE(grey_guides)) {
    p <- p + geom_vline(xintercept = c(-1, 1), col = "grey")
  }
  p <- p +
    geom_text(
      data = label_data,
      aes(label = !!sym(label_col)),
      hjust = 0.5,
      vjust = -0.5,
      size = label_size
    )
  if (isTRUE(facet)) {
    p <- p + facet_wrap(~group, scales = "free")
  }
  if (isTRUE(expand_axes)) {
    p <- p +
      scale_x_continuous(
        limits = c(
          min(plot_data$effect_size, na.rm = TRUE),
          max(plot_data$effect_size, na.rm = TRUE)
        ),
        expand = expansion(0.1)
      ) +
      scale_y_continuous(expand = expansion(c(0, 0.1)))
  }
  p +
    scale_color_gradientn(
      colors = gradient_palette,
      limits = c(-1, 1) * max(abs(plot_data$effect_size), na.rm = TRUE)
    ) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    labs(
      title = title,
      subtitle = subtitle,
      x = x_lab,
      y = "-log10(adjusted p-value)",
      color = "Effect size"
    )
}

# Clustered heatmap of effect sizes across cell types
pseudobulk_effect_heatmap <- function(mat) {
  effect_vals <- as.numeric(mat)
  effect_quantiles <- quantile(effect_vals, probs = c(0.01, 0.99), na.rm = TRUE)
  max_abs_effect <- max(abs(effect_quantiles))
  breaks_range <- seq(-max_abs_effect, max_abs_effect, length.out = 101)
  mat |>
    pheatmap(
      color = PixelgenGradient(100, "BluesCherry"),
      breaks = breaks_range,
      cellwidth = 8,
      cellheight = 8,
      cluster_rows = TRUE,
      cluster_cols = TRUE,
      heatmap_legend_param = list(title = "Effect size"),
      fontsize = 8
    ) |>
    as.ggplot()
}

# Matrix of effect sizes for significant features (rows) × cell types (columns)
pseudobulk_significant_effect_matrix <- function(
  plot_data,
  feature_col = "marker",
  abs_effect_min = diff_abs_effect_min
) {
  plot_data |>
    group_by(.data[[feature_col]]) |>
    filter(any(significant) & any(abs(effect_size) > abs_effect_min)) |>
    select(all_of(c(feature_col, "group", "effect_size"))) |>
    pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
    column_to_rownames(feature_col) |>
    as.matrix()
}

# Sample heatmap for pseudobulk clustering volcanoes (from the shared proximity table)
build_pb_summary_clustering_heatmap <- function(cell_type, top_features_df) {
  markers_vec <- unique(top_features_df$marker)
  if (
    length(markers_vec) < 1 ||
      !exists("summary_proximity_data_pseudobulk") ||
      !is.data.frame(summary_proximity_data_pseudobulk) ||
      nrow(summary_proximity_data_pseudobulk) < 1
  ) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }
  agg <- summary_proximity_data_pseudobulk |>
    filter(
      group == cell_type,
      marker_1 == marker_2,
      marker_1 %in% markers_vec
    ) |>
    rename(marker = marker_1)
  if (nrow(agg) < 1 || n_distinct(agg$sample_alias) < 2) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }
  sample_info_hm <- agg |>
    distinct(.data$sample_alias, .data$condition) |>
    arrange(.data$condition, .data$sample_alias)
  marker_order <- top_features_df |>
    arrange(desc(neg_log10_p_adj)) |>
    pull(marker) |>
    unique()
  if (length(marker_order) < 1) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }
  expr_mat <- agg_to_expr_mat(
    agg,
    "marker",
    "mean_log2_ratio",
    sample_info_hm,
    values_fill = 0
  )
  missing_markers <- setdiff(marker_order, rownames(expr_mat))
  if (length(missing_markers) > 0) {
    pad <- matrix(
      0,
      nrow = length(missing_markers),
      ncol = ncol(expr_mat),
      dimnames = list(missing_markers, colnames(expr_mat))
    )
    expr_mat <- rbind(expr_mat, pad)
  }
  expr_mat <- expr_mat[marker_order, , drop = FALSE]
  list(
    sample_hm = pseudobulk_sample_zscore_heatmap(
      expr_mat,
      sample_info_hm,
      legend_title = "log2_ratio\nZ-score"
    ),
    expr_mat = expr_mat
  )
}

# Sample heatmap for pseudobulk colocalization volcanoes (from the shared proximity table).
# Includes self-pairs when they are among the labeled features.
build_pb_summary_coloc_heatmap <- function(cell_type, top_pairs_df) {
  if (
    nrow(top_pairs_df) < 1 ||
      !exists("summary_proximity_data_pseudobulk") ||
      !is.data.frame(summary_proximity_data_pseudobulk) ||
      nrow(summary_proximity_data_pseudobulk) < 1
  ) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }
  pair_order <- top_pairs_df |>
    arrange(desc(neg_log10_p_adj)) |>
    pull(pair_id) |>
    unique()
  pair_info <- top_pairs_df |> distinct(pair_id, marker_1, marker_2)
  # Do not drop self-pairs: volcano labels can include marker__marker
  agg <- summary_proximity_data_pseudobulk |>
    filter(group == cell_type) |>
    mutate(pair_id = feature) |>
    filter(pair_id %in% pair_info$pair_id)
  if (nrow(agg) < 1 || n_distinct(agg$sample_alias) < 2) {
    return(list(sample_hm = NULL, expr_mat = NULL))
  }
  sample_info_hm <- agg |>
    distinct(.data$sample_alias, .data$condition) |>
    arrange(.data$condition, .data$sample_alias)
  expr_mat <- agg_to_expr_mat(
    agg,
    "pair_id",
    "mean_log2_ratio",
    sample_info_hm,
    values_fill = 0
  )
  missing_pairs <- setdiff(pair_order, rownames(expr_mat))
  if (length(missing_pairs) > 0) {
    pad <- matrix(
      0,
      nrow = length(missing_pairs),
      ncol = ncol(expr_mat),
      dimnames = list(missing_pairs, colnames(expr_mat))
    )
    expr_mat <- rbind(expr_mat, pad)
  }
  expr_mat <- expr_mat[pair_order, , drop = FALSE]
  list(
    sample_hm = pseudobulk_sample_zscore_heatmap(
      expr_mat,
      sample_info_hm,
      legend_title = "log2_ratio\nZ-score"
    ),
    expr_mat = expr_mat
  )
}

# Export abundance or clustering plots: faceted volcano, per–cell-type volcano
# (+ optional sample heatmap), and cross–cell-type effect heatmap
export_marker_diff_plots <- function(
  results_df,
  modality,
  effect_col,
  feature_col,
  analysis_label,
  level = "cell_level",
  n_labels = n_volcano_labels,
  volcano_subtitle = "Effect size vs statistical significance by cell type",
  append_volcano_label = TRUE,
  grey_guides_faceted = TRUE,
  expand_faceted = TRUE,
  grey_guides_cell = TRUE,
  expand_cell = TRUE,
  build_side_heatmap = NULL
) {
  if (is.null(build_side_heatmap)) {
    build_side_heatmap <- function(cell_type, top_features_df) {
      build_volcano_sample_heatmap(cell_type, top_features_df, modality = modality)
    }
  }

  plot_data_base <- prepare_diff_plot_data(results_df, effect_col, feature_col)
  target_conditions <- get_target_conditions(plot_data_base)
  plot_subdir <- paste0("differential_", modality, "_plots")

  for (target_condition in target_conditions) {
    plot_data <- plot_data_base |> filter(target == target_condition)
    if (nrow(plot_data) == 0) {
      warning(paste("No data found for target condition:", target_condition))
      next
    }

    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here(
      "results/05_statistical_testing",
      modality,
      level,
      plot_subdir,
      comparison_dir
    )

    top_markers_faceted <- top_features_by_significance(
      plot_data,
      n_labels,
      group_cols = "group"
    )

    faceted_title <- if (isTRUE(append_volcano_label)) {
      paste0(
        analysis_label, ": Volcano plot (",
        target_condition, " vs ", reference_condition, ")"
      )
    } else {
      paste0(analysis_label, " (", target_condition, " vs ", reference_condition, ")")
    }

    volcano_faceted <- pseudobulk_volcano(
      plot_data,
      top_markers_faceted,
      label_col = "marker",
      title = faceted_title,
      subtitle = volcano_subtitle,
      x_lab = paste0("Effect size (", effect_col, ")"),
      grey_guides = grey_guides_faceted,
      expand_axes = expand_faceted
    )

    export_plot(
      filename = file.path(base_output_dir, paste0("differential_", modality, "_volcano")),
      plot = volcano_faceted,
      width = 12,
      height = 8
    )

    cell_title <- if (isTRUE(append_volcano_label)) {
      paste0(analysis_label, ": Volcano plot")
    } else {
      analysis_label
    }

    for (cell_type in unique(plot_data$group)) {
      cell_type_dir <- file.path(base_output_dir, cell_type)
      dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)
      cell_type_plot_data <- plot_data |> filter(group == cell_type)
      top_markers_cell_type <- top_features_by_significance(
        cell_type_plot_data,
        n_labels
      )

      volcano_cell <- pseudobulk_volcano(
        cell_type_plot_data,
        top_markers_cell_type,
        label_col = "marker",
        title = cell_title,
        subtitle = paste0(
          "Cell type: ", cell_type, " (",
          target_condition, " vs ", reference_condition, ")"
        ),
        x_lab = paste0("Effect size (", effect_col, ")"),
        label_size = 3,
        grey_guides = grey_guides_cell,
        expand_axes = expand_cell
      )

      hm <- build_side_heatmap(cell_type, top_markers_cell_type)
      export_volcano_with_optional_heatmap(
        volcano_cell,
        hm$sample_hm,
        hm$expr_mat,
        file.path(cell_type_dir, "volcano_plot")
      )
    }

    significant_markers <- pseudobulk_significant_effect_matrix(plot_data)
    if (nrow(significant_markers) > 0 && ncol(significant_markers) > 0) {
      export_plot(
        filename = file.path(base_output_dir, paste0("differential_", modality, "_heatmap")),
        plot = pseudobulk_effect_heatmap(t(significant_markers)),
        width = max(6, nrow(significant_markers) * 0.15 + 1.5),
        height = max(4, ncol(significant_markers) * 0.2 + 1.5)
      )
    } else {
      warning(paste(
        "No significant markers found for heatmap visualization (p_adj <",
        diff_p_adj_cutoff, ") for",
        target_condition, "vs", reference_condition
      ))
    }
  }
  invisible(NULL)
}

# Export colocalization plots: volcanoes, optional sample heatmaps, pair heatmaps,
# and optionally a summary heatmap across cell types
export_colocalization_diff_plots <- function(
  results_df,
  level = "cell_level",
  analysis_label = "Differential colocalization analysis",
  volcano_subtitle = "Effect size vs statistical significance by cell type",
  fill_lab = "Difference\nin medians",
  n_labels = n_volcano_labels,
  append_volcano_label = TRUE,
  grey_guides_faceted = TRUE,
  expand_faceted = TRUE,
  expand_cell = TRUE,
  min_heatmap_pairs = 2L,
  include_cross_ct_heatmap = FALSE,
  build_side_heatmap = NULL
) {
  if (is.null(build_side_heatmap)) {
    build_side_heatmap <- function(cell_type, top_pairs_df) {
      build_volcano_sample_heatmap_pairs(cell_type, top_pairs_df)
    }
  }

  plot_data_base <- prepare_diff_plot_data(results_df, "diff_median") |>
    mutate(pair_id = paste(marker_1, marker_2, sep = "__"))

  target_conditions <- get_target_conditions(plot_data_base)
  unique_cell_types <- unique(plot_data_base$group)
  unique_cell_types <- unique_cell_types[!is.na(unique_cell_types)]

  for (target_condition in target_conditions) {
    plot_data_target <- plot_data_base |> filter(target == target_condition)
    if (nrow(plot_data_target) == 0) next

    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here(
      "results/05_statistical_testing/colocalization",
      level,
      "differential_colocalization_plots",
      comparison_dir
    )

    top_pairs_faceted <- top_features_by_significance(
      plot_data_target,
      n_labels,
      group_cols = "group"
    )

    faceted_title <- if (isTRUE(append_volcano_label)) {
      paste0(
        analysis_label, ": Volcano plot (",
        target_condition, " vs ", reference_condition, ")"
      )
    } else {
      paste0(analysis_label, " (", target_condition, " vs ", reference_condition, ")")
    }

    volcano_faceted <- pseudobulk_volcano(
      plot_data_target,
      top_pairs_faceted,
      label_col = "pair_id",
      title = faceted_title,
      subtitle = volcano_subtitle,
      x_lab = "Effect size (diff_median)",
      grey_guides = grey_guides_faceted,
      expand_axes = expand_faceted
    )

    export_plot(
      filename = file.path(base_output_dir, "differential_colocalization_volcano"),
      plot = volcano_faceted,
      width = 12,
      height = 8
    )

    if (isTRUE(include_cross_ct_heatmap)) {
      sig_pairs <- plot_data_target |> filter(p_adj < diff_p_adj_cutoff)
      if (nrow(sig_pairs) > 0) {
        pair_strength <- sig_pairs |>
          group_by(pair_id) |>
          summarise(max_neg_log10_p = max(neg_log10_p_adj), .groups = "drop") |>
          slice_max(
            order_by = max_neg_log10_p,
            n = coloc_cross_ct_heatmap_n,
            with_ties = FALSE
          )

        heatmap_mat <- sig_pairs |>
          filter(pair_id %in% pair_strength$pair_id) |>
          select(pair_id, group, effect_size) |>
          pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
          column_to_rownames("pair_id") |>
          as.matrix()

        if (nrow(heatmap_mat) > 0 && ncol(heatmap_mat) > 0) {
          export_plot(
            filename = file.path(base_output_dir, "differential_colocalization_heatmap"),
            plot = pseudobulk_effect_heatmap(heatmap_mat),
            width = max(6, nrow(heatmap_mat) * 0.15 + 1.5),
            height = max(4, ncol(heatmap_mat) * 0.2 + 1.5)
          )
        }
      }
    }

    cell_title <- paste0(analysis_label, ": Volcano plot")

    for (cell_type in unique_cell_types) {
      cell_type_dir <- file.path(base_output_dir, cell_type)
      dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

      plot_data <- plot_data_target |> filter(group == cell_type)
      if (nrow(plot_data) < 1) next

      top_pairs_cell <- top_features_by_significance(plot_data, n_labels)

      volcano_cell <- pseudobulk_volcano(
        plot_data,
        top_pairs_cell,
        label_col = "pair_id",
        title = cell_title,
        subtitle = paste0(
          "Cell type: ", cell_type, " (",
          target_condition, " vs ", reference_condition, ")"
        ),
        x_lab = "Effect size (diff_median)",
        label_size = 3,
        facet = FALSE,
        grey_guides = TRUE,
        expand_axes = expand_cell
      )

      hm <- build_side_heatmap(cell_type, top_pairs_cell)
      export_volcano_with_optional_heatmap(
        volcano_cell,
        hm$sample_hm,
        hm$expr_mat,
        file.path(cell_type_dir, "volcano_plot")
      )

      plot_data_hm <- plot_data |>
        filter(p_adj < diff_p_adj_cutoff) |>
        arrange(desc(neg_log10_p_adj))

      plot_markers <- plot_data_hm |>
        select(marker_1, marker_2) |>
        pivot_longer(cols = c(marker_1, marker_2), values_to = "marker") |>
        pull(marker) |>
        unique()

      if (nrow(plot_data_hm) < min_heatmap_pairs) {
        if (identical(level, "cell_level")) {
          warning(paste(
            "Insufficient significant pairs (p_adj <", diff_p_adj_cutoff,
            ") for colocalization heatmap for",
            target_condition, cell_type
          ))
        }
        next
      }

      colocalization_heatmap <-
        ColocalizationHeatmap(
          plot_data_hm,
          marker1_col = "marker_1",
          marker2_col = "marker_2",
          value_col = "diff_median",
          type = "dots",
          colors = gradient_palette,
          size_col_transform = \(x) {
            -log10(x)
          }
        ) &
          labs(size = "-log10(p_adj)") &
          scale_size(
            range = c(1, 5),
            limits = c(0, max(-log10(plot_data_hm$p_adj)))
          ) &
          labs(fill = fill_lab) &
          labs(
            title = analysis_label,
            subtitle = paste0(
              "Cell type: ", cell_type, " (",
              target_condition, " vs ", reference_condition, ")"
            )
          )

      export_plot(
        filename = file.path(cell_type_dir, "colocalization_heatmap"),
        plot = colocalization_heatmap,
        width = max(c(4, length(plot_markers) * 0.15 + 1.5)),
        height = max(c(4, length(plot_markers) * 0.15 + 3))
      )
    }
  }
  invisible(NULL)
}
```

## 5.1. Differential abundance analysis (cell-level)

Differential abundance analysis identifies markers that exhibit statistically significant differences in expression levels between experimental conditions. This analysis is essential for understanding how experimental treatments, disease states, or other conditions affect protein expression patterns within specific cell types.

The analysis is performed separately for each annotated cell type, comparing marker abundance between a target condition (e.g., stimulated, treated) and a reference condition (e.g., resting, control). For each marker within each cell type, a statistical test (typically the Wilcoxon rank-sum test, also known as the Mann-Whitney U test) is used to assess whether the distribution of marker abundance differs significantly between conditions. The Wilcoxon test is a non-parametric test that does not assume normal distribution of the data, making it robust for single-cell data which often exhibits non-normal distributions and high variability.

The test compares marker expression values (typically CLR-normalized values) between cells in the target condition versus cells in the reference condition. For each marker, the analysis reports an effect size (typically the difference in mean or median expression) and a p-value indicating the statistical significance of the difference. Multiple testing correction is applied using the false discovery rate (FDR) method to control for the large number of statistical tests performed across multiple markers and cell types.

Results are organized by cell type and include both the raw statistical outputs and summary visualizations. Markers with significant differences (typically p_adj < 0.05) and large effect sizes are of particular biological interest, as they represent markers whose expression is substantially altered by the experimental condition. This analysis enables identification of condition-specific markers, understanding of cellular responses to treatments, and discovery of biomarkers associated with different experimental states.

```{r}
#| label: statistical_testing_differential_abundance_analysis
#| eval: !expr isTRUE(run_cell_level)

differential_abundance_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      {
        daa_res <- RunDAA(subset(pg_data, cell_annotation == x),
          contrast_column = "condition",
          reference = reference_condition,
          features = markers_to_test[[x]]
        )
        daa_res$group <- x
        return(daa_res)
      },
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  adjust_global_fdr()

save_diff_results(
  differential_abundance_analysis,
  "abundance",
  "cell_level",
  "differential_abundance_analysis"
)
```    

### 5.1.1. Summarizing plots

Here we create summarizing plots for the differential abundance analysis results:

1. **Volcano plot**: Effect size vs −log10 adjusted p-value, faceted by cell type. The top markers labeled on each panel are selected by **adjusted significance** (not largest |effect|).

2. **Per–cell-type volcano + sample heatmap**: For each cell type, the volcano is exported beside a row-z-scored heatmap of sample-mean CLR for the labeled markers (same style as the pseudobulk path).

3. **Effect-size heatmap**: Significant markers (p_adj < `diff_p_adj_cutoff` and |effect| > `diff_abs_effect_min`) across cell types.

```{r}
#| label: statistical_testing_differential_abundance_plots
#| eval: !expr isTRUE(run_cell_level)

# Very small p-values are floored before −log10 so volcano axes stay finite
export_marker_diff_plots(
  differential_abundance_analysis,
  modality = "abundance",
  effect_col = "difference",
  feature_col = "marker",
  analysis_label = "Differential abundance analysis"
)

```

## 5.2. Differential Clustering Analysis (cell-level)

Differential clustering analysis identifies markers that exhibit statistically significant differences in spatial clustering patterns between experimental conditions. In this context, "clustering" refers to the spatial aggregation of a marker with itself—that is, the extent to which a marker tends to form clusters or aggregates within cells rather than being uniformly distributed.

The analysis is performed separately for each cell type, comparing clustering patterns between target and reference conditions. A statistical test (typically the Wilcoxon rank-sum test) is used to assess whether the distribution of clustering scores (log2 ratio of observed versus expected self-proximity) differs significantly between conditions. Multiple testing correction is applied using FDR to control for the number of statistical tests performed across markers and cell types.

```{r}
#| label: statistical_testing_differential_proximity
#| eval: !expr isTRUE(run_cell_level)

differential_clustering_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      run_cell_level_proximity_da(x, markers_to_test[[x]], pair_mode = "self"),
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  adjust_proximity_fdr()

save_diff_results(
  differential_clustering_analysis,
  "clustering",
  "cell_level",
  "differential_clustering_analysis"
)
```


### 5.2.1. Summarizing plots

Here we create summarizing plots for the differential clustering analysis results:

1. **Volcano plot**: Effect size vs −log10 adjusted p-value, faceted by cell type. Labeled markers are the top hits by **adjusted significance**.

2. **Per–cell-type volcano + sample heatmap**: Volcano beside a row-z-scored heatmap of sample-mean self-proximity (`log2_ratio`) for the labeled markers.

3. **Effect-size heatmap**: Significant markers across cell types.

```{r}
#| label: statistical_testing_differential_clustering_plots
#| eval: !expr isTRUE(run_cell_level)

export_marker_diff_plots(
  differential_clustering_analysis,
  modality = "clustering",
  effect_col = "diff_median",
  feature_col = "marker_1",
  analysis_label = "Differential clustering analysis"
)

```

## 5.3. Differential Colocalization Analysis (cell-level)

Differential colocalization analysis identifies markers that exhibit statistically significant differences in spatial proximity patterns between experimental conditions. The analysis is performed separately for each cell type, comparing colocalization patterns between target and reference conditions. A statistical test is used to assess whether the distribution of colocalization scores (log2 ratio of observed versus expected proximity) differs significantly between conditions. Multiple testing correction is applied using FDR to control for the number of statistical tests performed across markers and cell types.


```{r}
#| label: statistical_testing_differential_colocalization
#| eval: !expr isTRUE(run_cell_level)

differential_colocalization_analysis <-
  names(markers_to_test) |>
  map(function(x) {
    tryCatch(
      run_cell_level_proximity_da(x, markers_to_test[[x]], pair_mode = "all"),
      error = function(e) {
        message(paste("Error for cell type:", x, ":", e))
        return(NULL)
      }
    )
  }) |>
  bind_rows() |>
  adjust_proximity_fdr()

save_diff_results(
  differential_colocalization_analysis,
  "colocalization",
  "cell_level",
  "differential_colocalization_analysis"
)
```

### 5.3.1. Summarizing plots

Here we create summarizing plots for the differential colocalization analysis results:

1. **Volcano plot**: Effect size vs −log10 adjusted p-value for marker pairs. Labeled pairs are the top hits by **adjusted significance**.
2. **Per–cell-type volcano + sample heatmap**: Volcano beside a row-z-scored heatmap of sample-mean pair `log2_ratio` for the labeled pairs.
3. **Colocalization heatmap**: Difference in medians for marker pairs with p_adj < `diff_p_adj_cutoff` (dot size = −log10 adjusted p-value).

```{r}
#| label: statistical_testing_differential_colocalization_plot
#| eval: !expr isTRUE(run_cell_level)

export_colocalization_diff_plots(
  differential_colocalization_analysis,
  level = "cell_level",
  analysis_label = "Differential Colocalization Analysis",
  fill_lab = "Difference\nin medians",
  min_heatmap_pairs = 2L
)
```


## 5.4. Pseudobulk setup

Helpers and a replicate check for the optional limma path (sections 5.5–5.7).
If `run_pseudobulk` is FALSE, or there are fewer than two samples per condition,
those sections skip quietly. Clustering and colocalization share one sample-level
proximity table built at the start of section 5.6.

```{r}
#| label: statistical_testing_pseudobulk_helpers

# Check whether each condition has enough biological replicates for limma
check_pseudobulk_replicates <- function(meta_df, min_n = 2L) {
  counts <- meta_df |>
    distinct(.data$sample_alias, .data$condition) |>
    dplyr::count(.data$condition, name = "n_samples")
  list(ok = nrow(counts) > 0 && all(counts$n_samples >= min_n), counts = counts)
}

# Stop early if requested covariate columns are missing from metadata.csv
validate_pseudobulk_covariates <- function(covariate_cols, metadata) {
  if (is.null(covariate_cols) || length(covariate_cols) == 0) {
    return(invisible(NULL))
  }
  missing_cols <- setdiff(as.character(covariate_cols), colnames(metadata))
  if (length(missing_cols) > 0) {
    stop(
      "Missing covariate columns in metadata.csv: ",
      paste(missing_cols, collapse = ", ")
    )
  }
  invisible(NULL)
}

# Add covariate columns from metadata.csv onto the per-sample table for limma.
# Do not join condition here; it is already the contrast variable.
attach_pseudobulk_covariates <- function(sample_info, metadata, covariate_cols) {
  if (is.null(covariate_cols) || length(covariate_cols) == 0) {
    return(sample_info)
  }

  covariate_cols <- setdiff(
    as.character(covariate_cols),
    c("condition", "sample_alias")
  )
  if (length(covariate_cols) == 0) {
    return(sample_info)
  }

  out <- sample_info |>
    left_join(
      metadata |> select(sample_alias, all_of(covariate_cols)),
      by = "sample_alias"
    )

  if (!"condition" %in% names(out) && "condition.x" %in% names(out)) {
    out$condition <- out$condition.x
    out <- out |> select(-starts_with("condition."))
  }

  out
}

# List samples that have at least min_cells in this cell type
pseudobulk_keep_samples <- function(obj, min_cells) {
  obj[[]] |>
    as_tibble(rownames = "component") |>
    dplyr::count(.data$sample_alias, .data$condition, name = "n_cells") |>
    filter(n_cells >= min_cells)
}


# Mean proximity score per sample for one cell type (input to limma).
# Clustering keeps self-pairs (marker with itself); colocalization keeps
# pairs between different markers. Samples are kept separate so replicates
# are not pooled before testing.
pseudobulk_summarize_proximity <- function(
  obj,
  markers_vec,
  min_cells = pseudobulk_min_cells,
  group_vars = c("sample_alias", "condition"),
  background_threshold_pct = 0.001
) {
  markers_vec <- as.character(markers_vec)
  keep_samples <- pseudobulk_keep_samples(obj, min_cells)
  if (nrow(keep_samples) < 2L) {
    return(NULL)
  }

  proximity_data <- ProximityScores(
    obj,
    lazy = TRUE,
    meta_data_columns = group_vars,
    add_marker_proportions = TRUE
  ) |>
    filter(marker_1 %in% markers_vec) |>
    filter(marker_2 %in% markers_vec) |>
    filter(sample_alias %in% keep_samples$sample_alias) |>
    FilterProximityScores(background_threshold_pct = background_threshold_pct)

  SummarizeProximityScores(
    proximity_data,
    proximity_metric = "log2_ratio",
    group_vars = group_vars,
    summary_stat = "mean",
    include_missing_obs = TRUE
  ) |>
    mutate(feature = paste(marker_1, marker_2, sep = "__"))
}


# Convenience wrapper: save under the pseudobulk results folder
save_pseudobulk_results <- function(df, modality, basename) {
  save_diff_results(df, modality, "pseudobulk", basename)
}

# Fit limma on a features × samples matrix. Each non-reference condition is
# compared to the reference (optionally adjusting for covariates).
pseudobulk_limma_fit <- function(
  expr_mat,
  sample_info,
  reference_condition,
  covariate_cols = NULL
) {
  sample_info <- as.data.frame(sample_info, stringsAsFactors = FALSE)
  if (!"condition" %in% names(sample_info)) {
    stop(
      "sample_info has no condition column. Columns: ",
      paste(names(sample_info), collapse = ", ")
    )
  }
  sample_info$sample_alias <- as.character(sample_info[["sample_alias"]])
  sample_info$condition <- factor(as.character(sample_info[["condition"]]))
  stopifnot(identical(colnames(expr_mat), sample_info$sample_alias))
  if (!reference_condition %in% levels(sample_info$condition)) {
    stop("reference_condition not present in sample_info$condition")
  }
  sample_info$condition <- relevel(sample_info$condition, ref = reference_condition)

  # One coefficient per condition (no intercept) so contrasts are easy to write
  condition_design <- model.matrix(~ 0 + condition, data = sample_info)
  colnames(condition_design) <- levels(sample_info$condition)

  design <- condition_design
  if (!is.null(covariate_cols) && length(covariate_cols) > 0) {
    covariate_cols <- as.character(covariate_cols)
    missing_cols <- setdiff(covariate_cols, colnames(sample_info))
    if (length(missing_cols) > 0) {
      stop(
        "Missing pseudobulk covariate columns in sample_info: ",
        paste(missing_cols, collapse = ", ")
      )
    }

    # Drop the covariate intercept so the design is not over-parameterized
    covar_formula <- as.formula(paste0("~ ", paste(covariate_cols, collapse = " + ")))
    covar_design <- model.matrix(covar_formula, data = sample_info)[, -1, drop = FALSE]
    design <- cbind(condition_design, covar_design)
  }

  targets <- setdiff(levels(sample_info$condition), reference_condition)
  if (length(targets) == 0) {
    return(tibble())
  }
  # Not enough samples relative to model terms — skip this fit
  if (nrow(sample_info) <= ncol(design)) {
    return(tibble())
  }
  contrast_strings <- paste0(targets, "-", reference_condition)
  # Contrasts compare conditions only; covariate effects are held fixed
  contrast_matrix <- limma::makeContrasts(
    contrasts = contrast_strings,
    levels = design
  )
  fit <- limma::lmFit(expr_mat, design)
  fit2 <- limma::eBayes(limma::contrasts.fit(fit, contrast_matrix))
  map_dfr(seq_along(targets), function(i) {
    tt <- limma::topTable(
      fit2,
      coef = i,
      number = Inf,
      sort.by = "none"
    ) |>
      as_tibble(rownames = "feature")
    tt |>
      transmute(
        feature,
        difference = logFC,
        p = P.Value,
        target = targets[[i]],
        reference = reference_condition,
        method = "limma_pseudobulk"
      )
  })
}

# Build the expression matrix from sample-level means, then run limma
run_pseudobulk_limma <- function(
  agg,
  feature_col,
  value_col,
  reference_condition,
  metadata,
  covariate_cols = NULL,
  values_fill = NULL
) {
  sample_info <- agg |>
    distinct(.data$sample_alias, .data$condition) |>
    attach_pseudobulk_covariates(metadata, covariate_cols) |>
    arrange(.data$sample_alias)
  expr_mat <- agg_to_expr_mat(
    agg,
    feature_col,
    value_col,
    sample_info,
    values_fill = values_fill
  )
  list(
    res = pseudobulk_limma_fit(
      expr_mat,
      sample_info,
      reference_condition,
      covariate_cols = covariate_cols
    ),
    sample_info = sample_info,
    expr_mat = expr_mat
  )
}

# Limma on proximity pairs. Renames the effect column to diff_median so results
# match the cell-level schema used elsewhere (including module 07).
run_pseudobulk_proximity_limma <- function(agg, cell_type) {
  if (is.null(agg) || nrow(agg) == 0L) {
    return(NULL)
  }
  fit <- run_pseudobulk_limma(
    agg,
    feature_col = "feature",
    value_col = "mean_log2_ratio",
    reference_condition = reference_condition,
    metadata = metadata,
    covariate_cols = pseudobulk_covariate_cols,
    values_fill = 0
  )
  if (nrow(fit$res) == 0L) {
    return(NULL)
  }
  pair_map <- agg |> distinct(feature, marker_1, marker_2)
  fit$res |>
    left_join(pair_map, by = "feature") |>
    transmute(
      marker_1,
      marker_2,
      diff_median = difference,
      p,
      target,
      reference,
      method,
      group = cell_type
    )
}


# Run limma for all cell types, adjust FDR, and save. self_pairs = TRUE keeps
# marker–self scores (clustering); FALSE keeps different-marker pairs (coloc).
run_pseudobulk_proximity_modality <- function(
  summary_df,
  self_pairs,
  modality,
  basename
) {
  label <- if (isTRUE(self_pairs)) "clustering" else "coloc"
  differential <-
    names(markers_to_test) |>
    map(function(cell_type) {
      tryCatch(
        {
          agg <- summary_df |> filter(group == cell_type)
          if (isTRUE(self_pairs)) {
            agg <- agg |> filter(marker_1 == marker_2)
          } else {
            agg <- agg |> filter(marker_1 != marker_2)
          }
          run_pseudobulk_proximity_limma(agg, cell_type)
        },
        error = function(e) {
          message(paste("Pseudobulk", label, "error for", cell_type, ":", e$message))
          NULL
        }
      )
    }) |>
    bind_rows()

  if (nrow(differential) > 0) {
    differential <- adjust_global_fdr(differential)
  }
  save_diff_results(differential, modality, "pseudobulk", basename)
  differential
}

# Pseudobulk runs only when enabled and each condition has ≥2 samples
pseudobulk_rep_check <- check_pseudobulk_replicates(
  pg_data[[]],
  min_n = 2L
)
pseudobulk_ok <- isTRUE(run_pseudobulk) && isTRUE(pseudobulk_rep_check$ok)

if (isTRUE(run_pseudobulk) && !isTRUE(pseudobulk_rep_check$ok)) {
  message(
    "Skipping limma pseudobulk: need ≥2 sample_alias per condition. ",
    "Current counts:\n",
    paste(
      capture.output(print(pseudobulk_rep_check$counts)),
      collapse = "\n"
    )
  )
} else if (!isTRUE(run_pseudobulk)) {
  message("Skipping limma pseudobulk (run_pseudobulk = FALSE).")
}

if (isTRUE(pseudobulk_ok)) {
  validate_pseudobulk_covariates(pseudobulk_covariate_cols, metadata)
  suppressPackageStartupMessages(library(limma))
}
```

## 5.5. Differential abundance analysis (pseudobulk-level)

> **Decision checkpoint** (enable above: `run_pseudobulk`)  
> Pseudobulk aggregates CLR abundance to **mean per `sample_alias` × cell type**, then fits limma across samples. Needs ≥2 samples per `condition`. Sample-level tables are written as `summary_abundance_data.rds` (and `.csv`); limma results as `differential_abundance_analysis.rds`. Module 07 still reads the Wilcoxon results from §5.1 (`abundance/cell_level/`) unless you point it at `abundance/pseudobulk/`.

```{r}
#| label: statistical_testing_pseudobulk_abundance

if (isTRUE(pseudobulk_ok)) {
  # For each cell type: mean CLR per sample, then limma
  abundance_pseudobulk_parts <-
    names(markers_to_test) |>
    map(function(cell_type) {
      tryCatch(
        {
          markers_vec <- setdiff(
            as.character(unname(markers_to_test[[cell_type]])),
            c("sample_alias", "condition", "cell_id")
          )
          obj <- subset(pg_data, cell_annotation == cell_type)
          dat <- pseudobulk_fetch_abundance(obj, markers_vec)
          n_cells <- dat |>
            dplyr::count(.data$sample_alias, .data$condition, name = "n_cells") |>
            filter(n_cells >= pseudobulk_min_cells)
          if (nrow(n_cells) < 2) {
            return(NULL)
          }
          agg <- dat |>
            filter(.data$sample_alias %in% n_cells$sample_alias) |>
            pivot_longer(
              all_of(markers_vec),
              names_to = "marker",
              values_to = "clr"
            ) |>
            group_by(.data$sample_alias, .data$condition, .data$marker) |>
            summarise(mean_clr = mean(clr), .groups = "drop") |>
            mutate(group = cell_type)
          fit <- run_pseudobulk_limma(
            agg,
            feature_col = "marker",
            value_col = "mean_clr",
            reference_condition = reference_condition,
            metadata = metadata,
            covariate_cols = pseudobulk_covariate_cols
          )
          list(
            agg = agg,
            res = if (nrow(fit$res) == 0) {
              NULL
            } else {
              fit$res |>
                rename(marker = feature) |>
                mutate(group = cell_type)
            }
          )
        },
        error = function(e) {
          message(paste("Pseudobulk abundance error for", cell_type, ":", e$message))
          NULL
        }
      )
    })

  # Save the sample-level means used as limma input
  summary_abundance_data_pseudobulk <- abundance_pseudobulk_parts |>
    map("agg") |>
    bind_rows()
  if (nrow(summary_abundance_data_pseudobulk) > 0) {
    save_pseudobulk_results(
      summary_abundance_data_pseudobulk,
      "abundance",
      "summary_abundance_data"
    )
  }

  differential_abundance_analysis_pseudobulk <- abundance_pseudobulk_parts |>
    map("res") |>
    bind_rows()

  if (nrow(differential_abundance_analysis_pseudobulk) > 0) {
    differential_abundance_analysis_pseudobulk <-
      adjust_global_fdr(differential_abundance_analysis_pseudobulk)
  }

  save_diff_results(
    differential_abundance_analysis_pseudobulk,
    "abundance",
    "pseudobulk",
    "differential_abundance_analysis"
  )
}
```
### 5.5.1. Pseudobulk abundance plots

```{r}
#| label: statistical_testing_pseudobulk_abundance_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_abundance_analysis_pseudobulk") &&
    nrow(differential_abundance_analysis_pseudobulk) > 0
) {
  export_marker_diff_plots(
    differential_abundance_analysis_pseudobulk,
    modality = "abundance",
    effect_col = "difference",
    feature_col = "marker",
    analysis_label = "Pseudobulk differential abundance",
    level = "pseudobulk",
    volcano_subtitle = "limma on mean CLR per sample × cell type",
    append_volcano_label = FALSE,
    grey_guides_faceted = FALSE,
    expand_faceted = FALSE,
    grey_guides_cell = TRUE,
    expand_cell = TRUE
  )
}
```

## 5.6. Differential clustering analysis (pseudobulk-level)

Self-proximity (`marker_1 == marker_2`) from the shared sample-level table below, then limma. Column `diff_median` stores the limma logFC (mean difference) for schema compatibility with §5.2 / module 07.

The next chunk builds **one** proximity summary for both clustering and colocalization: `FilterProximityScores()` drops pairs below an isotype-derived background, then `SummarizeProximityScores()` takes the mean `log2_ratio` per `sample_alias` × cell type (missing cells as 0). That table is saved as `summary_proximity_data.rds` under both `clustering/pseudobulk/` and `colocalization/pseudobulk/`.

```{r}
#| label: statistical_testing_pseudobulk_proximity_summary

if (isTRUE(pseudobulk_ok)) {
  # Background threshold for weak pairs: median isotype fraction + 2 MAD
  # (minimum 0.001), matching the usual pixelatorR spatial-analysis rule
  if ("isotype_fraction" %in% colnames(pg_data[[]])) {
    pseudobulk_proximity_background_pct <- max(
      median(pg_data$isotype_fraction, na.rm = TRUE) +
        2 * mad(pg_data$isotype_fraction, na.rm = TRUE),
      0.001
    )
  } else {
    message(
      "isotype_fraction missing from object metadata; ",
      "using FilterProximityScores background_threshold_pct = 0.001"
    )
    pseudobulk_proximity_background_pct <- 0.001
  }
  message(
    "FilterProximityScores background_threshold_pct = ",
    signif(pseudobulk_proximity_background_pct, 3)
  )

  summary_proximity_data_pseudobulk <-
    names(markers_to_test) |>
    map(function(cell_type) {
      tryCatch(
        {
          agg <- pseudobulk_summarize_proximity(
            subset(pg_data, cell_annotation == cell_type),
            markers_to_test[[cell_type]],
            background_threshold_pct = pseudobulk_proximity_background_pct
          )
          if (is.null(agg)) {
            return(NULL)
          }
          agg |> mutate(group = cell_type)
        },
        error = function(e) {
          message(
            paste("Pseudobulk proximity summary error for", cell_type, ":", e$message)
          )
          NULL
        }
      )
    }) |>
    bind_rows()

  if (nrow(summary_proximity_data_pseudobulk) == 0) {
    message(
      "No sample-level proximity summaries; ",
      "clustering and colocalization pseudobulk tests will be skipped."
    )
  } else {
    # Same table is used for clustering and colocalization; save under both
    save_pseudobulk_results(
      summary_proximity_data_pseudobulk,
      "clustering",
      "summary_proximity_data"
    )
    save_pseudobulk_results(
      summary_proximity_data_pseudobulk,
      "colocalization",
      "summary_proximity_data"
    )
  }
}
```
```{r}
#| label: statistical_testing_pseudobulk_clustering

if (
  isTRUE(pseudobulk_ok) &&
    exists("summary_proximity_data_pseudobulk") &&
    nrow(summary_proximity_data_pseudobulk) > 0
) {
  differential_clustering_analysis_pseudobulk <-
    run_pseudobulk_proximity_modality(
      summary_proximity_data_pseudobulk,
      self_pairs = TRUE,
      modality = "clustering",
      basename = "differential_clustering_analysis"
    )
}
```

### 5.6.1. Pseudobulk clustering plots

```{r}
#| label: statistical_testing_pseudobulk_clustering_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_clustering_analysis_pseudobulk") &&
    nrow(differential_clustering_analysis_pseudobulk) > 0
) {
  export_marker_diff_plots(
    differential_clustering_analysis_pseudobulk,
    modality = "clustering",
    effect_col = "diff_median",
    feature_col = "marker_1",
    analysis_label = "Pseudobulk differential clustering",
    level = "pseudobulk",
    volcano_subtitle = "limma on mean log2_ratio per sample × cell type",
    append_volcano_label = FALSE,
    grey_guides_faceted = FALSE,
    expand_faceted = FALSE,
    grey_guides_cell = TRUE,
    expand_cell = TRUE,
    build_side_heatmap = build_pb_summary_clustering_heatmap
  )
}
```

## 5.7. Differential colocalization analysis (pseudobulk-level)

Off-diagonal pairs (`marker_1 != marker_2`) from the same sample-level table as §5.6, then limma. Same `diff_median` naming as §5.3 for schema compatibility. To use these results in module 07, point its coloc RDS path to `results/05_statistical_testing/colocalization/pseudobulk/differential_colocalization_analysis.rds` (and the abundance path similarly under `abundance/pseudobulk/`).

```{r}
#| label: statistical_testing_pseudobulk_colocalization

if (
  isTRUE(pseudobulk_ok) &&
    exists("summary_proximity_data_pseudobulk") &&
    nrow(summary_proximity_data_pseudobulk) > 0
) {
  differential_colocalization_analysis_pseudobulk <-
    run_pseudobulk_proximity_modality(
      summary_proximity_data_pseudobulk,
      self_pairs = FALSE,
      modality = "colocalization",
      basename = "differential_colocalization_analysis"
    )
}
```

### 5.7.1. Pseudobulk colocalization plots

```{r}
#| label: statistical_testing_pseudobulk_colocalization_plots

if (
  isTRUE(pseudobulk_ok) &&
    exists("differential_colocalization_analysis_pseudobulk") &&
    nrow(differential_colocalization_analysis_pseudobulk) > 0
) {
  export_colocalization_diff_plots(
    differential_colocalization_analysis_pseudobulk,
    level = "pseudobulk",
    analysis_label = "Pseudobulk differential colocalization",
    volcano_subtitle = "limma on mean log2_ratio per sample (pairs)",
    fill_lab = "Difference\n(mean)",
    append_volcano_label = FALSE,
    grey_guides_faceted = TRUE,
    expand_faceted = FALSE,
    expand_cell = FALSE,
    min_heatmap_pairs = 1L,
    include_cross_ct_heatmap = TRUE,
    build_side_heatmap = build_pb_summary_coloc_heatmap
  )
}
```
