# 6. Cell visualization

**Module 06 — Cell visualization.** 2D and 3D single-cell graph plots using
pixelatorR's `Plot2DGraphM` and `Plot3DGraph`. This shows *where* proteins sit on a
single cell's surface graph — clustering, polarization and colocalization patterns
that population-level scores (module `04`) summarise but don't show directly. Run
after module `02`. See the [cell visualization tutorial](https://software.pixelgen.com/pna-analysis/R/tutorials/cell_visualization/)
for more detail.

```{r}
#| label: module06_setup

library(here)
here::i_am("modules/06_cell_visualization.qmd")

source(here::here("modules", "common_setup.R"))

if (!exists("pg_data", inherits = TRUE)) {
  pg_data <- load_checkpoint("annotated_seurat_object.rds")
}
```

## 6.1. Parameters

> **Decision checkpoint** — set the cell type and markers to plot. The example uses
> CD8 T cells and the CD82/CD81 pair, which colocalize on the cell surface.

```{r}
#| label: cell_viz_parameters

cell_type_of_interest <- "CD8 T"            # one value from `cell_annotation`
markers_to_viz        <- c("CD82", "CD81")  # markers to render on the graph
n_cells               <- 4                   # cells to show in the 2D grid
```

## 6.2. Select cells

How you pick cells depends on what you want to show — the shape of typical cells, a
polarized marker, a colocalization pattern, or a cell–cell conjugate. Below we just
take a few cells of one type; change the selection to suit your question.

```{r}
#| label: cell_viz_select_cells

cells_of_interest <-
  WhichCells(pg_data, expression = cell_annotation == cell_type_of_interest) |>
  head(n_cells)

if (length(cells_of_interest) == 0) {
  stop("No cells found for '", cell_type_of_interest,
       "'. Check the values in `cell_annotation`.")
}
```

## 6.3. Load graphs and compute layouts

The 2D and 3D plots use different layout methods, so we compute both: `wpmds` for
the 2D grid and `cpmds` for the interactive 3D view.

```{r}
#| label: cell_viz_layout

pg_data <- LoadCellGraphs(pg_data, cells = cells_of_interest,
                          add_layouts = TRUE, verbose = FALSE)
pg_data <- ComputeLayout(pg_data, layout_method = "wpmds")  # for the 2D grid
pg_data <- ComputeLayout(pg_data, layout_method = "cpmds")  # for the 3D view
```

## 6.4. 2D graph (markers × cells)

`Plot2DGraphM` draws one row per marker and one column per cell, coloured by marker
abundance. Colocalizing markers light up the same regions across cells.

```{r}
#| label: cell_viz_2d

p_2d <- Plot2DGraphM(
  pg_data,
  cells         = cells_of_interest,
  markers       = markers_to_viz,
  layout_method = "wpmds_3d"
)
p_2d

export_plot(
  filename = here::here("results/06_cell_visualization",
                        paste0(cell_type_of_interest, "_2d_graph")),
  plot     = p_2d,
  width    = length(cells_of_interest) * 3 + 1,
  height   = length(markers_to_viz) * 3 + 0.5
)
```

## 6.5. Interactive 3D graph (one cell)

`Plot3DGraph` renders an interactive 3D view of one cell for one marker. Change
`cell_id` and `marker` to explore others.

```{r}
#| label: cell_viz_3d

Plot3DGraph(
  pg_data,
  cell_id       = cells_of_interest[1],
  marker        = markers_to_viz[1],
  layout_method = "cpmds_3d"
)
```
