# 7. Data interpretation

**Module 07 — Data interpretation (optional).** Map differential markers and colocalization partners to UniProt IDs from the PNA assay metadata and run panel-aware term enrichment to aid biological interpretation. Skip this module if you do not need enrichment or annotation context. Run after module `05` has produced differential abundance and colocalization results. Requires PXL / panel data that include a `uniprot_id` column in the PNA assay feature metadata (e.g. Proxiome Immuno 155 v2).

> **Note:** Overrepresentation analysis on a ~155-protein panel has very low statistical power. Treat enrichment tables as exploratory context, not definitive pathway calls.

```{r}
#| label: module07_setup

library(here)
here::i_am("modules/07_data_interpretation.qmd")

source(here::here("modules", "common_setup.R"))

if (!exists("pg_data", inherits = TRUE)) {
  pg_data <- load_checkpoint("annotated_seurat_object.rds")
}

dir.create(
  here::here("results", "07_data_interpretation"),
  recursive = TRUE,
  showWarnings = FALSE
)
```

## 7.0. Set interpretation parameters

> **Decision checkpoint** (significance cutoffs)  
> Adjust `p_adj` and effect-size thresholds for markers from module `05` differential abundance and colocalization. Protein sets are built for all cell types and contrasts present in those tables.

```{r}
#| label: module07_parameters

daa_p_adj_threshold <- 0.05
daa_effect_size_threshold <- 0 # minimum |difference|; 0 keeps all significant markers

coloc_p_adj_threshold <- 0.05
coloc_effect_size_threshold <- 0 # minimum |diff_median|; 0 keeps all significant pairs

# Used in a later chunk for term enrichment
ontology <- "GO_BP"
```

## 7.1. Load module 05 results

```{r}
#| label: module07_load_module05_results

daa_path <- here::here(
  "results/05_statistical_testing/abundance/differential_abundance_analysis.rds"
)
coloc_path <- here::here(
  "results/05_statistical_testing/colocalization/differential_colocalization_analysis.rds"
)

if (!file.exists(daa_path) || !file.exists(coloc_path)) {
  stop(
    "Module 05 results not found. Run statistical testing first.\n",
    "Expected:\n  ", daa_path, "\n  ", coloc_path
  )
}

if (!exists("differential_abundance_analysis", inherits = TRUE)) {
  differential_abundance_analysis <- readRDS(daa_path)
}
if (!exists("differential_colocalization_analysis", inherits = TRUE)) {
  differential_colocalization_analysis <- readRDS(coloc_path)
}
```

## 7.2. Build protein sets

Significant markers from differential abundance and unique markers from significant colocalization pairs are collected per cell type and contrast. These sets feed the UniProt mapping and enrichment steps below.

```{r}
#| label: module07_build_protein_sets

daa_sets <-
  differential_abundance_analysis |>
  filter(
    p_adj < daa_p_adj_threshold,
    abs(difference) >= daa_effect_size_threshold
  ) |>
  group_by(source = "differential_abundance", group, target) |>
  summarise(
    markers = list(sort(unique(marker))),
    n_markers = n_distinct(marker),
    .groups = "drop"
  )

coloc_sets <-
  differential_colocalization_analysis |>
  filter(
    p_adj < coloc_p_adj_threshold,
    abs(diff_median) >= coloc_effect_size_threshold
  ) |>
  group_by(source = "differential_colocalization", group, target) |>
  summarise(
    markers = list(sort(unique(c(marker_1, marker_2)))),
    .groups = "drop"
  ) |>
  mutate(n_markers = map_int(markers, length))

protein_sets <-
  bind_rows(daa_sets, coloc_sets) |>
  filter(n_markers > 0)

if (nrow(protein_sets) == 0) {
  warning(
    "No protein sets passed the current thresholds. ",
    "Relax p_adj / effect-size cutoffs or check module 05 outputs."
  )
} else {
  print(
    protein_sets |>
      select(source, group, target, n_markers) |>
      arrange(source, group, target)
  )
}
```

## 7.3. Map markers to UniProt IDs

This step reads `uniprot_id` from the PNA assay feature table for every panel marker. When a marker maps to several accessions (semicolon-separated), each accession becomes its own row. The protein sets from section 7.2 are then annotated with those UniProt IDs. The set of all panel UniProt IDs is kept as `panel_uniprot_universe` for enrichment.

```{r}
#| label: module07_id_mapping

pna_feature_meta <- pg_data@assays$PNA@meta.data
if (!"uniprot_id" %in% colnames(pna_feature_meta)) {
  stop(
    "Column 'uniprot_id' not found in pg_data@assays$PNA@meta.data.\n",
    "Use a PXL / panel that includes UniProt IDs (e.g. Proxiome Immuno 155 v2)."
  )
}

marker_uniprot_map <-
  pna_feature_meta |>
  as_tibble(rownames = "marker") |>
  select(marker, uniprot_id) |>
  mutate(uniprot_id = na_if(str_trim(as.character(uniprot_id)), "")) |>
  separate_rows(uniprot_id, sep = ";") |>
  mutate(uniprot_id = str_trim(uniprot_id)) |>
  filter(!is.na(uniprot_id), uniprot_id != "")

panel_uniprot_universe <- sort(unique(marker_uniprot_map$uniprot_id))

protein_sets <-
  protein_sets |>
  mutate(
    uniprot_ids = map(
      markers,
      \(m) {
        marker_uniprot_map |>
          filter(marker %in% m) |>
          pull(uniprot_id) |>
          unique() |>
          sort()
      }
    ),
    n_uniprot = map_int(uniprot_ids, length)
  )

message(
  "Mapped ", n_distinct(marker_uniprot_map$marker), " markers to ",
  length(panel_uniprot_universe), " UniProt IDs"
)
if (nrow(protein_sets) > 0) {
  print(
    protein_sets |>
      select(source, group, target, n_markers, n_uniprot) |>
      arrange(source, group, target)
  )
}

write_csv(
  protein_sets |>
    mutate(
      markers = map_chr(markers, paste, collapse = ";"),
      uniprot_ids = map_chr(uniprot_ids, paste, collapse = ";")
    ) |>
    select(source, group, target, n_markers, n_uniprot, markers, uniprot_ids),
  here::here("results/07_data_interpretation/protein_sets.csv")
)
```
