# 7. Data interpretation

**Module 07 — Data interpretation (optional).** Map differential markers and colocalization partners to UniProt IDs from the PNA assay metadata and run panel-aware term enrichment to aid biological interpretation. Skip this module if you do not need enrichment or annotation context. Run after module `05` has produced differential abundance and colocalization results. Requires PXL / panel data that include a `uniprot_id` column in the PNA assay feature metadata (e.g. Proxiome Immuno 155 v2).

> **Note:** Overrepresentation analysis on a ~155-protein panel has very low statistical power. Treat enrichment tables as exploratory context, not definitive pathway calls.

```{r}
#| label: module07_setup

library(here)
here::i_am("modules/07_data_interpretation.qmd")

source(here::here("modules", "common_setup.R"))

if (!exists("pg_data", inherits = TRUE)) {
  pg_data <- load_checkpoint("annotated_seurat_object.rds")
}

dir.create(
  here::here("results", "07_data_interpretation"),
  recursive = TRUE,
  showWarnings = FALSE
)
```

## 7.0. Set interpretation parameters

> **Decision checkpoint** (significance cutoffs)  
> Adjust `p_adj` and effect-size thresholds for markers from module `05` differential abundance and colocalization. Protein sets are built for all cell types and contrasts present in those tables.

```{r}
#| label: module07_parameters

daa_p_adj_threshold <- 0.05
daa_effect_size_threshold <- 0.5 # minimum |difference|; 0 keeps all significant markers

coloc_p_adj_threshold <- 0.05
coloc_effect_size_threshold <- 0.25 # minimum |diff_median|; 0 keeps all significant pairs
```

## 7.1. Load module 05 results

```{r}
#| label: module07_load_module05_results

daa_path <- here::here(
  "results/05_statistical_testing/abundance/differential_abundance_analysis.rds"
)
coloc_path <- here::here(
  "results/05_statistical_testing/colocalization/differential_colocalization_analysis.rds"
)

if (!file.exists(daa_path) || !file.exists(coloc_path)) {
  stop(
    "Module 05 results not found. Run statistical testing first.\n",
    "Expected:\n  ", daa_path, "\n  ", coloc_path
  )
}

if (!exists("differential_abundance_analysis", inherits = TRUE)) {
  differential_abundance_analysis <- readRDS(daa_path)
}
if (!exists("differential_colocalization_analysis", inherits = TRUE)) {
  differential_colocalization_analysis <- readRDS(coloc_path)
}
```

## 7.2. Build protein sets

Significant markers from differential abundance and unique markers from significant colocalization pairs are collected per cell type, contrast, and direction of change. Positive effect sizes (`difference` / `diff_median` > 0) are labeled `up` (higher in the target condition); negative values are labeled `down`. Enrichment is run separately for each direction.

```{r}
#| label: module07_build_protein_sets

daa_sets <-
  differential_abundance_analysis |>
  filter(
    p_adj < daa_p_adj_threshold,
    abs(difference) >= daa_effect_size_threshold,
    difference != 0
  ) |>
  mutate(direction = if_else(difference > 0, "up", "down")) |>
  group_by(source = "differential_abundance", group, target, direction) |>
  summarise(
    markers = list(sort(unique(marker))),
    n_markers = n_distinct(marker),
    .groups = "drop"
  )

coloc_sets <-
  differential_colocalization_analysis |>
  filter(
    p_adj < coloc_p_adj_threshold,
    abs(diff_median) >= coloc_effect_size_threshold,
    diff_median != 0
  ) |>
  mutate(direction = if_else(diff_median > 0, "up", "down")) |>
  group_by(source = "differential_colocalization", group, target, direction) |>
  summarise(
    markers = list(sort(unique(c(marker_1, marker_2)))),
    .groups = "drop"
  ) |>
  mutate(n_markers = map_int(markers, length))

protein_sets <-
  bind_rows(daa_sets, coloc_sets) |>
  filter(n_markers > 0)

if (nrow(protein_sets) == 0) {
  warning(
    "No protein sets passed the current thresholds. ",
    "Relax p_adj / effect-size cutoffs or check module 05 outputs."
  )
} else {
  print(
    protein_sets |>
      select(source, group, target, direction, n_markers) |>
      arrange(source, group, target, direction)
  )
}
```

## 7.3. Map markers to UniProt IDs

This step reads `uniprot_id` from the PNA assay feature metadata for every panel marker. When a marker maps to several accessions (semicolon-separated), each accession becomes its own row. The protein sets from section 7.2 are then annotated with those UniProt IDs. The set of all panel UniProt IDs is kept as `panel_uniprot_universe` for enrichment.

```{r}
#| label: module07_id_mapping

pna_feature_meta <- pg_data[["PNA"]][[]]
if (!"uniprot_id" %in% colnames(pna_feature_meta)) {
  stop(
    "Column 'uniprot_id' not found in PNA assay feature metadata.\n",
    "Use a PXL / panel that includes UniProt IDs (e.g. Proxiome Immuno 155 v2)."
  )
}

# Use the panel 'marker' column (meta rownames are not reliable on PNAAssay5).
# Seurat stores feature names with '-' instead of '_'.
marker_uniprot_map <-
  pna_feature_meta |>
  as_tibble(rownames = "marker") |>
  select(marker, uniprot_id) |>
  separate_rows(uniprot_id, sep = ";")

panel_uniprot_universe <- sort(unique(marker_uniprot_map$uniprot_id))

protein_sets <-
  protein_sets |>
  mutate(
    uniprot_ids = map(
      markers,
      \(m) {
        marker_uniprot_map |>
          filter(marker %in% m) |>
          pull(uniprot_id) |>
          unique() |>
          sort()
      }
    ),
    n_uniprot = map_int(uniprot_ids, length)
  )

message(
  "Mapped ", n_distinct(marker_uniprot_map$marker), " markers to ",
  length(panel_uniprot_universe), " UniProt IDs"
)
if (nrow(protein_sets) > 0) {
  print(
    protein_sets |>
      select(source, group, target, direction, n_markers, n_uniprot) |>
      arrange(source, group, target, direction)
  )
}

write_csv(
  protein_sets |>
    mutate(
      markers = map_chr(markers, paste, collapse = ";"),
      uniprot_ids = map_chr(uniprot_ids, paste, collapse = ";")
    ) |>
    select(
      source, group, target, direction,
      n_markers, n_uniprot, markers, uniprot_ids
    ),
  here::here("results/07_data_interpretation/protein_sets.csv")
)
```

## 7.4. GO Biological Process enrichment

For each protein set (cell type × contrast × direction), UniProt IDs are tested for overrepresentation in GO Biological Process terms against the panel UniProt universe using a Fisher / hypergeometric test (`clusterProfiler::enricher`).

Plots: a term overview bar chart, then a term × marker map where each point marks membership, colored by condition effect size and sized by significance.

```{r}
#| label: module07_enrichment

top_n_terms <- 10

safe_plot_stem <- function(...) {
  gsub("[^A-Za-z0-9._-]+", "_", paste(..., sep = "__"))
}

set_label <- function(group, target, source, direction) {
  paste(group, target, source, direction, sep = " | ")
}

go_ann <-
  suppressMessages(
    AnnotationDbi::select(
      org.Hs.eg.db::org.Hs.eg.db,
      keys = panel_uniprot_universe,
      columns = c("GO", "ONTOLOGY"),
      keytype = "UNIPROT"
    )
  ) |>
  as_tibble() |>
  filter(ONTOLOGY == "BP", !is.na(GO)) |>
  distinct(term = GO, gene = UNIPROT)

term2name <-
  clusterProfiler::go2term(unique(go_ann$term)) |>
  as_tibble() |>
  rename(term = go_id, name = Term)

enrich_one_set <- function(ids) {
  if (length(ids) < 2) {
    return(NULL)
  }
  res <- tryCatch(
    clusterProfiler::enricher(
      gene = ids,
      universe = panel_uniprot_universe,
      TERM2GENE = go_ann,
      TERM2NAME = term2name,
      pvalueCutoff = 0.05,
      qvalueCutoff = 0.2,
      minGSSize = 2,
      maxGSSize = 500
    ),
    error = function(e) NULL
  )
  if (is.null(res) || nrow(res) == 0) {
    return(NULL)
  }
  as_tibble(as.data.frame(res))
}

enrichment_results <-
  protein_sets |>
  filter(n_uniprot >= 2) |>
  mutate(enrich = map(uniprot_ids, enrich_one_set)) |>
  filter(!map_lgl(enrich, is.null)) |>
  select(source, group, target, direction, enrich) |>
  unnest(enrich)

write_csv(
  enrichment_results,
  here::here("results/07_data_interpretation/enrichment_results.csv")
)

if (nrow(enrichment_results) == 0) {
  term_marker_hits <- tibble(
    source = character(),
    group = character(),
    target = character(),
    direction = character(),
    ID = character(),
    Description = character(),
    p.adjust = double(),
    Count = integer(),
    uniprot_id = character(),
    marker = character()
  )
  message("No enriched GO BP terms found for the current protein sets.")
} else {
  set_markers_long <-
    protein_sets |>
    select(source, group, target, direction, markers) |>
    unnest_longer(markers, values_to = "marker")

  term_marker_hits <-
    enrichment_results |>
    select(
      source, group, target, direction,
      ID, Description, p.adjust, Count, geneID
    ) |>
    separate_rows(geneID, sep = "/") |>
    rename(uniprot_id = geneID) |>
    left_join(
      marker_uniprot_map |> distinct(uniprot_id, marker),
      by = "uniprot_id",
      relationship = "many-to-many"
    ) |>
    # Keep only markers that belonged to this directed protein set (avoids
    # pulling opposite-direction markers via shared UniProt IDs).
    inner_join(
      set_markers_long,
      by = c("source", "group", "target", "direction", "marker")
    )

  print(
    enrichment_results |>
      select(
        source, group, target, direction,
        ID, Description, Count, p.adjust
      ) |>
      arrange(p.adjust) |>
      head(20)
  )
}

write_csv(
  term_marker_hits,
  here::here("results/07_data_interpretation/term_marker_hits.csv")
)
```

### 7.4.1. Term overview

Horizontal bars of the top enriched GO BP terms per protein set (`-log10` adjusted *p*).

```{r}
#| label: module07_viz_term_overview

if (nrow(enrichment_results) == 0) {
  message("Skipping term-overview plots: no enrichment results.")
} else {
  enrichment_plot_dir <- here::here(
    "results/07_data_interpretation/go_bp_enrichment"
  )
  dir.create(enrichment_plot_dir, recursive = TRUE, showWarnings = FALSE)

  enrichment_sets <-
    enrichment_results |>
    distinct(source, group, target, direction)

  for (i in seq_len(nrow(enrichment_sets))) {
    set_source <- enrichment_sets$source[[i]]
    set_group <- enrichment_sets$group[[i]]
    set_target <- enrichment_sets$target[[i]]
    set_direction <- enrichment_sets$direction[[i]]

    set_plot_data <-
      enrichment_results |>
      filter(
        source == set_source,
        group == set_group,
        target == set_target,
        direction == set_direction
      ) |>
      slice_min(order_by = p.adjust, n = top_n_terms, with_ties = FALSE)

    enrichment_plot <-
      set_plot_data |>
      ggplot(aes(x = -log10(p.adjust), y = reorder(Description, p.adjust))) +
      geom_col() +
      labs(
        title = "GO Biological Process enrichment",
        subtitle = set_label(set_group, set_target, set_source, set_direction),
        x = expression(-log[10](adjusted ~ p)),
        y = NULL
      ) +
      theme_minimal()

    export_plot(
      enrichment_plot,
      filename = file.path(
        enrichment_plot_dir,
        safe_plot_stem(set_source, set_group, set_target, set_direction)
      ),
      width = 8,
      height = max(4, nrow(set_plot_data) * 0.35 + 1.5)
    )
  }
}
```

### 7.4.2. Term × marker map

Point map of top-term membership: a point means the marker belongs to the term; color is the condition effect size and size is `-log10(p_adj)` from module 05 (abundance `difference` for DAA sets; mean pair `diff_median` for coloc sets).

```{r}
#| label: module07_viz_term_marker

if (nrow(term_marker_hits) == 0) {
  message("Skipping term-marker maps: no term membership rows.")
} else {
  term_marker_plot_dir <- here::here(
    "results/07_data_interpretation/go_bp_term_marker"
  )
  dir.create(term_marker_plot_dir, recursive = TRUE, showWarnings = FALSE)

  enrichment_sets <-
    enrichment_results |>
    distinct(source, group, target, direction)

  term_marker_plot_sets <- list()

  for (i in seq_len(nrow(enrichment_sets))) {
    set_source <- enrichment_sets$source[[i]]
    set_group <- enrichment_sets$group[[i]]
    set_target <- enrichment_sets$target[[i]]
    set_direction <- enrichment_sets$direction[[i]]

    top_terms <-
      enrichment_results |>
      filter(
        source == set_source,
        group == set_group,
        target == set_target,
        direction == set_direction
      ) |>
      slice_min(order_by = p.adjust, n = top_n_terms, with_ties = FALSE)

    set_hits <-
      term_marker_hits |>
      filter(
        source == set_source,
        group == set_group,
        target == set_target,
        direction == set_direction,
        ID %in% top_terms$ID,
        !is.na(marker)
      ) |>
      distinct(Description, marker)

    if (nrow(set_hits) == 0) {
      next
    }

    if (identical(set_source, "differential_abundance")) {
      marker_effects <-
        differential_abundance_analysis |>
        filter(
          group == set_group,
          target == set_target,
          (set_direction == "up" & difference > 0) |
            (set_direction == "down" & difference < 0)
        ) |>
        select(marker, effect = difference, p_adj)
      effect_legend <- "Difference"
    } else {
      marker_effects <-
        differential_colocalization_analysis |>
        filter(
          group == set_group,
          target == set_target,
          (set_direction == "up" & diff_median > 0) |
            (set_direction == "down" & diff_median < 0)
        ) |>
        select(marker_1, marker_2, diff_median, p_adj) |>
        pivot_longer(
          c(marker_1, marker_2),
          names_to = NULL,
          values_to = "marker"
        ) |>
        group_by(marker) |>
        summarise(
          effect = mean(diff_median),
          p_adj = min(p_adj),
          .groups = "drop"
        )
      effect_legend <- "Mean diff_median"
    }

    plot_data <-
      set_hits |>
      inner_join(marker_effects, by = "marker") |>
      mutate(
        neg_log10_p_adj = -log10(pmax(p_adj, min_p_value_threshold)),
        Description = factor(Description, levels = rev(top_terms$Description)),
        marker = factor(marker, levels = sort(unique(marker)))
      )

    if (nrow(plot_data) == 0) {
      next
    }

    term_marker_plot_sets[[length(term_marker_plot_sets) + 1]] <- list(
      source = set_source,
      group = set_group,
      target = set_target,
      direction = set_direction,
      effect_legend = effect_legend,
      plot_data = plot_data
    )
  }

  if (length(term_marker_plot_sets) == 0) {
    message("Skipping term-marker maps: no plottable membership rows.")
  } else {
    all_plot_data <- map_dfr(term_marker_plot_sets, "plot_data")

    effect_limit <- max(abs(all_plot_data$effect), na.rm = TRUE)
    if (!is.finite(effect_limit) || effect_limit == 0) {
      effect_limit <- 1
    }
    color_limits <- c(-effect_limit, effect_limit)

    size_max <- max(all_plot_data$neg_log10_p_adj, na.rm = TRUE)
    if (!is.finite(size_max) || size_max <= 0) {
      size_max <- 1
    }
    size_limits <- c(0, size_max)

    for (plot_set in term_marker_plot_sets) {
      plot_data <- plot_set$plot_data

      term_marker_plot <-
        plot_data |>
        ggplot(aes(
          x = marker,
          y = Description,
          color = effect,
          size = neg_log10_p_adj
        )) +
        geom_point() +
        scale_color_gradientn(
          colors = gradient_palette,
          limits = color_limits
        ) +
        scale_size_continuous(
          range = c(1.5, 6),
          limits = size_limits
        ) +
        labs(
          title = "GO term × marker map",
          subtitle = set_label(
            plot_set$group,
            plot_set$target,
            plot_set$source,
            plot_set$direction
          ),
          x = NULL,
          y = NULL,
          color = plot_set$effect_legend,
          size = expression(-log[10](p[adj]))
        ) +
        theme_minimal() +
        theme(
          axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
          panel.grid.major = element_line(color = "grey90")
        )

      export_plot(
        term_marker_plot,
        filename = file.path(
          term_marker_plot_dir,
          safe_plot_stem(
            plot_set$source,
            plot_set$group,
            plot_set$target,
            plot_set$direction
          )
        ),
        width = min(15, n_distinct(plot_data$marker) * 0.35 + 8),
        height = max(4, n_distinct(plot_data$Description) * 0.35 + 2)
      )
    }
  }
}
```
