---
title: "Single Cell PNA Analysis SOP"
subtitle: "v0.1.0" 
format: 
    html:
      toc: true
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
execute:
  dev: ragg_png
  eval: false
  execute-dir: dir
  echo: true
  warning: false
  message: false
---

This SOP describes a standard workflow for analyzing single cell PNA data using the `pixelatorR` package in R. It covers data setup, quality control, data processing, and statistical testing. The workflow is interactive and modular, requiring manual intervention at several key steps. 

**Important:** Users must actively review and adjust quality control thresholds, select markers, and most critically, perform manual cell type annotation based on marker expression patterns. Automated annotation provides a starting point but should always be verified and refined through careful inspection of differential abundance patterns and marker distributions.

## Project Folder Structure

This analysis workflow creates the following folder structure for organizing input data and output results:

```
.
├── data/                                    # Input data directory
│   ├── metadata.csv                         # Sample metadata file (required)
│   └── *.pxl                                # PNA data files (one per sample)
│
├── scripts/                                 # Analysis scripts
│   └── helpers.R                            # Helper functions for plotting and data processing
│
└── results/                                 # Output results directory
    ├── 01_quality_control/                   # Section 1: Quality control metrics and plots
    │   ├── violin_plots/                     # Violin plots for QC metrics (read depth, k-core, etc.)
    │   ├── molecule_rank_plots/              # Molecule rank plots with thresholds
    │   ├── tau_plots/                        # Tau score plots for outlier detection
    │   └── marker_selection_plots/           # Marker selection plots (CPM-based filtering)
    │
    ├── 02_data_processing/                   # Section 2: Data processing outputs
    │   ├── dimensionality_reduction_plots/    # PCA, UMAP, and Harmony plots
    │   ├── clustering_plots/                  # Cluster visualization plots
    │   ├── marker_umap_plots/                 # Individual marker UMAP plots
    │   └── annotation_plots/                  # Cell type annotation plots and tables
    │       ├── automatic_annotation/          # Automatic annotation visualizations (L1, L2)
    │       ├── cell_type_composition/         # Cell type composition plots per sample
    │       ├── cluster_specific_markers/       # Cluster-specific marker analysis
    │       │   └── data_processing_differential_abundance_markers_per_cluster.csv
    │       ├── gating_strategy/                # Marker gating strategy plots
    │       │   ├── cd4_subtype_naive_memory_density_scatter
    │       │   ├── cd4_subtype_naive_memory_density_scatter_w_gates
    │       │   └── cd4_subtype_umap
    │       └── annotated_seurat_object.rds    # Saved annotated Seurat object
    │
    ├── 03_abundance/                          # Section 3: Abundance distribution analysis
    │   └── abundance_plots/
    │       ├── per_sample/                    # Abundance plots per sample
    │       ├── per_cell_type/                 # Abundance plots per cell type
    │       ├── per_cell_type_faceted_by_sample/  # Faceted plots by sample
    │       └── per_condition_faceted_by_celltype/ # Faceted plots by condition
    │
    ├── 04_raw_proximity/                      # Section 4: Colocalization score analysis
    │   ├── raw_colocalization_heatmap_per_sample/      # Heatmaps per sample
    │   ├── raw_colocalization_heatmap_per_cell_type/   # Heatmaps per cell type
    │   │   └── {cell_type}/                  # Subdirectories for each cell type
    │   └── colocalization_score_network/      # Network visualizations
    │       ├── colocalization_score_network_all_cell_types/
    │       └── colocalization_score_network_per_cell_type/
    │           └── {cell_type_of_interest}/
    │
    └── 05_statistical_testing/                # Section 5: Statistical testing results
        ├── abundance/                         # Differential abundance analysis
        │   ├── differential_abundance_analysis.csv
        │   ├── differential_abundance_analysis.rds
        │   └── differential_abundance_plots/
        │       └── {condition}_vs_{reference}/ # Comparison-specific plots
        │
        ├── clustering/                        # Differential clustering analysis
        │   ├── differential_clustering_analysis.csv
        │   ├── differential_clustering_analysis.rds
        │   └── differential_clustering_plots/
        │       └── {condition}_vs_{reference}/
        │
        └── colocalization/                    # Differential colocalization analysis
            ├── differential_colocalization_analysis.csv
            ├── differential_colocalization_analysis.rds
            └── differential_colocalization_plots/
                └── {condition}_vs_{reference}/
                    └── {cell_type}/          # Plots per cell type
```

**Note:** 

- The `{condition}`, `{reference}`, `{cell_type}`, and `{cell_type_of_interest}` placeholders represent dynamically created subdirectories based on your experimental conditions and cell types.

- All plots are saved in both PDF and PNG formats for easy viewing and publication.

- Folder numbering (01-05) corresponds to the analysis sections, ensuring consistent organization throughout the workflow.

# 0. Setup

## 0.1. Load packages and helper functions

A set of commonly used packages and helper functions are loaded here. The core packages include `tidyverse` for data manipulation and visualization, `Seurat` for single-cell analysis infrastructure, `pixelatorR` for PNA-specific analysis functions, `harmony` for batch correction and data harmonization, `ComplexHeatmap` for advanced heatmap visualizations, and `here` for robust file path management. Additionally, custom helper functions are sourced from `scripts/helpers.R` to provide reusable plotting and data processing utilities. These packages and functions are used throughout the analysis workflow. Users can customize the helper functions in `scripts/helpers.R` to adapt them to their specific needs or add additional utility functions as required.

```{r}
#| label: setup_packages

# Load commonly used packages
library(tidyverse)
library(Seurat)
library(pixelatorR)
library(harmony)
library(ComplexHeatmap)
library(ggplotify)
library(here)

# Anchor the project root directory
here::i_am("single_cell_analysis_sop.qmd")

# Load helper functions
source("scripts/helpers.R")

```

## 0.2. Create the PNA object

### 0.2.1. Read metadata file

To handle file paths and sample-specific data, a metadata file must be provided as a CSV file. The metadata file must contain the following required columns:

- **`sample_id`**: A short sample identifier (e.g., `S1`, `S2`, `S3`)
- **`sample_alias`**: A more descriptive sample name (e.g., `S1_resting`, `S2_PHA`, `S3_IL2`)
- **`file_path`**: The filename of the PXL file located in the `data` folder (e.g., `S1_resting.pxl`)
- **`condition`**: The sample treatment or experimental condition (e.g., `resting`, `stimulated`, `IL2_treated`)

Additional columns can be added as needed (e.g., `donor`, `time_point`, `batch`, `replicate`) and will automatically be incorporated into the PNA object as metadata columns for downstream analysis.

**Example metadata file:**

```csv
sample_id,sample_alias,file_path,condition,donor
S1,S1_resting,S1_resting.pxl,resting,D1
S2,S2_PHA,S2_PHA.pxl,stimulated,D2
S3,S3_IL2,S3_IL2.pxl,IL2_treated,D1
```

```{r}
#| label: setup_read_metadata

# Read metadata file into a dataframe
metadata <- read.csv(here::here("data", "metadata.csv")) |> as_tibble()

# Check that the metadata file contains the required columns - otherwise stop the script
check_metadata_columns(metadata)

# Print the metadata
print(metadata)
```

### 0.2.2. Load and merge data

This step loads each PXL file specified in the metadata and merges them into a single combined PNA object. First, each PXL file is read using `ReadPNA_Seurat()`, which creates a Seurat-compatible PNA object for each sample. Proximity scores are not loaded at this stage (`load_proximity_scores = FALSE`) to reduce memory usage, as they can be computed later if needed. The individual PNA objects are then merged using Seurat's `merge()` function, which combines all samples into a single object. During merging, the `sample_id` from the metadata is automatically appended to each cell's identifier to ensure unique cell names across samples. Finally, `JoinLayers()` consolidates any split layers in the merged object, ensuring all data is accessible in a unified structure.

```{r}
#| label: setup_merge_data

# Load the data from the file paths into a list of PNA objects
pg_data <-
    here::here("data", metadata$file_path) |>
    lapply(ReadPNA_Seurat, load_proximity_scores = FALSE)

# Merge the PNA objects, adding the sample_id to the cell names
pg_data <-
    merge(
        pg_data[[1]],
        y = pg_data[-1],
        add.cell.ids = metadata$sample_id
    ) |>
    JoinLayers()
```

### 0.2.3. Add metadata

This step adds sample-level metadata from the metadata file to each cell in the PNA object. The process extracts the existing cell metadata from the PNA object and converts it to a tibble format. The `sample_id` is then extracted from each cell's identifier by removing the last 18 characters. The metadata from the CSV file (excluding the `file_path` column, which is not needed in the cell metadata) is then left-joined to the cell metadata based on matching `sample_id` values. Finally, the enriched metadata is converted back to a data frame with cell identifiers as rownames and assigned back to the PNA object's metadata slot. A sanity check is performed to verify that the metadata was correctly added by displaying cross-tabulations of metadata columns.

```{r}
#| label: setup_add_metadata

# Add metadata to the PNA object
pg_data[[]] <-
    pg_data[[]] |>
    as_tibble(rownames = "cell_id") |>
    mutate(sample_id = str_sub(cell_id, end = -18)) |> # Remove the last 18 characters from the cell_id to get the sample_id
    left_join(metadata |> select(-file_path), by = "sample_id") |> # Do not add the file path to the PNA object
    column_to_rownames("cell_id")

# Sanity check the metadata
for (col in colnames(metadata |> select(-file_path))) {
    print(col)
    print(table(pg_data[[]]$sample_alias, pg_data[[]][, col]))
}
```

## 0.3. Theme

Here we specify the color palettes for clusters and conditions. These can be adjusted to the color combinations that you want for your project, by replacing either palette with a named vector containing color codes.  The color palettes are created using the `PixelgenAccentColors()` function from the `pixelatorR` package, which creates a set of 12 colors for each hue. The color palettes are then used to color the clusters and conditions in the plots.

```{r}
#| label: setup_theme

cluster_palette <-
    c(
        pixelatorR::PixelgenAccentColors(
            level = 7,
            hue = c(
                "blues",
                "reds",
                "cyans",
                "oranges",
                "pinks",
                "greens",
                "purples",
                "yellows",
                "greys",
                "beiges"
            )
        ),
        pixelatorR::PixelgenAccentColors(
            level = 9,
            hue = c(
                "blues",
                "pinks",
                "reds",
                "yellows",
                "greys",
                "beiges"
            )
        ),
        pixelatorR::PixelgenAccentColors(
            level = 5,
            hue = c(
                "purples",
                "blues",
                "cyans",
                "greens",
                "reds",
                "oranges"
            )
        ),
        pixelatorR::PixelgenAccentColors(
            level = 3,
            hue = c("blues", "reds")
        ),
        pixelatorR::PixelgenAccentColors(
            level = 12,
            hue = c("blues", "reds")
        )
    ) |>
    unname() |>
    rep(10)

# Create a condition palette for the conditions
unique_conditions <- unique(metadata$condition)
condition_palette <-
    set_names(
        pixelatorES::create_sample_palette(
            unique_conditions
        )[seq_len(length(unique_conditions))],
        unique_conditions
    )

# Create a sample palette for the samples
unique_samples <- unique(metadata$sample_alias)
sample_palette <-
    set_names(
        pixelatorES::create_sample_palette(
            unique_samples
        )[seq_len(length(unique_samples))],
        unique_samples
    )

cell_palette <-
    c(
        "CD4 T" = "#6D92D1",
        "CD4 Naive" = "#B9CDED",
        "Naive CD4 T" = "#B9CDED",
        "CD4 TSCM" = "#92B0E0",
        "CD4 TCM" = "#6D92D1",
        "CD4 TEM" = "#4A73C0",
        "CD4 TEFF" = "#224792",
        "CD4 TEMRA" = "#1C3A76",
        "Treg" = "#2955AE",
        "CD4 CTL" = "#2955AE",
        "CD8 T" = "#75C5B5",
        "CD8 Naive" = "#D0EDE6",
        "Naive CD8 T" = "#D0EDE6",
        "CD8 TSCM" = "#A2DACE",
        "CD8 TCM" = "#75C5B5",
        "CD8 TEM" = "#4AAF9D",
        "CD8 TEFF" = "#209785",
        "CD8 TEMRA" = "#1B7E6F",
        "other T" = "#1B7E9F",
        "Other T" = "#1B7E9F",
        "MAIT" = "#0D6D5E",
        "DPT" = "#1B7E9F",
        "DNT" = "#7B787F",
        "dnT" = "#7B787F",
        "NK" = "#A28EDB",
        "CD56dim NK" = "#A28EDB",
        "CD56bright NK" = "#866CCD",
        "NK_CD56bright" = "#866CCD",
        "NKT" = "#6C4ABD",
        "B" = "#C7A989", # "#917557",
        "B naive" = "#F0DAC4",
        "Naive B" = "#F0DAC4",
        "B intermediate" = "#F0DAC4",
        "Memory B" = "#C7A989",
        "B memory" = "#C7A989",
        "Plasma cells" = "#DE9982",
        "Plasmablast" = "#DE9982",
        "DC" = "#DA94C1",
        "cDC1" = "#DA94C1",
        "cDC2" = "#BB5391",
        "pDC" = "#9C4579",
        "Mono" = "#F7DFA0",
        "Classical Mono" = "#F7DFA0",
        "CD14 Mono" = "#F7DFA0",
        "Intermediate Mono" = "#E6BE4A",
        "CD16 Mono" = "#BD9315",
        "Non-classical Mono" = "#BD9315",
        "other" = "#797979",
        "Neutrophils" = "#CDCDCD",
        "Basophils" = "#797979",
        "Platelet" = "#DB6365",
        "Platelets" = "#DB6365",
        "gdT" = "#5A3E9E"
    )
```

# 1. Quality control

Quality control is a critical step that ensures only high-quality cells are included in downstream analyses. This section systematically evaluates multiple quality metrics at both the sample and cell level to identify and exclude low-quality cells, cell fragments, doublets, and technical artifacts. The quality metrics examined include:

- **Read depth**: The total number of reads per cell, which reflects sequencing depth and overall data quality
- **Graph connectivity**: Metrics such as average k-core and percentage of dangling nodes, which assess the quality of the protein interaction network structure
- **Unspecific background**: The fraction of isotype control signal, which quantifies non-specific binding and background noise
- **Tau score**: A measure of protein composition heterogeneity that helps identify cells with abnormal marker expression profiles

These metrics are visualized through violin plots and distribution plots to facilitate inspection across samples. Based on these visualizations and predefined thresholds, users can identify cells that should be excluded from the dataset. The actual filtering of cells based on these thresholds is performed during the Data Processing stage, allowing users to review the quality metrics and adjust thresholds as needed before applying filters.

## 1.1. Read Depth

First we inspect the read depth per cell. This is a measure of the total number of reads in the cell, and is a good indicator of the quality of the data. The higher the read depth, the better the data quality.

```{r}
#| label: quality_control_read_depth

read_depth_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_alias",
    var_y = "reads_in_component",
    palette = sample_palette,
    data_cutoff = 0.999 # Remove a few outlier cells that make the plot hard to read
)

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_read_depth"
    ),
    plot = read_depth_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)

```

## 1.2. Connectivity

Here we inspect the average k-core per cell. The k-core is a measure of the connectivity of the cell, and is another good indicator of the quality of the data. The higher the average k-core, the better the graph connectivity.

```{r}
#| label: quality_control_average_k_core

# Average k-core per cell
average_k_core_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_alias",
    var_y = "average_k_core",
    palette = sample_palette,
    data_cutoff = 1
)

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_average_k_core"
    ),
    plot = average_k_core_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)
```

Next, we plot the percentage of dangling nodes per cell. Dangling nodes belong to subgraphs that are connnected to the rest of the graph by a single edge, and are an indicator of the graph connectivity. The lower the percentage of dangling nodes, the better the graph connectivity.

```{r}
#| label: quality_control_dangling_nodes_plots

# Add percentage of dangling nodes per cell - i.e. K-core 1 fraction
pg_data <-
    AddMetaData(
        pg_data,
        FetchData(
            pg_data,
            c("sample_alias", "k_core_1", "n_umi")
        ) |>
            mutate(prc_dangling_nodes = 100 * k_core_1 / n_umi) |>
            select(prc_dangling_nodes),
        col.name = "prc_dangling_nodes"
    )

dangling_nodes_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_alias",
    var_y = "prc_dangling_nodes",
    palette = sample_palette,
    data_cutoff = 1
)

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_dangling_nodes"
    ),
    plot = dangling_nodes_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)

```



## 1.3. Select cells for downstream analysis

In the following steps, we will determine which cells to retain for downstream analysis. The objective is to remove cells that are likely to represent poor-quality data based on the following criteria:

- Unusually small cells, which may indicate cell fragments
- Unusually large cells, which may represent doublets or technical artifacts
- Cells with high levels of nonspecific background, as evidenced by elevated isotype marker signal
- Cells with abnormal marker expression profiles, as indicated by the tau score

We will begin by visually inspecting these quality metrics. Cells that fail to meet the defined thresholds will be excluded during the Data Processing stage.

### 1.3.1. Define thresholds

Revisit these thresholds according to your data. Sometimes it is necessary to use different thresholds for different samples. The thresholds are defined as follows:

- **`n_umi_min_threshold`**: The minimum number of UMI counts per cell. Cells with fewer than this number of UMI counts are likely to be cell fragments.
- **`n_umi_max_threshold`**: The maximum number of UMI counts per cell. Cells with more than this number of UMI counts are likely to be doublets or technical artifacts.
- **`isotype_fraction_threshold`**: The maximum fraction of UMI counts that can be attributed to isotype control antibodies. Cells with more than this fraction of UMI counts attributed to isotype control antibodies are likely to have high levels of nonspecific background.

```{r}
#| label: quality_control_define_thresholds

# Define threshold for minimum molecule counts per cell
n_umi_min_threshold <- 15000
n_umi_max_threshold <- 200000
isotype_fraction_threshold <- 0.0025
```


```{r}
#| label: quality_control_n_umi_plot

# Plot to visualize thresholds as a violin plot
n_umi_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_alias",
    var_y = "n_umi",
    palette = sample_palette,
    data_cutoff = 0.999
) &
    geom_hline(
        yintercept = n_umi_min_threshold,
        linetype = "dashed",
        color = "red"
    ) &
    geom_hline(
        yintercept = n_umi_max_threshold,
        linetype = "dashed",
        color = "red"
    ) &
    theme_bw()

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_n_umi"
    ),
    plot = n_umi_plot,
    width = 10,
    height = 8
)

```

### 1.3.2. Cell size filtering

This section generates molecule rank plots that visualize the distribution of UMI counts across cells, ranked by abundance. These plots are essential for evaluating the quality thresholds defined earlier and ensuring they appropriately capture the cell population distribution. The plots display cells ranked by their UMI count (from highest to lowest) on a log-log scale, with threshold lines overlaid to visualize how many cells would be retained or excluded based on the current threshold settings.

Setting correct thresholds is crucial for downstream analysis quality. Thresholds that are too lenient may retain low-quality cells, cell fragments, or doublets that can confound biological interpretation. Conversely, thresholds that are too stringent may exclude valid cells and reduce statistical power. The annotated plots generated here provide detailed per-sample statistics showing the number of cells that fall within the acceptable range versus those that would be excluded as outliers (both low and high UMI counts). Users should carefully review these plots and adjust thresholds as needed to balance data quality with cell retention, ensuring that the thresholds appropriately reflect the characteristics of their specific dataset.

```{r}
#| label: quality_control_select_cells

# Combined molecule rank plots
molecule_rank_plot_w_thresholds <- MoleculeRankPlot(pg_data, group_by = "sample_alias") +
    geom_hline(
        yintercept = n_umi_min_threshold,
        linetype = "dashed",
        color = "red"
    ) +
    geom_hline(
        yintercept = n_umi_max_threshold,
        linetype = "dashed",
        color = "red"
    )
save_plot(
    filename = here::here("results/01_quality_control/molecule_rank_plots/quality_control_molecule_rank_plot_w_thresholds"),
    plot = molecule_rank_plot_w_thresholds,
    width = 10,
    height = 8,
    create_dir = TRUE
)

molecule_rank_plot <- MoleculeRankPlot(pg_data, group_by = "sample_alias")
save_plot(
    filename = here::here("results/01_quality_control/molecule_rank_plots/quality_control_molecule_rank_plot"),
    plot = molecule_rank_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)

# Separate annotated molecule rank plot for each sample
pg_data[[]] <- pg_data[[]] |>
    group_by(sample_id) |>
    mutate(
        rank = rank(-n_umi, ties.method = "random")
    ) |>
    ungroup()

umi_normal <- pg_data[[]] |>
    mutate(
        umi_normal = case_when(
            n_umi > n_umi_min_threshold & n_umi < n_umi_max_threshold ~ TRUE,
            n_umi < n_umi_min_threshold | n_umi > n_umi_max_threshold ~ FALSE
        )
    ) |>
    filter(tau_type == "normal") |>
    select(sample_id, umi_normal)

umi_normal_count <- umi_normal |>
    filter(is.logical(umi_normal)) |>
    group_by(sample_id) |>
    summarise(normal_count = sum(umi_normal), .groups = "drop")

umi_outliers <- pg_data[[]] |>
    select(sample_id, n_umi, tau_type) |>
    filter(tau_type == "normal") |>
    mutate(
        umi_low = n_umi < n_umi_min_threshold,
        umi_high = n_umi > n_umi_max_threshold
    ) |>
    group_by(sample_id) |>
    summarise(
        low_count = sum(umi_low),
        high_count = sum(umi_high),
        .groups = "drop"
    )

hline_normal <- data.frame(
    sample_id = unique(pg_data[[]]$sample_id),
    cutoff = n_umi_min_threshold
)

point_colors <- pg_data[[]] |>
    rownames_to_column("component") |>
    select(sample_id, n_umi, rank, component) |>
    left_join(
        umi_normal_count |> rownames_to_column("component"),
        by = c("component", "sample_id")
    ) |>
    left_join(hline_normal, by = "sample_id") |>
    mutate(
        max_value = n_umi_max_threshold,
        col = case_when(
            isTRUE(umi_normal) ~ "#496389",
            n_umi < cutoff & !isTRUE(umi_normal) ~ "#FFA292",
            n_umi > max_value & !isTRUE(umi_normal) ~ "#FFA292",
            TRUE ~ "#496389"
        )
    )

annotated_molecule_rank_plot <- pg_data[[]] |>
    select(sample_id, rank, n_umi, tau_type) |>
    filter(tau_type == "normal") |>
    ggplot(aes(x = rank, y = n_umi)) +
    geom_point(
        data = point_colors,
        aes(x = rank, y = n_umi),
        color = point_colors$col,
        size = 1,
        alpha = 0.8
    ) +
    geom_hline(
        data = hline_normal,
        aes(yintercept = cutoff),
        linetype = "dashed",
        color = "red"
    ) +
    facet_wrap(
        ~ factor(sample_id, levels = sort(unique(sample_id))),
        nrow = 2
    ) +
    geom_text(
        data = umi_normal_count,
        aes(x = Inf, y = Inf, label = paste("Cells:", normal_count)),
        hjust = 1.2,
        vjust = 1.8,
        size = 4,
        color = "#496389",
        position = position_nudge(y = 0.5)
    ) +
    geom_text(
        data = umi_outliers,
        aes(x = 10, y = Inf, label = paste("High: ", high_count)),
        vjust = 1.8,
        hjust = 0.8,
        size = 4,
        color = "#FFA292",
        position = position_nudge(y = -0.5)
    ) +
    geom_text(
        data = umi_outliers,
        aes(x = 10, y = n_umi_min_threshold, label = paste("Low: ", low_count)),
        size = 4,
        vjust = 0.2,
        hjust = 0.7,
        color = "#FFA292",
        position = position_nudge(y = -0.5)
    ) +
    geom_rug(
        aes(x = rank),
        linewidth = 1,
        sides = "l",
        alpha = 0.05
    ) +
    scale_x_log10() +
    scale_y_log10() +
    coord_equal(ratio = 1.5) +
    theme_minimal(base_size = 18) +
    labs(x = "Component Rank", y = "UMIs", color = "UMI Normal") +
    theme(
        aspect.ratio = 1,
        legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(
            color = "black",
            fill = NA,
            linewidth = 0.5
        )
    )
save_plot(
    filename = here::here("results/01_quality_control/molecule_rank_plots/quality_control_annotated_molecule_rank_plot"),
    plot = annotated_molecule_rank_plot,
    width = 15,
    height = 12,
    create_dir = TRUE
)
```

### 1.3.3. Unspecific Background filtering

In PNA experiments, isotype control antibodies are included to measure unspecific binding and serve as important quality controls for background signal. The fraction of total signal that is derived from isotype controls in each cell provides a quantitative estimate of non-specific background. Generally, cells with a high proportion of isotype signal are more likely to be dominated by artefactual binding rather than true biological signal. Therefore, by examining the fraction of isotype control antibodies per cell, we can assess the overall level of background in each sample and identify cells potentially affected by excessive non-specific binding, which may need to be excluded from downstream analysis.

```{r}
#| label: quality_control_isotype_controls

isotype_fraction_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_alias",
    var_y = "isotype_fraction",
    palette = sample_palette,
    data_cutoff = 1
) +
  geom_hline(
    yintercept = isotype_fraction_threshold,
    linetype = "dashed",
    color = "red"
  ) 


save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_isotype_controls"
    ),
    plot = isotype_fraction_plot,
    width = 10,
    height = 8
)
```

### 1.3.4. Tau score filtering

The Tau score measures the heterogeneity of protein composition per cell, and can be used to detect various types of outliers. Cells with high Tau have high counts for few proteins, while cells with low Tau have a more heterogeneous composition of proteins than we would expect from a normal cell. As these outliers likely represent technical artefacts, we can remove them from the dataset.

```{r}
#| label: quality_control_tau_score

tau_score_plot <- TauPlot(pg_data, group_by = "sample_alias")

save_plot(
    filename = here::here(
        "results/01_quality_control/tau_plots/quality_control_tau_score"
    ),
    plot = tau_score_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)
```

## 1.4. Select markers for dimensionality reduction (optional)

This optional step allows users to filter markers based on expression levels, retaining only those markers that are likely to be positive in at least one cell population within the dataset. Marker selection is not a one-size-fits-all process; the optimal selection criteria depend on the specific dataset characteristics and experimental design.

This section demonstrates one approach using counts per million (CPM) thresholds. For each marker, the 50th percentile (median) and 95th percentile of CPM values are calculated across all cells. Markers are then selected based on whether their 95th percentile exceeds a defined threshold. This threshold should be adjusted according to the background signal levels in the dataset and the expected presence of specific markers based on the experimental design.

In the example provided here, the threshold is set to 500 CPM at the 95th percentile. This effectively means that markers are retained if they are expressed at detectable levels in cell populations comprising at least 5% of the sample. Comparing the 95th percentile to the 50th percentile (median) provides insight into the variability of marker expression across cells, helping to distinguish markers with consistent expression from those with highly variable or rare expression patterns.


```{r}
#| label: quality_control_select_markers

marker_cpm_p95_threshold <- 500

count_data <-
    LayerData(pg_data, "counts")
cpm_data <-
    count_data |>
    sweep(1, rowSums(count_data), "/")

cpm_data <- cpm_data * 1e6

cpm_quantiles <-
    cpm_data |>
    apply(1, function(x) quantile(x, c(0.5, 0.95))) |>
    t() |>
    as_tibble(rownames = "marker") |>
    mutate(kept = `95%` > marker_cpm_p95_threshold)

scatter_percentile_cpm_quantiles <- cpm_quantiles |>
    ggplot(aes(`50%`, `95%` / `50%`, color = kept)) +
    geom_point(size = 0.1) +
    geom_text(aes(label = marker), vjust = -0.5, size = 2) +
    scale_x_log10(expand = expansion(0.2)) +
    scale_y_log10() +
    theme_bw() +
    labs(
        title = "Marker selection based on CPM",
        x = "50th percentile of CPM (log10 scale)",
        y = "95th / 50th percentile of CPM (log10 scale)"
    )
save_plot(
    filename = here::here("results/01_quality_control/marker_selection_plots/quality_control_marker_selection_cpm"),
    plot = scatter_percentile_cpm_quantiles,
    width = 12,
    height = 4
)


barplot_percentile_cpm_quantiles <- cpm_quantiles |>
    arrange(`95%`) |>
    mutate(marker = factor(marker, levels = marker)) |>
    ggplot(aes(x = marker, y = `95%`, fill = kept)) +
    geom_col() +
    scale_y_log10() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5)) +
    labs(
        title = "95th percentile of CPM per marker",
        x = "Marker",
        y = "95th percentile of CPM (log10 scale)"
    )
save_plot(
    filename = here::here("results/01_quality_control/marker_selection_plots/quality_control_marker_selection_cpm_barplot"),
    width = 12,
    height = 4,
    plot = barplot_percentile_cpm_quantiles
)

selected_markers <-
    cpm_quantiles |>
    filter(kept) |>
    pull(marker)


```

# 2. Data processing

## 2.1. Normalize and scale data

PNA data are normalized using centered log-ratio (CLR) transformation over margin 2, which normalizes counts per cell by dividing by the geometric mean across all markers for that cell. This normalization approach effectively accounts for differences in sequencing depth between cells while preserving the relative differences in marker abundance within each cell. The CLR transformation is particularly well-suited for compositional data like PNA, where the total signal per cell can vary substantially.

Following normalization, the data are Z-score scaled per marker. This transformation centers each marker's distribution around zero and scales it to unit variance. Z-score scaling is essential for downstream dimensionality reduction and clustering analyses, as it ensures that all markers contribute equally to the analysis regardless of their absolute abundance levels. Without this scaling step, highly abundant markers would dominate the analysis, potentially obscuring important signals from less abundant but biologically relevant markers.

```{r}
#| label: data_processing_normalize_scale

pg_data <-
  pg_data |>
  # Filter cells based on quality control thresholds
  subset(tau_type == "normal") |> 
  subset(n_umi > n_umi_min_threshold & 
           n_umi < n_umi_max_threshold) |> 
  subset(isotype_fraction < isotype_fraction_threshold) |>
  # Data normalization
  NormalizeData(normalization.method = "CLR", margin = 2) |>
  # Z score scaling
  ScaleData()

```

## 2.2. Dimensionality reduction

Principal Component Analysis (PCA) is performed to extract the main axes of variation in the data. The first 10-20 principal components (PCs) are typically used for downstream analysis, though this number should be adjusted based on the specific dataset and experimental design. As a general guideline, the selected PCs should explain a substantial portion of the total variance while avoiding overfitting to noise. Principal components often capture technical effects, cell states, or experimental conditions, so the number of PCs can be adjusted according to the expected number of cell states or the complexity of the experimental design. The number of PCs is set to 20 by default, but can be adjusted as needed.

If marker selection was performed in the Quality Control section, those selected markers can be used for PCA instead of all markers. This optional step helps reduce noise from low-abundance or non-informative markers and can improve the interpretability of the results. 

Following PCA, Uniform Manifold Approximation and Projection (UMAP) is applied to visualize the data in a two-dimensional space. UMAP is a non-linear dimensionality reduction technique that preserves both local neighborhood structure and global relationships in the data. This makes it particularly well-suited for visualizing complex single-cell PNA datasets, where it can reveal cell type clusters and relationships that may not be apparent in higher-dimensional space. The number of neighbors is set to 30 by default, but can be adjusted as needed.

```{r}
#| label: data_processing_dimensionality_reduction

npcs <- 20
n_neighbors <- 30
pg_data <-
    pg_data |>
    RunPCA(npcs = npcs, features = selected_markers) |>
    RunUMAP(reduction = "pca", dims = 1:20, n.neighbors = n_neighbors)

pca_red <-
    Reductions(pg_data, "pca")

elbow_plot_pca <- pca_red@stdev |>
    as_tibble(rownames = "PC") |>
    mutate(PC = PC |> str_remove("PC_") |> as.integer()) |>
    ggplot(aes(PC, value)) +
    geom_line() +
    geom_point() +
    theme_bw() +
    labs(
        title = "PCA Elbow Plot",
        x = "Principal Component",
        y = "Standard Deviation"
    )
save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_pca_elbow_plot"),
    width = 6,
    height = 4,
    plot = elbow_plot_pca
)

cell_metadata <-
    FetchData(pg_data, c("sample_alias", "condition")) |>
    as_tibble(rownames = "cell_id")

pca_embeddings <- Embeddings(pg_data, "pca") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(PC_1, PC_2, color = sample_alias)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = sample_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_pca"),
    width = 6,
    height = 4,
    plot = pca_embeddings
)

umap_embeddings <-
    Embeddings(pg_data, "umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(umap_1, umap_2, color = sample_alias)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = sample_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_umap"),
    width = 6,
    height = 4,
    plot = umap_embeddings
)
```

## 2.3. Data harmonization

Data harmonization is an optional step that can be beneficial when experiments contain multiple samples with similar cell types that do not cluster together well due to technical or biological variability. The goal of data harmonization is to align similar cell types across samples so that they cluster together, facilitating the identification of shared cell types between samples even when they exhibit somewhat different protein expression profiles. This approach is particularly useful in experiments with batch effects, technical variability, or biological sources of variation such as donor differences or stimulation-induced perturbations.

The Harmony algorithm is used here to harmonize the data based on sample ID, though any sample metadata column can be specified for harmonization. The algorithm integrates information across samples while preserving biological variation within each sample.

**Important considerations**: Users should carefully inspect the harmonized data to ensure that harmonization has not removed important biological variation and that similar cell types are indeed clustering together across samples. In many cases, harmonization is not necessary, particularly when samples are already well-mixed and similar cell types cluster together naturally without harmonization.

The harmonization process generates a new dimensional reduction called "harmony", which is used for UMAP visualization and clustering. Importantly, the original data remains unchanged, so harmonization is used primarily to identify common cell identities across samples rather than to modify the underlying expression data.

```{r}
#| label: data_processing_data_harmonization
pg_data <-
    pg_data |>
    RunHarmony(
        group.by.vars = "sample_alias",
        plot_convergence = FALSE,
    ) |>
    RunUMAP(
        reduction = "harmony",
        dims = 1:20,
        reduction.name = "harmony_umap",
        n.neighbors = n_neighbors
    )

cell_metadata <-
    FetchData(pg_data, c("sample_alias", "condition")) |>
    as_tibble(rownames = "cell_id")

harmony_embeddings <- Embeddings(pg_data, "harmony") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(harmony_1, harmony_2, color = sample_alias)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = sample_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_harmony"),
    width = 6,
    height = 4,
    plot = harmony_embeddings
)

harmony_umap_embeddings <- Embeddings(pg_data, "harmony_umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = sample_alias)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = sample_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_harmonyumap"),
    width = 6,
    height = 4,
    plot = harmony_umap_embeddings
)

```

## 2.4. Clustering

Clustering groups cells with similar protein expression profiles into discrete clusters, which can subsequently be annotated as distinct cell types or cell states. This step is essential for identifying and characterizing the cellular composition of the dataset.

Leiden clustering is performed on a k-nearest neighbor graph constructed from the dimensionality-reduced space (either PCA or Harmony reduction). The algorithm identifies communities of cells that are more densely connected to each other than to cells in other communities, effectively partitioning the data into biologically meaningful groups.

The resolution parameter is a critical hyperparameter that controls the granularity of the clustering. Higher resolution values result in more fine-grained clusters (more clusters with fewer cells each), while lower values produce broader clusters (fewer clusters with more cells each). The optimal resolution depends on the dataset characteristics, experimental design, and the level of cellular heterogeneity expected. Users should test multiple resolution values (e.g., 0.3, 0.5, 0.7, 1.0) and visually inspect the resulting clusters on UMAP plots to determine which resolution best captures the expected cell type diversity without over-clustering or under-clustering the data. The resolution is set to 0.5 by default, but can be adjusted as needed.

```{r}
#| label: data_processing_clustering

pg_data <-
    pg_data |>
    FindNeighbors(reduction = "harmony", dims = 1:10) |>
    FindClusters(random.seed = 1, resolution = 0.5)

cell_metadata <-
    FetchData(pg_data, c("sample_alias", "condition", "seurat_clusters")) |>
    as_tibble(rownames = "cell_id")

harmony_clusters_embeddings <- Embeddings(pg_data, "harmony") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(harmony_1, harmony_2, color = seurat_clusters)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = cluster_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/clustering_plots/data_processing_harmony_clusters"),
    width = 6,
    height = 4,
    plot = harmony_clusters_embeddings
)

harmony_umap_clusters_embeddings <- Embeddings(pg_data, "harmony_umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = seurat_clusters)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = cluster_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/clustering_plots/data_processing_harmonyumap_clusters"),
    width = 6,
    height = 4,
    plot = harmony_umap_clusters_embeddings
)

```

## 2.5. Marker Distribution on UMAP

This section visualizes the distribution of all markers on the harmonized UMAP. This can help to identify which markers are characteristic of each cluster, and can guide manual annotation of cell types. 

```{r}
#| label: data_processing_marker_distribution_on_umap

for (marker in rownames(pg_data)) {
    feature_plot <- FeaturePlot(pg_data, features = marker, reduction = "harmony_umap") &
        theme(aspect.ratio = 1)
    save_plot(
        filename = glue::glue("results/02_data_processing/marker_umap_plots/{marker}_umap"),
        plot = feature_plot,
        width = 6,
        height = 6
    )
}
```

## 2.6. Annotation

### 2.6.1. Automatic annotation

Automatic cell annotation is performed by matching each cell's protein expression profile to annotated cells in a reference peripheral blood mononuclear cell (PBMC) dataset using label transfer methods. This approach provides an initial annotation framework that can accelerate the annotation process.

**Important limitations and recommendations**: Automated annotations should be interpreted with caution, particularly when the sample composition differs substantially from the PBMC reference dataset. For example, annotations may be unreliable for certain stimulated samples, cell lines, or tissues that are not well-represented in the reference. Additionally, automated methods may not capture cell states or subtypes that are specific to the experimental conditions.

All automated annotations should be inspected, verified, and adjusted as necessary. Automated annotation serves as a starting point, but manual review and refinement are essential to ensure accurate and context-appropriate cell type assignments. Users should carefully examine marker expression patterns, differential abundance results, and UMAP visualizations to validate and refine the automated annotations before proceeding with downstream analyses. 

```{r}
#| label: data_processing_annotation

# Load reference dataset
reference <- pixelatorES::read_annotation_reference()

# Set names of annotation columns in reference
reference$l1_annotation <- reference$celltype.l1
reference$l2_annotation <- reference$celltype.l2

# Annotation by label transfer
pg_data <-
    pg_data |>
    AnnotateCells(
        reference = reference,
        reference_groups = c("l1_annotation", "l2_annotation"),
        reference_assay = "PNA",
        query_assay = "PNA",
        # Summarise annotations by cluster, adding an extra column with majority vote
        summarize_by_column = "seurat_clusters",
        normalization_method = "CLR",
        method = "nmf"
    )


cell_metadata <-
    FetchData(pg_data, c(
        "sample_alias", "condition", "seurat_clusters",
        "l1_annotation_summary", "l2_annotation_summary",
        "l1_annotation", "l2_annotation"
    )) |>
    as_tibble(rownames = "cell_id")

# L1 automatic annotation - summarised by cluster
harmony_umap_l1_annotation_summary_plot <- Embeddings(pg_data, "harmony_umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata,
        by = "cell_id"
    ) |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = l1_annotation_summary)) +
    geom_point() +
    geom_text(
        data = . %>%
            group_by(l1_annotation_summary) %>%
            summarise(
                harmonyumap_1 = median(harmonyumap_1),
                harmonyumap_2 = median(harmonyumap_2)
            ),
        aes(label = l1_annotation_summary),
        size = 3,
        color = "black"
    ) +
    coord_fixed() +
    scale_color_manual(values = cell_palette) +
    theme_bw() +
    theme(panel.grid = element_blank())
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/automatic_annotation/data_processing_harmonyumap_l1_annotation_summary"),
    width = 6,
    height = 4,
    plot = harmony_umap_l1_annotation_summary_plot
)


# L1 automatic annotation - annotated by cell
harmony_umap_l1_annotation_plot <- Embeddings(pg_data, "harmony_umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata,
        by = "cell_id"
    ) |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = l1_annotation)) +
    geom_point() +
    geom_text(
        data = . %>%
            group_by(l1_annotation) %>%
            summarise(
                harmonyumap_1 = median(harmonyumap_1),
                harmonyumap_2 = median(harmonyumap_2)
            ),
        aes(label = l1_annotation),
        size = 3,
        color = "black"
    ) +
    coord_fixed() +
    scale_color_manual(values = cell_palette) +
    theme_bw() +
    theme(panel.grid = element_blank())
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/automatic_annotation/data_processing_harmonyumap_l1_annotation"),
    width = 6,
    height = 4,
    plot = harmony_umap_l1_annotation_plot
)


# L2 automatic annotation - summarised by cluster
harmony_umap_l2_annotation_summary_plot <- Embeddings(pg_data, "harmony_umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata,
        by = "cell_id"
    ) |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = l2_annotation_summary)) +
    geom_point() +
    geom_text(
        data = . %>%
            group_by(l2_annotation_summary) %>%
            summarise(
                harmonyumap_1 = median(harmonyumap_1),
                harmonyumap_2 = median(harmonyumap_2)
            ),
        aes(label = l2_annotation_summary),
        size = 3,
        color = "black"
    ) +
    coord_fixed() +
    scale_color_manual(values = cell_palette) +
    theme_bw() +
    theme(panel.grid = element_blank())
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/automatic_annotation/data_processing_harmonyumap_l2_annotation_summary"),
    width = 6,
    height = 4,
    plot = harmony_umap_l2_annotation_summary_plot
)

# L2 automatic annotation - annotated by cell
harmony_umap_l2_annotation_plot <- Embeddings(pg_data, "harmony_umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata,
        by = "cell_id"
    ) |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = l2_annotation)) +
    geom_point() +
    geom_text(
        data = . %>%
            group_by(l2_annotation) %>%
            summarise(
                harmonyumap_1 = median(harmonyumap_1),
                harmonyumap_2 = median(harmonyumap_2)
            ),
        aes(label = l2_annotation),
        size = 3,
        color = "black"
    ) +
    coord_fixed() +
    scale_color_manual(values = cell_palette) +
    theme_bw() +
    theme(panel.grid = element_blank())
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/automatic_annotation/data_processing_harmonyumap_l2_annotation"),
    width = 6,
    height = 4,
    plot = harmony_umap_l2_annotation_plot
)

# L1 cell type composition per sample (annotated by cell)
cell_type_composition_per_sample <- FetchData(pg_data, c("sample_alias", "l1_annotation_summary")) |>
    group_by(sample_alias, l1_annotation_summary) |>
    count(name = "n_cells") |>
    group_by(sample_alias) |>
    mutate(prc_cells = 100 * n_cells / sum(n_cells)) |>
    ggplot(aes(sample_alias, prc_cells, fill = l1_annotation_summary)) +
    geom_col(position = "fill") +
    geom_text(aes(label = paste0(round(prc_cells, 1), "%")),
        position = position_fill(vjust = 0.5),
        size = 3
    ) +
    scale_y_continuous(labels = scales::percent_format()) +
    scale_fill_manual(values = cell_palette) +
    theme_bw() +
    theme(
        axis.text.x = element_text(angle = 60, hjust = 1)
    ) +
    labs(
        title = "Cell type composition per sample",
        x = "Sample",
        y = "Cell type fraction"
    )
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/cell_type_composition/data_processing_cell_type_composition_per_sample"),
    width = 12,
    height = 8,
    plot = cell_type_composition_per_sample
)
```

### 2.6.2. Discover Cell Types

While automated annotation provides a useful starting point, manual inspection and refinement are essential to ensure accuracy. This section performs differential abundance analysis to identify cluster-specific marker expression patterns, which serve as the foundation for manual cell type identification.

Differential abundance analysis compares marker expression levels between each cluster and all other clusters, identifying markers that are significantly enriched or depleted in specific clusters. These cluster-specific markers are characteristic features that can be used to identify cell types based on known marker expression patterns. For example, clusters enriched for CD3, CD4, and CD45RA are likely CD4+ T cells, while clusters showing high CD19 and CD20 expression are likely B cells.

The analysis generates two key visualizations to aid interpretation:

1. **Heatmap of differential abundance**: Displays the average CLR difference for each marker across all clusters, allowing rapid identification of which markers distinguish each cluster. Markers with strong positive values (red) are enriched in that cluster, while negative values (blue) indicate depletion. This overview helps identify clusters with similar marker profiles and those with unique expression signatures.

2. **Volcano plots per cluster**: Show the relationship between statistical significance and effect size for each marker within each cluster. Markers in the upper right or upper left quadrants (high significance and large effect size) are the most reliable cluster-defining markers. The top markers for each cluster are automatically labeled, facilitating quick identification of key distinguishing features.

By examining these differential abundance patterns in conjunction with known cell type marker expression profiles, users can systematically identify cell types. The process involves: (1) identifying clusters with characteristic marker combinations, (2) comparing these patterns to established cell type markers, (3) confirming or adjusting automated annotations based on the observed expression patterns, and (4) identifying cell populations that may not be well-represented in the reference dataset. This evidence-based approach complements automated annotation and significantly improves confidence in cell type assignments.

```{r}
#| label: data_processing_manual_annotation

diff_abundant_markers_cluster <-
    FindAllMarkers(
        pg_data,
        logfc.threshold = 0,
        min_pct = 0,
        test.use = "wilcox",
        assay = "PNA",
        slot = "data",
        group.by = "seurat_clusters",
        mean.fxn = rowMeans,
        fc.name = "difference"
    ) |>
    as_tibble() |>
    rename(marker = gene)

# Heatmap of differential abundance markers per cluster
differential_abundance_markers_per_cluster_heatmap <- diff_abundant_markers_cluster |>
    group_by(marker) |>
    filter(any(p_val_adj < 0.05 & difference > 1)) |>
    select(marker, difference, cluster) |>
    pivot_wider(names_from = cluster, values_from = difference, values_fill = 0) |>
    column_to_rownames("marker") |>
    t() |>
    pheatmap(
        color = PixelgenGradient(100, "BluesCherry"),
        breaks = seq(-3, 3, length.out = 101),
        cellwidth = 8,
        cellheight = 8,
        heatmap_legend_param = list(title = "CLR Difference"),
    ) |>
    as.ggplot()
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/cluster_specific_markers/data_processing_differential_abundance_markers_per_cluster"),
    width = 10,
    height = 4,
    plot = differential_abundance_markers_per_cluster_heatmap
)

# Volcano plot of differential abundance markers per cluster
volcano_plot_differential_abundance_markers_per_cluster <- diff_abundant_markers_cluster |>
    ggplot(aes(difference, -log10(p_val))) +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) +
    geom_point(size = 0.1) +
    geom_text(
        data = . %>%
            group_by(cluster) %>%
            top_n(5, abs(difference)),
        aes(label = marker),
        vjust = -0.5,
        size = 2
    ) +
    facet_wrap(~cluster) +
    scale_fill_manual(values = cluster_palette) +
    theme_bw() +
    labs(
        title = "Differential abundance markers per cluster",
        x = "Average CLR Difference",
        y = "-log10(p value)"
    )

save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/cluster_specific_markers/data_processing_differential_abundance_markers_per_cluster_volcano"),
    width = 8,
    height = 6,
    plot = volcano_plot_differential_abundance_markers_per_cluster
)

# Write differential abundance markers per cluster to csv
write_csv(diff_abundant_markers_cluster, here::here("results/02_data_processing/annotation_plots/cluster_specific_markers/data_processing_differential_abundance_markers_per_cluster.csv"))
```

### 2.6.3. Marker abundance visualization

Here we visualize the abundance of **selected markers** across the harmonized UMAP and per cluster. This can help to identify which markers are characteristic of each cluster, and can guide manual annotation of cell types. Please change `markers_of_interest` according to which markers are most relevant for your dataset. 

```{r}
#| label: data_processing_marker_abundance_visualization

# Change markers_of_interest to the markers of interest for your dataset
markers_of_interest <- c("CD3e", "CD4", "CD8", "CD19", "CD14", "CD16", "CD56")

plot_data <-
    FetchData(
        pg_data,
        c(
            "harmonyumap_1", "harmonyumap_2",
            "sample_alias", "condition",
            "seurat_clusters",
            markers_of_interest
        ),
        slot = "data"
    ) |>
    pivot_longer(
        cols = all_of(markers_of_interest),
        names_to = "marker",
        values_to = "CLR"
    )

marker_abundance_harmonyumap_plot <- plot_data |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = CLR)) +
    geom_point(size = 0.1) +
    facet_wrap(~marker) +
    scale_color_gradient(low = "lightgrey", high = "darkblue") +
    coord_fixed() +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    labs(
        title = "Marker expression on Harmony UMAP",
        x = "Harmony UMAP 1",
        y = "Harmony UMAP 2",
        color = "CLR"
    )
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/cluster_specific_markers/data_processing_marker_abundance_harmonyumap"),
    width = 6,
    height = 4,
    plot = marker_abundance_harmonyumap_plot
)

marker_abundance_per_cluster_plot <- plot_data |>
    ggplot(aes(sample_alias, CLR, fill = condition)) +
    geom_violin(quantiles = 0.5, quantile.linetype = 1) +
    facet_grid(marker ~ seurat_clusters) +
    scale_fill_manual(values = condition_palette) +
    theme_bw() +
    theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        panel.border = element_rect(color = "gray"),
        panel.spacing = unit(0, "mm")
    ) +
    labs(
        title = "Marker abundance per cluster",
        x = "Sample",
        y = "CLR"
    )
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/data_processing_marker_abundance_per_cluster"),
    width = 8,
    height = 8,
    plot = marker_abundance_per_cluster_plot
)

mean_marker_abundance_per_cluster_heatmap <- plot_data |>
    group_by(marker, seurat_clusters) |>
    summarise(mean_CLR = mean(CLR)) |>
    pivot_wider(names_from = seurat_clusters, values_from = mean_CLR) |>
    column_to_rownames("marker") |>
    as.matrix() |>
    pheatmap(
        color = PixelgenGradient(100, "Cherry"),
        cellwidth = 8,
        cellheight = 8,
        heatmap_legend_param = list(title = "Mean CLR"),
    ) |>
    as.ggplot()
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/data_processing_mean_marker_abundance_per_cluster_heatmap"),
    width = 6,
    height = 4,
    plot = mean_marker_abundance_per_cluster_heatmap
)

```

### 2.6.4. Set manual annotation 

This step assigns cell type annotations to each cluster based on the comprehensive analysis performed in the previous sections. The annotations are determined by integrating information from multiple sources: (1) differential abundance analysis identifying cluster-specific markers, (2) marker abundance visualizations on UMAP and per cluster, (3) automated annotation results, and (4) knowledge of established cell type marker expression patterns.

Each cluster is assigned a cell type label through a named vector that maps cluster numbers to cell type annotations. Importantly, multiple clusters can share the same annotation when they represent different states or subtypes of the same cell type. For example, multiple clusters may all be annotated as "CD4 T" if they represent different CD4+ T cell subsets (e.g., naive, memory, effector) that are distinguished by additional markers but share core CD4+ T cell characteristics.

Users must carefully review all available evidence (differential abundance results, marker visualizations, automated annotations, and biological knowledge) before assigning annotations. The example annotations provided here are placeholders and must be customized according to the specific characteristics of your dataset. Incorrect annotations will propagate through all downstream analyses, so this step requires careful consideration and validation.

```{r}
#| label: data_processing_set_manual_annotation

# Specify cell annotation labels for each cluster
cluster_cell_annotation <-
    c(
        "0" = "CD4 T",
        "1" = "CD4 T",
        "2" = "Mono",
        "3" = "CD8 T",
        "4" = "CD4 T",
        "5" = "CD8 T",
        "6" = "B",
        "7" = "NK",
        "8" = "Platelets",
        "9" = "CD8 T",
        "10" = "Mono",
        "11" = "DC",
        "12" = "B",
        "13" = "DC"
    )

# Add annotation to Seurat object
pg_data[[]] <-
    pg_data[[]] |>
    mutate(
        cell_annotation = cluster_cell_annotation[as.character(seurat_clusters)]
    )

# Save annotated Seurat object
saveRDS(pg_data, here::here("results/02_data_processing/annotation_plots/annotated_seurat_object.rds"))

```

### 2.6.5. Comparison of automatic and manual annotations

Compare automatic and manual annotations. This can help to identify any discrepancies between the two methods, and can be used to refine the manual annotations.

```{r}
#| label: data_processing_comparison_automatic_manual_annotation

comparison_automatic_manual_annotation <- FetchData(pg_data, c("seurat_clusters", "l1_annotation", "cell_annotation")) |>
    group_by(seurat_clusters, l1_annotation, cell_annotation) |>
    count() |>
    group_by(seurat_clusters) |>
    mutate(frac = n / sum(n)) |>
    ggplot(aes(l1_annotation, cell_annotation, fill = frac)) +
    geom_tile() +
    facet_wrap(~seurat_clusters, scales = "free") +
    scale_fill_gradient(low = "white", high = "darkblue") +
    theme_bw() +
    theme(
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
    ) +
    labs(
        title = "Comparison of automatic and manual annotations",
        x = "Automatic annotation (L1)",
        y = "Manual annotation",
        fill = "Fraction of cells"
    )
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/data_processing_comparison_automatic_manual_annotation"),
    width = 8,
    height = 6,
    plot = comparison_automatic_manual_annotation
)

```

### 2.6.6. Alternative: Marker gating 

Marker gating provides an alternative approach to cell annotation that is analogous to flow cytometry gating strategies. This method uses density scatter plots to visualize the co-expression of two markers, allowing users to identify distinct cell populations based on marker expression patterns and define gates (thresholds) to classify cells.

The `DensityScatterPlot()` function creates a two-dimensional density plot showing the relationship between two markers, where each axis represents the expression level (typically CLR-normalized values) of one marker. Dense regions in the plot represent cell populations with similar marker expression profiles, while sparse regions indicate transitions between populations or rare cell types.

**How to use marker gating for annotation:**

1. **Select marker pairs**: Choose pairs of markers that are known to distinguish cell types of interest. For example:
   - CD3e vs CD4: Distinguishes T cells (CD3e+) and CD4+ T cells (CD3e+ CD4+)
   - CD3e vs CD8: Distinguishes T cells and CD8+ T cells
   - CD19 vs CD20: Identifies B cells
   - CD14 vs CD16: Distinguishes monocyte subsets

2. **Visualize marker pairs**: Generate density scatter plots for relevant marker combinations to identify distinct cell populations and their expression boundaries.

3. **Define gates**: Based on the density patterns, identify threshold values for each marker that separate different cell populations. These thresholds can be used to classify cells into categories.

4. **Apply gates to clusters**: Use the identified marker expression thresholds to validate or refine cluster annotations. Clusters that don't match expected marker expression patterns may need re-examination.

5. **Compare with other annotation methods**: Cross-reference gating-based annotations with automated annotation results, differential abundance analysis, and UMAP visualizations to ensure consistency.

This gating approach is particularly useful for validating annotations, identifying misclassified clusters, and understanding marker expression relationships. It complements the cluster-based annotation approach by providing a marker-centric perspective on cell type identification.

Below is an example of how to use the gating strategy to annotate CD4 T cell subtypes.

```{r}
#| label: data_processing_gating_strategy

# Define CD4 subtypes using memory/naive markers
# Naive CD4 T: CD45RA+ (high), CD45RO- (low), typically CCR7+ and CD62L+
# Memory CD4 T: CD45RA- (low), CD45RO+ (high)
# Note: Thresholds should be adjusted based on your data distribution
cd4_subtype_markers <- c("CD45RA", "CD45RO")

# subset to CD4 T cells: only these will be used for the gating strategy
pg_data_cd4_t <- subset(pg_data, subset = cell_annotation == "CD4 T")

# Create density scatter plot to distinguish CD4 subtypes
# CD45RA vs CD45RO is a classic combination for naive vs memory distinction
# CD4 T cell subtypes: Naive (CD45RA+ CD45RO-) vs Memory (CD45RA- CD45RO+)
cd4_subtype_density_plot <- DensityScatterPlot(
    pg_data_cd4_t,
    marker1 = "CD45RA",
    marker2 = "CD45RO"
) +
    labs(
        title = "CD4 T Cell Subtypes: Naive vs Memory",
        subtitle = "Naive: CD45RA+ CD45RO- | Memory: CD45RA- CD45RO+"
    )
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/gating_strategy/cd4_subtype_naive_memory_density_scatter"),
    width = 8,
    height = 6,
    plot = cd4_subtype_density_plot,
    create_dir = TRUE
)

# Define gates for CD4 subtypes
# Naive CD4 T cells: High CD45RA, Low CD45RO
naive_cd4_gate <- data.frame(
    xmin = 2.0, # CD45RA threshold (adjust based on data)
    xmax = 6.5,
    ymin = 0,
    ymax = 1.5 # CD45RO threshold (adjust based on data)
)
# Create plot with gates
cd4_subtype_density_plot_naive_cd4_gate <- DensityScatterPlot(
    pg_data_cd4_t,
    marker1 = "CD45RA",
    marker2 = "CD45RO",
    plot_gate = naive_cd4_gate
) +
    labs(
        title = "CD4 T Cell Subtypes with Gates",
        subtitle = "Gate 1: Naive CD4 T (CD45RA+ CD45RO-)"
    )

# Define gate for Memory CD4 T cells: Low CD45RA, High CD45RO
memory_cd4_gate <- data.frame(
    xmin = 0,
    xmax = 2.5, # CD45RA threshold (adjust based on data)
    ymin = 2.0, # CD45RO threshold (adjust based on data)
    ymax = 5
)
cd4_subtype_density_plot_memory_cd4_gate <- DensityScatterPlot(
    pg_data_cd4_t,
    marker1 = "CD45RA",
    marker2 = "CD45RO",
    plot_gate = memory_cd4_gate
) +
    labs(
        title = "CD4 T Cell Subtypes with Gates",
        subtitle = "Gate 2: Memory CD4 T (CD45RA- CD45RO+)"
    )
save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/gating_strategy/cd4_subtype_naive_memory_density_scatter_w_gates"),
    width = 16,
    height = 6,
    plot = cd4_subtype_density_plot_naive_cd4_gate + cd4_subtype_density_plot_memory_cd4_gate,
    create_dir = TRUE
)

# Classify CD4 cells into subtypes
cd4_cell_data <- FetchData(pg_data_cd4_t, vars = c("CD45RA", "CD45RO", "CD4", "CD3e"))

# Define subtypes based on marker expression thresholds
# Adjust thresholds based on your data distribution

cd4_subtype_classification <- cd4_cell_data |>
    mutate(
        cd4_subtype = case_when(
            CD45RA > naive_cd4_gate$xmin & CD45RA < naive_cd4_gate$xmax & CD45RO > naive_cd4_gate$ymin & CD45RO < naive_cd4_gate$ymax ~ "Naive CD4 T",
            CD45RA > memory_cd4_gate$xmin & CD45RA < memory_cd4_gate$xmax & CD45RO > memory_cd4_gate$ymin & CD45RO < memory_cd4_gate$ymax ~ "Memory CD4 T",
            TRUE ~ "CD4 T (other)"
        )
    )

# Add subtype annotation to pg_data
pg_data$cd4_subtype <- pg_data$cell_annotation
pg_data$cd4_subtype[rownames(cd4_subtype_classification)] <- cd4_subtype_classification$cd4_subtype

# Visualize CD4 subtypes on UMAP
cd4_subtype_umap <- DimPlot(
    pg_data,
    group.by = "cd4_subtype",
    reduction = "harmony_umap"
) +
    theme(aspect.ratio = 1) +
    labs(title = "CD4 T Cell Subtypes on UMAP")

save_plot(
    filename = here::here("results/02_data_processing/annotation_plots/gating_strategy/cd4_subtype_umap"),
    width = 8,
    height = 6,
    plot = cd4_subtype_umap,
    create_dir = TRUE
)
```

# 3. Abundance distribution

This section generates  abundance distribution plots for all markers in the dataset, providing detailed visualizations of marker expression patterns across different experimental dimensions. For each marker, plots are generated using both normalized CLR (centered log-ratio) values and raw count data, allowing users to assess expression patterns in both normalized and absolute terms.

The analysis creates multiple visualization perspectives to facilitate  exploration:

- **Per sample**: Shows marker abundance distributions across all samples, useful for identifying sample-specific expression patterns or batch effects
- **Per cell type**: Displays marker expression within each annotated cell type, helping to identify cell type-specific markers and expression heterogeneity
- **Per cell type faceted by sample**: Combines cell type and sample information, enabling identification of cell type-specific expression differences between samples
- **Per condition faceted by cell type**: Compares marker abundance across experimental conditions within each cell type, useful for identifying condition-specific changes in marker expression

These visualizations are valuable for quality control, marker validation, and exploratory data analysis. They help identify markers with high variability, detect potential batch effects, and understand marker expression patterns across the experimental design.

**Note**: This section generates a large number of plots (multiple plots per marker), which can take approximately 10-15 minutes to complete depending on the number of markers and cells in the dataset.

```{r}

# PLOTTING THIS TAKES AROUND 10-15 MIN!

all_markers <- rownames(pg_data)

# Collect CLR and count data
plot_data_clr <- FetchData(pg_data, c("sample_alias", "cell_annotation", "condition", all_markers), layer = "data")
plot_data_cts <- FetchData(pg_data, c("sample_alias", "cell_annotation", "condition", all_markers), layer = "counts")

# concatenate CLR and count data
plot_data <- plot_data_clr |>
    pivot_longer(cols = all_of(all_markers), names_to = "marker", values_to = "abundance") |>
    mutate(metric = "CLR") |>
    bind_rows(
        plot_data_cts |>
            pivot_longer(cols = all_of(all_markers), names_to = "marker", values_to = "abundance") |>
            mutate(metric = "Count")
    )

plot_data |>
    group_by(marker) |>
    group_map(~ {
        # Create 2 plots for the current marker: CLR per sample, Count per sample
        p_clr_per_sample <- .x |> # .x contains the grouped data for the current marker
            filter(metric == "CLR") |>
            ggplot(aes(sample_alias, abundance, color = sample_alias)) +
            ggbeeswarm::geom_quasirandom() +
            theme_bw() +
            theme(legend.position = "none") +
            scale_color_manual(values = sample_palette) +
            labs(title = "CLR per sample", x = "", y = "CLR")

        # Set y limits to the 99th percentile per marker
        count_99 <- quantile(.x$abundance[.x$metric == "Count"], 0.99, na.rm = TRUE)
        p_count_per_sample <- .x |>
            filter(metric == "Count") |>
            ggplot(aes(sample_alias, abundance, color = sample_alias)) +
            ggbeeswarm::geom_quasirandom() +
            theme_bw() +
            theme(legend.position = "none") +
            scale_color_manual(values = sample_palette) +
            labs(title = "Count per sample", x = "", y = "Count") +
            ylim(NA, count_99)

        # Save per sample plots
        p_abundance_distribution_per_sample <- (p_clr_per_sample + p_count_per_sample) + patchwork::plot_annotation(title = paste0("Marker: ", .y)) # .y contains the name of the current marker
        save_plot(
            filename = glue::glue(here::here("results/03_abundance/abundance_plots/per_sample/{.y}_abundance_distribution_per_sample")),
            width = 12,
            height = 8,
            plot = p_abundance_distribution_per_sample
        )

        # Create 2 plots for the current marker: CLR per cell type, Count per cell type
        p_clr_per_celltype <- .x |>
            filter(metric == "CLR") |>
            ggplot(aes(cell_annotation, abundance, color = cell_annotation)) +
            ggbeeswarm::geom_quasirandom() +
            theme_bw() +
            theme(legend.position = "none") +
            scale_color_manual(values = cell_palette) +
            labs(title = "CLR per cell type", x = "", y = "CLR")

        p_count_per_celltype <- .x |>
            filter(metric == "Count") |>
            ggplot(aes(cell_annotation, abundance, color = cell_annotation)) +
            ggbeeswarm::geom_quasirandom() +
            theme_bw() +
            theme(legend.position = "none") +
            scale_color_manual(values = cell_palette) +
            labs(title = "Count per cell type", x = "", y = "Count") +
            ylim(NA, count_99)

        # Save per cell type plots
        p_abundance_distribution_per_celltype <- (p_clr_per_celltype + p_count_per_celltype) + patchwork::plot_annotation(title = paste0("Marker: ", .y)) # .y contains the name of the current marker
        save_plot(
            filename = glue::glue(here::here("results/03_abundance/abundance_plots/per_cell_type/{.y}_abundance_distribution_per_celltype")),
            width = 12,
            height = 8,
            plot = p_abundance_distribution_per_celltype
        )

        # Save per cell type plots faceted by sample
        p_clr_per_celltype_faceted_by_sample <- .x |>
            filter(metric == "CLR") |>
            ggplot(aes(cell_annotation, abundance, color = cell_annotation)) +
            ggbeeswarm::geom_quasirandom() +
            theme_bw() +
            theme(legend.position = "none") +
            scale_color_manual(values = cell_palette) +
            labs(title = "CLR per cell type", x = "", y = "CLR") +
            facet_wrap(~sample_alias)

        p_count_per_celltype_faceted_by_sample <- .x |>
            filter(metric == "Count") |>
            ggplot(aes(cell_annotation, abundance, color = cell_annotation)) +
            ggbeeswarm::geom_quasirandom() +
            theme_bw() +
            theme(legend.position = "none") +
            scale_color_manual(values = cell_palette) +
            labs(title = "Count per cell type", x = "", y = "Count") +
            facet_wrap(~sample_alias) +
            ylim(NA, count_99)

        p_abundance_distribution_per_celltype_faceted_by_sample <- (p_clr_per_celltype_faceted_by_sample + p_count_per_celltype_faceted_by_sample) + patchwork::plot_annotation(title = paste0("Marker: ", .y)) # .y contains the name of the current marker
        save_plot(
            filename = glue::glue(here::here("results/03_abundance/abundance_plots/per_cell_type_faceted_by_sample/{.y}_abundance_distribution_per_celltype_faceted_by_sample")),
            width = 16,
            height = 12,
            plot = p_abundance_distribution_per_celltype_faceted_by_sample
        )

        # Create plots for the current marker: CLR per condition per cell type, Count per condition per cell type
        p_clr_per_condition_faceted_by_celltype <- .x |>
            filter(metric == "CLR") |>
            ggplot(aes(condition, abundance, color = condition)) +
            ggbeeswarm::geom_quasirandom() +
            theme_bw() +
            theme(legend.position = "none") +
            scale_color_manual(values = condition_palette) +
            labs(title = "CLR per condition per cell type", x = "", y = "CLR") +
            facet_wrap(~cell_annotation)

        p_count_per_condition_faceted_by_celltype <- .x |>
            filter(metric == "Count") |>
            ggplot(aes(condition, abundance, color = condition)) +
            ggbeeswarm::geom_quasirandom() +
            theme_bw() +
            theme(legend.position = "none") +
            scale_color_manual(values = condition_palette) +
            labs(title = "Count per condition per cell type", x = "", y = "Count") +
            facet_wrap(~cell_annotation) +
            ylim(NA, count_99)

        p_abundance_distribution_per_condition_faceted_by_celltype <- (p_clr_per_condition_faceted_by_celltype + p_count_per_condition_faceted_by_celltype) + patchwork::plot_annotation(title = paste0("Marker: ", .y)) # .y contains the name of the current marker
        save_plot(
            filename = glue::glue(here::here("results/03_abundance/abundance_plots/per_condition_faceted_by_celltype/{.y}_abundance_distribution_per_condition_faceted_by_celltype")),
            width = 16,
            height = 12,
            plot = p_abundance_distribution_per_condition_faceted_by_celltype
        )
    })

```


# 4. Colocalization score

## 4.1. Heatmap of colocalization score

This section generates colocalization heatmaps that visualize the spatial proximity patterns between marker pairs expressed in the dataset. Colocalization scores quantify the extent to which two markers are found in close proximity to each other within cells, providing insights into protein co-localization and potential molecular interactions or co-expression patterns.

### 4.1.1. Per Sample

For each sample, a heatmap is created showing the mean log2 ratio of observed versus expected colocalization (co-localization score) for all expressed marker pairs. The color intensity represents the strength of colocalization (positive values indicate higher-than-expected colocalization, negative values indicate lower-than-expected), while the dot size represents the percentage of cells in which the marker pair was detected. This dual encoding allows simultaneous assessment of both the magnitude and prevalence of colocalization events.

To enable direct comparison across samples, all heatmaps use a consistent marker ordering (determined by hierarchical clustering of a reference sample) and identical color and size scales. This standardization ensures that differences in colocalization patterns between samples reflect true biological or experimental differences rather than visualization artifacts. Individual heatmaps are saved for each sample, and a combined visualization showing all samples side-by-side is also generated to facilitate cross-sample comparison.

```{r}
#| label: colocalization_raw_heatmap_per_sample

expressed_markers <- unique(unlist(markers_to_test))

proximity_data <- ProximityScores(pg_data,
    lazy = TRUE,
    meta_data_columns = c("sample_alias")
) |>
    filter(marker_1 %in% expressed_markers) |>
    filter(marker_2 %in% expressed_markers) |>
    collect()

summary_proximity_data <- SummarizeProximityScores(proximity_data, group_vars = "sample_alias")
rm(proximity_data)
gc()

# Create reference heatmap to determine marker ordering
reference_heatmap <- ColocalizationHeatmap(
    summary_proximity_data |> filter(sample_alias == "S1_resting"),
    marker1_col = "marker_1",
    marker2_col = "marker_2",
    value_col = "mean_log2_ratio",
    size_col = "pct_detected",
    type = "dots",
    cluster_cols = TRUE,
    cluster_rows = TRUE,
    legend_range = c(-1, 1)
)

marker_order <- layer_scales(reference_heatmap)$y$range$range

# Calculate number of markers for dynamic plot sizing
n_markers <- length(marker_order)

# Calculate plot dimensions based on number of markers
# Base size + scaling factor to ensure readability across different marker counts
# Minimum size ensures small marker sets are still readable, scaling factor accommodates larger sets
base_width <- 6
base_height <- 6
marker_scaling_factor <- 0.15 # Additional inches per marker
plot_width <- max(base_width, base_width + (n_markers - 20) * marker_scaling_factor)
plot_height <- max(base_height, base_height + (n_markers - 20) * marker_scaling_factor)

# Calculate global scale ranges across all samples
# This ensures all heatmaps have consistent color and size scales for fair comparison
global_value_range <- range(summary_proximity_data$mean_log2_ratio, na.rm = TRUE)
global_size_range <- range(summary_proximity_data$pct_detected, na.rm = TRUE)

# Use symmetric range for value scale (centered around 0) to match legend_range behavior
# But ensure it covers the actual data range
max_abs_value <- max(abs(global_value_range))
global_value_limits <- c(-max_abs_value, max_abs_value)

# Get all unique sample_alias values
unique_samples <- unique(summary_proximity_data$sample_alias)

# Create heatmaps for each sample using the reference marker order
# Store plots in a list to combine them later
sample_heatmaps <- list()

for (sample in unique_samples) {
    # Filter data for current sample
    sample_data <- summary_proximity_data |> filter(sample_alias == sample)

    tmp_plot <- ColocalizationHeatmap(
        sample_data,
        marker1_col = "marker_1",
        marker2_col = "marker_2",
        value_col = "mean_log2_ratio",
        size_col = "pct_detected",
        type = "dots",
        cluster_cols = FALSE,
        cluster_rows = FALSE,
        legend_range = global_value_limits # Use global range for consistent scaling
    ) +
        ggtitle(glue::glue("Raw colocalization heatmap for {sample}"))

    # Apply the reference marker order to the ggplot object by modifying the scales
    # This ensures all heatmaps have consistent marker ordering
    tmp_plot <- tmp_plot +
        scale_x_discrete(limits = marker_order, drop = FALSE, position = "top") +
        scale_y_discrete(limits = rev(marker_order), drop = FALSE)

    # Ensure consistent scales across all samples for fair comparison
    # The legend_range parameter already handles the value scale, but we need to ensure
    # the size scale is also consistent. Try both fill and color aesthetics in case
    # ColocalizationHeatmap uses either one
    plot_layers <- tmp_plot$layers
    has_fill <- any(sapply(plot_layers, function(l) "fill" %in% names(l$mapping) || "fill" %in% names(l$aes_params)))
    has_color <- any(sapply(plot_layers, function(l) "colour" %in% names(l$mapping) || "color" %in% names(l$mapping) || "colour" %in% names(l$aes_params)))

    # Apply value scale limits (legend_range should handle this, but ensure it's applied)
    if (has_fill) {
        tmp_plot <- tmp_plot + scale_fill_continuous(limits = global_value_limits, oob = scales::oob_squish)
    }
    if (has_color && !has_fill) {
        tmp_plot <- tmp_plot + scale_color_continuous(limits = global_value_limits, oob = scales::oob_squish)
    }

    # Ensure consistent size scale across all samples
    tmp_plot <- tmp_plot + scale_size_continuous(limits = global_size_range, range = c(0.2, 3))

    # Save individual plot with dimensions scaled to number of markers
    save_plot(
        filename = glue::glue(here::here("results/04_raw_proximity/raw_colocalization_heatmap_per_sample/{sample}_raw_colocalization_heatmap")),
        plot = tmp_plot,
        width = plot_width,
        height = plot_height,
        create_dir = TRUE
    )

    # Store plot for combined visualization
    sample_heatmaps[[sample]] <- tmp_plot
}

# Create a combined plot with all samples side by side
# Use patchwork to combine plots (install with: install.packages("patchwork"))
n_samples <- length(unique_samples)
n_rows <- 2
n_cols <- ceiling(n_samples / n_rows)
combined_plot <- patchwork::wrap_plots(sample_heatmaps, nrow = n_rows)

# Calculate combined plot dimensions: scale width by number of columns, height by number of rows
# Each subplot maintains its aspect ratio while accommodating multiple samples
combined_width <- plot_width * n_cols
combined_height <- plot_height * n_rows

# Save the combined plot
save_plot(
    filename = glue::glue(here::here("results/04_raw_proximity/raw_colocalization_heatmap_per_sample/all_samples_combined_raw_colocalization_heatmap")),
    plot = combined_plot,
    width = combined_width,
    height = combined_height,
    create_dir = TRUE
)
```

### 4.1.2. Per cluster

This section generates cell type-specific colocalization heatmaps, analyzing colocalization patterns separately for each annotated cell type. For each cell type, colocalization heatmaps are created using a set of markers relevant to that specific cell type (defined in `markers_to_test`). The analysis is performed on cells belonging to each cell type separately, then heatmaps are generated for each sample within that cell type. This enables identification of cell type-specific colocalization patterns and comparison of how these patterns vary across samples within the same cell type.

The visualization follows the same principles as the per-sample analysis: color intensity represents the mean log2 ratio of observed versus expected colocalization, and dot size indicates the percentage of cells in which the marker pair was detected. All heatmaps for a given cell type use consistent marker ordering and scales to enable direct comparison across samples. Individual heatmaps are saved for each sample-cell type combination, and combined visualizations showing all samples side-by-side are generated for each cell type. Results are organized in separate subdirectories for each cell type to facilitate navigation and interpretation.

```{r}
#| label: colocalization_raw_heatmap_per_cell_type

cell_types <- unique(pg_data[[]]$cell_annotation)

for (cell_type in cell_types) {
    expressed_markers <- markers_to_test[[cell_type]]
    pg_data_cell_type <- subset(pg_data, cell_annotation == cell_type)

    proximity_data <- ProximityScores(pg_data_cell_type,
        lazy = TRUE,
        meta_data_columns = c("sample_alias")
    ) |>
        filter(marker_1 %in% expressed_markers) |>
        filter(marker_2 %in% expressed_markers) |>
        collect()

    summary_proximity_data <- SummarizeProximityScores(proximity_data, group_vars = "sample_alias")
    rm(proximity_data)
    gc()

    # Create reference heatmap to determine marker ordering
    reference_heatmap <- ColocalizationHeatmap(
        summary_proximity_data |> filter(sample_alias == "S1_resting"),
        marker1_col = "marker_1",
        marker2_col = "marker_2",
        value_col = "mean_log2_ratio",
        size_col = "pct_detected",
        type = "dots",
        cluster_cols = TRUE,
        cluster_rows = TRUE,
        legend_range = c(-1, 1)
    )

    marker_order <- layer_scales(reference_heatmap)$y$range$range

    # Calculate number of markers for dynamic plot sizing
    n_markers <- length(marker_order)

    # Calculate plot dimensions based on number of markers
    # Base size + scaling factor to ensure readability across different marker counts
    # Minimum size ensures small marker sets are still readable, scaling factor accommodates larger sets
    base_width <- 6
    base_height <- 6
    marker_scaling_factor <- 0.15 # Additional inches per marker
    plot_width <- max(base_width, base_width + (n_markers - 20) * marker_scaling_factor)
    plot_height <- max(base_height, base_height + (n_markers - 20) * marker_scaling_factor)

    # Calculate global scale ranges across all samples for this cell type
    # This ensures all heatmaps have consistent color and size scales for fair comparison
    global_value_range <- range(summary_proximity_data$mean_log2_ratio, na.rm = TRUE)
    global_size_range <- range(summary_proximity_data$pct_detected, na.rm = TRUE)

    # Use symmetric range for value scale (centered around 0) to match legend_range behavior
    # But ensure it covers the actual data range
    max_abs_value <- max(abs(global_value_range))
    global_value_limits <- c(-max_abs_value, max_abs_value)

    # Get all unique sample_alias values
    unique_samples <- unique(summary_proximity_data$sample_alias)

    # Create heatmaps for each sample using the reference marker order
    # Store plots in a list to combine them later
    sample_heatmaps <- list()

    for (sample in unique_samples) {
        # Filter data for current sample
        sample_data <- summary_proximity_data |> filter(sample_alias == sample)

        tmp_plot <- ColocalizationHeatmap(
            sample_data,
            marker1_col = "marker_1",
            marker2_col = "marker_2",
            value_col = "mean_log2_ratio",
            size_col = "pct_detected",
            type = "dots",
            cluster_cols = FALSE,
            cluster_rows = FALSE,
            legend_range = global_value_limits # Use global range for consistent scaling
        ) +
            ggtitle(glue::glue("Raw colocalization heatmap for {cell_type} in {sample}"))

        # Apply the reference marker order to the ggplot object by modifying the scales
        # This ensures all heatmaps have consistent marker ordering
        tmp_plot <- tmp_plot +
            scale_x_discrete(limits = marker_order, drop = FALSE, position = "top") +
            scale_y_discrete(limits = rev(marker_order), drop = FALSE)

        # Ensure consistent scales across all samples for fair comparison
        # The legend_range parameter already handles the value scale, but we need to ensure
        # the size scale is also consistent. Try both fill and color aesthetics in case
        # ColocalizationHeatmap uses either one
        plot_layers <- tmp_plot$layers
        has_fill <- any(sapply(plot_layers, function(l) "fill" %in% names(l$mapping) || "fill" %in% names(l$aes_params)))
        has_color <- any(sapply(plot_layers, function(l) "colour" %in% names(l$mapping) || "color" %in% names(l$mapping) || "colour" %in% names(l$aes_params)))

        # Apply value scale limits (legend_range should handle this, but ensure it's applied)
        if (has_fill) {
            tmp_plot <- tmp_plot + scale_fill_continuous(limits = global_value_limits, oob = scales::oob_squish)
        }
        if (has_color && !has_fill) {
            tmp_plot <- tmp_plot + scale_color_continuous(limits = global_value_limits, oob = scales::oob_squish)
        }

        # Ensure consistent size scale across all samples
        tmp_plot <- tmp_plot + scale_size_continuous(limits = global_size_range, range = c(0.5, 5))

        # Save individual plot with dimensions scaled to number of markers
        save_plot(
            filename = glue::glue(here::here("results/04_raw_proximity/raw_colocalization_heatmap_per_cell_type/{cell_type}/{sample}_{cell_type}_raw_colocalization_heatmap")),
            plot = tmp_plot,
            width = plot_width,
            height = plot_height,
            create_dir = TRUE
        )

        # Store plot for combined visualization
        sample_heatmaps[[sample]] <- tmp_plot
    }

    # Create a combined plot with all samples side by side
    # Use patchwork to combine plots (install with: install.packages("patchwork"))
    n_samples <- length(unique_samples)
    n_rows <- 2
    n_cols <- ceiling(n_samples / n_rows)
    combined_plot <- patchwork::wrap_plots(sample_heatmaps, nrow = n_rows)

    # Calculate combined plot dimensions: scale width by number of columns, height by number of rows
    # Each subplot maintains its aspect ratio while accommodating multiple samples
    combined_width <- plot_width * n_cols
    combined_height <- plot_height * n_rows

    # Save the combined plot
    save_plot(
        filename = glue::glue(here::here("results/04_raw_proximity/raw_colocalization_heatmap_per_cell_type/{cell_type}/{cell_type}_all_samples_combined_raw_colocalization_heatmap")),
        plot = combined_plot,
        width = combined_width,
        height = combined_height,
        create_dir = TRUE
    )
}
```


## 4.2. Colocalization Score network

Network visualizations provide an alternative representation of colocalization patterns that can reveal complex relationships and connectivity structures between markers. Unlike heatmaps, which display all pairwise interactions in a matrix format, network graphs focus on the most significant colocalization relationships by filtering based on statistical and biological criteria. This filtering approach helps reduce visual clutter and highlights the most biologically relevant interactions.

In network visualizations, markers are represented as nodes, and colocalization relationships are shown as edges connecting pairs of markers. Edge color encodes the strength of colocalization (mean log2 ratio), while edge width represents the magnitude of the colocalization effect. The network layout algorithm positions nodes to minimize edge crossings and optimize visual clarity, with markers that share similar colocalization patterns tending to cluster together spatially.

The network is constructed by filtering marker pairs based on two key criteria: (1) a minimum colocalization threshold (`metric_threshold`), which can be applied to either positive colocalization only or absolute values, and (2) a minimum detection threshold (`pct_detected_threshold`), ensuring that only marker pairs detected in a sufficient proportion of cells are included. These filters help focus the visualization on robust, biologically meaningful colocalization relationships while reducing noise from weak or rare interactions.

This section provides two complementary analyses: a whole-dataset network visualization that captures global colocalization patterns across all cell types, and cell type-specific network visualizations that reveal how colocalization patterns differ between cell types. Users should select a focused set of markers of interest (`network_markers`) for network visualization, as including too many markers can make the network difficult to interpret.

### 4.2.1. All Cell Types

This section generates network visualizations that capture colocalization patterns across all cell types in the dataset, with separate network plots created for each sample. This whole-dataset approach provides a global view of marker colocalization relationships and enables comparison of how these relationships vary across samples. The analysis includes all cells regardless of cell type annotation, making it useful for identifying sample-wide colocalization patterns and detecting potential batch effects or experimental condition-specific changes.

Users should specify a focused set of `network_markers` for visualization, as network graphs become difficult to interpret with too many nodes. The filtering thresholds (`metric_threshold` and `pct_detected_threshold`) help focus the visualization on the most robust colocalization relationships. The `only_positive_mean_log2_ratio` parameter can be set to filter for only positive colocalization (markers that co-occur more frequently than expected) or to include both positive and negative colocalization relationships.

```{r}
#| label: colocalization_score_network_all_cell_types

network_markers <- c("CD52", "CD59", "CD81", "CD82", "CD43", "CD44", "CD45", "CD45RA", "CD45RB", "CD8")
only_positive_mean_log2_ratio <- TRUE
# Define the threshold for the metric and the percentage of detected cells - this will be used to filter the data for the network plot
metric_threshold <- 0.1
pct_detected_threshold <- 0.5

proximity_data <- ProximityScores(pg_data,
    lazy = TRUE,
    meta_data_columns = c("sample_alias")
) |>
    filter(marker_1 %in% network_markers) |>
    filter(marker_2 %in% network_markers) |>
    collect()

summary_proximity_data <- SummarizeProximityScores(proximity_data, group_vars = "sample_alias")
rm(proximity_data)
gc()

# Create the network plot: filter the data for the network plot based on the metric and the percentage of detected cells, then create the network plot - this is also where you can include additional filtering options, such as for example to only include positive mean_log2_ratio.
df_net <- summary_proximity_data %>%
    filter(if (only_positive_mean_log2_ratio) mean_log2_ratio > metric_threshold else abs(mean_log2_ratio) > metric_threshold) %>%
    filter(pct_detected > pct_detected_threshold) %>%
    mutate(from = marker_1, to = marker_2) %>%
    select(from, to, mean_log2_ratio, sample_alias)

network_plot_all_cell_types <- tidygraph::as_tbl_graph(df_net) %>%
    ggraph(layout = "kk") +
    geom_edge_link(aes(color = mean_log2_ratio, width = abs(mean_log2_ratio))) +
    geom_node_label(aes(label = name), size = 2.5) +
    scale_edge_color_gradientn(
        colours = c(
            "#1F395F", "#647DA3", "#BBC8DA",
            "#FFFFFF", "#F0BCCD", "#BE5F7D",
            "#781534"
        ),
        limits = c(-abs(max(df_net$mean_log2_ratio)), abs(max(df_net$mean_log2_ratio)))
    ) +
    scale_x_continuous(expand = c(0.2, 0.2)) +
    scale_y_continuous(expand = c(0.2, 0.2)) +
    facet_edges(~sample_alias, nrow = 2) +
    theme(
        aspect.ratio = 1,
        panel.background = element_rect(fill = "white", color = "black"),
        panel.spacing = unit(1, "lines")
    ) +
    labs(edge_color = "Mean\nlog2 ratio", title = "Colocalization Score Network All Cell Types", subtitle = glue::glue("Metric threshold{if (only_positive_mean_log2_ratio) ' (only positive)' else ' (absolute value)'}: {metric_threshold}, Percentage of cells above threshold: {pct_detected_threshold}")) +
    guides(edge_width = "none")

save_plot(
    filename = glue::glue(here::here("results/04_raw_proximity/colocalization_score_network/colocalization_score_network_all_cell_types/all_cell_types_colocalization_score_network")),
    plot = network_plot_all_cell_types,
    width = 12,
    height = 8,
    create_dir = TRUE
)

```


### 4.2.2. Per Cell Type

This section generates cell type-specific network visualizations, analyzing colocalization patterns within a single cell type of interest. By focusing on a specific cell type, this analysis can reveal cell type-specific colocalization signatures that may be masked in whole-dataset analyses. The network visualization helps identify which markers form strong colocalization relationships within the selected cell type and how these relationships vary across samples.

Users should specify the `cell_type_of_interest` variable to select which cell type to analyze. The same filtering criteria and visualization principles apply as in the all-cell-types analysis, but the proximity data is filtered to include only cells belonging to the specified cell type. This enables identification of cell type-specific marker interactions and comparison of how colocalization networks differ between cell types.

```{r}
#| label: colocalization_score_network_per_cell_type

network_markers <- c("CD52", "CD59", "CD81", "CD82", "CD43", "CD44", "CD45", "CD45RA", "CD45RB", "CD8")
cell_type_of_interest <- "CD8 T"
only_positive_mean_log2_ratio <- TRUE
# Define the threshold for the metric and the percentage of detected cells - this will be used to filter the data for the network plot
metric_threshold <- 0.1
pct_detected_threshold <- 0.5

proximity_data <- ProximityScores(pg_data,
    lazy = TRUE,
    meta_data_columns = c("sample_alias", "cell_annotation")
) |>
    filter(cell_annotation == cell_type_of_interest) |>
    filter(marker_1 %in% network_markers) |>
    filter(marker_2 %in% network_markers) |>
    filter(marker_1 != marker_2) |>
    collect()

summary_proximity_data <- SummarizeProximityScores(proximity_data, group_vars = "sample_alias")
rm(proximity_data)
gc()

# Create the network plot: filter the data for the network plot based on the metric and the percentage of detected cells, then create the network plot - this is also where you can include additional filtering options, such as for example to only include positive mean_log2_ratio.
df_net <- summary_proximity_data %>%
    filter(if (only_positive_mean_log2_ratio) mean_log2_ratio > metric_threshold else abs(mean_log2_ratio) > metric_threshold) %>%
    filter(pct_detected > pct_detected_threshold) %>%
    mutate(from = marker_1, to = marker_2) %>%
    select(from, to, mean_log2_ratio, sample_alias)

network_plot_cell_type_of_interest <- tidygraph::as_tbl_graph(df_net) %>%
    ggraph(layout = "kk") +
    geom_edge_link(aes(color = mean_log2_ratio, width = abs(mean_log2_ratio))) +
    geom_node_label(aes(label = name), size = 2.5) +
    scale_edge_color_gradientn(
        colours = c(
            "#1F395F", "#647DA3", "#BBC8DA",
            "#FFFFFF", "#F0BCCD", "#BE5F7D",
            "#781534"
        ),
        limits = c(-abs(max(df_net$mean_log2_ratio)), abs(max(df_net$mean_log2_ratio)))
    ) +
    scale_x_continuous(expand = c(0.2, 0.2)) +
    scale_y_continuous(expand = c(0.2, 0.2)) +
    facet_edges(~sample_alias, nrow = 2) +
    theme(
        aspect.ratio = 1,
        panel.background = element_rect(fill = "white", color = "black"),
        panel.spacing = unit(1, "lines")
    ) +
    labs(
        edge_color = "Mean\nlog2 ratio",
        title = glue::glue("Colocalization Score Network {cell_type_of_interest}"),
        subtitle = glue::glue(
            "Metric threshold{if (only_positive_mean_log2_ratio) ' (only positive)' else ' (absolute value)'}: {metric_threshold}, Percentage of cells above threshold: {pct_detected_threshold}"
        )
    ) +
    guides(edge_width = "none")

save_plot(
    filename = glue::glue(here::here("results/04_raw_proximity/colocalization_score_network/colocalization_score_network_per_cell_type/{cell_type_of_interest}/{cell_type_of_interest}_colocalization_score_network")),
    plot = network_plot_cell_type_of_interest,
    width = 12,
    height = 8,
    create_dir = TRUE
)

```




# 5. Statistical testing

## 5.1. Marker selection

```{r}
#| label: statistical_testing_marker_selection


```



## 5.2. Differential abundance analysis

Differential abundance analysis identifies markers that exhibit statistically significant differences in expression levels between experimental conditions. This analysis is essential for understanding how experimental treatments, disease states, or other conditions affect protein expression patterns within specific cell types.

The analysis is performed separately for each annotated cell type, comparing marker abundance between a target condition (e.g., stimulated, treated) and a reference condition (e.g., resting, control). For each marker within each cell type, a statistical test (typically the Wilcoxon rank-sum test, also known as the Mann-Whitney U test) is used to assess whether the distribution of marker abundance differs significantly between conditions. The Wilcoxon test is a non-parametric test that does not assume normal distribution of the data, making it robust for single-cell data which often exhibits non-normal distributions and high variability.

The test compares marker expression values (typically CLR-normalized values) between cells in the target condition versus cells in the reference condition. For each marker, the analysis reports an effect size (typically the difference in mean or median expression) and a p-value indicating the statistical significance of the difference. Multiple testing correction is applied using the false discovery rate (FDR) method to control for the large number of statistical tests performed across multiple markers and cell types.

Results are organized by cell type and include both the raw statistical outputs and summary visualizations. Markers with significant differences (typically p_adj < 0.05) and large effect sizes are of particular biological interest, as they represent markers whose expression is substantially altered by the experimental condition. This analysis enables identification of condition-specific markers, understanding of cellular responses to treatments, and discovery of biomarkers associated with different experimental states.

```{r}
#| label: statistical_testing_differential_abundance

## REMOVE THIS CHUNK AFTER TESTING

# Select 10 random markers to test for each cell type [TESTING]
markers_to_test <- lapply(unique(pg_data[[]]$cell_annotation), function(x) {
    sample(rownames(pg_data), size = 10)
})
names(markers_to_test) <- unique(pg_data[[]]$cell_annotation)

# create an additional condition factor to test multiple comparisons
pg_data$condition <- ifelse(pg_data$sample_alias == "S6_PHA", "PMA", pg_data$condition)
```


```{r}

differential_abundance_analysis <-
    names(markers_to_test) |>
    map(function(x) {
        tryCatch(
            {
                daa_res <- RunDAA(subset(pg_data, cell_annotation == x),
                    contrast_column = "condition",
                    reference = "resting",
                    features = markers_to_test[[x]]
                )
                daa_res$group <- x
                return(daa_res)
            },
            error = function(e) {
                message(paste("Error for cell type:", x, ":", e))
                return(NULL)
            }
        )
    }) |>
    bind_rows() |>
    mutate(p_adj = p.adjust(p, method = "fdr")) # recalculate the adjusted p values after multiple testing correction for all comparisons, not only per cell type

dir.create(here::here("results/05_statistical_testing/abundance"), recursive = TRUE, showWarnings = FALSE)

# Save the differential abundance analysis results to a file
saveRDS(differential_abundance_analysis, here::here("results/05_statistical_testing/abundance/differential_abundance_analysis.rds"))

# Save the differential abundance analysis results to a csv file
write.csv(differential_abundance_analysis, here::here("results/05_statistical_testing/abundance/differential_abundance_analysis.csv"), row.names = FALSE)

```    

### 5.2.1. Summarizing plots

Here we create two types of summarizing plots for the differential abundance analysis results:

1. **Volcano plot**: Shows the relationship between effect size (log fold change) and statistical significance (-log10 adjusted p-value) for each marker, faceted by cell type. This helps identify markers that are both significantly different and have large effect sizes.

2. **Heatmap**: Displays significant markers (adjusted p-value < 0.05) across all cell types, with colors representing the effect size. This provides an overview of which markers are differentially abundant in which cell types.

```{r}
#| label: statistical_testing_differential_abundance_plots

# Determine the effect size column name
effect_size_col <- "difference"

# Determine the feature/marker column name
feature_col <- "marker"

# Extract target and reference condition names from the dataframe
# Handle multiple target conditions by identifying all comparisons
reference_condition <- "resting" # Known from RunDAA call

# Determine target conditions - check if dataframe has target/contrast columns
if ("target" %in% colnames(differential_abundance_analysis)) {
    # If target column exists, get all unique targets
    target_conditions <- unique(differential_abundance_analysis$target)
    target_conditions <- target_conditions[!is.na(target_conditions)]
} else {
    stop("Could not find target column. Available columns: ", paste(colnames(differential_abundance_analysis), collapse = ", "))
}

# Prepare base data for plotting (add target column if it doesn't exist)
plot_data_base <- differential_abundance_analysis |>
    mutate(
        neg_log10_p_adj = -log10(p_adj),
        significant = p_adj < 0.05,
        effect_size = !!sym(effect_size_col),
        marker = !!sym(feature_col)
    )


# Loop through each target condition and create plots
for (target_condition in target_conditions) {
    # Filter data for this target (if target column exists)
    if ("target" %in% colnames(plot_data_base)) {
        plot_data <- plot_data_base |>
            filter(target == target_condition)
    }

    # Skip if no data for this target
    if (nrow(plot_data) == 0) {
        warning(paste("No data found for target condition:", target_condition))
        next
    }

    # Create directory name based on target and reference
    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here("results/05_statistical_testing/differential_abundance_plots", comparison_dir)

    # Plot 1: Volcano plot showing effect size vs significance, faceted by cell type
    # This plot helps identify markers that are both significantly different and have large effect sizes
    # Identify top 5 most extreme effect sizes per cell type for labeling
    top_markers_faceted <- plot_data |>
        group_by(group) |>
        slice_max(order_by = abs(effect_size), n = 5) |>
        ungroup()

    volcano_plot_differential_abundance <- plot_data |>
        ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
        geom_point() +
        scale_color_gradient2(
            low = "#0066FFFF",
            mid = "#74D055FF",
            high = "#FF0000FF",
            midpoint = 0
        ) +
        geom_hline(yintercept = 0) +
        geom_vline(xintercept = 0) +
        geom_vline(xintercept = c(-1, 1), col = "grey") +
        coord_cartesian(xlim = c(1.1 * min(plot_data$effect_size, na.rm = TRUE), 1.1 * max(plot_data$effect_size, na.rm = TRUE))) +
        facet_wrap(~group, scales = "free") +
        geom_text(
            data = top_markers_faceted,
            aes(label = marker),
            hjust = -0.2,
            vjust = 0
        ) +
        theme_bw() +
        labs(
            title = paste("Differential abundance analysis: Volcano plot (", target_condition, " vs ", reference_condition, ")", sep = ""),
            subtitle = "Effect size vs statistical significance by cell type",
            x = paste("Effect size (", effect_size_col, ")", sep = ""),
            y = "-log10(adjusted p-value)",
            color = "Effect size"
        ) +
        theme(
            strip.background = element_rect(fill = "lightblue", color = "black"),
            panel.grid.minor = element_blank(),
            panel.grid.major = element_blank()
        )

    # Save faceted volcano plot in the comparison directory
    save_plot(
        filename = file.path(base_output_dir, "differential_abundance_volcano"),
        plot = volcano_plot_differential_abundance,
        width = 12,
        height = 8,
        create_dir = TRUE
    )

    # Plot 2: Non-faceted volcano plots per cell type
    # Create individual folders per cell type and save non-faceted volcano plots for each cell type
    # Get unique cell types (groups)
    unique_cell_types <- unique(plot_data$group)

    # For each cell type, create a non-faceted volcano plot
    for (cell_type in unique_cell_types) {
        # Create directory for this cell type
        cell_type_dir <- file.path(base_output_dir, cell_type)
        dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

        # Filter data for this cell type
        cell_type_plot_data <- plot_data |>
            filter(group == cell_type)

        # Create non-faceted volcano plot for this cell type
        # Identify top 5 most extreme effect sizes for labeling
        top_markers_cell_type <- cell_type_plot_data |>
            slice_max(order_by = abs(effect_size), n = 5)

        volcano_plot_cell_type <- cell_type_plot_data |>
            ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
            geom_point() +
            scale_color_gradient2(
                low = "#0066FFFF",
                mid = "#74D055FF",
                high = "#FF0000FF",
                midpoint = 0
            ) +
            geom_hline(yintercept = 0) +
            geom_vline(xintercept = 0) +
            geom_vline(xintercept = c(-1, 1), col = "grey") +
            coord_cartesian(xlim = c(1.1 * min(cell_type_plot_data$effect_size, na.rm = TRUE), 1.1 * max(cell_type_plot_data$effect_size, na.rm = TRUE))) +
            geom_text(
                data = top_markers_cell_type,
                aes(label = marker),
                hjust = -0.2,
                vjust = 0
            ) +
            theme_bw() +
            labs(
                title = paste("Differential abundance analysis: Volcano plot", sep = ""),
                subtitle = paste("Cell type: ", cell_type, " (", target_condition, " vs ", reference_condition, ")", sep = ""),
                x = paste("Effect size (", effect_size_col, ")", sep = ""),
                y = "-log10(adjusted p-value)",
                color = "Effect size"
            ) +
            theme(
                plot.title = element_text(size = 14, face = "bold"),
                plot.subtitle = element_text(size = 12),
                panel.grid.minor = element_blank(),
                panel.grid.major = element_blank()
            )

        # Save non-faceted volcano plot
        save_plot(
            filename = file.path(cell_type_dir, "volcano_plot"),
            plot = volcano_plot_cell_type,
            width = 8,
            height = 6,
            create_dir = FALSE # Directory already created
        )
    }

    # Plot 3: Heatmap showing significant markers across cell types
    # This provides an overview of which markers are differentially abundant in which cell types
    # Filter for significant markers and prepare data for heatmap
    significant_markers <- plot_data |>
        filter(significant) |>
        select(marker, group, effect_size) |>
        # Pivot to wide format for heatmap
        pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
        column_to_rownames("marker") |>
        as.matrix()

    # Only create heatmap if there are significant markers
    if (nrow(significant_markers) > 0 && ncol(significant_markers) > 0) {
        # Determine color scale breaks based on robust data range (use 1st and 99th percentiles to reduce outlier influence)
        effect_vals <- as.numeric(significant_markers)
        effect_quantiles <- quantile(effect_vals, probs = c(0.01, 0.99), na.rm = TRUE)
        max_abs_effect <- max(abs(effect_quantiles))
        breaks_range <- seq(-max_abs_effect, max_abs_effect, length.out = 101)

        heatmap_differential_abundance <- significant_markers |>
            pheatmap(
                color = PixelgenGradient(100, "BluesCherry"),
                breaks = breaks_range,
                cellwidth = 12,
                cellheight = 8,
                cluster_rows = TRUE,
                cluster_cols = TRUE,
                heatmap_legend_param = list(title = "Effect size"),
                fontsize = 8,
                fontsize_row = 6,
                fontsize_col = 8
            ) |>
            as.ggplot()

        save_plot(
            filename = file.path(base_output_dir, "differential_abundance_heatmap"),
            plot = heatmap_differential_abundance,
            width = 10,
            height = max(6, nrow(significant_markers) * 0.3),
            create_dir = TRUE
        )
    } else {
        warning(paste("No significant markers found for heatmap visualization (p_adj < 0.05) for", target_condition, "vs", reference_condition))
    }
}
```

## 5.3. Differential Clustering Analysis

Differential clustering analysis identifies markers that exhibit statistically significant differences in spatial clustering patterns between experimental conditions. In this context, "clustering" refers to the spatial aggregation of a marker with itself—that is, the extent to which a marker tends to form clusters or aggregates within cells rather than being uniformly distributed.

The analysis is performed separately for each cell type, comparing clustering patterns between target and reference conditions. A statistical test (typically the Wilcoxon rank-sum test) is used to assess whether the distribution of clustering scores (log2 ratio of observed versus expected self-proximity) differs significantly between conditions. Multiple testing correction is applied using FDR to control for the number of statistical tests performed across markers and cell types.

```{r}
#| label: statistical_testing_differential_proximity

differential_clustering_analysis <-
    names(markers_to_test) |>
    map(function(x) {
        tryCatch(
            {
                # Store markers in a variable to ensure proper evaluation in lazy query context
                # Convert to character vector to ensure proper SQL translation
                markers_vec <- as.character(markers_to_test[[x]])
                proximity_data <- ProximityScores(subset(pg_data, cell_annotation == x),
                    lazy = TRUE,
                    meta_data_columns = c("sample_alias", "condition")
                ) |>
                    filter(marker_1 == marker_2) |>
                    filter(marker_1 %in% markers_vec) |>
                    filter(marker_2 %in% markers_vec) |>
                    collect()

                res <- DifferentialProximityAnalysis(proximity_data,
                    contrast_column = "condition",
                    reference = "resting",
                    proximity_metric = "log2_ratio",
                    min_n_obs = 5
                )
                res$group <- x
                return(res)
            },
            error = function(e) {
                message(paste("Error for cell type:", x, ":", e))
                return(NULL)
            }
        )
    }) |>
    bind_rows() |>
    mutate(p_adj = p.adjust(p, method = "fdr")) # recalculate the adjusted p values after multiple testing correction for all comparisons, not only per cell type

dir.create(here::here("results/05_statistical_testing/clustering"), recursive = TRUE, showWarnings = FALSE)

# Save the differential abundance analysis results to a file
saveRDS(differential_clustering_analysis, here::here("results/05_statistical_testing/clustering/differential_clustering_analysis.rds"))

# Save the differential abundance analysis results to a csv file
write.csv(differential_clustering_analysis, here::here("results/05_statistical_testing/clustering/differential_clustering_analysis.csv"), row.names = FALSE)

```


### 5.3.1. Summarizing plots

Here we create two types of summarizing plots for the differential clustering analysis results:

1. **Volcano plot**: Shows the relationship between effect size (log fold change) and statistical significance (-log10 adjusted p-value) for each marker, faceted by cell type. This helps identify markers that are both significantly different and have large effect sizes.

2. **Heatmap**: Displays significant markers (adjusted p-value < 0.05) across all cell types, with colors representing the effect size. This provides an overview of which markers are differentially clustered in which cell types.

```{r}
#| label: statistical_testing_differential_clustering_plots

# Determine the effect size column name
effect_size_col <- "diff_median"

# Determine the feature/marker column name
feature_col <- "marker_1"

# Extract target and reference condition names from the dataframe
# Handle multiple target conditions by identifying all comparisons
reference_condition <- "resting" # Known from DifferentialProximityAnalysis call

# Determine target conditions - check if dataframe has target/contrast columns
if ("target" %in% colnames(differential_clustering_analysis)) {
    # If target column exists, get all unique targets
    target_conditions <- unique(differential_clustering_analysis$target)
    target_conditions <- target_conditions[!is.na(target_conditions)]
} else {
    stop("Could not find target column. Available columns: ", paste(colnames(differential_clustering_analysis), collapse = ", "))
}

# Prepare base data for plotting (add target column if it doesn't exist)
plot_data_base <- differential_clustering_analysis |>
    mutate(
        neg_log10_p_adj = -log10(p_adj),
        significant = p_adj < 0.05,
        effect_size = !!sym(effect_size_col),
        marker = !!sym(feature_col)
    )



# Loop through each target condition and create plots
for (target_condition in target_conditions) {
    # Filter data for this target (if target column exists)
    if ("target" %in% colnames(plot_data_base)) {
        plot_data <- plot_data_base |>
            filter(target == target_condition)
    }

    # Skip if no data for this target
    if (nrow(plot_data) == 0) {
        warning(paste("No data found for target condition:", target_condition))
        next
    }

    # Create directory name based on target and reference
    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here("results/05_statistical_testing/clustering/differential_clustering_plots", comparison_dir)

    # Plot 1: Volcano plot showing effect size vs significance, faceted by cell type
    # This plot helps identify markers that are both significantly different and have large effect sizes
    # Identify top 5 most extreme effect sizes per cell type for labeling
    top_markers_faceted <- plot_data |>
        group_by(group) |>
        slice_max(order_by = abs(effect_size), n = 5) |>
        ungroup()

    volcano_plot_differential_clustering <- plot_data |>
        ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
        geom_point() +
        scale_color_gradient2(
            low = "#0066FFFF",
            mid = "#74D055FF",
            high = "#FF0000FF",
            midpoint = 0
        ) +
        geom_hline(yintercept = 0) +
        geom_vline(xintercept = 0) +
        geom_vline(xintercept = c(-1, 1), col = "grey") +
        coord_cartesian(xlim = c(1.1 * min(plot_data$effect_size, na.rm = TRUE), 1.1 * max(plot_data$effect_size, na.rm = TRUE))) +
        facet_wrap(~group, scales = "free") +
        geom_text(
            data = top_markers_faceted,
            aes(label = marker),
            hjust = -0.2,
            vjust = 0
        ) +
        theme_bw() +
        labs(
            title = paste("Differential clustering analysis: Volcano plot (", target_condition, " vs ", reference_condition, ")", sep = ""),
            subtitle = "Effect size vs statistical significance by cell type",
            x = paste("Effect size (", effect_size_col, ")", sep = ""),
            y = "-log10(adjusted p-value)",
            color = "Effect size"
        ) +
        theme(
            strip.background = element_rect(fill = "lightblue", color = "black"),
            panel.grid.minor = element_blank(),
            panel.grid.major = element_blank()
        )

    # Save faceted volcano plot in the comparison directory
    save_plot(
        filename = file.path(base_output_dir, "differential_clustering_volcano"),
        plot = volcano_plot_differential_clustering,
        width = 12,
        height = 8,
        create_dir = TRUE
    )

    # Plot 2: Non-faceted volcano plots per cell type
    # Create individual folders per cell type and save non-faceted volcano plots for each cell type
    # Get unique cell types (groups)
    unique_cell_types <- unique(plot_data$group)

    # For each cell type, create a non-faceted volcano plot
    for (cell_type in unique_cell_types) {
        # Create directory for this cell type
        cell_type_dir <- file.path(base_output_dir, cell_type)
        dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

        # Filter data for this cell type
        cell_type_plot_data <- plot_data |>
            filter(group == cell_type)

        # Create non-faceted volcano plot for this cell type
        # Identify top 5 most extreme effect sizes for labeling
        top_markers_cell_type <- cell_type_plot_data |>
            slice_max(order_by = abs(effect_size), n = 5)

        volcano_plot_cell_type <- cell_type_plot_data |>
            ggplot(aes(x = effect_size, y = neg_log10_p_adj, color = effect_size)) +
            geom_point() +
            scale_color_gradient2(
                low = "#0066FFFF",
                mid = "#74D055FF",
                high = "#FF0000FF",
                midpoint = 0
            ) +
            geom_hline(yintercept = 0) +
            geom_vline(xintercept = 0) +
            geom_vline(xintercept = c(-1, 1), col = "grey") +
            coord_cartesian(xlim = c(1.1 * min(cell_type_plot_data$effect_size, na.rm = TRUE), 1.1 * max(cell_type_plot_data$effect_size, na.rm = TRUE))) +
            geom_text(
                data = top_markers_cell_type,
                aes(label = marker),
                hjust = -0.2,
                vjust = 0
            ) +
            theme_bw() +
            labs(
                title = paste("Differential clustering analysis: Volcano plot", sep = ""),
                subtitle = paste("Cell type: ", cell_type, " (", target_condition, " vs ", reference_condition, ")", sep = ""),
                x = paste("Effect size (", effect_size_col, ")", sep = ""),
                y = "-log10(adjusted p-value)",
                color = "Effect size"
            ) +
            theme(
                plot.title = element_text(size = 14, face = "bold"),
                plot.subtitle = element_text(size = 12),
                panel.grid.minor = element_blank(),
                panel.grid.major = element_blank()
            )

        # Save non-faceted volcano plot
        save_plot(
            filename = file.path(cell_type_dir, "volcano_plot"),
            plot = volcano_plot_cell_type,
            width = 8,
            height = 6,
            create_dir = FALSE # Directory already created
        )
    }

    # Plot 3: Heatmap showing significant markers across cell types
    # This provides an overview of which markers are differentially clustered in which cell types
    # Filter for significant markers and prepare data for heatmap
    significant_markers <- plot_data |>
        filter(significant) |>
        select(marker, group, effect_size) |>
        # Pivot to wide format for heatmap
        pivot_wider(names_from = group, values_from = effect_size, values_fill = 0) |>
        column_to_rownames("marker") |>
        as.matrix()

    # Only create heatmap if there are significant markers
    if (nrow(significant_markers) > 0 && ncol(significant_markers) > 0) {
        # Determine color scale breaks based on robust data range (use 1st and 99th percentiles to reduce outlier influence)
        effect_vals <- as.numeric(significant_markers)
        effect_quantiles <- quantile(effect_vals, probs = c(0.01, 0.99), na.rm = TRUE)
        max_abs_effect <- max(abs(effect_quantiles))
        breaks_range <- seq(-max_abs_effect, max_abs_effect, length.out = 101)

        heatmap_differential_clustering <- significant_markers |>
            pheatmap(
                color = PixelgenGradient(100, "BluesCherry"),
                breaks = breaks_range,
                cellwidth = 12,
                cellheight = 8,
                cluster_rows = TRUE,
                cluster_cols = TRUE,
                heatmap_legend_param = list(title = "Effect size"),
                fontsize = 8,
                fontsize_row = 6,
                fontsize_col = 8
            ) |>
            as.ggplot()

        save_plot(
            filename = file.path(base_output_dir, "differential_clustering_heatmap"),
            plot = heatmap_differential_clustering,
            width = 10,
            height = max(6, nrow(significant_markers) * 0.3),
            create_dir = TRUE
        )
    } else {
        warning(paste("No significant markers found for heatmap visualization (p_adj < 0.05) for", target_condition, "vs", reference_condition))
    }
}
```

## 5.4. Differential Colocalization Analysis

Differential colocalization analysis identifies markers that exhibit statistically significant differences in spatial proximity patterns between experimental conditions. The analysis is performed separately for each cell type, comparing colocalization patterns between target and reference conditions. A statistical test is used to assess whether the distribution of colocalization scores (log2 ratio of observed versus expected proximity) differs significantly between conditions. Multiple testing correction is applied using FDR to control for the number of statistical tests performed across markers and cell types.


```{r}
#| label: statistical_testing_differential_colocalization


differential_colocalization_analysis <-
    names(markers_to_test) |>
    map(function(x) {
        tryCatch(
            {
                # Store markers in a variable to ensure proper evaluation in lazy query context
                # Convert to character vector to ensure proper SQL translation
                markers_vec <- as.character(markers_to_test[[x]])
                proximity_data <- ProximityScores(subset(pg_data, cell_annotation == x),
                    lazy = TRUE,
                    meta_data_columns = c("sample_alias", "condition")
                ) |>
                    filter(marker_1 %in% markers_vec) |>
                    filter(marker_2 %in% markers_vec) |>
                    collect()

                res <- DifferentialProximityAnalysis(proximity_data,
                    contrast_column = "condition",
                    reference = "resting",
                    proximity_metric = "log2_ratio",
                    min_n_obs = 5
                )
                res$group <- x
                return(res)
            },
            error = function(e) {
                message(paste("Error for cell type:", x, ":", e))
                return(NULL)
            }
        )
    }) |>
    bind_rows() |>
    mutate(p_adj = p.adjust(p, method = "fdr")) # recalculate the adjusted p values after multiple testing correction for all comparisons, not only per cell type

dir.create(here::here("results/05_statistical_testing/colocalization"), recursive = TRUE, showWarnings = FALSE)

# Save the differential abundance analysis results to a file
saveRDS(differential_colocalization_analysis, here::here("results/05_statistical_testing/colocalization/differential_colocalization_analysis.rds"))

# Save the differential abundance analysis results to a csv file
write.csv(differential_colocalization_analysis, here::here("results/05_statistical_testing/colocalization/differential_colocalization_analysis.csv"), row.names = FALSE)


```

### 5.4.1. Summarizing plots

Here we create summarizing plots for the differential colocalization analysis results:

1. **Colocalization Heatmap**: Displays the difference in medians for marker pairs, with dot size representing statistical significance (-log10 adjusted p-value). This provides an overview of which marker pairs are differentially colocalized for each cell type and condition comparison.

```{r}
#| label: statistical_testing_differential_colocalization_plot

# Extract target and reference condition names from the dataframe
reference_condition <- "resting" # Known from DifferentialProximityAnalysis call

# Determine target conditions - check if dataframe has target/contrast columns
if ("target" %in% colnames(differential_colocalization_analysis)) {
    # If target column exists, get all unique targets
    target_conditions <- unique(differential_colocalization_analysis$target)
    target_conditions <- target_conditions[!is.na(target_conditions)]
} else {
    stop("Could not find target column. Available columns: ", paste(colnames(differential_colocalization_analysis), collapse = ", "))
}

# Get unique cell types (groups)
unique_cell_types <- unique(differential_colocalization_analysis$group)
unique_cell_types <- unique_cell_types[!is.na(unique_cell_types)]

# Loop through each target condition and cell type combination
for (target_condition in target_conditions) {
    # Create directory name based on target and reference
    comparison_dir <- paste0(target_condition, "_vs_", reference_condition)
    base_output_dir <- here::here("results/05_statistical_testing/colocalization/differential_colocalization_plots", comparison_dir)

    for (cell_type in unique_cell_types) {
        # Create directory for this cell type
        cell_type_dir <- file.path(base_output_dir, cell_type)
        dir.create(cell_type_dir, recursive = TRUE, showWarnings = FALSE)

        # Filter data for this target and cell type
        # Set diff_median to 0 for non-significant results (p_adj >= 0.001) to focus on highly significant changes
        plot_data <- differential_colocalization_analysis %>%
            filter(group == cell_type & target == target_condition) %>%
            mutate(diff_median = case_when(
                p_adj < 0.001 ~ diff_median,
                TRUE ~ 0
            ))

        # Skip if no data for this combination
        if (nrow(plot_data) == 0) {
            warning(paste("No data found for target condition:", target_condition, "and cell type:", cell_type))
            next
        }

        # Create the colocalization heatmap plot
        # Add title indicating the target and cell type
        colocalization_heatmap <- ColocalizationHeatmap(
            plot_data,
            marker1_col = "marker_1",
            marker2_col = "marker_2",
            value_col = "diff_median",
            type = "dots",
            size_col_transform = \(x) {
                -log10(x)
            }
        ) &
            labs(size = "-log10(p_adj)") &
            scale_size(range = c(1, 5), limits = c(0, max(-log10(plot_data$p_adj)))) &
            labs(fill = "Difference\nin medians") &
            labs(
                title = paste("Differential Colocalization Analysis", sep = ""),
                subtitle = paste("Cell type: ", cell_type, " (", target_condition, " vs ", reference_condition, ")", sep = "")
            )

        # Save the plot
        save_plot(
            filename = file.path(cell_type_dir, "colocalization_heatmap"),
            plot = colocalization_heatmap,
            width = 10,
            height = 8,
            create_dir = FALSE # Directory already created
        )
    }
}
```
