---
title: "Single Cell PNA Analysis SOP"
subtitle: "v0.1.0" 
format: html
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
execute:
  dev: ragg_png
  execute-dir: dir
  echo: false
  warning: false
  message: false
---

This SOP describes a standard workflow for analyzing single cell PNA data using the `pixelatorR` package in R. It covers data setup, quality control, data processing, and statistical testing. The workflow is meant to be interactive and modular, allowing users to adapt it to their specific experimental designs and analysis needs.

# Setup

## Load packages and helper functions

A set of commonly used packages and helper functions are loaded here. These are used throughout the analysis and can be customized as needed.

```{r}
#| label: setup_packages

# Load commonly used packages
library(tidyverse)
library(Seurat)
library(pixelatorR)
library(harmony)
library(ComplexHeatmap)
library(ggplotify)
library(here)

# Anchor the project root directory
here::i_am("single_cell_analysis_sop.qmd")

# Load helper functions
source("scripts/helpers.R")

```

## Create the PNA object

### Read metadata file

To handle file paths and sample-specific data, we need to read in a metadata file. This is a csv file with at least the following columns: sample_id, sample_alias, file_path and condition. Sample_id is a short sample ID (e.g. S1, S2, S3), sample_alias is a more descriptive sample alias (e.g. S1_resting, S2_PHA), file_path is the full path to the PNA file and condition describes the sample treatment or condition. Additional columns can be added as needed and will be added to the PNA object as metadata columns. For example, you might want to add a column for the condition, the donor, the time point, the treatment, etc.

Here is an example of what your metadata file should look like:

sample_id,sample_alias,file_path,condition,donor
S1,S1_resting,Users/username/data/S1_resting.pna,resting,D1
S2,S2_PHA,Users/username/data/S2_PHA.pna,stimulated,D2
S3,S3_IL2,Users/username/data/S3_IL2.pna,IL2_treated,D1


```{r}
#| label: setup_read_metadata

# Read metadata file into a dataframe
metadata <- read.csv(here::here("data", "metadata.csv")) |> as_tibble()

# Check that the metadata file contains the required columns - otherwise stop the script
check_metadata_columns(metadata)

# Print the metadata
print(metadata)
```

### Load and merge data

```{r}
#| label: setup_merge_data

# Load the data from the file paths into a list of PNA objects
pg_data <-
    metadata$file_path |>
    lapply(ReadPNA_Seurat, load_proximity_scores = FALSE)

# Merge the PNA objects, adding the sample_id to the cell names
pg_data <-
    merge(
        pg_data[[1]],
        y = pg_data[-1],
        add.cell.ids = metadata$sample_id
    ) |>
    JoinLayers()
```

### Add metadata

```{r}
#| label: setup_add_metadata

# Add metadata to the PNA object
pg_data[[]] <-
    pg_data[[]] |>
    as_tibble(rownames = "cell_id") |>
    mutate(sample_id = str_sub(cell_id, end = -18)) |> # Remove the last 17 characters from the cell_id to get the sample_id
    left_join(metadata |> select(-file_path), by = "sample_id") |> # Do not add the file path to the PNA object
    column_to_rownames("cell_id")

# Sanity check the metadata
for (col in colnames(metadata |> select(-file_path))) {
    print(col)
    print(table(pg_data[[]]$sample_alias, pg_data[[]][, col]))
}
```


## Theme

Here we specify the color palettes for clusters and conditions. These can be adjusted to the color combinations that you want for your project, by replacing either palette with a named vector containing color codes.  

```{r}
#| label: setup_theme

cluster_palette <-
    c(
        pixelatorR::PixelgenAccentColors(
            level = 7,
            hue = c(
                "blues",
                "reds",
                "cyans",
                "oranges",
                "pinks",
                "greens",
                "purples",
                "yellows",
                "greys",
                "beiges"
            )
        ),
        pixelatorR::PixelgenAccentColors(
            level = 9,
            hue = c(
                "blues",
                "pinks",
                "reds",
                "yellows",
                "greys",
                "beiges"
            )
        ),
        pixelatorR::PixelgenAccentColors(
            level = 5,
            hue = c(
                "purples",
                "blues",
                "cyans",
                "greens",
                "reds",
                "oranges"
            )
        ),
        pixelatorR::PixelgenAccentColors(
            level = 3,
            hue = c("blues", "reds")
        ),
        pixelatorR::PixelgenAccentColors(
            level = 12,
            hue = c("blues", "reds")
        )
    ) |>
    unname() |>
    rep(10)

# Create a condition palette for the conditions
unique_conditions <- unique(metadata$condition)
condition_palette <-
    set_names(
        pixelatorES::create_sample_palette(
            unique_conditions
        )[seq_len(length(unique_conditions))],
        unique_conditions
    )

# Create a sample palette for the samples
unique_samples <- unique(metadata$sample_alias)
sample_palette <-
    set_names(
        pixelatorES::create_sample_palette(
            unique_samples
        )[seq_len(length(unique_samples))],
        unique_samples
    )


#' Cell palette
#'
#' @export
#'
cell_palette <-
  c(
    "CD4 T" = "#6D92D1",
    "CD4 Naive" = "#B9CDED",
    "Naive CD4 T" = "#B9CDED",
    "CD4 TSCM" = "#92B0E0",
    "CD4 TCM" = "#6D92D1",
    "CD4 TEM" = "#4A73C0",
    "CD4 TEFF" = "#224792",
    "CD4 TEMRA" = "#1C3A76",
    "Treg" = "#2955AE",
    "CD4 CTL" = "#2955AE",
    "CD8 T" = "#75C5B5",
    "CD8 Naive" = "#D0EDE6",
    "Naive CD8 T" = "#D0EDE6",
    "CD8 TSCM" = "#A2DACE",
    "CD8 TCM" = "#75C5B5",
    "CD8 TEM" = "#4AAF9D",
    "CD8 TEFF" = "#209785",
    "CD8 TEMRA" = "#1B7E6F",
    "other T" = "#1B7E9F",
    "Other T" = "#1B7E9F",
    "MAIT" = "#0D6D5E",
    "DPT" = "#1B7E9F",
    "DNT" = "#7B787F",
    "dnT" = "#7B787F",
    "NK" = "#A28EDB",
    "CD56dim NK" = "#A28EDB",
    "CD56bright NK" = "#866CCD",
    "NK_CD56bright" = "#866CCD",
    "NKT" = "#6C4ABD",
    "B" = "#C7A989", # "#917557",
    "B naive" = "#F0DAC4",
    "Naive B" = "#F0DAC4",
    "B intermediate" = "#F0DAC4",
    "Memory B" = "#C7A989",
    "B memory" = "#C7A989",
    "Plasma cells" = "#DE9982",
    "Plasmablast" = "#DE9982",
    "DC" = "#DA94C1",
    "cDC1" = "#DA94C1",
    "cDC2" = "#BB5391",
    "pDC" = "#9C4579",
    "Mono" = "#F7DFA0",
    "Classical Mono" = "#F7DFA0",
    "CD14 Mono" = "#F7DFA0",
    "Intermediate Mono" = "#E6BE4A",
    "CD16 Mono" = "#BD9315",
    "Non-classical Mono" = "#BD9315",
    "other" = "#797979",
    "Neutrophils" = "#CDCDCD",
    "Basophils" = "#797979",
    "Platelet" = "#DB6365",
    "Platelets" = "#DB6365",
    "gdT" = "#5A3E9E"
  )
```

# Quality Control

In Quality Control we inspect certain metrics that are indicative of the quality of the data per sample and/or per cell. Based on these metrics we can decide whether to remove cells from the dataset before proceeding with data processing.

## Read Depth

First we inspect the read depth per cell. This is a measure of the total number of reads in the cell, and is a good indicator of the quality of the data. The higher the read depth, the better the data quality.

```{r}
#| label: quality_control_read_depth

read_depth_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_id",
    var_y = "reads_in_component",
    palette = sample_palette,
    data_cutoff = 0.999 # Remove a few outlier cells that make the plot hard to read
)

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_read_depth"
    ),
    plot = read_depth_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)

```

## Connectivity

Here we inspect the average k-core per cell. The k-core is a measure of the connectivity of the cell, and is another good indicator of the quality of the data. The higher the average k-core, the better the graph connectivity.

```{r}
#| label: quality_control_average_k_core

# Average k-core per cell
average_k_core_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_id",
    var_y = "average_k_core",
    palette = sample_palette,
    data_cutoff = 1
)

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_average_k_core"
    ),
    plot = average_k_core_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)
```

Next, we plot the percentage of dangling nodes per cell. Dangling nodes belong to subgraphs that are connnected to the rest of the graph by a single edge, and are an indicator of the graph connectivity. The lower the percentage of dangling nodes, the better the graph connectivity.

```{r}
#| label: quality_control_dangling_nodes_plots

# Add percentage of dangling nodes per cell - i.e. K-core 1 fraction
pg_data <-
    AddMetaData(
        pg_data,
        FetchData(
            pg_data,
            c("sample_id", "k_core_1", "n_umi")
        ) |>
            mutate(prc_dangling_nodes = 100 * k_core_1 / n_umi) |>
            select(prc_dangling_nodes),
        col.name = "prc_dangling_nodes"
    )

dangling_nodes_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_id",
    var_y = "prc_dangling_nodes",
    palette = sample_palette,
    data_cutoff = 1
)

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_dangling_nodes"
    ),
    plot = dangling_nodes_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)

```

## Unspecific Background

In PNA experiments, isotype control antibodies are included to measure unspecific binding and serve as important quality controls for background signal. The fraction of total signal that is derived from isotype controls in each cell provides a quantitative estimate of non-specific background. Generally, cells with a high proportion of isotype signal are more likely to be dominated by artefactual binding rather than true biological signal. Therefore, by examining the fraction of isotype control antibodies per cell, we can assess the overall level of background in each sample and identify cells potentially affected by excessive non-specific binding, which may need to be excluded from downstream analysis.


```{r}
#| label: quality_control_isotype_controls

isotype_fraction_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_id",
    var_y = "isotype_fraction",
    palette = sample_palette,
    data_cutoff = 1
)

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_isotype_controls"
    ),
    plot = isotype_fraction_plot,
    width = 10,
    height = 8
)
```

## Tau score

The Tau score measures the heterogeneity of protein composition per cell, and can be used to detect various types of outliers. Cells with high Tau have high counts for few proteins, while cells with low Tau have a more heterogeneous composition of proteins than we would expect from a normal cell. As these outliers likely represent technical artefacts, we can remove them from the dataset.

```{r}
#| label: quality_control_tau_score

tau_score_plot <- TauPlot(pg_data, group_by = "sample_id")

save_plot(
    filename = here::here(
        "results/01_quality_control/tau_plots/quality_control_tau_score"
    ),
    plot = tau_score_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)
```


## Select cells

In the next steps, we will decide which cells to retain for downstream analysis. Our goal is to exclude cells that:

- Are unusually small, which could indicate cell fragments
- Are unusually large, which may represent cell doublets or technical artefacts
- Show high levels of nonspecific background, as indicated by elevated isotype marker signal
- Display abnormal marker expression profiles, as identified by the tau score

We will start by visually inspecting these quality metrics. Cells that do not meet our defined thresholds will be excluded during the Data Processing stage.

### Define thresholds

Revisit these thresholds according to your data. Sometimes it is necessary to use different thresholds for different samples.

```{r}
#| label: quality_control_define_thresholds

# Define threshold for minimum molecule counts per cell
n_umi_min_threshold <- 15000
n_umi_max_threshold <- 200000
isotype_fraction_threshold <- 0.005

# Plot to visualize thresholds as a violin plot
n_umi_plot <- create_violin_plot(
    data = pg_data,
    var_x = "sample_id",
    var_y = "n_umi",
    palette = sample_palette,
    data_cutoff = 0.999
) &
    geom_hline(
        yintercept = n_umi_min_threshold,
        linetype = "dashed",
        color = "red"
    ) &
    geom_hline(
        yintercept = n_umi_max_threshold,
        linetype = "dashed",
        color = "red"
    ) &
    theme_bw()

save_plot(
    filename = here::here(
        "results/01_quality_control/violin_plots/quality_control_n_umi"
    ),
    plot = n_umi_plot,
    width = 10,
    height = 8
)

```

### Molecule Rank plots

```{r}
#| label: quality_control_select_cells

# Combined molecule rank plots
molecule_rank_plot_w_thresholds <- MoleculeRankPlot(pg_data, group_by = "sample_alias") +
    geom_hline(
        yintercept = n_umi_min_threshold,
        linetype = "dashed",
        color = "red"
    ) +
    geom_hline(
        yintercept = n_umi_max_threshold,
        linetype = "dashed",
        color = "red"
    )
save_plot(
    filename = here::here("results/01_quality_control/molecule_rank_plots/quality_control_molecule_rank_plot_w_thresholds"),
    plot = molecule_rank_plot_w_thresholds,
    width = 10,
    height = 8,
    create_dir = TRUE
)

molecule_rank_plot <- MoleculeRankPlot(pg_data, group_by = "sample_alias")
save_plot(
    filename = here::here("results/01_quality_control/molecule_rank_plots/quality_control_molecule_rank_plot"),
    plot = molecule_rank_plot,
    width = 10,
    height = 8,
    create_dir = TRUE
)

# Separate annotated molecule rank plot for each sample
pg_data[[]] <- pg_data[[]] |>
    group_by(sample_id) |>
    mutate(
        rank = rank(-n_umi, ties.method = "random")
    ) |>
    ungroup()

umi_normal <- pg_data[[]] |>
    mutate(
        umi_normal = case_when(
            n_umi > n_umi_min_threshold & n_umi < n_umi_max_threshold ~ TRUE,
            n_umi < n_umi_min_threshold | n_umi > n_umi_max_threshold ~ FALSE
        )
    ) |>
    filter(tau_type == "normal") |>
    select(sample_id, umi_normal)

umi_normal_count <- umi_normal |>
    filter(is.logical(umi_normal)) |>
    group_by(sample_id) |>
    summarise(normal_count = sum(umi_normal), .groups = "drop")

umi_outliers <- pg_data[[]] |>
    select(sample_id, n_umi, tau_type) |>
    filter(tau_type == "normal") |>
    mutate(
        umi_low = n_umi < n_umi_min_threshold,
        umi_high = n_umi > n_umi_max_threshold
    ) |>
    group_by(sample_id) |>
    summarise(
        low_count = sum(umi_low),
        high_count = sum(umi_high),
        .groups = "drop"
    )

hline_normal <- data.frame(
    sample_id = unique(pg_data[[]]$sample_id),
    cutoff = n_umi_min_threshold
)

point_colors <- pg_data[[]] |>
    rownames_to_column("component") |>
    select(sample_id, n_umi, rank, component) |>
    left_join(
        umi_normal_count |> rownames_to_column("component"),
        by = c("component", "sample_id")
    ) |>
    left_join(hline_normal, by = "sample_id") |>
    mutate(
        max_value = n_umi_max_threshold,
        col = case_when(
            isTRUE(umi_normal) ~ "#496389",
            n_umi < cutoff & !isTRUE(umi_normal) ~ "#FFA292",
            n_umi > max_value & !isTRUE(umi_normal) ~ "#FFA292",
            TRUE ~ "#496389"
        )
    )

annotated_molecule_rank_plot <- pg_data[[]] |>
    select(sample_id, rank, n_umi, tau_type) |>
    filter(tau_type == "normal") |>
    ggplot(aes(x = rank, y = n_umi)) +
    geom_point(
        data = point_colors,
        aes(x = rank, y = n_umi),
        color = point_colors$col,
        size = 1,
        alpha = 0.8
    ) +
    geom_hline(
        data = hline_normal,
        aes(yintercept = cutoff),
        linetype = "dashed",
        color = "red"
    ) +
    facet_wrap(
        ~ factor(sample_id, levels = sort(unique(sample_id))),
        nrow = 2
    ) +
    geom_text(
        data = umi_normal_count,
        aes(x = Inf, y = Inf, label = paste("Cells:", normal_count)),
        hjust = 1.2,
        vjust = 1.8,
        size = 4,
        color = "#496389",
        position = position_nudge(y = 0.5)
    ) +
    geom_text(
        data = umi_outliers,
        aes(x = 10, y = Inf, label = paste("High: ", high_count)),
        vjust = 1.8,
        hjust = 0.8,
        size = 4,
        color = "#FFA292",
        position = position_nudge(y = -0.5)
    ) +
    geom_text(
        data = umi_outliers,
        aes(x = 10, y = n_umi_min_threshold, label = paste("Low: ", low_count)),
        size = 4,
        vjust = 0.2,
        hjust = 0.7,
        color = "#FFA292",
        position = position_nudge(y = -0.5)
    ) +
    geom_rug(
        aes(x = rank),
        linewidth = 1,
        sides = "l",
        alpha = 0.05
    ) +
    scale_x_log10() +
    scale_y_log10() +
    coord_equal(ratio = 1.5) +
    theme_minimal(base_size = 18) +
    labs(x = "Component Rank", y = "UMIs", color = "UMI Normal") +
    theme(
        legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_rect(
            color = "black",
            fill = NA,
            linewidth = 0.5
        )
    )
save_plot(
    filename = here::here("results/01_quality_control/molecule_rank_plots/quality_control_annotated_molecule_rank_plot"),
    plot = annotated_molecule_rank_plot,
    width = 15,
    height = 12,
    create_dir = TRUE
)
```

## Select markers (optional)

In this step we select markers that we believe are positive in at least one cell population in the dataset. There is not a single correct way to do this, and the selection criteria will depend on the dataset and experimental design. 

Here we provide one example approach based on counts per million (CPM) thresholds. We calculate the 50th percentile (the median) and 95th percentile of CPM per marker across all cells, and select markers that have a 95th percentile above a certain threshold. The threshold can be adjusted according to the background levels of the dataset, and according to expectations of which markers should be present in the data.

In this example, we set the threshold to 500 CPM at the 95th percentile. This means, in effect, that we let cell populations composing at least 5% of the sample dictate which markers we use in later analysis steps. Comparing the 95th percentile to the 50th percentile also gives us an idea of how much the marker expression varies across cells.


```{r}
#| label: quality_control_select_markers

marker_cpm_p95_threshold <- 500

count_data <-
    LayerData(pg_data, "counts")
cpm_data <-
    count_data |>
    sweep(1, rowSums(count_data), "/")

cpm_data <- cpm_data * 1e6

cpm_quantiles <-
    cpm_data |>
    apply(1, function(x) quantile(x, c(0.5, 0.95))) |>
    t() |>
    as_tibble(rownames = "marker") |>
    mutate(kept = `95%` > marker_cpm_p95_threshold)

scatter_percentile_cpm_quantiles <- cpm_quantiles |>
    ggplot(aes(`50%`, `95%` / `50%`, color = kept)) +
    geom_point(size = 0.1) +
    geom_text(aes(label = marker), vjust = -0.5, size = 2) +
    scale_x_log10(expand = expansion(0.2)) +
    scale_y_log10() +
    theme_bw() +
    labs(
        title = "Marker selection based on CPM",
        x = "50th percentile of CPM (log10 scale)",
        y = "95th / 50th percentile of CPM (log10 scale)"
    )
save_plot(
    filename = here::here("results/01_quality_control/marker_selection_plots/quality_control_marker_selection_cpm"),
    plot = scatter_percentile_cpm_quantiles,
    width = 6,
    height = 4
)


barplot_percentile_cpm_quantiles <- cpm_quantiles |>
    arrange(`95%`) |>
    mutate(marker = factor(marker, levels = marker)) |>
    ggplot(aes(x = marker, y = `95%`, fill = kept)) +
    geom_col() +
    scale_y_log10() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(
        title = "95th percentile of CPM per marker",
        x = "Marker",
        y = "95th percentile of CPM (log10 scale)"
    )
save_plot(
    filename = here::here("results/01_quality_control/marker_selection_plots/quality_control_marker_selection_cpm_barplot"),
    width = 6,
    height = 4,
    plot = barplot_percentile_cpm_quantiles
)

selected_markers <-
    cpm_quantiles |>
    filter(kept) |>
    pull(marker)


```

# Data processing

## Normalize and scale data

We typically normalize PNA data using CLR over margin 2, which means that we normalize the counts per cell by the geometric mean across all markers. This normalization method accounts for differences in sequencing depth between cells, while preserving relative differences in marker abundance within each cell.

Next, we perform Z-score scaling of the normalized data per marker, which centers the data around zero and scales it to unit variance. This step is important for downstream dimensionality reduction and clustering, as it ensures that all markers contribute equally to the analysis, regardless of their absolute abundance levels.

```{r}
#| label: data_processing_normalize_scale

pg_data <-
    pg_data |>
    # Data normalization
    NormalizeData(normalization.method = "CLR", margin = 2) |>
    # Z score scaling
    ScaleData()

```

## Dimensionality reduction

Next, we perform a Principal Component Analysis (PCA) to extract the main axes of variation in the data. We typically use the first 10-20 principal components (PCs) for downstream analysis, but this number can be adjusted based on the dataset and experimental design. A rule of thumb is that the selected PCs should explain a substantial portion of the total variance in the data, while avoiding overfitting or noise. Oftentimes PCs correspond to technical effects, cell states, or experiment effects in the data, and so this number can be adjusted e.g. according to expectations on the number of cell states one expects.

Here we use the marker selection we previously performed to select markers for PCA. This step is optional, but can help to reduce noise and improve the interpretability of the results.

After PCA, we perform Uniform Manifold Approximation and Projection (UMAP) to visualize the data in a low-dimensional space. UMAP is a non-linear dimensionality reduction technique that preserves both local and global structure in the data, making it well-suited for visualizing complex datasets such as single-cell PNA data.

```{r}
#| label: data_processing_dimensionality_reduction

npcs <- 20
n_neighbors <- 30
pg_data <-
    pg_data |>
    RunPCA(npcs = npcs, features = selected_markers) |>
    RunUMAP(reduction = "pca", dims = 1:20, n.neighbors = n_neighbors)

pca_red <-
    Reductions(pg_data, "pca")

elbow_plot_pca <- pca_red@stdev |>
    as_tibble(rownames = "PC") |>
    mutate(PC = PC |> str_remove("PC_") |> as.integer()) |>
    ggplot(aes(PC, value)) +
    geom_line() +
    geom_point() +
    theme_bw() +
    labs(
        title = "PCA Elbow Plot",
        x = "Principal Component",
        y = "Standard Deviation"
    )
save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_pca_elbow_plot"),
    width = 6,
    height = 4,
    plot = elbow_plot_pca
)

cell_metadata <-
    FetchData(pg_data, c("sample_alias", "condition")) |>
    as_tibble(rownames = "cell_id")

pca_embeddings <- Embeddings(pg_data, "pca") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(PC_1, PC_2, color = sample_alias)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = sample_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_pca"),
    width = 6,
    height = 4,
    plot = pca_embeddings
)

umap_embeddings <-
    Embeddings(pg_data, "umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(umap_1, umap_2, color = sample_alias)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = sample_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_umap"),
    width = 6,
    height = 4,
    plot = umap_embeddings
)
```

## Data harmonization (optional)

If your experiment contains multiple samples with similar cell types, but these cell types for one reason or other do not cluster together well, it can be beneficial to harmonize the data. Data harmonization aims to align similar cell types across samples such that they cluster together, which can help to understand which cell types are shared between samples, even if they have somewhat different protein profile. This is particularly useful in experiments with batch effects or technical variability, or biological effects such as donor variability or stimulation perturbations. 

Here we use the Harmony algorithm to harmonize the data based on the sample ID, but any sample metadata can be used. 

We recommend inspecting the harmonized data carefully to ensure that the harmonization has not removed important biological variation, and that similar cell types are indeed clustering together across samples. In many cases it is not necessary to harmonize the data, especially if the samples are already well-mixed and similar cell types cluster together without harmonization.

The harmonization we perform here results in a harmonized dimensional reduction called "harmony", which we use for UMAP visualization and clustering. However, the original data remains unchanged, so we really only use the harmonization to find common cell identities between samples.

```{r}
#| label: data_processing_data_harmonization
pg_data <-
    pg_data |>
    RunHarmony(
        group.by.vars = "sample_alias",
        plot_convergence = FALSE,
    ) |>
    RunUMAP(
        reduction = "harmony",
        dims = 1:20,
        reduction.name = "harmony_umap",
        n.neighbors = n_neighbors
    )

cell_metadata <-
    FetchData(pg_data, c("sample_alias", "condition")) |>
    as_tibble(rownames = "cell_id")

harmony_embeddings <- Embeddings(pg_data, "harmony") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(harmony_1, harmony_2, color = sample_alias)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = sample_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_harmony"),
    width = 6,
    height = 4,
    plot = harmony_embeddings
)

harmony_umap_embeddings <- Embeddings(pg_data, "harmony_umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = sample_alias)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = sample_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/dimensionality_reduction_plots/data_processing_harmonyumap"),
    width = 6,
    height = 4,
    plot = harmony_umap_embeddings
)

```

## Clustering

After dimensionality reduction, we perform clustering. We essentially group cells that are similar in their protein profiles into clusters, which we can later annotate as cell types or cell states. We will perform Leiden clustering on a nearest neighbor graph constructed from the PCA or Harmony reduction. The resolution parameter controls the granularity of the clustering, with higher values resulting in more clusters. The optimal resolution depends on the dataset and experimental design, and can be determined by inspecting the clustering results at different resolutions.

```{r}
#| label: data_processing_clustering

pg_data <-
    pg_data |>
    FindNeighbors(reduction = "harmony", dims = 1:10) |>
    FindClusters(random.seed = 1, resolution = 0.5)

cell_metadata <-
    FetchData(pg_data, c("sample_alias", "condition", "seurat_clusters")) |>
    as_tibble(rownames = "cell_id")

harmony_clusters_embeddings <- Embeddings(pg_data, "harmony") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(harmony_1, harmony_2, color = seurat_clusters)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = cluster_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/clustering_plots/data_processing_harmony_clusters"),
    width = 6,
    height = 4,
    plot = harmony_clusters_embeddings
)

harmony_umap_clusters_embeddings <- Embeddings(pg_data, "harmony_umap") |>
    as_tibble(rownames = "cell_id") |>
    left_join(cell_metadata, by = "cell_id") |>
    ggplot(aes(harmonyumap_1, harmonyumap_2, color = seurat_clusters)) +
    geom_point() +
    coord_fixed() +
    scale_color_manual(values = cluster_palette) +
    theme_bw()

save_plot(
    filename = here::here("results/02_data_processing/clustering_plots/data_processing_harmonyumap_clusters"),
    width = 6,
    height = 4,
    plot = harmony_umap_clusters_embeddings
)

```

## Annotation

### Automatic annotation

Here we perform an automatic cell annotation by matching each cell’s protein profile to annotated 
cells in a reference PBMC dataset. These automated labels should be interpreted with caution, particularly when the 
sample composition differs substantially from the PBMC reference. In such cases, for example in certain stimulated 
samples or cell lines, the annotations may be unreliable. We therefore recommend performing cell annotation as part of 
downstream analysis to ensure accurate and context-appropriate results.

Automated annotation can be a great head start in annotation, but all annotations should at least be inspected and verified, and adjusted when necessary. 

```{r}
#| label: data_processing_annotation

# Load reference dataset
reference <- pixelatorES::read_annotation_reference()

# Set names of annotation columns in reference
reference$l1_annotation <- reference$celltype.l1
reference$l2_annotation <- reference$celltype.l2

# Annotation by label transfer
pg_data <-
  pg_data |>
  AnnotateCells(
    reference = reference,
    reference_groups = c("l1_annotation", "l2_annotation"),
    reference_assay = "PNA",
    query_assay = "PNA",
    # Summarise annotations by cluster, adding an extra column with majority vote
    summarize_by_column = "seurat_clusters",
    normalization_method = "CLR",
    method = "nmf"
  )


cell_metadata <- 
  FetchData(pg_data, c("sample_alias", "condition", "seurat_clusters", 
                       "l1_annotation_summary", "l2_annotation_summary", 
                       "l1_annotation", "l2_annotation")) |> 
  as_tibble(rownames = "cell_id")


Embeddings(pg_data, "harmony_umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = l1_annotation_summary)) +
  geom_point() +
  geom_text(data = . %>% 
              group_by(l1_annotation_summary) %>% 
              summarise(harmonyumap_1 = median(harmonyumap_1),
                        harmonyumap_2 = median(harmonyumap_2)),
            aes(label = l1_annotation_summary),
            size = 3,
            color = "black") +
  coord_fixed() +
  scale_color_manual(values = cell_palette) +
  theme_bw() +
  theme(panel.grid = element_blank())
save_plot(
  filename = "data_processing_harmonyumap_l1_annotation_summary",
  width = 6,
  height = 4
)


Embeddings(pg_data, "harmony_umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = l1_annotation)) +
  geom_point() +
  geom_text(data = . %>% 
              group_by(l1_annotation) %>% 
              summarise(harmonyumap_1 = median(harmonyumap_1),
                        harmonyumap_2 = median(harmonyumap_2)),
            aes(label = l1_annotation),
            size = 3,
            color = "black") +
  coord_fixed() +
  scale_color_manual(values = cell_palette) +
  theme_bw() +
  theme(panel.grid = element_blank())
save_plot(
  filename = "data_processing_harmonyumap_l1_annotation",
  width = 6,
  height = 4
)



Embeddings(pg_data, "harmony_umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = l2_annotation)) +
  geom_point() +
  geom_text(data = . %>% 
              group_by(l2_annotation) %>% 
              summarise(harmonyumap_1 = median(harmonyumap_1),
                        harmonyumap_2 = median(harmonyumap_2)),
            aes(label = l2_annotation),
            size = 3,
            color = "black") +
  coord_fixed() +
  scale_color_manual(values = cell_palette) +
  theme_bw() +
  theme(panel.grid = element_blank())
save_plot(
  filename = "data_processing_harmonyumap_l2_annotation",
  width = 6,
  height = 4
)


FetchData(pg_data, c("sample_alias", "l1_annotation_summary")) |> 
  group_by(sample_alias, l1_annotation_summary) |> 
  count(name = "n_cells") |> 
  group_by(sample_alias) |> 
  mutate(prc_cells = 100 * n_cells / sum(n_cells)) |>
  ggplot(aes(sample_alias, prc_cells, fill = l1_annotation_summary)) +
  geom_col(position = "fill") +
  geom_text(aes(label = paste0(round(prc_cells, 1), "%")),
            position = position_fill(vjust = 0.5),
            size = 3) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = cell_palette) +
  theme_bw() +
  labs(
    title = "Cell type composition per sample",
    x = "Sample",
    y = "Cell type fraction"
  )
save_plot(
  filename = "data_processing_cell_type_composition_per_sample",
  width = 6,
  height = 4
)
  
  
```

### Manual annotation

While automated annotation provides a useful starting point, manual inspection and refinement are essential to ensure accuracy. Manual annotation can be guided by examining patterns of protein expression across clusters, for example by identifying differential abundance between clusters or evaluating the average abundance of specific markers. By carefully considering these expression patterns, you can adjust or confirm automated labels and capture cell populations that may not be well represented in the reference dataset. In this section, we  perform manual annotation based on these principles, complementing the automated approach and improving overall confidence in cell type assignments.

```{r}
#| label: data_processing_manual_annotation

diff_abundant_markers_cluster <- 
  FindAllMarkers(
    pg_data, 
    logfc.threshold = 0,
    min_pct = 0, 
    test.use = "wilcox",
    assay = "PNA",
    slot = "data",
    group.by = "seurat_clusters",
    mean.fxn = rowMeans,
    fc.name = "difference"
  ) |> 
  as_tibble() |> 
  rename(marker = gene)

diff_abundant_markers_cluster |> 
  group_by(marker) |> 
  filter(any(p_val_adj < 0.05 & difference  > 1)) |> 
  select(marker, difference, cluster) |> 
  pivot_wider(names_from = cluster, values_from = difference, values_fill = 0) |>
  column_to_rownames("marker") |>
  t() |> 
  pheatmap(
    color = PixelgenGradient(100, "BluesCherry"),
    breaks = seq(-3, 3, length.out = 101),
    cellwidth = 8, 
    cellheight = 8,
    heatmap_legend_param = list(title = "CLR Difference"),
  ) |> 
  as.ggplot()

save_plot(
  filename = "data_processing_differential_abundance_markers_per_cluster",
  width = 10,
  height = 4
)


diff_abundant_markers_cluster |> 
  ggplot(aes(difference, -log10(p_val))) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(size = 0.1) +
  geom_text(data = . %>%
              group_by(cluster) %>%
              top_n(5, abs(difference)),
            aes(label = marker),
            vjust = -0.5,
            size = 2) +
  facet_wrap(~ cluster) +
  scale_fill_manual(values = cluster_palette) +
  theme_bw() +
  labs(
    title = "Differential abundance markers per cluster",
    x = "Average CLR Difference",
    y = "-log10(p value)"
  )

save_plot(
  filename = "data_processing_differential_abundance_markers_per_cluster_volcano",
  width = 8,
  height = 6
)



```

#### Marker abundance visualization

Here we visualize the abundance of selected markers across the harmonized UMAP and per cluster. This can help to identify which markers are characteristic of each cluster, and can guide manual annotation of cell types. Please change `markers_of_interest` according to which markers are most relevant for your dataset. 

```{r}
markers_of_interest <- c("CD3e", "CD4", "CD8", "CD19", "CD14", "CD16", "CD56")
plot_data <- 
  FetchData(
    pg_data,
    c("harmonyumap_1", "harmonyumap_2",
      "sample_alias", "condition",
      "seurat_clusters",
      markers_of_interest),
    slot = "data"
  ) |> 
  pivot_longer(
    cols = all_of(markers_of_interest),
    names_to = "marker",
    values_to = "CLR"
  )

plot_data |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = CLR)) +
  geom_point(size = 0.1) +
  facet_wrap(~marker) +
  scale_color_gradient(low = "lightgrey", high = "darkblue") +
  coord_fixed() +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(
    title = "Marker expression on Harmony UMAP",
    x = "Harmony UMAP 1",
    y = "Harmony UMAP 2",
    color = "CLR"
  )
save_plot(
  filename = "data_processing_marker_abundance_harmonyumap",
  width = 6,
  height = 4
)

plot_data |> 
  ggplot(aes(sample_alias, CLR, fill = condition)) +
  geom_violin(quantiles = 0.5, quantile.linetype = 1) +
  facet_grid(marker ~ seurat_clusters) +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.border = element_rect(color = "gray"),
    panel.spacing = unit(0, "mm")
  ) +
  labs(
    title = "Marker abundance per cluster",
    x = "Sample",
    y = "CLR"
  )
save_plot(
  filename = "data_processing_marker_abundance_per_cluster",
  width = 8,
  height = 8
)


plot_data |> 
  group_by(marker, seurat_clusters) |> 
  summarise(mean_CLR = mean(CLR)) |> 
  pivot_wider(names_from = seurat_clusters, values_from = mean_CLR) |>
  column_to_rownames("marker") |>
  as.matrix() |> 
  pheatmap(
    color = PixelgenGradient(100, "Cherry"),
    cellwidth = 8, 
    cellheight = 8,
    heatmap_legend_param = list(title = "Mean CLR"),
  ) |> 
  as.ggplot()
save_plot(
  filename = "data_processing_mean_marker_abundance_per_cluster_heatmap",
  width = 6,
  height = 4
)

```

#### Set manual annotation 

Here we set the actual annotations based on the previous inspection of marker abundance. Each cluster gets its own annotation, and oftentimes several clusters can have the same annotation. 

Please change the annotations according to your data.

```{r}
#| label: data_processing_set_manual_annotation

# Specify cell annotation labels for each cluster
cluster_cell_annotation <- 
  c("0" = "CD4 T",
    "1" = "CD4 T", 
    "2" = "Mono", 
    "3" = "B", 
    "4" = "CD8 T", 
    "5" = "CD4 T",
    "6" = "NK",
    "7" = "CD8 T",
    "8" = "CD4 T",
    "9" = "B",
    "10" = "CD4 T", 
    "11" = "Platelets", 
    "12" = "Mono")

# Add annotation to Seurat object
pg_data[[]] <- 
  pg_data[[]] |>
  mutate(
    cell_annotation = cluster_cell_annotation[as.character(seurat_clusters)]
  )


  
```

### Comparison 

```{r}


FetchData(pg_data, c("seurat_clusters", "l1_annotation", "cell_annotation")) |>
  group_by(seurat_clusters, l1_annotation, cell_annotation) |>
  count() |> 
  group_by(seurat_clusters) |>
  mutate(frac = n / sum(n)) |> 
  ggplot(aes(l1_annotation, cell_annotation, fill = frac)) +
  geom_tile() +
  facet_wrap(~seurat_clusters, scales = "free") +
  scale_fill_gradient(low = "white", high = "darkblue") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  labs(
    title = "Comparison of automatic and manual annotations",
    x = "Automatic annotation (L1)",
    y = "Manual annotation",
    fill = "Fraction of cells"
  )
save_plot(
  filename = "data_processing_comparison_automatic_manual_annotation",
  width = 8,
  height = 6
)

```


# Statistical testing

## Marker selection

Select markers to evaluate per cell type, based on:

Abundance level and variance (e.g. mean vs standard deviation)

Deviation from background levels (isotypes or ambient negative levels)

Colocalization to isotypes

### Colocalization with isotypes

All filters can be used together or separately, but it's recommended to only use the background_threshold_pct filter. This filter is applied to the fraction of UMI counts which does not depend on the cell size (total number of molecules).

The filters are applied in the following order:

Remove protein pairs with a minimum UMI count fraction below background_threshold_pct

Remove protein pairs with a minimum UMI count below background_threshold_count

Remove protein pairs detected in fewer than min_cells_count cells

```{r}
#| label: statistical_testing_marker_selection

cell_annotation_column <- 
  "l1_annotation_summary"
isotype_markers <- 
  c("mIgG1", "mIgG2a", "mIgG2b")

abundance_data <- 
  LayerData(pg_data, "data") |> 
  as_tibble(rownames = "marker") |> 
  pivot_longer(-marker, 
               names_to = "component", 
               values_to = "CLR") |> 
  left_join(
    FetchData(pg_data, c("sample_alias", cell_annotation_column)) |> 
      as_tibble(rownames = "component"),
    by = "component"
  ) 

# Recalculate colocalization scores between each marker and any control
marker_isotype_coloc <- 
  ProximityScores(pg_data, lazy = TRUE, 
                  add_marker_proportions = TRUE,
                  meta_data_columns = c("sample_alias", cell_annotation_column)) |> 
  filter(marker_1 %in% isotype_markers | marker_2 %in% isotype_markers) |> 
  mutate(marker = ifelse(marker_1 %in% isotype_markers, marker_2, marker_1)) |> 
  group_by(sample_alias, !!sym(cell_annotation_column), component, marker) |>
  summarise(join_count = sum(join_count), 
            join_count_expected_mean = sum(join_count_expected_mean)) |> 
  mutate(log2_ratio = log2((join_count + 1) / (join_count_expected_mean + 1))) |> 
  collect()


marker_isotype_coloc |> 
  left_join(abundance_data) |> 
  ggplot(aes(CLR, log2_ratio, color = log2_ratio > log2(1.5))) +
  geom_point(size = 0.1) +
  geom_hline(yintercept = 0) +
  facet_wrap(~sample_alias) +
  theme_bw() +
  scale_color_manual(values = c("black", "orangered")) +
  labs(
    title = "Marker colocalization with isotypes",
    x = "CLR abundance",
    y = "Log2 ratio of observed vs expected colocalization"
  )
save_plot(
  filename = "statistical_testing_marker_selection_marker_isotype_coloc",
  width = 6,
  height = 4
)

marker_isotype_coloc |> 
  group_by(sample_alias, marker) |> 
  summarise(join_count = sum(join_count), 
            join_count_expected_mean = sum(join_count_expected_mean)) |> 
  mutate(log2_ratio = log2((join_count + 1) / (join_count_expected_mean + 1))) |> 
  ungroup() |>
  left_join(abundance_data |> 
              group_by(marker, sample_alias) |> 
              summarise(CLR = mean(CLR)),
            by = c("marker", "sample_alias")) |>
  ggplot(aes(CLR, log2_ratio, color = log2_ratio > log2(1.5))) +
  geom_hline(yintercept = 0) +
  geom_point(size = 0.1) +
  geom_text(aes(label = marker),
            vjust = -0.5,
            size = 2) +
  facet_wrap(~sample_alias) +
  scale_color_manual(values = c("black", "orangered")) +
  theme_bw() +
  labs(
    title = "Marker colocalization with isotypes per sample",
    x = "Mean CLR abundance",
    y = "Sample-wide log2 ratio of observed vs expected colocalization"
  ) 
save_plot(
  filename = "statistical_testing_marker_selection_marker_isotype_coloc_per_sample",
  width = 8,
  height = 6
)



plot_data <- 
  marker_isotype_coloc |> 
  group_by(sample_alias, !!sym(cell_annotation_column), marker) |> 
  summarise(join_count = sum(join_count), 
            join_count_expected_mean = sum(join_count_expected_mean)) |> 
  mutate(log2_ratio = log2((join_count + 1) / (join_count_expected_mean + 1))) |> 
  ungroup() |>
  left_join(abundance_data |> 
              group_by(marker, sample_alias, !!sym(cell_annotation_column)) |> 
              summarise(CLR = mean(CLR)),
            by = c("marker", "sample_alias", cell_annotation_column))

plot_data |>
  ggplot(aes(CLR, log2_ratio, color = log2_ratio > log2(1.5))) +
  geom_hline(yintercept = 0) +
  geom_point(size = 0.1) +
  
  geom_text(aes(label = marker),
            vjust = -0.5,
            size = 2) +
  facet_grid(rows = vars(!!sym(cell_annotation_column)),
             cols = vars(sample_alias)) +
  scale_color_manual(values = c("black", "orangered")) +
  theme_bw() +
  labs(
    title = "Marker colocalization with isotypes per cluster",
    x = "Mean CLR abundance",
    y = "Cluster-wide log2 ratio of observed vs expected colocalization"
  ) 
save_plot(
  filename = "statistical_testing_marker_selection_marker_isotype_coloc_per_sample_cluster",
  width = 8,
  height = 6
)

plot_data |> 
  group_by(marker) |> 
  filter(any(log2_ratio > log2(1.5))) |>
  mutate(total_log2_ratio = sum(log2_ratio)) |>
  arrange(desc(total_log2_ratio)) |>
  ungroup() |> 
  mutate(marker = factor(marker, 
                         levels = unique(marker))) |>
  ggplot(aes(marker, seurat_clusters, fill = log2_ratio)) +
  geom_tile() +
  scale_fill_gradientn(colors = PixelgenGradient(100, "BluesCherry")) +
  coord_fixed() +
  facet_grid(sample_alias ~ .) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Marker colocalization with isotypes per cluster",
    x = "Marker",
    y = "Cluster"
  )
save_plot(
  filename = "statistical_testing_marker_selection_marker_isotype_coloc_heatmap",
  width = 12,
  height = 6
)

```

## Differential abundance analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_abundance

print(1)

```

## Differential proximity analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_proximity

print(1)

```

