---
title: "Single Cell PNA Analysis SOP"
subtitle: "v0.1.0" 
format: html
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
execute:
  dev: ragg_png
  execute-dir: dir
  echo: false
  warning: false
  message: false
---

This SOP describes a standard workflow for analyzing single cell PNA data using the `pixelatorR` package in R. It covers data setup, quality control, data processing, and statistical testing. The workflow is meant to be interactive and modular, allowing users to adapt it to their specific experimental designs and analysis needs.

# Setup

## Packages

```{r}
#| label: setup_packages

library(tidyverse)
library(Seurat)
library(pixelatorR)
library(harmony)
library(ComplexHeatmap)
library(ggplotify)
library(here)

```

## Functions

```{r}
#| label: setup_functions

save_plot <-
  function(
    filename,
    plot = last_plot(),
    width = 10,
    height = 10,
    create_dir = TRUE,
    file_formats = c("png", "pdf")
  ) {
    for (format in file_formats) {
      ggsave(
        filename = file.path("results", paste0(filename, ".", format)),
        plot = plot,
        width = width,
        height = height,
        create.dir = create_dir
      )
    }
  }

```

## Create the PNA object

### Read metadata file

```{r}
#| label: setup_read_metadata

# This is a csv file with at least the following columns: sample_id, sample_alias and file_path.
# sample_id is a short sample ID (e.g. S1, S2, S3), sample_alias is a more descriptive sample alias (e.g. S1_resting, S2_PHA), and file_path is the path to the PNA file.
# Additional columns can be added as needed and will be added to the PNA object as metadata columns.
metadata <- read.csv(here::here("data", "metadata.csv")) |> as_tibble()

# Print the metadata
print(metadata)
```

### Load and merge data

```{r}
#| label: setup_merge_data

# Load the data from the file paths into a list of PNA objects
pg_data <-
  metadata$file_path |>
  lapply(ReadPNA_Seurat, load_proximity_scores = FALSE)

# Merge the PNA objects, adding the sample_id to the cell names
pg_data <-
  merge(
    pg_data[[1]],
    y = pg_data[-1],
    add.cell.ids = metadata$sample_id
  ) |>
  JoinLayers()
```

### Add metadata

```{r}
#| label: setup_add_metadata

# Add metadata to the PNA object
pg_data[[]] <-
  pg_data[[]] |>
  as_tibble(rownames = "cell_id") |>
  mutate(sample_id = str_sub(cell_id, end = -18)) |> # Remove the last 17 characters
  left_join(metadata |> select(-file_path), by = "sample_id") |> # Do not add the file path to the PNA object
  column_to_rownames("cell_id")

# Sanity check the metadata
for (col in colnames(metadata |> select(-file_path))) {
  print(col)
  print(table(pg_data[[]]$sample_alias, pg_data[[]][, col]))
}
```


## Theme

Here we specify the color palettes for clusters and conditions. These can be adjusted to the color combinations that you want for your project, by replacing either palette with a named vector containing color codes.  

```{r}
#| label: setup_theme

cluster_palette <-
  c(
    pixelatorR::PixelgenAccentColors(
      level = 7,
      hue = c(
        "blues",
        "reds",
        "cyans",
        "oranges",
        "pinks",
        "greens",
        "purples",
        "yellows",
        "greys",
        "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 9,
      hue = c(
        "blues",
        "pinks",
        "reds",
        "yellows",
        "greys",
        "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 5,
      hue = c(
        "purples",
        "blues",
        "cyans",
        "greens",
        "reds",
        "oranges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 3,
      hue = c("blues", "reds")
    ),
    pixelatorR::PixelgenAccentColors(
      level = 12,
      hue = c("blues", "reds")
    )
  ) |>
  unname() |>
  rep(10)


condition_palette <-
  set_names(
    pixelatorES::create_sample_palette(
      unique(conditions)
    )[seq_len(n_distinct(conditions))],
    unique(conditions)
  )


sample_palette <-
  set_names(
    pixelatorES::create_sample_palette(
      names(data_files)
    )[seq_len(length(data_files))],
    names(data_files)
  )

```

# Quality Control

In Quality Control we inspect certain metrics that are indicative of the quality of the data per sample and/or per cell. Based on these metrics we can decide whether to remove cells from the dataset before proceeding with data processing.

## Read depth (n_reads)

```{r}
#| label: quality_control_read_depth

FetchData(pg_data, c("sample_alias", "condition", "reads_in_component")) |>
  ggplot(aes(sample_alias, reads_in_component, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(
    data = function(x) {
      x |>
        group_by(sample_alias) |>
        summarise(median = median(reads_in_component))
    },
    aes(sample_alias, median, label = round(median)),
    vjust = -0.5,
    inherit.aes = FALSE
  ) +
  scale_y_log10() +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Read depth per cell",
    x = "Sample",
    y = "Reads per cell (log10 scale)"
  )
save_plot(
  filename = "quality_control_read_depth",
  width = 6,
  height = 4
)

```

## Sequencing saturation

Here we run simulations of 6 cells to estimate the sequencing saturation curves per sample.

```{r}
#| label: quality_control_sequencing_saturation

object_meta <-
  FetchData(pg_data, c("sample_alias", "reads_in_component")) |>
  group_by(sample_alias) |>
  summarise(
    n_comps = n(),
    reads_per_component = mean(reads_in_component)
  )

seqsat_curve_data <-
  data_files |>
  names() |>
  set_names() |>
  parallel::mclapply(
    function(sampl) {
      n_comps <-
        object_meta$n_comps[which(object_meta$sample_alias == sampl)]
      n_comps <- pmin(n_comps, 6)
      sampled_comps <- sample(
        pg_data[[]] |>
          filter(sample_alias == sampl) |>
          rownames() |>
          sample(n_comps)
      )
      el <- pg_data |>
        subset(cells = sampled_comps) |>
        Edgelists(lazy = FALSE)

      el |>
        SequenceSaturationCurve(n_comps = n_comps)
    },
    mc.cores = 1
  ) |>
  bind_rows(.id = "sample_alias") |>
  left_join(
    object_meta,
    by = join_by(sample_alias)
  ) |>
  mutate(
    reads_per_component = reads_per_component * sample_frac
  ) |>
  select(
    sample_alias,
    reads_per_component,
    graph_node_saturation,
    graph_edge_saturation
  ) |>
  pivot_longer(
    cols = c("graph_node_saturation", "graph_edge_saturation"),
    names_to = "type",
    values_to = "saturation"
  ) |>
  mutate(
    type = str_remove(type, "graph_") |>
      str_remove("_saturation") |>
      str_to_sentence()
  )

seqsat_curve_data |>
  ggplot(aes(reads_per_component, saturation, color = sample_alias)) +
  geom_line(
    data = seqsat_curve_data |>
      group_by(sample_alias, type, reads_per_component) |>
      summarise(saturation = mean(saturation)) |>
      ungroup(),
    aes(group = sample_alias)
  ) +
  geom_point(size = 0.1) +
  facet_wrap(~type) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_manual(values = sample_palette) +
  theme_bw() +
  labs(
    x = "Reads per component",
    y = "Sequencing saturation [%]",
    title = "Sequencing saturation curve",
    color = "Sample ID"
  )
save_plot(
  filename = "quality_control_sequencing_saturation",
  width = 6,
  height = 4
)
```

## Connectivity (average kcore and dangling nodes)

TODO: Add text here

```{r}
#| label: quality_control_connectivity

plot_data <-
  FetchData(
    pg_data,
    c("sample_alias", "condition", "average_k_core", "k_core_1", "n_umi")
  ) |>
  mutate(prc_dangling_nodes = 100 * k_core_1 / n_umi)

plot_data |>
  ggplot(aes(sample_alias, average_k_core, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(
    data = function(x) {
      x |>
        group_by(sample_alias) |>
        summarise(median = median(average_k_core))
    },
    aes(sample_alias, median, label = round(median, 2)),
    vjust = -0.5,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Average k-core per cell",
    x = "Sample",
    y = "Average k-core"
  )

save_plot(
  filename = "quality_control_average_k_core",
  width = 6,
  height = 4
)

plot_data |>
  ggplot(aes(sample_alias, prc_dangling_nodes, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(
    data = function(x) {
      x |>
        group_by(sample_alias) |>
        summarise(median = median(prc_dangling_nodes))
    },
    aes(sample_alias, median, label = paste(round(median, 1), "%")),
    vjust = -0.5,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Dangling nodes per cell",
    x = "Sample",
    y = "Dangling nodes [%]"
  )

save_plot(
  filename = "quality_control_dangling_nodes",
  width = 6,
  height = 4
)

```



## Select cells

Next, we will select cells to keep for downstream analysis. We want to remove cells that:
- Are too small, possibly representing cell fragments
- Are noisy, using isotype marker background level
- Have an abnormal marker distribution, using the tau score

We will begin with inspecting the data, and later remove cells in Data Processing that do not meet our quality criteria.

### Define thresholds

Revisit these thresholds according to your data. Sometimes it is necessary to use different thresholds for different samples.

```{r}
#| label: quality_control_define_thresholds

# Define threshold for minimum molecule counts per cell

n_umi_threshold <- 10000
isotype_fraction_threshold <- 0.005

```

### Cell size / Antibody molecule counts (n_umi)

```{r}
#| label: quality_control_select_cells

MoleculeRankPlot(pg_data, group_by = "sample_alias") +
  geom_hline(
    yintercept = n_umi_threshold,
    linetype = "dashed",
    color = "red"
  )
save_plot(
  filename = "quality_control_molecule_rank_plot",
  width = 6,
  height = 4
)


FetchData(pg_data, c("sample_alias", "condition", "n_umi")) |>
  ggplot(aes(sample_alias, n_umi, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(
    data = function(x) {
      x |>
        group_by(sample_alias) |>
        summarise(median = median(n_umi))
    },
    aes(sample_alias, median, label = round(median)),
    vjust = -0.5,
    inherit.aes = FALSE
  ) +
  geom_hline(
    yintercept = n_umi_threshold,
    linetype = "dashed",
    color = "red"
  ) +
  scale_y_log10() +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Antibody molecule counts per cell",
    x = "Sample",
    y = "Molecule counts per cell (log10 scale)"
  )
save_plot(
  filename = "quality_control_antibody_molecule_counts",
  width = 6,
  height = 4
)

```

### Isotype controls

TODO: Add text here

```{r}
#| label: quality_control_isotype_controls

FetchData(pg_data, c("sample_alias", "condition", "isotype_fraction")) |>
  ggplot(aes(sample_alias, isotype_fraction * 100, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(
    data = function(x) {
      x |>
        group_by(sample_alias) |>
        summarise(median = median(isotype_fraction * 100))
    },
    aes(sample_alias, median, label = paste(round(median, 2), "%")),
    vjust = -0.5,
    inherit.aes = FALSE
  ) +
  geom_hline(
    yintercept = isotype_fraction_threshold * 100,
    linetype = "dashed",
    color = "red"
  ) +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Isotype control fraction per cell",
    x = "Sample",
    y = "Isotype control fraction per cell [%]"
  )

save_plot(
  filename = "quality_control_isotype_controls",
  width = 6,
  height = 4
)
```

### Tau score

The Tau score measures the heterogeneity of protein composition per cell, and can be used to detect various types of outliers. Cells with high Tau high counts for few proteins, while cells with low Tau have a more heterogeneous composition of proteins than we would expect from a normal cell. As these outliers likely represent technical artefacts, we can remove them from the dataset.

```{r}
#| label: quality_control_tau_score

TauPlot(pg_data)
save_plot(
  filename = "quality_control_tau_score",
  width = 6,
  height = 4
)

```

## Select markers (optional)

In this step we select markers that we believe are positive in at least one cell population in the dataset. There is not a single correct way to do this, and the selection criteria will depend on the dataset and experimental design. 

Here we provide one example approach based on counts per million (CPM) thresholds. We calculate the 50th percentile (the median) and 95th percentile of CPM per marker across all cells, and select markers that have a 95th percentile above a certain threshold. The threshold can be adjusted according to the background levels of the dataset, and according to expectations of which markers should be present in the data.

In this example, we set the threshold to 1000 CPM at the 95th percentile. This means, in effect, that we let cell populations composing at least 5% of the sample dictate which markers we use in later analysis steps. Comparing the 95th percentile to the 50th percentile also gives us an idea of how much the marker expression varies across cells.


```{r}
#| label: quality_control_select_markers

marker_cpm_p95_threshold <- 1000

count_data <-
  LayerData(pg_data, "counts")
cpm_data <-
  count_data |>
  sweep(1, colSums(count_data), "/")

cpm_data <- cpm_data * 1e6

cpm_quantiles <-
  cpm_data |>
  apply(1, function(x) quantile(x, c(0.5, 0.95))) |>
  t() |>
  as_tibble(rownames = "marker") |>
  mutate(kept = `95%` > marker_cpm_p95_threshold)

cpm_quantiles |>
  ggplot(aes(`50%`, `95%` / `50%`, color = kept)) +
  geom_point(size = 0.1) +
  geom_text(aes(label = marker), vjust = -0.5, size = 2) +
  scale_x_log10(expand = expansion(0.2)) +
  scale_y_log10() +
  theme_bw() +
  labs(
    title = "Marker selection based on CPM",
    x = "50th percentile of CPM (log10 scale)",
    y = "95th / 50th percentile of CPM (log10 scale)"
  )
save_plot(
  filename = "quality_control_marker_selection_cpm",
  width = 6,
  height = 4
)


cpm_quantiles |>
  arrange(`95%`) |>
  mutate(marker = factor(marker, levels = marker)) |>
  ggplot(aes(x = marker, y = `95%`, fill = kept)) +
  geom_col() +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "95th percentile of CPM per marker",
    x = "Marker",
    y = "95th percentile of CPM (log10 scale)"
  )
save_plot(
  filename = "quality_control_marker_selection_cpm_barplot",
  width = 6,
  height = 4
)

selected_markers <-
  cpm_quantiles |>
  filter(kept) |>
  pull(marker)


```

# Data processing

## Normalize and scale data

We typically normalize PNA data using CLR over margin 2, which means that we normalize the counts per cell by the geometric mean across all markers. This normalization method accounts for differences in sequencing depth between cells, while preserving relative differences in marker abundance within each cell.

Next, we perform Z-score scaling of the normalized data per marker, which centers the data around zero and scales it to unit variance. This step is important for downstream dimensionality reduction and clustering, as it ensures that all markers contribute equally to the analysis, regardless of their absolute abundance levels.

```{r}
#| label: data_processing_normalize_scale

pg_data <-
  pg_data |>
  # Data normalization
  NormalizeData(normalization.method = "CLR", margin = 2) |>
  # Z score scaling
  ScaleData()

```

## Dimensionality reduction

Next, we perform a Principal Component Analysis (PCA) to extract the main axes of variation in the data. We typically use the first 10-20 principal components (PCs) for downstream analysis, but this number can be adjusted based on the dataset and experimental design. A rule of thumb is that the selected PCs should explain a substantial portion of the total variance in the data, while avoiding overfitting or noise. Oftentimes PCs correspond to technical effects, cell states, or experiment effects in the data, and so this number can be adjusted e.g. according to expectations on the number of cell states one expects.

Here we use the marker selection we previously performed to select markers for PCA. This step is optional, but can help to reduce noise and improve the interpretability of the results.

After PCA, we perform Uniform Manifold Approximation and Projection (UMAP) to visualize the data in a low-dimensional space. UMAP is a non-linear dimensionality reduction technique that preserves both local and global structure in the data, making it well-suited for visualizing complex datasets such as single-cell PNA data.

```{r}
#| label: data_processing_dimensionality_reduction

npcs <- 20
n_neighbors <- 30
pg_data <-
  pg_data |>
  RunPCA(npcs = npcs, features = selected_markers) |>
  RunUMAP(reduction = "pca", dims = 1:20, n.neighbors = n_neighbors)

pca_red <-
  Reductions(pg_data, "pca")

pca_red@stdev |>
  as_tibble(rownames = "PC") |>
  mutate(PC = PC |> str_remove("PC_") |> as.integer()) |>
  ggplot(aes(PC, value)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(
    title = "PCA Elbow Plot",
    x = "Principal Component",
    y = "Standard Deviation"
  )

ggsave(
  filename = "results/data_processing_pca_elbow_plot.png",
  width = 6,
  height = 4
)

cell_metadata <-
  FetchData(pg_data, c("sample_alias", "condition")) |>
  as_tibble(rownames = "cell_id")

Embeddings(pg_data, "pca") |>
  as_tibble(rownames = "cell_id") |>
  left_join(cell_metadata, by = "cell_id") |>
  ggplot(aes(PC_1, PC_2, color = sample_alias)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = sample_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_pca",
  width = 6,
  height = 4
)

Embeddings(pg_data, "umap") |>
  as_tibble(rownames = "cell_id") |>
  left_join(cell_metadata, by = "cell_id") |>
  ggplot(aes(umap_1, umap_2, color = sample_alias)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = sample_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_umap",
  width = 6,
  height = 4
)
```

## Data harmonization (optional)

If your experiment contains multiple samples with similar cell types, but these cell types for one reason or other do not cluster together well, it can be beneficial to harmonize the data. Data harmonization aims to align similar cell types across samples such that they cluster together, which can help to understand which cell types are shared between samples, even if they have somewhat different protein profile. This is particularly useful in experiments with batch effects or technical variability, or biological effects such as donor variability or stimulation perturbations. 

Here we use the Harmony algorithm to harmonize the data based on the sample ID, but any sample metadata can be used. 

We recommend inspecting the harmonized data carefully to ensure that the harmonization has not removed important biological variation, and that similar cell types are indeed clustering together across samples. In many cases it is not necessary to harmonize the data, especially if the samples are already well-mixed and similar cell types cluster together without harmonization.

The harmonization we perform here results in a harmonized dimensional reduction called "harmony", which we use for UMAP visualization and clustering. However, the original data remains unchanged, so we really only use the harmonization to find common cell identities between samples.

```{r}
#| label: data_processing_data_harmonization
pg_data <-
  pg_data |>
  RunHarmony(
    group.by.vars = "sample_alias",
    plot_convergence = FALSE,
  ) |>
  RunUMAP(
    reduction = "harmony",
    dims = 1:20,
    reduction.name = "harmony_umap",
    n.neighbors = n_neighbors
  )

cell_metadata <-
  FetchData(pg_data, c("sample_alias", "condition")) |>
  as_tibble(rownames = "cell_id")

Embeddings(pg_data, "harmony") |>
  as_tibble(rownames = "cell_id") |>
  left_join(cell_metadata, by = "cell_id") |>
  ggplot(aes(harmony_1, harmony_2, color = sample_alias)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = sample_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_harmony",
  width = 6,
  height = 4
)

Embeddings(pg_data, "harmony_umap") |>
  as_tibble(rownames = "cell_id") |>
  left_join(cell_metadata, by = "cell_id") |>
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = sample_alias)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = sample_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_harmonyumap",
  width = 6,
  height = 4
)

```

## Clustering

After dimensionality reduction, we perform clustering. We essentially group cells that are similar in their protein profiles into clusters, which we can later annotate as cell types or cell states. We will perform Leiden clustering on a nearest neighbor graph constructed from the PCA or Harmony reduction. The resolution parameter controls the granularity of the clustering, with higher values resulting in more clusters. The optimal resolution depends on the dataset and experimental design, and can be determined by inspecting the clustering results at different resolutions.

```{r}
#| label: data_processing_clustering

pg_data <-
  pg_data |>
  FindNeighbors(reduction = "harmony", dims = 1:10) |>
  FindClusters(random.seed = 1, resolution = 0.5)

cell_metadata <-
  FetchData(pg_data, c("sample_alias", "condition", "seurat_clusters")) |>
  as_tibble(rownames = "cell_id")

Embeddings(pg_data, "harmony") |>
  as_tibble(rownames = "cell_id") |>
  left_join(cell_metadata, by = "cell_id") |>
  ggplot(aes(harmony_1, harmony_2, color = seurat_clusters)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = cluster_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_harmony_clusters",
  width = 6,
  height = 4
)

Embeddings(pg_data, "harmony_umap") |>
  as_tibble(rownames = "cell_id") |>
  left_join(cell_metadata, by = "cell_id") |>
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = seurat_clusters)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = cluster_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_harmonyumap_clusters",
  width = 6,
  height = 4
)

```

## Annotation

### Automatic annotation

TODO: Fill this text out 
Annotate cells by:

Manual annotation from differential abundance between clusters or average abundance across clusters.

Automatic annotation using label transfer from a reference dataset.

```{r}
#| label: data_processing_annotation

# Automatic annotation
# reference <- pixelatorES::read_annotation_reference()
# reference$l1_annotation <- reference$celltype.l1
# reference$l2_annotation <- reference$celltype.l2
# object <-
# object |>
# AnnotateCells(
# reference = reference,
# reference_groups = c("l1_annotation", "l2_annotation"),
# reference_assay = "PNA",
# query_assay = "PNA",
# summarize_by_column = "seurat_clusters",
# normalization_method = ifelse(params$annotation_method == "Seurat", "LogNormalize", "CLR"),
# method = params$annotation_method
# )

```

### Manual annotation

```{r}
#| label: data_processing_manual_annotation

diff_abundant_markers_cluster <-
  FindAllMarkers(
    pg_data,
    logfc.threshold = 0,
    min_pct = 0,
    return.thresh = 1
  ) |>
  as_tibble() |>
  rename(marker = gene)

diff_abundant_markers_cluster |>
  group_by(marker) |>
  filter(any(p_val_adj < 0.05 & avg_log2FC > 1)) |>
  select(marker, avg_log2FC, cluster) |>
  pivot_wider(
    names_from = cluster,
    values_from = avg_log2FC,
    values_fill = 0
  ) |>
  column_to_rownames("marker") |>
  pheatmap(
    cellwidth = 8,
    cellheight = 8,
    heatmap_legend_param = list(title = "Avg. log2 FC"),
  ) |>
  as.ggplot()

```

# Statistical testing

## Marker selection

Select markers to evaluate per cell type, based on:

Abundance level and variance (e.g. mean vs standard deviation)

Deviation from background levels (isotypes or ambient negative levels)

Colocalization to isotypes

Colocalize with isotypes

```{r}
#| label: statistical_testing_marker_selection

```

## Differential abundance analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_abundance

```

## Differential proximity analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_proximity

```

