---
title: "Single Cell PNA Analysis SOP"
format: html
---

# Setup

## Packages

```{r}
# | 
```


## Read data

```{r}

```

# Merge data

# Quality Control

Investigate levels and/or distribution of:

## Read depth (n_reads)

## Sequencing saturation

## Antibody molecule counts (n_umi)

## Isotype controls

## Connectivity (average kcore and dangling nodes)

## Select cells

Remove cells that:

Are small, using the molecule rank plot

Are noisy, using isotype background level 

Have an abnormal marker distribution, using the tau score

## Select markers (optional)

Select markers for dimensionality and clustering (optional) 

Remove markers that: 

Are not positive in any cell population

Colocalize with isotypes

# Data processing

## Normalize and scale data

Normalize using CLR, margin 2

Z-score scale normalized data

## Dimensionality reduction

PCA (optionally using a pre-selected subset of markers)

Selection of dimensions 

UMAP 

Check sample distribution

## Data harmonization (optional) 

Harmonize data using Harmony

Inspect the result compared to the non-harmonized data and decide whether it is necessary

## Clustering

Cluster using the Leiden algorithm

## Annotation

Annotate cells by: 

Manual annotation from differential abundance between clusters or average abundance across clusters.

Automatic annotation using label transfer from a reference dataset. 


# Statistical testing

## Marker selection

Select markers to evaluate per cell type, based on: 

Abundance level and variance (e.g. mean vs standard deviation)

Deviation from background levels (isotypes or ambient negative levels)

Colocalization to isotypes

## Differential abundance analysis

Statistical test (typically Wilcoxon) between conditions

## Differential proximity analysis

Statistical test (typically Wilcoxon) between conditions
