---
title: "Single Cell PNA Analysis SOP"
subtitle: "v0.1.0" 
format: html
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
execute:
  dev: ragg_png
  execute-dir: dir
  echo: false
  warning: false
  message: false
params: 
  
  sample_sheet: "../../../ES_test_data/test_samplesheet.csv"
  data_folder: "../../../ES_test_data/"
  control_markers: ["mIgG1", "mIgG2a", "mIgG2b"]
  mc_cores: 1
---

This SOP describes a standard workflow for analyzing single cell PNA data using the `pixelatorR` package in R. It covers data setup, quality control, data processing, and statistical testing. The workflow is meant to be interactive and modular, allowing users to adapt it to their specific experimental designs and analysis needs.

# Setup

## Packages

```{r}
#| label: setup_packages

library(tidyverse)
library(Seurat)
library(pixelatorR)

```

## Read data

We will begin with reading data. Specify the data file IDs, paths, and conditions below.

```{r}
#| label: setup_read_data

data_files <- 
  c(S1 = "C:/Users/max/data/PNACC_data/pixelator/PNA109_S04_Activated_CD4_cells.layout.pxl",
    S2 = "C:/Users/max/data/PNACC_data/pixelator/PNA109_S05_Raji.layout.pxl",
    S3 = "C:/Users/max/data/PNACC_data/pixelator/PNA109_S07_Raji_CD4_with_SEE.layout.pxl")

conditions <- 
  c(S1 = "ctrl", 
    S2 = "stim1", 
    S3 = "stim2")

pg_data <-
  data_files |> 
  lapply(ReadPNA_Seurat,
         load_proximity_scores = FALSE
  )

```

## Merge data

We merge all data into a single Seurat object, adding metadata for sample alias and condition.

```{r}
#| label: setup_merge_data

# Merge data
pg_data <-
  merge(
    pg_data[[1]],
    y = pg_data[-1],
    add.cell.ids = names(pg_data)
  ) |> 
  JoinLayers()

# Add metadata
pg_data <-
  pg_data |> 
  AddMetaData(colnames(pg_data) |> 
                enframe("i", "comp_id") |> 
                mutate(
                  sample_alias = str_extract(comp_id, ".*_") |> 
                    str_remove("_$"),
                  condition = conditions[sample_alias]
                ) |> 
                column_to_rownames("comp_id"))

```


## Theme

Here we specify the color palettes for clusters and conditions. These can be adjusted to the color combinations that you want for your project, by replacing either palette with a named vector containing color codes.  

```{r}
#| label: setup_theme

cluster_palette <-
  c(
    pixelatorR::PixelgenAccentColors(
      level = 7,
      hue = c(
        "blues", "reds",
        "cyans", "oranges",
        "pinks", "greens",
        "purples", "yellows",
        "greys", "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 9,
      hue = c(
        "blues", "pinks",
        "reds", "yellows",
        "greys", "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 5,
      hue = c(
        "purples", "blues",
        "cyans", "greens",
        "reds", "oranges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 3,
      hue = c("blues", "reds")
    ),
    pixelatorR::PixelgenAccentColors(
      level = 12,
      hue = c("blues", "reds")
    )
  ) |>
  unname() |>
  rep(10)


condition_palette <- 
  set_names(
    pixelatorES::create_sample_palette(
      unique(conditions)
    )[seq_len(n_distinct(conditions))], 
    unique(conditions))


sample_palette <- 
  set_names(
    pixelatorES::create_sample_palette(
      names(data_files)
    )[seq_len(length(data_files))], 
    names(data_files))

```

# Quality Control

In Quality Control we inspect certain metrics that are indicative of the quality of the data per sample and/or per cell. Based on these metrics we can decide whether to remove cells from the dataset before proceeding with data processing.

## Read depth (n_reads)

```{r}
#| label: quality_control_read_depth

FetchData(pg_data, 
          c("sample_alias", "condition", "reads_in_component")) |> 
  ggplot(aes(sample_alias, reads_in_component, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(reads_in_component)),
            aes(sample_alias, median, label = round(median)),
            vjust = -0.5,
            inherit.aes = FALSE)+
  scale_y_log10() +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Read depth per cell",
    x = "Sample",
    y = "Reads per cell (log10 scale)"
  )
ggsave(
  filename = file.path("results", "quality_control_read_depth.png"),
  width = 6,
  height = 4
)

```

## Sequencing saturation

Here we run simulations of 6 cells to estimate the sequencing saturation curves per sample.

```{r}
#| label: quality_control_sequencing_saturation

object_meta <-
  FetchData(pg_data, c("sample_alias", "reads_in_component")) |>
  group_by(sample_alias) |>
  summarise(
    n_comps = n(),
    reads_per_component = mean(reads_in_component)
  )

seqsat_curve_data <-
  data_files |>
  names() |> 
  set_names() |>
  parallel::mclapply(function(sampl) {
    n_comps <-
      object_meta$n_comps[which(object_meta$sample_alias == sampl)]
    n_comps <- pmin(n_comps, 6)
    sampled_comps <- sample(
      pg_data[[]] |>
        filter(sample_alias == sampl) |>
        rownames() |>
        sample(n_comps)
    )
    el <- pg_data |>
      subset(cells = sampled_comps) |>
      Edgelists(lazy = FALSE)
    
    el |>
      SequenceSaturationCurve(n_comps = n_comps)
  }, mc.cores = mc_cores) |>
  bind_rows(.id = "sample_alias") |>
  left_join(
    object_meta,
    by = join_by(sample_alias)
  ) |>
  mutate(
    reads_per_component = reads_per_component * sample_frac
  ) |>
  select(
    sample_alias, reads_per_component, graph_node_saturation,
    graph_edge_saturation
  ) |>
  pivot_longer(
    cols = c("graph_node_saturation", "graph_edge_saturation"),
    names_to = "type", values_to = "saturation"
  ) |>
  mutate(type = str_remove(type, "graph_") |>
           str_remove("_saturation") |>
           str_to_sentence())

seqsat_curve_data |>
  ggplot(aes(reads_per_component, saturation, color = sample_alias)) +
  geom_line(
    data = seqsat_curve_data |>
      group_by(sample_alias, type, reads_per_component) |>
      summarise(saturation = mean(saturation)) |>
      ungroup(),
    aes(group = sample_alias)
  ) +
  geom_point(size = 0.1) +
  facet_wrap(~type) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_manual(values = sample_palette) +
  theme_bw() +
  labs(
    x = "Reads per component",
    y = "Sequencing saturation [%]",
    title = "Sequencing saturation curve",
    color = "Sample ID"
  )
ggsave(
  filename = file.path("results", "quality_control_sequencing_saturation.png"),
  width = 6,
  height = 4
)
```

## Connectivity (average kcore and dangling nodes)

TODO: Add text here

```{r}
#| label: quality_control_connectivity

plot_data <- 
FetchData(pg_data, 
c("sample_alias", "condition", "average_k_core", "k_core_1", "n_umi")) |> 
mutate(prc_dangling_nodes = 100 * k_core_1 / n_umi)
  
plot_data |> 
  ggplot(aes(sample_alias, average_k_core, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(average_k_core)),
            aes(sample_alias, median, label = round(median, 2)),
            vjust = -0.5,
            inherit.aes = FALSE)+
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Average k-core per cell",
    x = "Sample",
    y = "Average k-core"
  )

ggsave(
  filename = file.path("results", "quality_control_average_k_core.png"),
  width = 6,
  height = 4
)

plot_data |> 
  ggplot(aes(sample_alias, prc_dangling_nodes, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(prc_dangling_nodes)),
            aes(sample_alias, median, label = paste(round(median, 1), "%")),
            vjust = -0.5,
            inherit.aes = FALSE)+
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Dangling nodes per cell",
    x = "Sample",
    y = "Dangling nodes [%]"
  )

ggsave(
  filename = file.path("results", "quality_control_dangling_nodes.png"),
  width = 6,
  height = 4
)

```



## Select cells

Next, we will select cells to keep for downstream analysis. We want to remove cells that:
- Are too small, possibly representing cell fragments
- Are noisy, using isotype marker background level
- Have an abnormal marker distribution, using the tau score

We will begin with inspecting the data, and later remove cells in Data Processing that do not meet our quality criteria.

### Define thresholds

Revisit these thresholds according to your data. Sometimes it is necessary to use different thresholds for different samples.

```{r}
#| label: quality_control_define_thresholds

# Define threshold for minimum molecule counts per cell

n_umi_threshold <- 10000
isotype_fraction_threshold <- 0.005

```

### Cell size / Antibody molecule counts (n_umi)

```{r}
#| label: quality_control_select_cells

MoleculeRankPlot(pg_data, group_by = "sample_alias") +
  geom_hline(
    yintercept = n_umi_threshold,
    linetype = "dashed",
    color = "red"
  )
ggsave(
  filename = file.path("results", "quality_control_molecule_rank_plot.png"),
  width = 6,
  height = 4
)


FetchData(pg_data, 
          c("sample_alias", "condition", "n_umi")) |> 
  ggplot(aes(sample_alias, n_umi, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(n_umi)),
            aes(sample_alias, median, label = round(median)),
            vjust = -0.5,
            inherit.aes = FALSE)+
  geom_hline(
    yintercept = n_umi_threshold,
    linetype = "dashed",
    color = "red"
  ) +
  scale_y_log10() +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Antibody molecule counts per cell",
    x = "Sample",
    y = "Molecule counts per cell (log10 scale)"
  )
ggsave(
  filename = file.path("results", "quality_control_antibody_molecule_counts.png"),
  width = 6,
  height = 4
)

```

### Isotype controls

TODO: Add text here

```{r}
#| label: quality_control_isotype_controls


FetchData(pg_data, 
          c("sample_alias", "condition", "isotype_fraction")) |> 
  ggplot(aes(sample_alias, isotype_fraction * 100, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(isotype_fraction * 100)),
            aes(sample_alias, median, label = paste(round(median, 2), "%")),
            vjust = -0.5,
            inherit.aes = FALSE)+
  geom_hline(
    yintercept = isotype_fraction_threshold * 100,
    linetype = "dashed",
    color = "red"
  ) +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Isotype control fraction per cell",
    x = "Sample",
    y = "Isotype control fraction per cell [%]"
  )
      
ggsave(
  filename = file.path("results", "quality_control_isotype_controls.png"),
  width = 6,
  height = 4
)
```

### Tau score

The Tau score measures the heterogeneity of protein composition per cell, and can be used to detect various types of outliers. Cells with high Tau high counts for few proteins, while cells with low Tau have a more heterogeneous composition of proteins than we would expect from a normal cell. As these outliers likely represent technical artefacts, we can remove them from the dataset.

```{r}
#| label: quality_control_tau_score

TauPlot(pg_data)  
ggsave(
  filename = file.path("results", "quality_control_tau_score.png"),
  width = 6,
  height = 4
)

```

## Select markers (optional)

Select markers for dimensionality and clustering (optional) 

Remove markers that: 

Are not positive in any cell population

Colocalize with isotypes

```{r}
#| label: quality_control_select_markers


```

# Data processing

## Normalize and scale data

Normalize using CLR, margin 2

Z-score scale normalized data

```{r}
#| label: data_processing_normalize_scale
```

## Dimensionality reduction

PCA (optionally using a pre-selected subset of markers)

Selection of dimensions 

UMAP 

Check sample distribution

```{r}
#| label: data_processing_dimensionality_reduction
```

## Data harmonization (optional) 

Harmonize data using Harmony

Inspect the result compared to the non-harmonized data and decide whether it is necessary

```{r}
#| label: data_processing_data_harmonization
```

## Clustering

Cluster using the Leiden algorithm

```{r}
#| label: data_processing_clustering
```

## Annotation

Annotate cells by: 

Manual annotation from differential abundance between clusters or average abundance across clusters.

Automatic annotation using label transfer from a reference dataset. 

```{r}
#| label: data_processing_annotation
```

# Statistical testing

## Marker selection

Select markers to evaluate per cell type, based on: 

Abundance level and variance (e.g. mean vs standard deviation)

Deviation from background levels (isotypes or ambient negative levels)

Colocalization to isotypes

```{r}
#| label: statistical_testing_marker_selection
```

## Differential abundance analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_abundance
```

## Differential proximity analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_proximity
```

