---
title: "Single Cell PNA Analysis SOP"
subtitle: "v0.1.0" 
format: html
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
execute:
  dev: ragg_png
  execute-dir: dir
  echo: false
  warning: false
  message: false
params: 
  
  sample_sheet: "../../../ES_test_data/test_samplesheet.csv"
  data_folder: "../../../ES_test_data/"
  control_markers: ["mIgG1", "mIgG2a", "mIgG2b"]
---

# Setup

## Packages

```{r}
#| label: setup_packages

library(tidyverse)
library(Seurat)
library(pixelatorR)

```

## Read data

```{r}
#| label: setup_read_data

data_files <- 
  c(S1 = "C:/Users/max/data/PNACC_data/pixelator/PNA109_S04_Activated_CD4_cells.layout.pxl",
    S2 = "C:/Users/max/data/PNACC_data/pixelator/PNA109_S05_Raji.layout.pxl",
    S3 = "C:/Users/max/data/PNACC_data/pixelator/PNA109_S07_Raji_CD4_with_SEE.layout.pxl")

conditions <- 
  c(S1 = "ctrl", 
    S2 = "stim1", 
    S3 = "stim2")

pg_data <-
  data_files |> 
  lapply(ReadPNA_Seurat,
         load_proximity_scores = FALSE
  )

```

## Merge data

```{r}
#| label: setup_merge_data

pg_data <-
  merge(
    pg_data[[1]],
    y = pg_data[-1],
    add.cell.ids = names(pg_data)
  ) |> 
  JoinLayers()


pg_data <-
  pg_data |> 
  AddMetaData(colnames(pg_data) |> 
                enframe("i", "comp_id") |> 
                mutate(
                  sample_alias = str_extract(comp_id, ".*_") |> 
                    str_remove("_$"),
                  condition = conditions[sample_alias]
                ) |> 
                column_to_rownames("comp_id"))

```


## Theme

```{r}
#| label: setup_theme

cluster_palette <-
  c(
    pixelatorR::PixelgenAccentColors(
      level = 7,
      hue = c(
        "blues", "reds",
        "cyans", "oranges",
        "pinks", "greens",
        "purples", "yellows",
        "greys", "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 9,
      hue = c(
        "blues", "pinks",
        "reds", "yellows",
        "greys", "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 5,
      hue = c(
        "purples", "blues",
        "cyans", "greens",
        "reds", "oranges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 3,
      hue = c("blues", "reds")
    ),
    pixelatorR::PixelgenAccentColors(
      level = 12,
      hue = c("blues", "reds")
    )
  ) %>%
  unname() %>%
  rep(10)


condition_palette <- 
  set_names(
    pixelatorES::create_sample_palette(
      unique(conditions)
      )[seq_len(n_distinct(conditions))], 
    unique(conditions))

```

# Quality Control

Investigate levels and/or distribution of:

## Read depth (n_reads)

```{r}
#| label: quality_control_read_depth

FetchData(pg_data, 
          c("sample_alias", "condition", "reads_in_component")) |> 
  ggplot(aes(sample_alias, reads_in_component, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(reads_in_component)),
            aes(sample_alias, median, label = round(median)),
            vjust = -0.5,
            inherit.aes = FALSE)+
  scale_y_log10() +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Read depth per cell",
    x = "Sample",
    y = "Reads per cell (log10 scale)"
  )
ggsave(
  filename = file.path("results", "quality_control_read_depth.png"),
  width = 6,
  height = 4
)

```

## Sequencing saturation

```{r}
#| label: quality_control_sequencing_saturation



```

## Antibody molecule counts (n_umi)

```{r}
#| label: quality_control_antibody_molecule_counts
```

## Isotype controls

```{r}
#| label: quality_control_isotype_controls
```

## Connectivity (average kcore and dangling nodes)

```{r}
#| label: quality_control_connectivity
```

## Select cells

Remove cells that:

Are small, using the molecule rank plot

Are noisy, using isotype background level 

Have an abnormal marker distribution, using the tau score

```{r}
#| label: quality_control_select_cells
```


## Select markers (optional)

Select markers for dimensionality and clustering (optional) 

Remove markers that: 

Are not positive in any cell population

Colocalize with isotypes

```{r}
#| label: quality_control_select_markers
```

# Data processing

## Normalize and scale data

Normalize using CLR, margin 2

Z-score scale normalized data

```{r}
#| label: data_processing_normalize_scale
```

## Dimensionality reduction

PCA (optionally using a pre-selected subset of markers)

Selection of dimensions 

UMAP 

Check sample distribution

```{r}
#| label: data_processing_dimensionality_reduction
```

## Data harmonization (optional) 

Harmonize data using Harmony

Inspect the result compared to the non-harmonized data and decide whether it is necessary

```{r}
#| label: data_processing_data_harmonization
```

## Clustering

Cluster using the Leiden algorithm

```{r}
#| label: data_processing_clustering
```

## Annotation

Annotate cells by: 

Manual annotation from differential abundance between clusters or average abundance across clusters.

Automatic annotation using label transfer from a reference dataset. 

```{r}
#| label: data_processing_annotation
```

# Statistical testing

## Marker selection

Select markers to evaluate per cell type, based on: 

Abundance level and variance (e.g. mean vs standard deviation)

Deviation from background levels (isotypes or ambient negative levels)

Colocalization to isotypes

```{r}
#| label: statistical_testing_marker_selection
```

## Differential abundance analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_abundance
```

## Differential proximity analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_proximity
```

