---
title: "Single Cell PNA Analysis SOP"
subtitle: "v0.1.0" 
format: html
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
execute:
  dev: ragg_png
  execute-dir: dir
  echo: false
  warning: false
  message: false
---

This SOP describes a standard workflow for analyzing single cell PNA data using the `pixelatorR` package in R. It covers data setup, quality control, data processing, and statistical testing. The workflow is meant to be interactive and modular, allowing users to adapt it to their specific experimental designs and analysis needs.

# Setup

## Packages

```{r}
#| label: setup_packages

library(tidyverse)
library(Seurat)
library(pixelatorR)
library(harmony)
library(ComplexHeatmap)
library(ggplotify)

```

## Functions

```{r}
#| label: setup_functions

save_plot <- 
  function(
    filename, 
    plot = last_plot(),
    width = 6, 
    height = 4, 
    file_formats = c("png", "pdf")
  ) {
    for (format in file_formats) {
      ggsave(
        filename = file.path("results", 
                             paste0(filename, ".", format)),
        plot = plot,
        width = width,
        height = height
      )
    }
  }

```

## Read data

We will begin with reading data. Specify the data file IDs, paths, and conditions below.

```{r}
#| label: setup_read_data

data_files <- 
  c(S1 = "C:/Users/max/data/sop_dev_data/PNA056_Maud_PBMCs_PHA_sample02_S2.layout.dataset.pxl",
    S2 = "C:/Users/max/data/sop_dev_data/PNA056_Maud_PBMCs_unstimulated_sample01_S1.layout.dataset.pxl",
    S3 = "C:/Users/max/data/sop_dev_data/PNA056_Maud_PBMCs_unstimulated_sample06_S6.layout.dataset.pxl")

conditions <- 
  c(S1 = "PHA", 
    S2 = "PBMC", 
    S3 = "PBMC")

pg_data <-
  data_files |> 
  lapply(ReadPNA_Seurat,
         load_proximity_scores = FALSE
  )

```

## Merge data

We merge all data into a single Seurat object, adding metadata for sample alias and condition.

```{r}
#| label: setup_merge_data

# Merge data
pg_data <-
  merge(
    pg_data[[1]],
    y = pg_data[-1],
    add.cell.ids = names(pg_data)
  ) |> 
  JoinLayers()

# Add metadata
pg_data <-
  pg_data |> 
  AddMetaData(colnames(pg_data) |> 
                enframe("i", "comp_id") |> 
                mutate(
                  sample_alias = str_extract(comp_id, ".*_") |> 
                    str_remove("_$"),
                  condition = conditions[sample_alias]
                ) |> 
                column_to_rownames("comp_id"))

```


## Theme

Here we specify the color palettes for clusters and conditions. These can be adjusted to the color combinations that you want for your project, by replacing either palette with a named vector containing color codes.  

```{r}
#| label: setup_theme

cluster_palette <-
  c(
    pixelatorR::PixelgenAccentColors(
      level = 7,
      hue = c(
        "blues", "reds",
        "cyans", "oranges",
        "pinks", "greens",
        "purples", "yellows",
        "greys", "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 9,
      hue = c(
        "blues", "pinks",
        "reds", "yellows",
        "greys", "beiges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 5,
      hue = c(
        "purples", "blues",
        "cyans", "greens",
        "reds", "oranges"
      )
    ),
    pixelatorR::PixelgenAccentColors(
      level = 3,
      hue = c("blues", "reds")
    ),
    pixelatorR::PixelgenAccentColors(
      level = 12,
      hue = c("blues", "reds")
    )
  ) |>
  unname() |>
  rep(10)


condition_palette <- 
  set_names(
    pixelatorES::create_sample_palette(
      unique(conditions)
    )[seq_len(n_distinct(conditions))], 
    unique(conditions))


sample_palette <- 
  set_names(
    pixelatorES::create_sample_palette(
      names(data_files)
    )[seq_len(length(data_files))], 
    names(data_files))


#' Cell palette
#'
#' @export
#'
cell_palette <-
  c(
    "CD4 T" = "#6D92D1",
    "CD4 Naive" = "#B9CDED",
    "Naive CD4 T" = "#B9CDED",
    "CD4 TSCM" = "#92B0E0",
    "CD4 TCM" = "#6D92D1",
    "CD4 TEM" = "#4A73C0",
    "CD4 TEFF" = "#224792",
    "CD4 TEMRA" = "#1C3A76",
    "Treg" = "#2955AE",
    "CD4 CTL" = "#2955AE",
    "CD8 T" = "#75C5B5",
    "CD8 Naive" = "#D0EDE6",
    "Naive CD8 T" = "#D0EDE6",
    "CD8 TSCM" = "#A2DACE",
    "CD8 TCM" = "#75C5B5",
    "CD8 TEM" = "#4AAF9D",
    "CD8 TEFF" = "#209785",
    "CD8 TEMRA" = "#1B7E6F",
    "other T" = "#1B7E9F",
    "Other T" = "#1B7E9F",
    "MAIT" = "#0D6D5E",
    "DPT" = "#1B7E9F",
    "DNT" = "#7B787F",
    "dnT" = "#7B787F",
    "NK" = "#A28EDB",
    "CD56dim NK" = "#A28EDB",
    "CD56bright NK" = "#866CCD",
    "NK_CD56bright" = "#866CCD",
    "NKT" = "#6C4ABD",
    "B" = "#C7A989", # "#917557",
    "B naive" = "#F0DAC4",
    "Naive B" = "#F0DAC4",
    "B intermediate" = "#F0DAC4",
    "Memory B" = "#C7A989",
    "B memory" = "#C7A989",
    "Plasma cells" = "#DE9982",
    "Plasmablast" = "#DE9982",
    "DC" = "#DA94C1",
    "cDC1" = "#DA94C1",
    "cDC2" = "#BB5391",
    "pDC" = "#9C4579",
    "Mono" = "#F7DFA0",
    "Classical Mono" = "#F7DFA0",
    "CD14 Mono" = "#F7DFA0",
    "Intermediate Mono" = "#E6BE4A",
    "CD16 Mono" = "#BD9315",
    "Non-classical Mono" = "#BD9315",
    "other" = "#797979",
    "Neutrophils" = "#CDCDCD",
    "Basophils" = "#797979",
    "Platelet" = "#DB6365",
    "Platelets" = "#DB6365",
    "gdT" = "#5A3E9E"
  )
```

# Quality Control

In Quality Control we inspect certain metrics that are indicative of the quality of the data per sample and/or per cell. Based on these metrics we can decide whether to remove cells from the dataset before proceeding with data processing.

## Read depth (n_reads)

```{r}
#| label: quality_control_read_depth

FetchData(pg_data, 
          c("sample_alias", "condition", "reads_in_component")) |> 
  ggplot(aes(sample_alias, reads_in_component, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(reads_in_component)),
            aes(sample_alias, median, label = round(median)),
            vjust = -0.5,
            inherit.aes = FALSE)+
  scale_y_log10() +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Read depth per cell",
    x = "Sample",
    y = "Reads per cell (log10 scale)"
  )
save_plot(
  filename = "quality_control_read_depth",
  width = 6,
  height = 4
)

```

## Sequencing saturation

Here we run simulations of 6 cells to estimate the sequencing saturation curves per sample.

```{r}
#| label: quality_control_sequencing_saturation

object_meta <-
  FetchData(pg_data, c("sample_alias", "reads_in_component")) |>
  group_by(sample_alias) |>
  summarise(
    n_comps = n(),
    reads_per_component = mean(reads_in_component)
  )

seqsat_curve_data <-
  data_files |>
  names() |> 
  set_names() |>
  parallel::mclapply(function(sampl) {
    n_comps <-
      object_meta$n_comps[which(object_meta$sample_alias == sampl)]
    n_comps <- pmin(n_comps, 6)
    sampled_comps <- sample(
      pg_data[[]] |>
        filter(sample_alias == sampl) |>
        rownames() |>
        sample(n_comps)
    )
    el <- pg_data |>
      subset(cells = sampled_comps) |>
      Edgelists(lazy = FALSE)
    
    el |>
      SequenceSaturationCurve(n_comps = n_comps)
  }, mc.cores = 1) |>
  bind_rows(.id = "sample_alias") |>
  left_join(
    object_meta,
    by = join_by(sample_alias)
  ) |>
  mutate(
    reads_per_component = reads_per_component * sample_frac
  ) |>
  select(
    sample_alias, reads_per_component, graph_node_saturation,
    graph_edge_saturation
  ) |>
  pivot_longer(
    cols = c("graph_node_saturation", "graph_edge_saturation"),
    names_to = "type", values_to = "saturation"
  ) |>
  mutate(type = str_remove(type, "graph_") |>
           str_remove("_saturation") |>
           str_to_sentence())

seqsat_curve_data |>
  ggplot(aes(reads_per_component, saturation, color = sample_alias)) +
  geom_line(
    data = seqsat_curve_data |>
      group_by(sample_alias, type, reads_per_component) |>
      summarise(saturation = mean(saturation)) |>
      ungroup(),
    aes(group = sample_alias)
  ) +
  geom_point(size = 0.1) +
  facet_wrap(~type) +
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_manual(values = sample_palette) +
  theme_bw() +
  labs(
    x = "Reads per component",
    y = "Sequencing saturation [%]",
    title = "Sequencing saturation curve",
    color = "Sample ID"
  )
save_plot(
  filename = "quality_control_sequencing_saturation",
  width = 6,
  height = 4
)
```

## Connectivity (average kcore and dangling nodes)

TODO: Add text here

```{r}
#| label: quality_control_connectivity

plot_data <- 
  FetchData(pg_data, 
            c("sample_alias", "condition", "average_k_core", "k_core_1", "n_umi")) |> 
  mutate(prc_dangling_nodes = 100 * k_core_1 / n_umi)

plot_data |> 
  ggplot(aes(sample_alias, average_k_core, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(average_k_core)),
            aes(sample_alias, median, label = round(median, 2)),
            vjust = -0.5,
            inherit.aes = FALSE)+
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Average k-core per cell",
    x = "Sample",
    y = "Average k-core"
  )

save_plot(
  filename = "quality_control_average_k_core",
  width = 6,
  height = 4
)

plot_data |> 
  ggplot(aes(sample_alias, prc_dangling_nodes, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(prc_dangling_nodes)),
            aes(sample_alias, median, label = paste(round(median, 1), "%")),
            vjust = -0.5,
            inherit.aes = FALSE)+
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Dangling nodes per cell",
    x = "Sample",
    y = "Dangling nodes [%]"
  )

save_plot(
  filename = "quality_control_dangling_nodes",
  width = 6,
  height = 4
)

```



## Select cells

Next, we will select cells to keep for downstream analysis. We want to remove cells that:
- Are too small, possibly representing cell fragments
- Are noisy, using isotype marker background level
- Have an abnormal marker distribution, using the tau score

We will begin with inspecting the data, and later remove cells in Data Processing that do not meet our quality criteria.

### Define thresholds

Revisit these thresholds according to your data. Sometimes it is necessary to use different thresholds for different samples.

```{r}
#| label: quality_control_define_thresholds

# Define threshold for minimum molecule counts per cell

n_umi_threshold <- 10000
isotype_fraction_threshold <- 0.005

```

### Cell size / Antibody molecule counts (n_umi)

```{r}
#| label: quality_control_select_cells

MoleculeRankPlot(pg_data, group_by = "sample_alias") +
  geom_hline(
    yintercept = n_umi_threshold,
    linetype = "dashed",
    color = "red"
  )
save_plot(
  filename = "quality_control_molecule_rank_plot",
  width = 6,
  height = 4
)


FetchData(pg_data, 
          c("sample_alias", "condition", "n_umi")) |> 
  ggplot(aes(sample_alias, n_umi, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(n_umi)),
            aes(sample_alias, median, label = round(median)),
            vjust = -0.5,
            inherit.aes = FALSE)+
  geom_hline(
    yintercept = n_umi_threshold,
    linetype = "dashed",
    color = "red"
  ) +
  scale_y_log10() +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Antibody molecule counts per cell",
    x = "Sample",
    y = "Molecule counts per cell (log10 scale)"
  )
save_plot(
  filename = "quality_control_antibody_molecule_counts",
  width = 6,
  height = 4
)

```

### Isotype controls

TODO: Add text here

```{r}
#| label: quality_control_isotype_controls


FetchData(pg_data, 
          c("sample_alias", "condition", "isotype_fraction")) |> 
  ggplot(aes(sample_alias, isotype_fraction * 100, fill = condition)) +
  geom_violin(draw_quantiles = 0.5, scale = "width") +
  geom_text(data = function(x) x |> 
              group_by(sample_alias) |>  
              summarise(median = median(isotype_fraction * 100)),
            aes(sample_alias, median, label = paste(round(median, 2), "%")),
            vjust = -0.5,
            inherit.aes = FALSE)+
  geom_hline(
    yintercept = isotype_fraction_threshold * 100,
    linetype = "dashed",
    color = "red"
  ) +
  scale_fill_manual(values = condition_palette) +
  theme_bw() +
  labs(
    title = "Isotype control fraction per cell",
    x = "Sample",
    y = "Isotype control fraction per cell [%]"
  )

save_plot(
  filename = "quality_control_isotype_controls",
  width = 6,
  height = 4
)
```

### Tau score

The Tau score measures the heterogeneity of protein composition per cell, and can be used to detect various types of outliers. Cells with high Tau high counts for few proteins, while cells with low Tau have a more heterogeneous composition of proteins than we would expect from a normal cell. As these outliers likely represent technical artefacts, we can remove them from the dataset.

```{r}
#| label: quality_control_tau_score

TauPlot(pg_data)  
save_plot(
  filename = "quality_control_tau_score",
  width = 6,
  height = 4
)

```

## Select markers (optional)

In this step we select markers that we believe are positive in at least one cell population in the dataset. There is not a single correct way to do this, and the selection criteria will depend on the dataset and experimental design. 

Here we provide one example approach based on counts per million (CPM) thresholds. We calculate the 50th percentile (the median) and 95th percentile of CPM per marker across all cells, and select markers that have a 95th percentile above a certain threshold. The threshold can be adjusted according to the background levels of the dataset, and according to expectations of which markers should be present in the data.

In this example, we set the threshold to 1000 CPM at the 95th percentile. This means, in effect, that we let cell populations composing at least 5% of the sample dictate which markers we use in later analysis steps. Comparing the 95th percentile to the 50th percentile also gives us an idea of how much the marker expression varies across cells.


```{r}
#| label: quality_control_select_markers

marker_cpm_p95_threshold <- 1000

count_data <- 
  LayerData(pg_data, "counts")
cpm_data <- 
  count_data |> 
  sweep(1, colSums(count_data), "/") 

cpm_data <- cpm_data * 1e6

cpm_quantiles <- 
  cpm_data |> 
  apply(1, function(x) quantile(x, c(0.5, 0.95))) |> 
  t() |> 
  as_tibble(rownames = "marker") |> 
  mutate(kept = `95%` > marker_cpm_p95_threshold)

cpm_quantiles|> 
  ggplot(aes(`50%`, `95%` / `50%`, color = kept)) +
  geom_point(size = 0.1) +
  geom_text(aes(label = marker),
            vjust = -0.5, 
            size = 2) +
  scale_x_log10(expand = expansion(0.2)) +
  scale_y_log10() +
  theme_bw() +
  labs(
    title = "Marker selection based on CPM",
    x = "50th percentile of CPM (log10 scale)",
    y = "95th / 50th percentile of CPM (log10 scale)"
  )
save_plot(
  filename = "quality_control_marker_selection_cpm",
  width = 6,
  height = 4
)


cpm_quantiles |> 
  arrange(`95%`) |>
  mutate(marker = factor(marker, levels = marker)) |>
  ggplot(aes(x = marker, y = `95%`, fill = kept)) +
  geom_col() +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "95th percentile of CPM per marker",
    x = "Marker",
    y = "95th percentile of CPM (log10 scale)"
  )
save_plot(
  filename = "quality_control_marker_selection_cpm_barplot",
  width = 6,
  height = 4
)

selected_markers <- 
  cpm_quantiles |> 
  filter(kept) |> 
  pull(marker)


```

# Data processing

## Normalize and scale data

We typically normalize PNA data using CLR over margin 2, which means that we normalize the counts per cell by the geometric mean across all markers. This normalization method accounts for differences in sequencing depth between cells, while preserving relative differences in marker abundance within each cell.

Next, we perform Z-score scaling of the normalized data per marker, which centers the data around zero and scales it to unit variance. This step is important for downstream dimensionality reduction and clustering, as it ensures that all markers contribute equally to the analysis, regardless of their absolute abundance levels.

```{r}
#| label: data_processing_normalize_scale

pg_data <-
  pg_data |>
  # Data normalization
  NormalizeData(normalization.method = "CLR", margin = 2) |>
  # Z score scaling
  ScaleData() 

```

## Dimensionality reduction

Next, we perform a Principal Component Analysis (PCA) to extract the main axes of variation in the data. We typically use the first 10-20 principal components (PCs) for downstream analysis, but this number can be adjusted based on the dataset and experimental design. A rule of thumb is that the selected PCs should explain a substantial portion of the total variance in the data, while avoiding overfitting or noise. Oftentimes PCs correspond to technical effects, cell states, or experiment effects in the data, and so this number can be adjusted e.g. according to expectations on the number of cell states one expects.

Here we use the marker selection we previously performed to select markers for PCA. This step is optional, but can help to reduce noise and improve the interpretability of the results.

After PCA, we perform Uniform Manifold Approximation and Projection (UMAP) to visualize the data in a low-dimensional space. UMAP is a non-linear dimensionality reduction technique that preserves both local and global structure in the data, making it well-suited for visualizing complex datasets such as single-cell PNA data.

```{r}
#| label: data_processing_dimensionality_reduction

npcs <- 20
n_neighbors <- 30
pg_data <- 
  pg_data |> 
  RunPCA(npcs = npcs, features = selected_markers) |>
  RunUMAP(reduction = "pca", dims = 1:20, n.neighbors = n_neighbors)

pca_red <- 
  Reductions(pg_data, "pca")

pca_red@stdev |> 
  as_tibble(rownames = "PC") |> 
  mutate(PC = PC |> str_remove("PC_") |> as.integer()) |> 
  ggplot(aes(PC, value)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(
    title = "PCA Elbow Plot",
    x = "Principal Component",
    y = "Standard Deviation"
  )

ggsave(
  filename = "results/data_processing_pca_elbow_plot.png",
  width = 6,
  height = 4
)

cell_metadata <- 
  FetchData(pg_data, c("sample_alias", "condition")) |> 
  as_tibble(rownames = "cell_id")

Embeddings(pg_data, "pca") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(PC_1, PC_2, color = sample_alias)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = sample_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_pca",
  width = 6,
  height = 4
)

Embeddings(pg_data, "umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(umap_1, umap_2, color = sample_alias)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = sample_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_umap",
  width = 6,
  height = 4
)
```

## Data harmonization (optional)

If your experiment contains multiple samples with similar cell types, but these cell types for one reason or other do not cluster together well, it can be beneficial to harmonize the data. Data harmonization aims to align similar cell types across samples such that they cluster together, which can help to understand which cell types are shared between samples, even if they have somewhat different protein profile. This is particularly useful in experiments with batch effects or technical variability, or biological effects such as donor variability or stimulation perturbations. 

Here we use the Harmony algorithm to harmonize the data based on the sample ID, but any sample metadata can be used. 

We recommend inspecting the harmonized data carefully to ensure that the harmonization has not removed important biological variation, and that similar cell types are indeed clustering together across samples. In many cases it is not necessary to harmonize the data, especially if the samples are already well-mixed and similar cell types cluster together without harmonization.

The harmonization we perform here results in a harmonized dimensional reduction called "harmony", which we use for UMAP visualization and clustering. However, the original data remains unchanged, so we really only use the harmonization to find common cell identities between samples.

```{r}
#| label: data_processing_data_harmonization
pg_data <-
  pg_data |>
  RunHarmony(
    group.by.vars = "sample_alias",
    plot_convergence = FALSE,
  ) |>
  RunUMAP(
    reduction = "harmony", dims = 1:20,
    reduction.name = "harmony_umap",
    n.neighbors = n_neighbors
  ) 

cell_metadata <- 
  FetchData(pg_data, c("sample_alias", "condition")) |> 
  as_tibble(rownames = "cell_id")

Embeddings(pg_data, "harmony") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmony_1, harmony_2, color = sample_alias)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = sample_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_harmony",
  width = 6,
  height = 4
)

Embeddings(pg_data, "harmony_umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = sample_alias)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = sample_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_harmonyumap",
  width = 6,
  height = 4
)

```

## Clustering

After dimensionality reduction, we perform clustering. We essentially group cells that are similar in their protein profiles into clusters, which we can later annotate as cell types or cell states. We will perform Leiden clustering on a nearest neighbor graph constructed from the PCA or Harmony reduction. The resolution parameter controls the granularity of the clustering, with higher values resulting in more clusters. The optimal resolution depends on the dataset and experimental design, and can be determined by inspecting the clustering results at different resolutions.

```{r}
#| label: data_processing_clustering

pg_data <- 
  pg_data |>
  FindNeighbors(reduction = "harmony", dims = 1:10) |>
  FindClusters(random.seed = 1, resolution = 0.5)

cell_metadata <- 
  FetchData(pg_data, c("sample_alias", "condition", "seurat_clusters")) |> 
  as_tibble(rownames = "cell_id")

Embeddings(pg_data, "harmony") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmony_1, harmony_2, color = seurat_clusters)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = cluster_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_harmony_clusters",
  width = 6,
  height = 4
)

Embeddings(pg_data, "harmony_umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = seurat_clusters)) +
  geom_point() +
  coord_fixed() +
  scale_color_manual(values = cluster_palette) +
  theme_bw()

save_plot(
  filename = "data_processing_harmonyumap_clusters",
  width = 6,
  height = 4
)

```

## Annotation

### Automatic annotation


Here we perform an automatic cell annotation by matching each cell’s protein profile to annotated 
cells in a reference PBMC dataset. These automated labels should be interpreted with caution, particularly when the 
sample composition differs substantially from the PBMC reference. In such cases, for example in certain stimulated 
samples or cell lines, the annotations may be unreliable. We therefore recommend performing cell annotation as part of 
downstream analysis to ensure accurate and context-appropriate results.

Automated annotation can be a great head start in annotation, but all annotations should at least be inspected and verified, and adjusted when necessary. 

```{r}
#| label: data_processing_annotation

# Load reference dataset
reference <- pixelatorES::read_annotation_reference()

# Set names of annotation columns in reference
reference$l1_annotation <- reference$celltype.l1
reference$l2_annotation <- reference$celltype.l2

# Annotation by label transfer
pg_data <-
  pg_data |>
  AnnotateCells(
    reference = reference,
    reference_groups = c("l1_annotation", "l2_annotation"),
    reference_assay = "PNA",
    query_assay = "PNA",
    # Summarise annotations by cluster, adding an extra column with majority vote
    summarize_by_column = "seurat_clusters",
    normalization_method = "CLR",
    method = "nmf"
  )


cell_metadata <- 
  FetchData(pg_data, c("sample_alias", "condition", "seurat_clusters", 
                       "l1_annotation_summary", "l2_annotation_summary", 
                       "l1_annotation", "l2_annotation")) |> 
  as_tibble(rownames = "cell_id")


Embeddings(pg_data, "harmony_umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = l1_annotation_summary)) +
  geom_point() +
  geom_text(data = . %>% 
              group_by(l1_annotation_summary) %>% 
              summarise(harmonyumap_1 = median(harmonyumap_1),
                        harmonyumap_2 = median(harmonyumap_2)),
            aes(label = l1_annotation_summary),
            size = 3,
            color = "black") +
  coord_fixed() +
  scale_color_manual(values = cell_palette) +
  theme_bw() +
  theme(panel.grid = element_blank())
save_plot(
  filename = "data_processing_harmonyumap_l1_annotation_summary",
  width = 6,
  height = 4
)


Embeddings(pg_data, "harmony_umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = l1_annotation)) +
  geom_point() +
  geom_text(data = . %>% 
              group_by(l1_annotation) %>% 
              summarise(harmonyumap_1 = median(harmonyumap_1),
                        harmonyumap_2 = median(harmonyumap_2)),
            aes(label = l1_annotation),
            size = 3,
            color = "black") +
  coord_fixed() +
  scale_color_manual(values = cell_palette) +
  theme_bw() +
  theme(panel.grid = element_blank())
save_plot(
  filename = "data_processing_harmonyumap_l1_annotation",
  width = 6,
  height = 4
)



Embeddings(pg_data, "harmony_umap") |> 
  as_tibble(rownames = "cell_id") |> 
  left_join(cell_metadata, 
            by = "cell_id") |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = l2_annotation)) +
  geom_point() +
  geom_text(data = . %>% 
              group_by(l2_annotation) %>% 
              summarise(harmonyumap_1 = median(harmonyumap_1),
                        harmonyumap_2 = median(harmonyumap_2)),
            aes(label = l2_annotation),
            size = 3,
            color = "black") +
  coord_fixed() +
  scale_color_manual(values = cell_palette) +
  theme_bw() +
  theme(panel.grid = element_blank())
save_plot(
  filename = "data_processing_harmonyumap_l2_annotation",
  width = 6,
  height = 4
)


FetchData(pg_data, c("sample_alias", "l1_annotation_summary")) |> 
  group_by(sample_alias, l1_annotation_summary) |> 
  count(name = "n_cells") |> 
  group_by(sample_alias) |> 
  mutate(prc_cells = 100 * n_cells / sum(n_cells)) |>
  ggplot(aes(sample_alias, prc_cells, fill = l1_annotation_summary)) +
  geom_col(position = "fill") +
  geom_text(aes(label = paste0(round(prc_cells, 1), "%")),
            position = position_fill(vjust = 0.5),
            size = 3) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = cell_palette) +
  theme_bw() +
  labs(
    title = "Cell type composition per sample",
    x = "Sample",
    y = "Cell type fraction"
  )
save_plot(
  filename = "data_processing_cell_type_composition_per_sample",
  width = 6,
  height = 4
)
  
  
```

### Manual annotation

Manual annotation from differential abundance between clusters or average abundance across clusters.


```{r}
#| label: data_processing_manual_annotation

diff_abundant_markers_cluster <- 
  FindAllMarkers(
    pg_data, 
    logfc.threshold = 0,
    min_pct = 0, 
    return.thresh = 1
  ) |> 
  as_tibble() |> 
  rename(marker = gene)

diff_abundant_markers_cluster |> 
  group_by(marker) |> 
  filter(any(p_val_adj < 0.05 & avg_log2FC > 1)) |> 
  select(marker, avg_log2FC, cluster) |> 
  pivot_wider(names_from = cluster, values_from = avg_log2FC, values_fill = 0) |>
  column_to_rownames("marker") |>
  t() |> 
  pheatmap(
    color = PixelgenGradient(100, "BluesCherry"),
    breaks = seq(-3, 3, length.out = 101),
    cellwidth = 8, 
    cellheight = 8,
    heatmap_legend_param = list(title = "Avg. log2 FC"),
  ) |> 
  as.ggplot()

save_plot(
  filename = "data_processing_differential_abundance_markers_per_cluster",
  width = 6,
  height = 4
)

markers_of_interest <- c("CD3e", "CD4", "CD8", "CD19", "CD14", "CD16", "CD56")
plot_data <- 
  FetchData(
    pg_data,
    c("harmonyumap_1", "harmonyumap_2",
      markers_of_interest),
    slot = "data"
  ) |> 
  pivot_longer(
    cols = all_of(markers_of_interest),
    names_to = "marker",
    values_to = "CLR"
  )

plot_data |> 
  ggplot(aes(harmonyumap_1, harmonyumap_2, color = CLR)) +
  geom_point(size = 0.1) +
  facet_wrap(~marker) +
  scale_color_gradient(low = "lightgrey", high = "darkblue") +
  coord_fixed() +
  theme_bw() +
  theme(panel.grid = element_blank()) +
  labs(
    title = "Marker expression on Harmony UMAP",
    x = "Harmony UMAP 1",
    y = "Harmony UMAP 2",
    color = "CLR"
  )
save_plot(
  filename = "data_processing_marker_expression_harmonyumap",
  width = 6,
  height = 4
)
```

# Statistical testing

## Marker selection

Select markers to evaluate per cell type, based on:

Abundance level and variance (e.g. mean vs standard deviation)

Deviation from background levels (isotypes or ambient negative levels)

Colocalization to isotypes

### Colocalization with isotypes

All filters can be used together or separately, but it's recommended to only use the background_threshold_pct filter. This filter is applied to the fraction of UMI counts which does not depend on the cell size (total number of molecules).

The filters are applied in the following order:

Remove protein pairs with a minimum UMI count fraction below background_threshold_pct

Remove protein pairs with a minimum UMI count below background_threshold_count

Remove protein pairs detected in fewer than min_cells_count cells

```{r}
#| label: statistical_testing_marker_selection

isotype_markers <- 
  c("mIgG1", "mIgG2a", "mIgG2b")

abundance_data <- 
  LayerData(pg_data, "data") |> 
  as_tibble(rownames = "marker") |> 
  pivot_longer(-marker, 
               names_to = "component", 
               values_to = "CLR") |> 
  left_join(
    FetchData(pg_data, c("sample_alias", "seurat_clusters")) |> 
      as_tibble(rownames = "component"),
    by = "component"
  ) 

marker_isotype_coloc <- 
  ProximityScores(pg_data, lazy = TRUE, 
                  add_marker_proportions = TRUE,
                  meta_data_columns = c("sample_alias", "seurat_clusters")) |> 
  filter(marker_1 %in% isotype_markers | marker_2 %in% isotype_markers) |> 
  mutate(marker = ifelse(marker_1 %in% isotype_markers, marker_2, marker_1)) |> 
  group_by(sample_alias, seurat_clusters, component, marker) |>
  summarise(join_count = sum(join_count), 
            join_count_expected_mean = sum(join_count_expected_mean)) |> 
  mutate(log2_ratio = log2((join_count + 1) / (join_count_expected_mean + 1))) |> 
  collect()


marker_isotype_coloc |> 
  left_join(abundance_data) |> 
  ggplot(aes(CLR, log2_ratio, color = log2_ratio > log2(1.5))) +
  geom_point(size = 0.1) +
  geom_hline(yintercept = 0) +
  facet_wrap(~sample_alias) +
  theme_bw() +
  scale_color_manual(values = c("black", "orangered")) +
  labs(
    title = "Marker colocalization with isotypes",
    x = "CLR abundance",
    y = "Log2 ratio of observed vs expected colocalization"
  )
save_plot(
  filename = "statistical_testing_marker_selection_marker_isotype_coloc",
  width = 6,
  height = 4
)

marker_isotype_coloc |> 
  group_by(sample_alias, marker) |> 
  summarise(join_count = sum(join_count), 
            join_count_expected_mean = sum(join_count_expected_mean)) |> 
  mutate(log2_ratio = log2((join_count + 1) / (join_count_expected_mean + 1))) |> 
  ungroup() |>
  left_join(abundance_data |> 
              group_by(marker, sample_alias) |> 
              summarise(CLR = mean(CLR)),
            by = c("marker", "sample_alias")) |>
  ggplot(aes(CLR, log2_ratio, color = log2_ratio > log2(1.5))) +
  geom_hline(yintercept = 0) +
  geom_point(size = 0.1) +
  geom_text(aes(label = marker),
            vjust = -0.5,
            size = 2) +
  facet_wrap(~sample_alias) +
  scale_color_manual(values = c("black", "orangered")) +
  theme_bw() +
  labs(
    title = "Marker colocalization with isotypes per sample",
    x = "Mean CLR abundance",
    y = "Sample-wide log2 ratio of observed vs expected colocalization"
  ) 
save_plot(
  filename = "statistical_testing_marker_selection_marker_isotype_coloc_per_sample",
  width = 8,
  height = 6
)



plot_data <- 
  marker_isotype_coloc |> 
  group_by(sample_alias, seurat_clusters, marker) |> 
  summarise(join_count = sum(join_count), 
            join_count_expected_mean = sum(join_count_expected_mean)) |> 
  mutate(log2_ratio = log2((join_count + 1) / (join_count_expected_mean + 1))) |> 
  ungroup() |>
  left_join(abundance_data |> 
              group_by(marker, sample_alias, seurat_clusters) |> 
              summarise(CLR = mean(CLR)),
            by = c("marker", "sample_alias", "seurat_clusters"))

plot_data |>
  ggplot(aes(CLR, log2_ratio, color = log2_ratio > log2(1.5))) +
  geom_hline(yintercept = 0) +
  geom_point(size = 0.1) +
  
  geom_text(aes(label = marker),
            vjust = -0.5,
            size = 2) +
  facet_grid(seurat_clusters~sample_alias) +
  scale_color_manual(values = c("black", "orangered")) +
  theme_bw() +
  labs(
    title = "Marker colocalization with isotypes per cluster",
    x = "Mean CLR abundance",
    y = "Cluster-wide log2 ratio of observed vs expected colocalization"
  ) 
save_plot(
  filename = "statistical_testing_marker_selection_marker_isotype_coloc_per_sample_cluster",
  width = 8,
  height = 6
)

plot_data |> 
  group_by(marker) |> 
  filter(any(log2_ratio > log2(1.5))) |>
  mutate(total_log2_ratio = sum(log2_ratio)) |>
  arrange(desc(total_log2_ratio)) |>
  ungroup() |> 
  mutate(marker = factor(marker, 
                         levels = unique(marker))) |>
  ggplot(aes(marker, seurat_clusters, fill = log2_ratio)) +
  geom_tile() +
  scale_fill_gradientn(colors = PixelgenGradient(100, "BluesCherry")) +
  coord_fixed() +
  facet_grid(sample_alias ~ .) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "Marker colocalization with isotypes per cluster",
    x = "Marker",
    y = "Cluster"
  )
save_plot(
  filename = "statistical_testing_marker_selection_marker_isotype_coloc_heatmap",
  width = 12,
  height = 6
)

```

## Differential abundance analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_abundance

print(1)

```

## Differential proximity analysis

Statistical test (typically Wilcoxon) between conditions

```{r}
#| label: statistical_testing_differential_proximity

print(1)

```

